# Positron: The Other IDE {#sec-positron}

::: {.callout-note}
## What this chapter covers
RStudio is not the only way to write R any more. This chapter explains what
Positron is, how it differs from RStudio, how to set it up, and how to decide
whether to switch. Nothing else in this book depends on which IDE you use.
:::

```{r}
#| label: setup-positron
#| include: false
library(webexercises)
```

## What Positron is

[Positron](https://positron.posit.co/) is a free IDE from Posit — the same
company that makes RStudio — built for Python and R. It was released in
public beta in 2024 and has had stable releases since.

The important architectural difference: RStudio was written from scratch for R,
while Positron is built on Code OSS, the open-source foundation of Visual Studio
Code. That single fact explains most of the differences below. Positron inherits
VS Code's extension ecosystem, its editor, its keyboard shortcuts and its
workspace model, and Posit has added the data-science pieces on top — a real
console, a Variables pane, a Plots pane, a Help pane and a Data Explorer.

::: {.callout-important}
## You do not need Positron for this course
Everything in this book works identically in RStudio, Positron, VS Code or plain
R in a terminal. R is the language; the IDE is just an editor with conveniences
attached. If you are new to R, stay in RStudio until you are comfortable, then
try Positron when you are curious. Learning two new things at once is harder
than learning one.
:::

## How it differs from RStudio

### Projects and workspaces

This is the difference that trips up RStudio users first.

In RStudio, a project is marked by an `.Rproj` file, and opening it sets the
working directory and loads project-specific settings.

Positron has no equivalent of `.Rproj`. It uses VS Code's **workspace** model:
you open a *folder*, and that folder becomes your working directory. There is no
marker file unless you create project-specific settings, in which case Positron
writes `.vscode/settings.json`.

The practical consequences:

- Positron will happily open a folder containing an `.Rproj` file and ignore it.
- You can move between the two IDEs on the same project without anything
  breaking.
- The `here` package works in both. `here()` does not look only for `.Rproj`; it
  recognises a whole list of root markers, including `.git`,
  `.vscode/settings.json`, `_quarto.yml`, `renv.lock` and `DESCRIPTION`. If your
  project is under version control — and it should be — `here()` finds the root
  in Positron without you doing anything.

::: {.callout-tip}
## Add a `.here` file if you need one
If a project has no `.Rproj` and no `.git`, `here()` has nothing to anchor to.
Run `here::set_here()` once in the project root: it creates an empty `.here`
file that both IDEs respect.
:::

### What RStudio still does that Positron does not

Posit's own comparison lists three:

1. **Saving and restoring the workspace** when you restart R. Positron does not
   do this — which, given the advice in @sec-tidy to turn that feature off
   anyway, is arguably a feature rather than a gap.
2. **Dedicated panes and buttons for specialised tasks**, such as code profiling
   or the package-development tools (`Build`, `Check`, `Install`). These are
   available in Positron through the terminal and extensions, but not as
   buttons.
3. **The data import wizard** (**File → Import Dataset**). There is an open
   feature request; for now you write the `read_csv()` call yourself. Since
   @sec-import recommends copying the generated code into your script anyway,
   this costs you a preview rather than a capability.

### What Positron adds

- **The Data Explorer.** Click a data frame in the Variables pane and it opens
  in a sortable, filterable, scrollable viewer with column summaries and
  missing-value counts. It is noticeably better than RStudio's `View()`.
- **Multiple R versions**, selectable per session from the interpreter picker in
  the top right.
- **Python and R with equal support**, in the same window, in the same project.
  If you collaborate with people who use Python, this matters.
- **The VS Code extension ecosystem** — Quarto, Git tooling, spell checking,
  and so on.
- **Multiple consoles** running at once, so a long model fit does not block you
  from doing anything else.
- **Posit Assistant**, the built-in AI assistant (@sec-ai).

### Things that will feel different on day one

| | RStudio | Positron |
|---|---|---|
| Run current line | `Ctrl/Cmd + Enter` | `Ctrl/Cmd + Enter` |
| Run whole chunk | `Ctrl/Cmd + Shift + Enter` | `Ctrl/Cmd + Shift + Enter` |
| Insert pipe | `Ctrl/Cmd + Shift + M` | `Ctrl/Cmd + Shift + M` |
| Insert chunk | `Ctrl/Cmd + Alt + I` | `Ctrl/Cmd + Alt + I` |
| Command palette | `Ctrl/Cmd + Shift + P` | `Ctrl/Cmd + Shift + P` |
| Restart R | `Ctrl/Cmd + Shift + F10` | Command palette → "Restart R" |
| Project switching | `.Rproj` files | **File → Open Folder** |
| Environment | Environment pane | Variables pane |

The most useful habit in Positron is the **command palette**
(`Ctrl/Cmd + Shift + P`). Almost everything is reachable by typing part of its
name, which matters because there are fewer buttons than in RStudio.

::: {.callout-tip}
## Keep the RStudio keybindings
If your fingers know RStudio, install the **RStudio Keymap** extension in
Positron. It maps the shortcuts you already have to the equivalent commands.
:::

## Setting up Positron

1. **Download** it from
   [positron.posit.co/download](https://positron.posit.co/download.html).
   It is free, and available for Windows, macOS and Linux.
2. **Install R first**, if you have not already (@sec-install). Positron detects
   installed R versions; it does not bundle one.
3. **Open a folder**, not a file: **File → Open Folder**, and pick your project
   directory. Opening a single `.qmd` file gives you an editor with no project
   context, and relative paths will not behave as you expect.
4. **Check the interpreter** in the top right. It shows which R version the
   console is running. Click it to switch.
5. **Install the extensions you want.** The Quarto extension is worth having
   immediately; the RStudio Keymap extension if you are migrating.
6. **Install Quarto** separately if you intend to render documents — Positron
   does not bundle it either.

::: {.callout-warning}
## Packages are per R version, not per IDE
Positron and RStudio share the same R installation and the same library, so
every package you have installed is already available. But if you switch
Positron to a *different* R version, that version has its own library and will
appear to have no packages installed. This is expected behaviour, not a bug —
and it is the reason `renv` exists (@sec-workflow).
:::

## Which should you use?

An honest summary:

**Stay with RStudio if** you are new to R, you develop packages, you rely on the
import wizard or the profiler, or your department has documented workflows built
around it. RStudio is not deprecated and Posit continues to develop it.

**Try Positron if** you also write Python, you work with several R versions, you
already know VS Code, you frequently inspect large datasets, or you want the
built-in AI assistant.

**Use both** — this is entirely reasonable, and Posit says so explicitly. The
files are just files. Nothing about opening a project in one IDE prevents you
from opening it in the other tomorrow.

::: {.callout-note}
## What actually makes your work reproducible
Neither IDE. Reproducibility comes from an R project with a defined root, paths
built with `here()`, a script that runs top to bottom in a clean session,
recorded package versions, and version control. Those are properties of your
code, not of the window it was typed into — which is why this chapter is short
and the rest of the book does not mention the IDE again.
:::

::: {.webex-check .webex-box}
**Check yourself**

Positron uses `.Rproj` files to define a project. `r torf(FALSE)`

Which feature does RStudio have that Positron currently lacks?
`r mcq(c("A console", answer = "A data import wizard", "A plots pane", "Git integration"))`

Positron and RStudio can be used on the same project. `r torf(TRUE)`

If a project has no `.Rproj` and no `.git`, how do you give `here()` an anchor?
`r mcq(c("setwd()", answer = "here::set_here()", "You cannot", "Rename the folder"))`
:::

::: {.callout-note collapse="true"}
## Exercise: try it on a project you already have

If you want to try Positron, do it on a project that already works.

1. Install Positron and open an existing project folder with **Open Folder**.
2. Run `here::here()` in the console. Does it find the right root? If not, why?
3. Load a dataset and click it in the Variables pane. Compare the Data Explorer
   with what `View()` gives you in RStudio.
4. Render a Quarto document. Did it work, or is something missing?
5. Open the command palette and search for "Restart R".

::: {.callout-tip collapse="true"}
## What to expect

**Step 2** is the interesting one. `here()` looks upward from the current
directory for any of a list of marker files. You can see the full list with:

```{r}
#| eval: false
rprojroot::criteria
```

It includes `.here`, `.Rproj`, `.git`, `.vscode/settings.json`, `_quarto.yml`,
`renv.lock` and `DESCRIPTION`, among others. So:

- A project with an `.Rproj` file works in Positron even though Positron itself
  ignores that file — `here` still reads it.
- A project with only a `.git` directory works in both.
- A project with none of these markers falls back to the current working
  directory, which may not be what you want. `here::set_here()` fixes it by
  creating an empty `.here` file, which takes priority over everything else.

**Step 4**: if rendering fails, the usual cause is that Quarto is not installed
system-wide. RStudio bundles a copy; Positron expects you to install it from
[quarto.org](https://quarto.org/docs/get-started/).

Whatever you conclude, notice what did *not* need to change: not one line of
your R code. That is the point of writing code that depends on the project
structure rather than on the editor.
:::
:::

## More information

- [Positron documentation](https://positron.posit.co/)
- [Migrating from RStudio](https://positron.posit.co/migrate-rstudio.html)
- [Comparing RStudio and Positron features](https://positron.posit.co/migrate-rstudio-compare.html)
- [Positron FAQs](https://positron.posit.co/faqs.html)
