# Projects and File Paths {#sec-projects}

Managing working directories and file paths is a crucial aspect of data analysis and project organization in R. This section will cover how to set up and manage projects, the difference between absolute and relative paths, and the benefits of using the `here` package.

## Projects in R

An R project is a way to organize your work into a self-contained, reproducible, and shareable format. Working within a project ensures that your scripts, data, and outputs are well-organized and easy to manage.

### Setting Up a Project

How you create a project depends on your editor, but the *result* is the same: a
folder that acts as the root of your work, with every path inside it written
relative to that root.

::: {.panel-tabset}
## RStudio

Go to **File → New Project** and follow the prompts:

- **New Directory** to create a new folder for the project.
- **Existing Directory** if you already have one.

RStudio writes an `.Rproj` file into the folder. Opening that file sets the
working directory to the folder and restores project-specific settings.

## Positron

Positron has no `.Rproj` equivalent. It uses VS Code's **workspace** model: you
open a *folder*, and that folder is the project.

- **File → Open Folder**, and pick the project directory.
- The working directory is set to that folder.
- Nothing is written into the folder unless you change a project-specific
  setting, in which case Positron creates `.vscode/settings.json`.

**Open the folder, not a file.** Double-clicking a single `.qmd` gives you an
editor with no project context, and relative paths will not resolve the way you
expect.

## Either, from the console

`usethis` works identically in both, and does slightly more than the menus do —
it creates the folder, the `.Rproj` file and a `.gitignore` in one step:

```{r}
#| eval: false
usethis::create_project("~/projects/my_study")
```

To turn a folder you already have into a project:

```{r}
#| eval: false
usethis::use_git() # initialise version control
usethis::use_readme_md() # add a README
```
:::

::: {.callout-tip}
## An `.Rproj` file is not the only project marker
A folder can be a perfectly good project with no `.Rproj` in it. What matters
for your code is that `here()` can find the root, and it recognises several
markers — `.here`, `.Rproj`, `.git`, `.vscode/settings.json`, `_quarto.yml`,
`renv.lock` and `DESCRIPTION` among others.

So a project created in Positron and put under version control works in both
editors without anything extra. If a folder has none of those markers, run
`here::set_here()` once to place an empty `.here` file in the root.

See @sec-positron for the fuller comparison.
:::


### Benefits of Working in Projects

- **Organization**: Keeps all related files and scripts in one place.  
- **Reproducibility**: Ensures that the project can be run on different systems without issues.  
- **Collaboration**: Makes it easier to share and collaborate with others.   
- **Version Control**: Integrates seamlessly with Git for version tracking.  

Check [this](https://bookdown.org/daniel_dauber_io/r4np_book/starting-your-r-projects.html) for a more in depth tutorial of projects  



## Absolute vs. Relative Paths

Understanding the difference between absolute and relative paths is essential for managing file locations in your projects.

### Absolute Path

An absolute path provides the complete directory path from the root of the file system. It specifies the exact location of a file or directory.

Example:
```
# Absolute path
"C:/Users/YourName/Documents/project/data/raw/your_data_file.csv"
```

### Relative Path

A relative path specifies the location of a file or directory relative to the current working directory. It is shorter and more flexible than an absolute path.

Example:
```
# Relative path
"data/raw/your_data_file.csv"
```

## Working Directories

The working directory is the default location where R looks for files to read and where it saves files. You can check your current working directory using the `getwd()` function.

```
# Get current working directory
getwd()
```

Traditionally, setting the working directory in R involved using `setwd()`, which takes an absolute file path. You would then use `getwd()` to verify the current working directory. 

However, using `setwd()` with absolute paths can cause issues. If the directory is moved or shared, the paths can break, making it difficult for others to run the script. This method often limits the script’s usability to the original author’s environment.


```
# Set a new working directory
setwd("path/to/your/directory")
```

the solution is to use the `here` Package:  

## The `here` Package

The `here` package simplifies the management of file paths in your project. It constructs paths relative to the top-level directory of your project, making your code more portable and easier to share.

### Why Use the `here` Package?

- **Consistency**: Avoids issues with setting and resetting the working directory.
- **Portability**: Makes your code easier to share and run on different systems.
- **Clarity**: Improves code readability by using relative paths from the project root.

### Using the `here` Function

First, install and load the `here` package:

```
# Install here package
install.packages("here")

# Load here package
library(here)
```

The `here()` function constructs file paths relative to the project root. For example, to access a data file in the `data/raw/` directory:

```
# Construct a path to a raw data file
data_path <- here("data", "raw", "your_data_file.csv")

# Read the data file
data <- read.csv(data_path)
```

## Recommended Folder Structure

A well-organized folder structure helps keep your project files organized and easily accessible. A typical structure might look like this:

```
project/
├── my_project.Rproj      <- if created in RStudio
├── .vscode/              <- if you changed a setting in Positron
│   └── settings.json
├── .git/                 <- if under version control
├── .gitignore
├── data/
│   ├── raw/
│   ├── processed/
├── scripts/
├── results/
│   ├── figures/
│   ├── tables/
├── docs/
```
Where:  
- `data/`: Contains all data files.  
  - `raw/`: Raw data files.  
  - `processed/`: Processed data files.  
- `scripts/`: Contains all R scripts.  
- `results/`: Contains output files such as figures and tables.  
  - `figures/`: Output figures.  
  - `tables/`: Output tables.  
- `docs/`: Contains documentation files.  


## R File Types

| File Type | Description |
|-----------|-------------|
| `.Rproj`  | R Project file, used by RStudio to manage project settings. Positron ignores it, but `here()` still reads it as a root marker. |
| `.vscode/settings.json` | Workspace settings written by Positron (and VS Code) when you change a project-specific setting. Also a root marker for `here()`. |
| `.R`      | R script file, containing R code for data analysis, visualization, and other tasks. |
| `.Rmd`    | R Markdown file, combining R code and Markdown for dynamic documents, reports, and presentations. |
| `.qmd`    | Quarto Markdown file, similar to R Markdown but used with the Quarto publishing system for reproducible documents. |
| `.Rhistory` | R history file, containing a record of commands that have been executed in the R console. |
| `.RData`  | R data file, used to save and load R objects such as data frames, lists, and other R data structures. |
| `.rds`    | R serialized data file, used to save a single R object to disk and load it back. |

### Version Control

Integrating version control with Git ensures that your project files are tracked and versioned, making collaboration and version management easier. Tools like GitHub provide platforms to host and collaborate on projects.



## Conclusion

Managing working directories and file paths is essential for maintaining an organized and efficient workflow in R. The `here` package offers a robust solution for handling file paths, enhancing the portability and readability of your code. By following recommended practices for project organization and leveraging tools like `here`, you can streamline your data analysis projects.
