# Classes and Objects {#sec-classes}

::: {.callout-note}
## What this chapter covers
This chapter follows the lecture *Klasser och objekt*. You will use R as a
calculator, learn the operators that show up later in data transformation, and
build up the object types that everything else in the book depends on: vectors,
factors, data frames, tibbles and functions.
:::

```{r}
#| label: setup-classes
#| include: false
library(webexercises)
```

## R as an advanced calculator

Before R is a statistics tool it is a calculator. Everything you type in the
console is evaluated and the result is printed. Learning the arithmetic rules
first pays off later: the same operators appear inside `mutate()` when you
derive BMI, z-scores or ratios.

### Arithmetic operators

```{r}
1 + 2 # addition
3 - 1 # subtraction
2 * 3 # multiplication
5 / 2 # division
2^3 # exponent
```

Two operators are less familiar but genuinely useful. Integer division `%/%`
returns the whole-number part, and the modulus `%%` returns the remainder.

```{r}
5 %/% 2 # integer division: how many whole 2s fit in 5?
5 %% 2 # remainder after integer division
```

The modulus is the standard way to test divisibility, for example to find even
numbers or to convert minutes into hours and minutes.

```{r}
# an even number has no remainder when divided by two
c(4, 7, 10) %% 2 == 0
```

### Operator precedence

R follows the usual mathematical order of operations, not the order you typed
things in. When in doubt, add parentheses — they cost nothing and make the
intent explicit for whoever reads the code next.

```{r}
1 / 2 * 3 # same precedence, evaluated left to right
1 + 2 * 3 # * before +
(1 + 2) * 3 # parentheses change the order
2^1 / 2 # ^ before /
2^(1 / 2) # the square root of 2
```

### Comparisons

Comparisons return a *logical* value, `TRUE` or `FALSE`. These are the building
blocks of `filter()`, `if_else()` and `case_when()`.

```{r}
2 > 1 # greater than
2 >= 2 # greater than or equal to
2 < 1 # less than
2 <= 1 # less than or equal to
1 == 1 # equal to (two equals signs!)
1 != 1 # not equal to
```

::: {.callout-warning}
## `=` is not `==`
A single `=` assigns a value; a double `==` asks a question. Writing
`filter(penguins, species = "Gentoo")` is one of the most common beginner errors
in R, and the error message it produces is not especially helpful.
:::

### Logical operators

```{r}
(2 > 1) & (1 > 2) # & is AND: both sides must be TRUE
(2 < 1) | (1 < 2) # | is OR: at least one side must be TRUE
(1 < 1) | !(1 < 1) # ! negates
```

::: {.webex-check .webex-box}
**Check yourself**

What does `7 %% 3` return? `r fitb(1)`

What does `2 + 3 * 4` return? `r fitb(14)`

`(TRUE | FALSE) & FALSE` evaluates to TRUE. `r torf(FALSE)`

Which operator tests whether two values are equal?
`r mcq(c("=", answer = "==", "!=", "<-"))`
:::

## Objects

An expression typed in the console is calculated and then forgotten. To reuse a
result you have to *store* it in an object using the assignment arrow `<-`.

```{r}
x <- 3 * 4
x
```

`=` also works for assignment, but the convention in the R community — and the
one used throughout this book — is `<-`. Keeping `=` reserved for function
arguments makes it obvious at a glance which lines create objects.

Every object has a class, a length, and a structure. Three functions tell you
what you are holding:

```{r}
class(x) # what kind of object is this?
length(x) # how many elements does it contain?
is.vector(x) # a single value is still a vector of length 1
```

::: {.callout-tip}
## The class tells you what is possible
Knowing the class is not trivia. It determines which functions will work. You
cannot take the mean of a character vector, and a statistical model will treat a
factor completely differently from a numeric variable with the same values. When
something behaves unexpectedly, `class()` is usually the first diagnostic.
:::

### Naming objects

R is case sensitive, so `R_code` and `r_code` are two different objects.

```{r}
R_code <- 666
```

```{r}
#| eval: false
r_code
#> Error: object 'r_code' not found
```

Rules and conventions for names:

- No spaces in names.
- A name cannot start with a digit.
- Avoid `!`, `@`, `#`, `$`, `%` and similar characters. Underscore `_` and
  period `.` are allowed.
- Use readable names: `body_mass_kg` beats `bmk` and `x2`.
- This book uses `snake_case` (lower case, words separated by underscores),
  which is the tidyverse convention.
- The decimal separator is always `.`; the comma `,` separates arguments.

::: {.callout-important}
## Swedish characters in names
`å`, `ä` and `ö` are technically allowed in object and column names, but they
cause trouble the moment a file crosses between Windows and macOS, or when a
collaborator opens your script with a different encoding. Use `a` and `o` in
code, and keep the Swedish spelling for labels in figures and tables where it is
actually read by humans.
:::

## Vectors

A vector is the fundamental data structure in R: an ordered collection of
elements **of the same type**. Build one with `c()` (concatenate, though
"combine" is easier to remember).

```{r}
primes <- c(2, 3, 5, 7, 11, 13)
primes
```

### Vectorised arithmetic

Operations apply to every element at once. This is why R rarely needs loops for
data work, and it is exactly what happens inside `mutate()`.

```{r}
primes * 2 # every element is multiplied by 2
primes * primes # element-wise multiplication, position by position
```

### Recycling

If the two vectors have different lengths, R *recycles* the shorter one.

```{r}
primes * c(1, 2)
```

The shorter vector `c(1, 2)` is repeated to match the length of `primes`:
`2 * 1`, `3 * 2`, `5 * 1`, `7 * 2`, and so on. When the longer length is not a
multiple of the shorter one, R still recycles but warns you:

```{r}
#| warning: true
primes + c(1, 2, 3, 4)
```

::: {.callout-warning}
## Recycling silently hides bugs
Recycling is convenient with a single value (`primes * 2`) and dangerous with
anything longer. If you ever see the "longer object length is not a multiple"
warning in a data-cleaning script, treat it as an error: something is not the
length you thought it was.
:::

### Character vectors

Character vectors store text. Each element is a *string* — a sequence of
characters written between quotes.

```{r}
x <- "a new value for x"
x

fruits <- c("Apple", "Banana", "Orange", "Grape")
fruits
```

A vector can only hold one type. If you mix numbers and text, R silently
converts everything to the most flexible type, which is character:

```{r}
y <- c("eighteen", 23, 27)
class(y)
y
```

Note the quotes around `"23"` — it is no longer a number, and `mean(y)` will
fail. This coercion is a frequent source of confusion when a stray text value
such as `"missing"` or `"<0.01"` sits in an otherwise numeric column of an
imported Excel file.

### Sorting

```{r}
sort(fruits)
```

Sorting text is alphabetical, which is rarely what you want for something like
months:

```{r}
months <- c(
  "August", "February", "January", "December",
  "March", "May", "September", "April",
  "November", "June", "October", "July"
)

sort(months)
```

December before February is correct alphabetically and useless analytically.
This is precisely the problem factors solve.

::: {.webex-check .webex-box}
**Check yourself**

What is the class of `c(1, 2, "3")`?
`r mcq(c("numeric", answer = "character", "logical", "factor"))`

What does `c(1, 2, 3, 4) * c(10, 100)` return?
`r mcq(c("10 200 3 4", answer = "10 200 30 400", "an error", "10 200"))`

`length(c("a", "b", "c"))` is `r fitb(3)`
:::

::: {.callout-note collapse="true"}
## Exercise 1: build and inspect vectors

1. Create a numeric vector `heights_cm` with the values 165, 175, 162, 180 and
   168.
2. Create a character vector `sex` with the values "female", "male", "female",
   "male", "female".
3. Convert `heights_cm` to metres in a new object.
4. Check the class and length of all three objects.
5. What happens if you run `mean(sex)`? Read the error message carefully.

::: {.callout-tip collapse="true"}
## Solution

```{r}
heights_cm <- c(165, 175, 162, 180, 168)
sex <- c("female", "male", "female", "male", "female")

# dividing by 100 applies to every element (vectorised)
heights_m <- heights_cm / 100

class(heights_cm)
class(sex)
class(heights_m)
length(heights_cm)
```

```{r}
#| eval: false
mean(sex)
#> Warning: argument is not numeric or logical: returning NA
```

`mean()` has no meaningful definition for text, so R warns and returns `NA`
rather than guessing.
:::
:::

## Factors

A **factor** is R's data structure for categorical data. It stores a limited set
of unique values, called **levels**, and remembers the order you gave them.

Factors matter for three reasons:

- **Statistical analysis**: models treat factors as categories, and the first
  level becomes the reference (comparison) group.
- **Order**: you can define a meaningful order such as low < medium < high.
- **Storage**: levels are stored once and the data as integer codes.

### Unordered factors

Categories with no inherent ranking — colours, cities, treatment arms.

```{r}
colours <- factor(c("Red", "Green", "Blue", "Green", "Red"))
colours

cities <- factor(c("Stockholm", "Gothenburg", "Malmo", "Gothenburg"))
cities
```

Notice the `Levels:` line in the output. By default R sorts levels
alphabetically.

### Ordered factors

Categories with a logical order — education level, dose, pain rating.

```{r}
factor(
  c("Upper secondary", "Bachelor", "Master"),
  ordered = TRUE
)
```

### Fixing the month problem

Give `factor()` an explicit `levels` argument and sorting follows *your* order
rather than the alphabet.

```{r}
months_factor <- factor(
  months,
  levels = c(
    "January", "February", "March", "April",
    "May", "June", "July", "August",
    "September", "October", "November", "December"
  )
)

sort(months_factor)
```

Adding `ordered = TRUE` additionally allows comparisons between levels:

```{r}
months_ordered <- factor(
  months,
  levels = c(
    "January", "February", "March", "April",
    "May", "June", "July", "August",
    "September", "October", "November", "December"
  ),
  ordered = TRUE
)

months_ordered[1] > months_ordered[3] # August is later in the year than January
```

### When does it matter?

- Ordered factors can be compared with `<` and `>`.
- Sorting follows the level order for both ordered and unordered factors.
- In models, the level order decides the reference group.
- Contrasts are constructed differently for ordered and unordered factors.
- For the ordering of groups in a figure, either type works — the level order is
  what `ggplot2` reads.

::: {.callout-tip}
## `forcats` does this more comfortably
Writing out twelve levels by hand is tedious and error prone. The `forcats`
package (part of the tidyverse) gives you `fct_relevel()`, `fct_infreq()`,
`fct_reorder()` and friends. See @sec-strings-factors.
:::

::: {.callout-note collapse="true"}
## Exercise 2: pain ratings

A small study records pain as "None", "Mild", "Moderate" and "Severe".

```{r}
pain <- c("Moderate", "None", "Severe", "Mild", "Moderate", "None")
```

1. Turn `pain` into an unordered factor and run `sort()`. Is the order sensible?
2. Turn it into an ordered factor with the levels in clinical order.
3. Use `table()` to count how many observations fall in each level. What happens
   to levels with zero observations?

::: {.callout-tip collapse="true"}
## Solution

```{r}
# alphabetical order is clinically meaningless here
sort(factor(pain))

# an explicit level order encodes the clinical severity
pain_ordered <- factor(
  pain,
  levels = c("None", "Mild", "Moderate", "Severe"),
  ordered = TRUE
)

sort(pain_ordered)
table(pain_ordered)
```

`table()` reports every level, including empty ones. That is a feature: an empty
category is information, and it keeps tables comparable across subgroups.
:::
:::

## Rectangular data: data frames and tibbles

A data frame is a set of equal-length vectors placed side by side as columns. It
is the R equivalent of a spreadsheet, with one important difference: each column
has a type, and the type is enforced.

```{r}
set.seed(123) # a seed makes the random draw reproducible

name <- sample(
  c("Anna", "Bjorn", "Cecilia", "David", "Elin", "Fredrik"),
  30,
  replace = TRUE
)
age <- sample(20:40, 30, replace = TRUE)
city <- sample(
  c("Stockholm", "Gothenburg", "Malmo", "Uppsala", "Lund"),
  30,
  replace = TRUE
)
```

From these vectors we can build a data frame:

```{r}
df <- data.frame(name = name, age = age, city = city)
head(df)
```

Or a tibble, the modern tidyverse version:

```{r}
library(tibble)

tb <- tibble(name = name, age = age, city = city)
tb
```

Differences worth knowing:

- A data frame prints everything; a tibble prints the first ten rows plus the
  dimensions, so you do not flood the console by accident.
- A tibble shows the type of each column (`<chr>`, `<int>`, `<dbl>`, `<fct>`).
- A tibble allows column names with spaces and special characters. Avoid these
  in analysis code, but they are useful in the final table or figure.
- Subsetting a tibble with `[` always returns a tibble; a data frame sometimes
  drops to a vector.

::: {.callout-tip}
## Reproducibility: always set a seed
Any time your code involves randomness — `sample()`, `rnorm()`, bootstrapping,
cross-validation — call `set.seed()` first and note the value. Without it, your
results change every time the script runs, and no one, including you, can
reproduce the numbers in your manuscript.
:::

## Functions

Functions are objects too, and they are the ones you will use most. A function
is called by name followed by parentheses, with any arguments inside.

```{r}
mean(c(10, 20, 30))
sum(c(10, 20, 30))
length(c(10, 20, 30))
sort(c(30, 10, 20))
seq(1, 10, by = 2)
rep(5, times = 3)
round(3.14159, digits = 2)
```

### Arguments

Arguments are the inputs a function needs. Some are required, some are optional
and fall back to a default value.

`mean()` takes among others:

- `x`: a numeric vector (required)
- `na.rm`: whether missing values should be dropped (optional, defaults to
  `FALSE`)

```{r}
a_vector <- c(10, 20, 30, NA)

mean(a_vector) # NA propagates: the answer is unknown
mean(a_vector, na.rm = TRUE) # drop the NA and average the rest
```

::: {.callout-important}
## `NA` is contagious on purpose
`mean(a_vector)` returning `NA` is not a bug. R is telling you that it cannot
know the average without knowing the missing value. `na.rm = TRUE` is you making
the decision to ignore it — which is a real analytical choice with real
consequences, so make it deliberately and say so in a comment.
:::

Two functions help you inspect any function you meet:

```{r}
args(mean) # which arguments does it take?
```

```{r}
#| eval: false
?mean # the full help page, with examples at the bottom
```

### Arguments by name or by position

If you supply arguments in the documented order you can omit the names.
Otherwise, name them.

```{r}
substr(x = fruits, start = 1, stop = 3) # explicit and readable
substr(fruits, 1, 3) # same call, relying on position
```

Naming is safer. Position works fine for the first, obvious argument and gets
risky after that.

### Functions on text

```{r}
toupper(fruits)
nchar(fruits) # number of characters in each string
```

### Changing the class of a vector

The `as.*()` family converts between types: `as.character()`, `as.numeric()`,
`as.integer()`, `as.logical()`, `as.factor()`.

```{r}
numeric_vector <- c(1, 2, 3, 4, 5)

text_vector <- as.character(numeric_vector)
text_vector

as.numeric(text_vector) # back again
as.factor(text_vector)
```

::: {.callout-warning}
## The factor-to-numeric trap
Converting a factor straight to numeric returns the *level codes*, not the
values. This silently corrupts data and is one of the classic R bugs.

```{r}
f <- factor(c("10", "20", "30"))

as.numeric(f) # WRONG: 1, 2, 3 - these are level positions
as.numeric(as.character(f)) # RIGHT: go via character first
```
:::

::: {.webex-check .webex-box}
**Check yourself**

`mean(c(1, 2, NA))` returns
`r mcq(c("1.5", answer = "NA", "3", "an error"))`

To convert a factor of numbers to real numbers you should use
`r mcq(c("as.numeric(f)", answer = "as.numeric(as.character(f))", "as.integer(f)", "round(f)"))`

A tibble prints all rows by default. `r torf(FALSE)`
:::

## Base R subsetting

The tidyverse gives you `filter()` and `select()`, but base R subsetting still
appears constantly — in other people's code, in error messages, and in places
where the tidyverse has no equivalent.

### The dollar sign `$`

Extracts one column from a data frame or tibble, as a vector.

```{r}
data <- data.frame(A = c(1, 2, 3), B = c("X", "Y", "Z"))
data

data$A
```

### Single brackets `[ ]`

Preserve the structure. For a data frame the syntax is `[rows, columns]`.

```{r}
data[1, ] # first row, all columns
data[, 1] # all rows, first column
data[1, 1] # a single cell
```

### Double brackets `[[ ]]`

Extract the contents of an element, dropping the surrounding structure.

```{r}
list_data <- list(A = 1:3, B = c("X", "Y", "Z"))

list_data[1] # a list containing A
list_data[[1]] # the vector inside A
```

The same distinction holds for data frames:

```{r}
data["A"] # a data frame with one column
data[["A"]] # a plain vector
```

::: {.callout-tip}
## A mnemonic
Think of a list as a train of wagons. `[` gives you a shorter train; `[[` gives
you the cargo inside one wagon.
:::

::: {.callout-note collapse="true"}
## Exercise 3: putting it together

Using the `penguins` dataset from `palmerpenguins`:

1. Load the package and the data.
2. What class is the `species` column? What are its levels?
3. Use `$` to pull out `body_mass_g` and calculate its mean, ignoring missing
   values.
4. Use `[ , ]` to display the first three rows and the first two columns.
5. How many penguins are missing a value for `sex`?

::: {.callout-tip collapse="true"}
## Solution

```{r}
library(palmerpenguins)

class(penguins$species)
levels(penguins$species)

# na.rm = TRUE is needed because two penguins lack a body mass measurement
mean(penguins$body_mass_g, na.rm = TRUE)

penguins[1:3, 1:2]

sum(is.na(penguins$sex))
```

`is.na()` returns a logical vector, and `sum()` counts `TRUE` values as 1. This
is the standard idiom for counting missing values.
:::
:::

::: {.callout-note collapse="true"}
## Exercise 4: think before you run

Predict the output of each line, then run it and check.

```{r}
#| eval: false
c(1, 2, 3) + c(10, 20, 30)
c(1, 2, 3) + 10
c(TRUE, FALSE, TRUE) + 0
class(TRUE)
length(NA)
c("a", "b") == c("a", "c")
```

::: {.callout-tip collapse="true"}
## Solution

```{r}
c(1, 2, 3) + c(10, 20, 30) # element-wise
c(1, 2, 3) + 10 # 10 is recycled
c(TRUE, FALSE, TRUE) + 0 # logicals coerce to 1/0 in arithmetic
class(TRUE)
length(NA) # NA is a length-1 vector, not "nothing"
c("a", "b") == c("a", "c") # comparison is also element-wise
```

The third line is worth remembering: because `TRUE` counts as 1, `sum()` and
`mean()` on a logical vector give you a count and a proportion respectively.
:::
:::
