# Quarto {#sec-quarto}

::: {.callout-note}
## What this chapter covers
This chapter follows the lecture *Quarto*. It covers what a Quarto document is
made of, the chunk options you will actually use, how to produce a
self-contained report that someone else can open, and how the same document
becomes a Word file, a PDF, slides or a website. It was written for Quarto
1.10.
:::

```{r}
#| label: setup-quarto
#| include: false
library(webexercises)
```

## Reproducible research

Using Quarto for your analyses puts code, results and documentation in a single
document. The idea goes back to Donald Knuth's *literate programming*
[@knuth1984]: a program written primarily to be read by humans, with the code
embedded in the explanation rather than the other way round. The numbers in
your text are computed from your data every time the document is rendered, so they cannot drift out of date. This is the practical
core of reproducibility: not that you *could* reproduce the analysis, but that
you do so automatically on every render.

::: {.callout-tip}
## Set the output location first
When you run a chunk while writing, where does the output go?

- **RStudio** shows it inline, underneath each chunk, by default. **Tools →
  Global Options → R Markdown → Show output inline** switches it to the console
  and the plot pane instead.
- **Positron** sends it to the console and the Plots pane by default. Turn on
  the setting `quarto.inlineOutput.enabled` to show it underneath each cell
  instead. With inline output on, each `.qmd` file gets its own R session, which
  keeps documents from accidentally sharing objects.

Neither is right. Inline output is good for a document you are writing; the
console and plot pane are better when you are debugging and want a large plot.
Try both and pick one.
:::

{{< video https://youtu.be/hgpL-sppw7E start="10" >}}

[Quarto.org](https://quarto.org) is the go-to place for full documentation and
tutorials.

## The three parts of a Quarto document

1. **YAML header** — metadata about the document
2. **Markdown content** — the text
3. **Code chunks** — the code that produces results

### 1. YAML header

The YAML header sits at the top, enclosed in `---` lines.

``` yaml
---
title: "Introduction to Quarto"
author: "Your Name"
date: "2026-09-08"
format: html
---
```

For a file you can email to someone, everything needs to be embedded in the
single HTML file:

``` yaml
---
title: "Self-contained report"
author: "Your Name"
format:
  html:
    embed-resources: true
---
```

A more complete header for a report:

``` yaml
---
title: "Physical activity and cardiorespiratory fitness"
subtitle: "Preliminary analysis"
author:
  name: "Your Name"
  affiliation: "GIH"
date: today
format:
  html:
    toc: true              # table of contents
    toc-depth: 3
    number-sections: true
    code-fold: true        # readers can hide the code
    code-tools: true       # a button to view the source
    embed-resources: true
    theme: cosmo
    df-print: kable
execute:
  echo: true               # show code
  warning: false           # hide warnings
  message: false           # hide messages
  cache: false
bibliography: references.bib
---
```

::: {.callout-warning}
## YAML is fussy about indentation
YAML uses spaces, never tabs, and indentation is meaningful. `format: html` and
`format:` followed by an indented `html:` mean different things. If rendering
fails with an unhelpful error, check the header first — it is the most common
cause.
:::

### 2. Markdown content

Markdown is a lightweight way of formatting plain text.

``` markdown
# A level 1 heading
## A level 2 heading

Plain text, with *italics*, **bold**, and `inline code`.

- A bullet
- Another bullet

1. A numbered item
2. Another one

[A link](https://quarto.org)

> A block quote.
```

::: {.callout-important}
## `#` means two different things
In **Markdown text**, `#` at the start of a line creates a heading.
In **R code**, `#` starts a comment.

Same character, completely different meaning depending on whether you are
inside a code chunk. This confuses nearly everyone at least once.
:::

### 3. Code chunks

Code chunks are enclosed in triple backticks with the language in braces. To
insert one, press <kbd>Ctrl/Cmd</kbd>+<kbd>Alt</kbd>+<kbd>I</kbd> in RStudio;
in Positron, run **Quarto: Insert Code Cell** from the command palette, which
also shows its shortcut.

```{{r}}
summary(cars)
```

Chunk options go on lines starting with `#|` at the top of the chunk:

```{{r}}
#| label: fig-example
#| fig-cap: "A caption for the figure."
#| echo: false
#| warning: false
plot(cars)
```

### Chunk options worth knowing

| Option | Effect |
|--------|--------|
| `echo: false` | Run the code, show the output, hide the code |
| `eval: false` | Show the code, do not run it |
| `include: false` | Run the code, show nothing (for setup chunks) |
| `warning: false` / `message: false` | Hide warnings / messages |
| `fig-cap:` | Caption the figure |
| `fig-width:` / `fig-height:` | Size in inches |
| `label:` | Name the chunk, so errors point somewhere useful |
| `cache: true` | Reuse the result unless the code changes |

::: {.callout-tip}
## Label your chunks
An unlabelled chunk that fails reports an error in "chunk 14". A labelled one
reports an error in `clean-outliers`. Labels beginning `fig-` also enable
cross-referencing.
:::

## Inline code

The single most useful Quarto feature for a report. Instead of typing a number
into your text, compute it:

``` markdown
The dataset contains `r knitr::inline_expr("nrow(penguins)")` observations.
```

```{r}
#| label: inline-demo
#| message: false
library(palmerpenguins)
```

The dataset contains `r nrow(penguins)` observations, of which
`r sum(is.na(penguins$sex))` are missing a recorded sex.

Change the data, re-render, and the sentence updates itself. A number typed by
hand does not, and sooner or later your text and your table will disagree.

## Callouts

Quarto has five callout types, useful for structuring a report:

::: {.callout-note}
## Note
Background information.
:::

::: {.callout-tip}
## Tip
A recommendation.
:::

::: {.callout-warning}
## Warning
Something that could go wrong.
:::

::: {.callout-important}
## Important
Something that must not be missed.
:::

::: {.callout-caution collapse="true"}
## Caution (collapsed by default)
Written as `::: {.callout-caution collapse="true"}`. The reader clicks to open
it — which is how the exercises in this book hide their solutions.
:::

## Cross-references

Label a figure or table with a `fig-` or `tbl-` prefix and refer to it with `@`:

````
```{{r}}
#| label: fig-mass
#| fig-cap: "Body mass by species."
...
```

As shown in @fig-mass, Gentoo penguins are heaviest.
````

The number is generated automatically and stays correct when you reorder your
document.

## Rendering

- Click **Render** in RStudio, or **Preview** in Positron, or press
  <kbd>Ctrl/Cmd</kbd>+<kbd>Shift</kbd>+<kbd>K</kbd> in either. Both editors
  bundle their own copy of Quarto, so there is nothing extra to install.
- Or run `quarto::quarto_render("my_document.qmd")` from the console.
- Or run `quarto render my_document.qmd` in the terminal; `quarto preview`
  re-renders automatically every time you save.

Rendering runs the document in a **fresh R session**. This is the single most
useful thing about it: anything that works in your console but not on render is
relying on an object you created by hand and forgot about. If it renders, it is
reproducible.

::: {.callout-important}
## Render often
If you only render at the end, you get all of your errors at once, in a document
that has grown large enough that finding them is hard. Render after every
substantial addition.
:::

## One document, many formats

The format in the YAML header decides what comes out. The same text and code
can produce an HTML report, a Word document for co-authors who want to use
track changes, a PDF for a journal, or slides.

``` yaml
format:
  html:
    embed-resources: true
  docx:
    reference-doc: gih-template.docx   # fonts, margins and styles from a Word file
  typst: default                       # PDF, without installing LaTeX
```

With several formats listed, **Render** produces the first one; to get them
all, run `quarto render report.qmd` in the terminal, or
`quarto::quarto_render("report.qmd", output_format = "all")`.

| Format | Produces | Notes |
|--------|----------|-------|
| `html` | A web page | The default. Add `embed-resources: true` for one file to email. |
| `docx` | Word | `reference-doc` copies the styles of a Word file you provide, such as a department or journal template. |
| `typst` | PDF | Fast and needs no LaTeX. The easiest way to a PDF today. |
| `pdf` | PDF via LaTeX | Needed by some journal templates. Install LaTeX once with `quarto install tinytex`. |
| `revealjs` | HTML slides | The course lecture slides are made this way. |

Word output is often what collaborators actually need. Write and render in
Quarto, send the `.docx`, and bring their comments back into the `.qmd` by hand,
so the numbers in the text are still computed rather than typed.

## Citations

Put references in a BibTeX file, point to it in the header, and cite with `@`:

``` yaml
bibliography: references.bib
csl: vancouver.csl        # optional: a citation style, from zotero.org/styles
```

``` markdown
Physical activity is associated with lower mortality [@ekelund2019].
@ekelund2019 found a dose-response relationship.
```

Quarto formats the citations and adds a reference list at the end. You do not
have to write BibTeX by hand. In RStudio's **visual editor** (the
**Visual** button above the document), **Insert → Citation** looks up a DOI or
searches your Zotero library and adds the entry to `references.bib` for you.
Zotero's Better BibTeX extension can also keep a `.bib` file up to date
automatically. This book's own reference list is built the same way.

## Parameters: one report, many versions

A **parameter** is a value in the header that the document uses like a
variable. It lets you render the same report for each clinic, region or year
without copying the file.

``` yaml
---
title: "Report for `r knitr::inline_expr("params$region")`"
params:
  region: "Stockholm"
---
```

Inside the document, `params$region` holds the value:

```{{r}}
region_data <- visits |>
  filter(region == params$region)
```

Render one version per region from R:

```{r}
#| eval: false
regions <- c("Stockholm", "Göteborg", "Malmö")

walk(regions, \(r) {
  quarto::quarto_render(
    "region_report.qmd",
    execute_params = list(region = r),
    output_file = str_glue("report_{r}.html")
  )
})
```

`walk()` is `map()` for functions you call for their side effect, here writing
files (@sec-functions).

## Projects, books and manuscripts

A single `.qmd` file is a document. Add a `_quarto.yml` file to a folder and it
becomes a **Quarto project**, with shared settings and the option to render
everything at once with `quarto render`. The project type decides what comes
out:

- `type: website` — a set of linked pages, like the course slide site;
- `type: book` — chapters with numbering and cross-references between them,
  like this book;
- `type: manuscript` — a scientific article, with its notebooks, rendered to
  HTML, Word and PDF together, ready to share alongside a submission.

A website or book can be published to GitHub Pages or Quarto Pub with
`quarto publish`. This book is rendered and published automatically by GitHub
every time a change is pushed (@sec-git).

::: {.callout-warning}
## A rendered report contains your results
Output is part of the document: every printed table, `head()` and summary ends
up in the HTML or Word file. A report on registry data is as sensitive as the
data it shows until you have checked that it contains only aggregates, and small
counts may need to be suppressed. Never publish it to a public site to "just
share it with a colleague".
:::

## Example Quarto sites

- [Champions Lessons Series](https://openscapes.github.io/series)
- [Openscapes Approach Guide](https://openscapes.github.io/approach-guide/)
- [NASA Earthdata Cloud Cookbook](https://nasa-openscapes.github.io/earthdata-cloud-cookbook/)
- [Faylab Lab Manual](https://thefaylab.github.io/lab-manual/)
- [A Quarto tip a day](https://mine-cetinkaya-rundel.github.io/quarto-tip-a-day/)

::: {.webex-check .webex-box}
**Check yourself**

Which chunk option runs the code but hides it from the reader?
`r mcq(c("eval: false", answer = "echo: false", "include: false", "output: false"))`

Rendering a Quarto document runs it in your current R session.
`r torf(FALSE)`

To make a single HTML file you can email, you add
`r mcq(c("self-contained-html: true", answer = "embed-resources: true", "standalone: yes", "format: file"))`

Which format gives you a PDF without installing LaTeX?
`r mcq(c("pdf", answer = "typst", "docx", "revealjs"))`

To render one report per region from the same `.qmd` file, you use
`r mcq(c("a for loop that copies the file", answer = "parameters", "cross-references", "callouts"))`

In a code chunk, `# something` is
`r mcq(c("a heading", answer = "a comment", "a chunk option", "an error"))`
:::

::: {.callout-note collapse="true"}
## Exercise 1: build a small report

Create a new Quarto document and make it into a short report on the `penguins`
data.

1. Write a YAML header with a title, your name, a table of contents, and
   `embed-resources: true`.
2. Add a setup chunk with `include: false` that loads `tidyverse` and
   `palmerpenguins`.
3. Write a sentence that uses inline R code to report the number of penguins and
   the number of species.
4. Add a labelled, captioned figure and refer to it from the text with `@`.
5. Add a table of mean body mass per species.
6. Render it and check that the figure number in the text is correct.

::: {.callout-tip collapse="true"}
## Solution

````markdown
---
title: "The Palmer penguins"
author: "Your Name"
date: today
format:
  html:
    toc: true
    embed-resources: true
execute:
  warning: false
  message: false
---

```{{r}}
#| label: setup
#| include: false
library(tidyverse)
library(palmerpenguins)
theme_set(theme_minimal(base_size = 13))
```

## Introduction

The dataset contains `r knitr::inline_expr("nrow(penguins)")` penguins from
`r knitr::inline_expr("n_distinct(penguins$species)")` species, measured on
`r knitr::inline_expr("n_distinct(penguins$island)")` islands.

## Body mass

```{{r}}
#| label: fig-mass
#| fig-cap: "Body mass by species."
penguins |>
  filter(!is.na(body_mass_g)) |>
  ggplot(aes(species, body_mass_g, fill = species)) +
  geom_boxplot(show.legend = FALSE) +
  labs(x = NULL, y = "Body mass (g)")
```

As @fig-mass shows, Gentoo penguins are considerably heavier than the other
two species.

```{{r}}
#| label: tbl-mass
#| tbl-cap: "Mean body mass by species."
penguins |>
  summarise(
    n = sum(!is.na(body_mass_g)),
    mean_mass = round(mean(body_mass_g, na.rm = TRUE), 0),
    .by = species
  ) |>
  knitr::kable(col.names = c("Species", "n", "Mean body mass (g)"))
```
````

Two things to notice when you render it. First, the figure number in the text is
generated, not typed — reorder the document and it follows. Second, if you
change `filter()` to something else, both the figure and the sentence above it
update together. That coupling between text and data is the whole point.
:::
:::

::: {.callout-note collapse="true"}
## Exercise 2: make it render

A colleague's document fails to render. Their setup looks like this:

````markdown
---
title: My Analysis
author: Anna
format:
html:
  toc: true
---

```{{r}}
library(tidyverse)
d <- read_csv("C:/Users/anna/Desktop/data.csv")
summary(d)
```
````

Identify at least three problems.

::: {.callout-tip collapse="true"}
## Solution

1. **The YAML indentation is wrong.** `html:` must be indented under `format:`.
   As written, `format:` is empty and `html:` is a separate top-level key.

2. **The file path is absolute** and points at one specific computer. Nobody
   else can render this, and neither can Anna after she gets a new laptop. Use
   an R project and `here("data", "data.csv")`.

3. **The chunk is unlabelled**, so any error reports an unhelpful chunk number.

Two more worth mentioning:

4. **`library(tidyverse)` prints attaching messages** into the report. Add
   `#| message: false`, or set it once in the YAML `execute:` block.

5. **There is no `embed-resources: true`**, so the rendered HTML depends on a
   folder of supporting files. Emailing just the `.html` gives the recipient an
   unstyled page with no figures.

```` markdown
---
title: "My Analysis"
author: "Anna"
format:
  html:
    toc: true
    embed-resources: true
execute:
  message: false
  warning: false
---

```{{r}}
#| label: import-data
library(tidyverse)
library(here)

d <- read_csv(here("data", "data.csv"))
glimpse(d)
```
````
:::
:::
