# Quarto {#sec-quarto}

::: {.callout-note}
## What this chapter covers
This chapter follows the lecture *Quarto*. It covers what a Quarto document is
made of, the chunk options you will actually use, and how to produce a
self-contained report that someone else can open.
:::

```{r}
#| label: setup-quarto
#| include: false
library(webexercises)
```

## Reproducible research

Using Quarto for your analyses puts code, results and documentation in a single
document. The idea goes back to Donald Knuth's *literate programming*
[@knuth1984]: a program written primarily to be read by humans, with the code
embedded in the explanation rather than the other way round. The numbers in your text are computed from your data every time the
document is rendered, so they cannot drift out of date. This is the practical
core of reproducibility: not that you *could* reproduce the analysis, but that
you do so automatically on every render.

::: {.callout-tip}
## Set the output location first
By default, plots and printed data frames appear inline underneath each code
chunk. You can change this in **Tools → Global Options → R Markdown** so that
output goes to the console and the plot pane instead.

Neither is right. Inline output is good for a document you are writing; the
console and plot pane are better when you are debugging and want a large plot.
Try both and pick one.
:::

{{< video https://youtu.be/hgpL-sppw7E start="10" >}}

[Quarto.org](https://quarto.org) is the go-to place for full documentation and
tutorials.

## The three parts of a Quarto document

1. **YAML header** — metadata about the document
2. **Markdown content** — the text
3. **Code chunks** — the code that produces results

### 1. YAML header

The YAML header sits at the top, enclosed in `---` lines.

``` yaml
---
title: "Introduction to Quarto"
author: "Your Name"
date: "2026-09-08"
format: html
---
```

For a file you can email to someone, everything needs to be embedded in the
single HTML file:

``` yaml
---
title: "Self-contained report"
author: "Your Name"
format:
  html:
    embed-resources: true
---
```

A more complete header for a report:

``` yaml
---
title: "Physical activity and cardiorespiratory fitness"
subtitle: "Preliminary analysis"
author:
  name: "Your Name"
  affiliation: "GIH"
date: today
format:
  html:
    toc: true              # table of contents
    toc-depth: 3
    number-sections: true
    code-fold: true        # readers can hide the code
    code-tools: true       # a button to view the source
    embed-resources: true
    theme: cosmo
    df-print: kable
execute:
  echo: true               # show code
  warning: false           # hide warnings
  message: false           # hide messages
  cache: false
bibliography: references.bib
---
```

::: {.callout-warning}
## YAML is fussy about indentation
YAML uses spaces, never tabs, and indentation is meaningful. `format: html` and
`format:` followed by an indented `html:` mean different things. If rendering
fails with an unhelpful error, check the header first — it is the most common
cause.
:::

### 2. Markdown content

Markdown is a lightweight way of formatting plain text.

``` markdown
# A level 1 heading
## A level 2 heading

Plain text, with *italics*, **bold**, and `inline code`.

- A bullet
- Another bullet

1. A numbered item
2. Another one

[A link](https://quarto.org)

> A block quote.
```

::: {.callout-important}
## `#` means two different things
In **Markdown text**, `#` at the start of a line creates a heading.
In **R code**, `#` starts a comment.

Same character, completely different meaning depending on whether you are
inside a code chunk. This confuses nearly everyone at least once.
:::

### 3. Code chunks

Code chunks are enclosed in triple backticks with the language in braces. The
keyboard shortcut to insert one is `Ctrl/Cmd + Alt + I`.

```{{r}}
summary(cars)
```

Chunk options go on lines starting with `#|` at the top of the chunk:

```{{r}}
#| label: fig-example
#| fig-cap: "A caption for the figure."
#| echo: false
#| warning: false
plot(cars)
```

### Chunk options worth knowing

| Option | Effect |
|--------|--------|
| `echo: false` | Run the code, show the output, hide the code |
| `eval: false` | Show the code, do not run it |
| `include: false` | Run the code, show nothing (for setup chunks) |
| `warning: false` / `message: false` | Hide warnings / messages |
| `fig-cap:` | Caption the figure |
| `fig-width:` / `fig-height:` | Size in inches |
| `label:` | Name the chunk, so errors point somewhere useful |
| `cache: true` | Reuse the result unless the code changes |

::: {.callout-tip}
## Label your chunks
An unlabelled chunk that fails reports an error in "chunk 14". A labelled one
reports an error in `clean-outliers`. Labels beginning `fig-` also enable
cross-referencing.
:::

## Inline code

The single most useful Quarto feature for a report. Instead of typing a number
into your text, compute it:

``` markdown
The dataset contains `r knitr::inline_expr("nrow(penguins)")` observations.
```

```{r}
#| label: inline-demo
#| message: false
library(palmerpenguins)
```

The dataset contains `r nrow(penguins)` observations, of which
`r sum(is.na(penguins$sex))` are missing a recorded sex.

Change the data, re-render, and the sentence updates itself. A number typed by
hand does not, and sooner or later your text and your table will disagree.

## Callouts

Quarto has five callout types, useful for structuring a report:

::: {.callout-note}
## Note
Background information.
:::

::: {.callout-tip}
## Tip
A recommendation.
:::

::: {.callout-warning}
## Warning
Something that could go wrong.
:::

::: {.callout-important}
## Important
Something that must not be missed.
:::

::: {.callout-caution collapse="true"}
## Caution (collapsed by default)
Written as `::: {.callout-caution collapse="true"}`. The reader clicks to open
it — which is how the exercises in this book hide their solutions.
:::

## Cross-references

Label a figure or table with a `fig-` or `tbl-` prefix and refer to it with `@`:

````
```{{r}}
#| label: fig-mass
#| fig-cap: "Body mass by species."
...
```

As shown in @fig-mass, Gentoo penguins are heaviest.
````

The number is generated automatically and stays correct when you reorder your
document.

## Rendering

- Click **Render** in RStudio or Positron, or press `Ctrl/Cmd + Shift + K`.
  Positron needs the Quarto CLI installed separately (@sec-positron).
- Or run `quarto::quarto_render("my_document.qmd")`.

Rendering runs the document in a **fresh R session**. This is the single most
useful thing about it: anything that works in your console but not on render is
relying on an object you created by hand and forgot about. If it renders, it is
reproducible.

::: {.callout-important}
## Render often
If you only render at the end, you get all of your errors at once, in a document
that has grown large enough that finding them is hard. Render after every
substantial addition.
:::

## Example Quarto sites

- [Champions Lessons Series](https://openscapes.github.io/series)
- [Openscapes Approach Guide](https://openscapes.github.io/approach-guide/)
- [NASA Earthdata Cloud Cookbook](https://nasa-openscapes.github.io/earthdata-cloud-cookbook/)
- [Faylab Lab Manual](https://thefaylab.github.io/lab-manual/)
- [A Quarto tip a day](https://mine-cetinkaya-rundel.github.io/quarto-tip-a-day/)

::: {.webex-check .webex-box}
**Check yourself**

Which chunk option runs the code but hides it from the reader?
`r mcq(c("eval: false", answer = "echo: false", "include: false", "output: false"))`

Rendering a Quarto document runs it in your current R session.
`r torf(FALSE)`

To make a single HTML file you can email, you add
`r mcq(c("self-contained-html: true", answer = "embed-resources: true", "standalone: yes", "format: file"))`

In a code chunk, `# something` is
`r mcq(c("a heading", answer = "a comment", "a chunk option", "an error"))`
:::

::: {.callout-note collapse="true"}
## Exercise 1: build a small report

Create a new Quarto document and make it into a short report on the `penguins`
data.

1. Write a YAML header with a title, your name, a table of contents, and
   `embed-resources: true`.
2. Add a setup chunk with `include: false` that loads `tidyverse` and
   `palmerpenguins`.
3. Write a sentence that uses inline R code to report the number of penguins and
   the number of species.
4. Add a labelled, captioned figure and refer to it from the text with `@`.
5. Add a table of mean body mass per species.
6. Render it and check that the figure number in the text is correct.

::: {.callout-tip collapse="true"}
## Solution

````markdown
---
title: "The Palmer penguins"
author: "Your Name"
date: today
format:
  html:
    toc: true
    embed-resources: true
execute:
  warning: false
  message: false
---

```{{r}}
#| label: setup
#| include: false
library(tidyverse)
library(palmerpenguins)
theme_set(theme_minimal(base_size = 13))
```

## Introduction

The dataset contains `r knitr::inline_expr("nrow(penguins)")` penguins from
`r knitr::inline_expr("n_distinct(penguins$species)")` species, measured on
`r knitr::inline_expr("n_distinct(penguins$island)")` islands.

## Body mass

```{{r}}
#| label: fig-mass
#| fig-cap: "Body mass by species."
penguins |>
  filter(!is.na(body_mass_g)) |>
  ggplot(aes(species, body_mass_g, fill = species)) +
  geom_boxplot(show.legend = FALSE) +
  labs(x = NULL, y = "Body mass (g)")
```

As @fig-mass shows, Gentoo penguins are considerably heavier than the other
two species.

```{{r}}
#| label: tbl-mass
#| tbl-cap: "Mean body mass by species."
penguins |>
  summarise(
    n = sum(!is.na(body_mass_g)),
    mean_mass = round(mean(body_mass_g, na.rm = TRUE), 0),
    .by = species
  ) |>
  knitr::kable(col.names = c("Species", "n", "Mean body mass (g)"))
```
````

Two things to notice when you render it. First, the figure number in the text is
generated, not typed — reorder the document and it follows. Second, if you
change `filter()` to something else, both the figure and the sentence above it
update together. That coupling between text and data is the whole point.
:::
:::

::: {.callout-note collapse="true"}
## Exercise 2: make it render

A colleague's document fails to render. Their setup looks like this:

````markdown
---
title: My Analysis
author: Anna
format:
html:
  toc: true
---

```{{r}}
library(tidyverse)
d <- read_csv("C:/Users/anna/Desktop/data.csv")
summary(d)
```
````

Identify at least three problems.

::: {.callout-tip collapse="true"}
## Solution

1. **The YAML indentation is wrong.** `html:` must be indented under `format:`.
   As written, `format:` is empty and `html:` is a separate top-level key.

2. **The file path is absolute** and points at one specific computer. Nobody
   else can render this, and neither can Anna after she gets a new laptop. Use
   an R project and `here("data", "data.csv")`.

3. **The chunk is unlabelled**, so any error reports an unhelpful chunk number.

Two more worth mentioning:

4. **`library(tidyverse)` prints attaching messages** into the report. Add
   `#| message: false`, or set it once in the YAML `execute:` block.

5. **There is no `embed-resources: true`**, so the rendered HTML depends on a
   folder of supporting files. Emailing just the `.html` gives the recipient an
   unstyled page with no figures.

```` markdown
---
title: "My Analysis"
author: "Anna"
format:
  html:
    toc: true
    embed-resources: true
execute:
  message: false
  warning: false
---

```{{r}}
#| label: import-data
library(tidyverse)
library(here)

d <- read_csv(here("data", "data.csv"))
glimpse(d)
```
````
:::
:::
