# Data Visualisation with ggplot2 {#sec-visualisation}

::: {.callout-note}
## What this chapter covers
This chapter follows the lecture *Datavisualisering*. It works through the seven
layers of the grammar of graphics — data, aesthetics, geometries, facets,
statistics, coordinates and themes — and finishes with saving figures and a set
of design guidelines.
:::

```{r}
#| label: setup-viz
#| include: false
library(webexercises)
```

> Data visualisation is the graphical representation of information and data.

A useful way to think about why you are making a figure at all:

**Explore → Understand → Explain.**

The first plots you make are for yourself: quick, ugly and diagnostic. The last
one is for your reader: deliberate, labelled and polished. Confusing these two
modes is the most common reason people waste time on visualisation.

### Before you visualise

Three questions, every time:

- **What is the purpose?** Are you checking an assumption or making an argument?
- **Who is the audience?** A supervisor, a reviewer, a clinician, the public?
- **What is the best way to show this message?** The answer is often simpler
  than the plot you first imagined.

## The grammar of graphics

The grammar of graphics was formulated by [Leland
Wilkinson](https://en.wikipedia.org/wiki/Leland_Wilkinson) in 1999 and
implemented in R by [Hadley Wickham](https://en.wikipedia.org/wiki/Hadley_Wickham)
in 2005 as `ggplot2`. The core idea is that any statistical graphic can be
decomposed into independent components. Instead of choosing from a fixed menu of
chart types, you compose a figure from layers.

The layers, in the order this chapter treats them:

| Layer | Component | Question it answers |
|-------|-----------|---------------------|
| 1 | **Data** | What is being plotted? |
| 2 | **Aesthetics** (`aes()`) | Which variable maps to which visual property? |
| 3 | **Geometries** (`geom_*`) | What shapes represent the data? |
| 4 | **Facets** (`facet_*`) | Should this be split into small multiples? |
| 5 | **Statistics** (`stat_*`) | Is the data summarised before drawing? |
| 6 | **Coordinates** (`coord_*`) | How is the 2D space laid out? |
| 7 | **Themes** (`theme_*`) | How does everything non-data look? |

```{r}
#| label: viz-data
library(palmerpenguins)
library(tidyverse)

glimpse(penguins)
```

```{r}
#| label: viz-theme
# set a readable default for every figure in this chapter
theme_set(theme_minimal(base_size = 13))
```

## Layer 1: Data

`ggplot()` on its own produces an empty canvas. Every layer you add draws
something onto it.

```{r}
#| label: fig-layer1
#| fig-cap: "Layer 1 -- data only, so nothing is drawn yet."
ggplot(data = penguins)
```

The data can be supplied in four equivalent places. Which you choose matters
once you start combining several geoms.

::: {.panel-tabset}
## Inside `ggplot()`

Most common. The data and mapping are inherited by every geom.

```{r}
#| label: fig-data-a
#| fig-cap: "Data and mapping declared globally."
ggplot(data = penguins, aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point()
```

## Inside the geom

Useful when different layers use different datasets, for example raw points plus
a summarised mean.

```{r}
#| label: fig-data-b
#| fig-cap: "Data declared locally in the geom."
ggplot() +
  geom_point(data = penguins, aes(x = bill_length_mm, y = bill_depth_mm))
```

## Split

```{r}
#| label: fig-data-c
#| fig-cap: "Data global, mapping local."
ggplot(data = penguins) +
  geom_point(aes(x = bill_length_mm, y = bill_depth_mm))
```

## Piped in

The tidyverse-idiomatic form: wrangle, then plot, in one readable chain.

```{r}
#| label: fig-data-d
#| fig-cap: "Data piped into ggplot()."
penguins |>
  ggplot() +
  geom_point(aes(x = bill_length_mm, y = bill_depth_mm))
```
:::

::: {.callout-warning}
## `|>` before the plot, `+` inside it
This trips up almost everyone once. You pipe data *into* `ggplot()` with `|>`,
but you add layers *within* a plot with `+`. Using `|>` between layers produces
an error; using `+` before `ggplot()` produces a stranger one.
:::

## Layer 2: Aesthetics

**Aesthetics** link variables in your data to visual properties of the plot.
`aes()` is where you say which variable determines position, colour, fill or
shape.

| Aesthetic | Controls |
|-----------|----------|
| `x`, `y` | position along the axes |
| `colour` / `color` | the colour of lines and points (the outline) |
| `fill` | the colour inside bars, areas and boxes |
| `group` | which observations belong together, e.g. one line per participant |
| `shape` | the plotting symbol for points |
| `linetype` | solid, dashed, dotted |
| `size` / `linewidth` | point size / line thickness |
| `alpha` | transparency, from 0 (invisible) to 1 (solid) |

### Inside or outside `aes()`

This is the single most important distinction in `ggplot2`.

- **Inside `aes()`**: the property is *mapped* to a variable, and varies across
  observations. A legend appears.
- **Outside `aes()`**: the property is *set* to a fixed value for the whole
  layer. No legend.

```{r}
#| label: fig-aes-inout
#| fig-cap: "Left: colour mapped to a variable. Right: colour set to a constant."
#| fig-width: 9
#| fig-height: 3.5
library(patchwork)

mapped <- ggplot(penguins, aes(bill_length_mm, bill_depth_mm, colour = species)) +
  geom_point() +
  labs(title = "colour inside aes()")

set_value <- ggplot(penguins, aes(bill_length_mm, bill_depth_mm)) +
  geom_point(colour = "steelblue") +
  labs(title = "colour outside aes()")

mapped + set_value
```

::: {.callout-warning}
## The classic mistake
`geom_point(aes(colour = "steelblue"))` does not give you blue points. It
creates a one-level variable literally called "steelblue", maps it to the colour
scale, and gives you red points plus a pointless legend. If the colour is a
constant, it belongs outside `aes()`.
:::

### Global versus local mapping

Aesthetics in `ggplot(aes())` apply to every layer. Aesthetics in a specific geom
apply only to that layer.

```{r}
#| label: fig-global-local
#| fig-cap: "Global colour affects both the points and the smoother; local colour affects only the points."
#| fig-width: 9
#| fig-height: 3.5
global <- ggplot(penguins, aes(bill_length_mm, bill_depth_mm, colour = species)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Global: three fitted lines")

local <- ggplot(penguins, aes(bill_length_mm, bill_depth_mm)) +
  geom_point(aes(colour = species), alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Local: one fitted line")

global + local
```

That difference is not cosmetic — it is the difference between fitting one model
and fitting three. We return to it under Simpson's paradox in @sec-eda.

### More aesthetics at once

::: {.panel-tabset}
## colour

```{r}
#| label: fig-aes-colour
ggplot(penguins, aes(bill_length_mm, bill_depth_mm, colour = species)) +
  geom_point()
```

## + size

```{r}
#| label: fig-aes-size
ggplot(penguins, aes(bill_length_mm, bill_depth_mm,
  colour = species, size = body_mass_g
)) +
  geom_point()
```

## + alpha

```{r}
#| label: fig-aes-alpha
ggplot(penguins, aes(bill_length_mm, bill_depth_mm,
  colour = species, size = body_mass_g
)) +
  geom_point(alpha = 0.7)
```

## + shape

```{r}
#| label: fig-aes-shape
ggplot(penguins, aes(bill_length_mm, bill_depth_mm,
  colour = species, size = body_mass_g, shape = island
)) +
  geom_point(alpha = 0.5)
```
:::

The last plot encodes four variables at once, and it is close to unreadable.
That is the lesson: just because an aesthetic is available does not mean it
should be used.

::: {.webex-check .webex-box}
**Check yourself**

To make all points in a layer semi-transparent you should write
`r mcq(c("geom_point(aes(alpha = 0.4))", answer = "geom_point(alpha = 0.4)", "aes(alpha = 0.4)", "scale_alpha(0.4)"))`

`fill` controls the outline colour of a bar. `r torf(FALSE)`

Which aesthetic would you map to draw one line per participant over time?
`r mcq(c("colour", "shape", answer = "group", "facet"))`
:::

## Scales and guides

A **scale** controls how data values are translated into an aesthetic. Every
aesthetic has a corresponding scale, and each aesthetic can have only one — a
later scale silently overrides an earlier one.

**Guides** are how the reader reverses that translation: axes are the guides for
position scales, legends are the guides for colour, shape and size.

Scale functions are named `scale_<aesthetic>_<type>()`. The three main types:

1. **Continuous** — `scale_x_continuous()`, `scale_colour_gradient()`
2. **Discrete** — `scale_colour_discrete()`, `scale_shape_manual()`
3. **Binned** — `scale_fill_binned()`

::: {.panel-tabset}
## Continuous colour

```{r}
#| label: fig-scale-cont
ggplot(penguins, aes(flipper_length_mm, body_mass_g, colour = bill_length_mm)) +
  geom_point() +
  scale_colour_gradient(low = "blue", high = "red")
```

## Manual colours

```{r}
#| label: fig-scale-manual
ggplot(penguins, aes(flipper_length_mm, body_mass_g, colour = species)) +
  geom_point() +
  scale_colour_manual(values = c(
    "Adelie" = "darkorange",
    "Chinstrap" = "purple",
    "Gentoo" = "cyan4"
  ))
```

## Manual breaks

```{r}
#| label: fig-scale-breaks
ggplot(penguins, aes(flipper_length_mm, body_mass_g, colour = species)) +
  geom_point() +
  scale_x_continuous(
    breaks = c(180, 200, 220),
    labels = c("Short", "Medium", "Long")
  )
```

## Log scale

```{r}
#| label: fig-scale-log
ggplot(penguins, aes(flipper_length_mm, body_mass_g, colour = species)) +
  geom_point() +
  scale_x_log10()
```

## Axis names

```{r}
#| label: fig-scale-names
ggplot(penguins, aes(flipper_length_mm, body_mass_g)) +
  geom_point() +
  scale_x_continuous(name = "Flipper length (mm)") +
  scale_y_continuous(name = "Body mass (g)")
```
:::

### Labels with `labs()`

`labs()` is the quickest way to set every label at once, and the label for an
aesthetic is also the legend title.

```{r}
#| label: fig-labs
ggplot(penguins, aes(bill_length_mm, bill_depth_mm, colour = species)) +
  geom_point(alpha = 0.7) +
  labs(
    title = "Bill dimensions differ between penguin species",
    subtitle = "Palmer Archipelago, Antarctica, 2007-2009",
    x = "Bill length (mm)",
    y = "Bill depth (mm)",
    colour = "Species",
    caption = "Data: palmerpenguins"
  )
```

::: {.callout-tip}
## Units belong on the axis
An axis labelled `body_mass_g` tells the reader you did not finish the figure.
An axis labelled "Body mass (g)" tells them what they are looking at. This is
the cheapest quality improvement available in any figure.
:::

### Customising the legend with `guides()`

```{r}
#| label: fig-guides
ggplot(penguins, aes(bill_length_mm, bill_depth_mm,
  colour = species, shape = species
)) +
  geom_point(alpha = 0.7) +
  labs(x = "Bill length (mm)", y = "Bill depth (mm)") +
  guides(
    colour = guide_legend(title = "Species", override.aes = list(size = 4)),
    shape = guide_legend(title = "Species")
  )
```

Note that `colour` and `shape` are given the *same* title, which merges them
into one legend instead of two. `override.aes` makes the legend keys larger than
the plotted points, which is almost always an improvement when the points are
small or transparent.

## Layer 3: Geometries

Different geoms require different aesthetics. Several geoms can be stacked in
one figure, and **order matters** — later layers are drawn on top of earlier
ones.

| Geom | Use | Common aesthetics |
|------|-----|-------------------|
| `geom_point()` | scatterplots | `x`, `y`, `colour`, `size`, `shape` |
| `geom_line()` | line plots, trends over time | `x`, `y`, `colour`, `linetype` |
| `geom_bar()` | bar chart of counts | `x`, `fill` |
| `geom_col()` | bar chart of pre-summarised values | `x`, `y`, `fill` |
| `geom_histogram()` | distribution of a continuous variable | `x`, `fill` |
| `geom_density()` | smoothed distribution | `x`, `fill`, `colour` |
| `geom_boxplot()` | five-number summary by group | `x`, `y`, `fill` |
| `geom_violin()` | full distribution by group | `x`, `y`, `fill` |
| `geom_jitter()` | points with added noise | `x`, `y`, `colour` |
| `geom_smooth()` | fitted trend line | `x`, `y`, `colour` |
| `geom_errorbar()` | uncertainty intervals | `x`, `ymin`, `ymax` |
| `geom_text()` / `geom_label()` | annotate points | `x`, `y`, `label` |
| `geom_tile()` | heatmaps | `x`, `y`, `fill` |
| `geom_ribbon()` | confidence bands | `x`, `ymin`, `ymax`, `fill` |

### Building up a distribution plot

::: {.panel-tabset}
## Violin

```{r}
#| label: fig-geom-violin
ggplot(penguins, aes(species, body_mass_g, fill = species)) +
  geom_violin(alpha = 0.5)
```

## + boxplot

```{r}
#| label: fig-geom-box
ggplot(penguins, aes(species, body_mass_g, fill = species)) +
  geom_violin(alpha = 0.5, trim = FALSE) +
  geom_boxplot(width = 0.2, alpha = 0.7)
```

## + raw points

```{r}
#| label: fig-geom-points
ggplot(penguins, aes(species, body_mass_g, fill = species)) +
  geom_violin(alpha = 0.5, trim = FALSE) +
  geom_boxplot(width = 0.2, alpha = 0.7, outlier.shape = NA) +
  geom_point(
    position = position_jitter(width = 0.2),
    colour = "black", alpha = 0.5, size = 1
  )
```
:::

Note `outlier.shape = NA` in the last panel. Without it the boxplot draws its
outliers *and* the point layer draws them again — the same observation appears
twice.

::: {.callout-tip}
## Show the data
A bar chart of group means hides everything that matters: sample size, spread,
skew, outliers. With the group sizes typical in sport and health science, plot
the raw observations. The violin-box-point combination above shows the summary
and the data at the same time.
:::

### Position adjustments

Position is the third required part of a layer, alongside the geom and the stat.
It resolves overlap between groups.

::: {.panel-tabset}
## stack

```{r}
#| label: fig-pos-stack
ggplot(penguins, aes(species, fill = sex)) +
  geom_bar(position = "stack")
```

## dodge

```{r}
#| label: fig-pos-dodge
ggplot(penguins, aes(species, fill = sex)) +
  geom_bar(position = "dodge")
```

## fill

```{r}
#| label: fig-pos-fill
# proportions instead of counts - good for comparing composition,
# but it hides how many observations each bar is based on
ggplot(penguins, aes(species, fill = sex)) +
  geom_bar(position = "fill")
```

## jitter

```{r}
#| label: fig-pos-jitter
ggplot(penguins, aes(species, body_mass_g, colour = sex)) +
  geom_jitter(width = 0.2, alpha = 0.6)
```

## nudge

```{r}
#| label: fig-pos-nudge
ggplot(penguins, aes(species, body_mass_g, colour = sex)) +
  geom_point(position = position_nudge(x = 0.2), alpha = 0.6)
```
:::

::: {.callout-note collapse="true"}
## Exercise 1: your first figure

Using `penguins`:

1. Make a scatterplot of `flipper_length_mm` against `body_mass_g`.
2. Colour the points by `species`.
3. Add a linear trend line per species.
4. Give the figure proper axis labels with units, a title and a legend title.

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: fig-ex1-solution
#| warning: false
ggplot(penguins, aes(flipper_length_mm, body_mass_g, colour = species)) +
  geom_point(alpha = 0.7) +
  # colour is mapped globally, so geom_smooth() fits one line per species
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Heavier penguins have longer flippers",
    x = "Flipper length (mm)",
    y = "Body mass (g)",
    colour = "Species"
  )
```

Because `colour = species` sits in `ggplot(aes())` it is inherited by
`geom_smooth()`, which is why three lines appear rather than one.
:::
:::

## Layer 4: Facets

Faceting splits one plot into small multiples by a categorical variable. It is
often a better answer than adding a fourth aesthetic.

::: {.panel-tabset}
## facet_wrap()

One variable, wrapped into a grid.

```{r}
#| label: fig-facet-wrap
ggplot(penguins, aes(bill_length_mm, bill_depth_mm)) +
  geom_point(aes(colour = species)) +
  facet_wrap(~island)
```

## facet_grid()

Two variables, as rows and columns.

```{r}
#| label: fig-facet-grid
ggplot(penguins, aes(bill_length_mm, bill_depth_mm)) +
  geom_point(aes(colour = species)) +
  facet_grid(sex ~ island)
```

## Free scales

```{r}
#| label: fig-facet-free
# free scales let each panel zoom to its own data, which is useful for
# exploration but makes panels no longer directly comparable
ggplot(penguins, aes(bill_length_mm, bill_depth_mm)) +
  geom_point(aes(colour = species)) +
  facet_grid(sex ~ island, scales = "free")
```
:::

::: {.callout-warning}
## `scales = "free"` breaks comparability
Small multiples work because every panel shares the same ruler. The moment you
free the scales, a difference in panel appearance no longer means a difference
in the data. Use free scales when exploring, fixed scales when explaining.
:::

## Layer 5: Statistics

A statistical transformation, or **stat**, processes the data before it is
drawn — usually by summarising it. Every geom has a default stat, and every stat
has a default geom.

- `stat_bin()` — used by `geom_histogram()`, `geom_bar()`, `geom_freqpoly()`
- `stat_boxplot()` — used by `geom_boxplot()`
- `stat_smooth()` — used by `geom_smooth()`
- `stat_summary()` — arbitrary summaries such as means with error bars

There are two equivalent ways to use them: add a `stat_*()` and override its
geom, or add a `geom_*()` and override its stat.

::: {.panel-tabset}
## stat_summary()

```{r}
#| label: fig-stat-summary
ggplot(penguins, aes(species, bill_length_mm)) +
  geom_point(alpha = 0.4, position = position_jitter(width = 0.15)) +
  stat_summary(geom = "point", fun = "mean", colour = "darkorange", size = 5)
```

## geom with stat

```{r}
#| label: fig-geom-stat
# identical result, written the other way round
ggplot(penguins, aes(species, bill_length_mm)) +
  geom_point(alpha = 0.4, position = position_jitter(width = 0.15)) +
  geom_point(stat = "summary", fun = "mean", colour = "darkorange", size = 5)
```

## Mean with CI

```{r}
#| label: fig-stat-ci
ggplot(penguins, aes(species, bill_length_mm)) +
  geom_point(alpha = 0.3, position = position_jitter(width = 0.15)) +
  stat_summary(
    fun.data = mean_cl_normal,
    geom = "errorbar", width = 0.2, colour = "darkorange", linewidth = 1
  )
```
:::

### Binning

`stat_bin()` is what makes a histogram. It creates new variables you can access
with `after_stat()`:

- `count` — number of observations per bin (the default)
- `density` — proportion per unit of the x axis
- `x` — the midpoint of each bin

```{r}
#| label: fig-hist-count
ggplot(penguins, aes(body_mass_g)) +
  geom_histogram(binwidth = 100)
```

```{r}
#| label: fig-hist-density
# density instead of count: the total area sums to 1, so histograms
# with different sample sizes become comparable
ggplot(penguins, aes(body_mass_g)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 100)
```

::: {.callout-important}
## Always set `binwidth`
The default 30 bins is a placeholder, and `ggplot2` says so in a message every
time. Bin width is an analytical choice: too wide hides structure, too narrow
turns the histogram into noise. Try a few and pick one you can defend.
:::

## Layer 6: Coordinates

Scales decide which values appear; the **coordinate system** decides where they
are physically drawn. Coordinate systems can be linear or non-linear.

```{r}
#| label: fig-coord-base
base_plot <- ggplot(penguins, aes(bill_length_mm, bill_depth_mm)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", formula = y ~ x)

base_plot
```

::: {.panel-tabset}
## coord_cartesian()

```{r}
#| label: fig-coord-cart
# zooms without discarding data - the smoother is still fitted on everything
base_plot + coord_cartesian(xlim = c(40, 50))
```

## coord_flip()

```{r}
#| label: fig-coord-flip
base_plot + coord_flip()
```

## coord_fixed()

```{r}
#| label: fig-coord-fixed
# one unit on x takes the same physical space as one unit on y
base_plot + coord_fixed(ratio = 1)
```

## coord_polar()

```{r}
#| label: fig-coord-polar
# a stacked bar in polar coordinates is a pie chart
ggplot(penguins, aes(x = "", fill = species)) +
  geom_bar(width = 1) +
  coord_polar(theta = "y") +
  labs(x = NULL, y = NULL)
```
:::

::: {.callout-warning}
## Zooming: two ways, two meanings
`coord_cartesian(xlim = ...)` zooms the view and keeps all the data.
`scale_x_continuous(limits = ...)` *removes* observations outside the range
before anything is computed, which silently changes your means, smoothers and
regression lines. When you mean "zoom", use `coord_cartesian()`.
:::

## Layer 7: Themes

Themes control everything that is not data: fonts, background, grid lines,
spacing, legend placement.

::: {.panel-tabset}
## theme_minimal()

```{r}
#| label: fig-theme-min
ggplot(penguins, aes(bill_length_mm, bill_depth_mm, colour = species)) +
  geom_point() +
  theme_minimal()
```

## theme_classic()

```{r}
#| label: fig-theme-classic
ggplot(penguins, aes(bill_length_mm, bill_depth_mm, colour = species)) +
  geom_point() +
  theme_classic()
```

## Custom theme()

```{r}
#| label: fig-theme-custom
ggplot(penguins, aes(bill_length_mm, bill_depth_mm, colour = species)) +
  geom_point(size = 2) +
  theme(
    panel.background = element_rect(fill = "grey95", colour = NA),
    panel.grid.major = element_line(colour = "white"),
    panel.grid.minor = element_blank(),
    axis.title = element_text(size = 12, face = "bold"),
    legend.position = "bottom" # use "none" to drop the legend entirely
  )
```
:::

The `element_*()` functions are how you address theme components:
`element_text()` for anything written, `element_line()` for lines,
`element_rect()` for backgrounds and borders, and `element_blank()` to remove a
component entirely.

### Setting a theme once

Rather than repeating `theme_minimal()` in every plot, set it once at the top of
your document:

```{r}
#| eval: false
theme_set(
  theme_minimal(base_size = 13) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title.position = "plot" # left-align the title with the plot edge
    )
)
```

Every subsequent figure inherits it, which is both less typing and more
consistent.

## Combining plots with patchwork

```{r}
#| label: fig-patchwork
#| fig-width: 9
#| fig-height: 7
p1 <- ggplot(penguins, aes(body_mass_g)) +
  geom_histogram(binwidth = 200, fill = "steelblue", colour = "white")

p2 <- ggplot(penguins, aes(species, body_mass_g, fill = species)) +
  geom_boxplot(show.legend = FALSE)

p3 <- ggplot(penguins, aes(flipper_length_mm, body_mass_g, colour = species)) +
  geom_point(alpha = 0.6)

# / stacks vertically, | places side by side
(p1 | p2) / p3 +
  plot_annotation(
    title = "Body mass in three penguin species",
    tag_levels = "A"
  )
```

## Saving figures

```{r}
#| eval: false
ggsave(
  filename = "figures/my_figure.pdf", # extension decides the format
  plot = my_figure, # the plot object to save
  width = 11, # width
  height = 8.5, # height
  units = "cm" # "cm", "mm" or "in"
)

ggsave(
  filename = "figures/my_figure.png",
  plot = my_figure,
  width = 11,
  height = 8.5,
  units = "cm",
  dpi = 300 # resolution; journals typically require 300 or more
)
```

::: {.callout-tip}
## Vector or raster?
Save as **PDF or SVG** for anything with points, lines and text — the figure
stays sharp at any size, which is what journals want. Save as **PNG** when the
figure contains very many overlapping points (tens of thousands) and the vector
file would be enormous, or when you need to paste into PowerPoint. Never save a
final figure as JPEG: the compression smears text and thin lines.
:::

Note also that `ggsave()` does not take a screenshot of the RStudio plot pane —
it redraws the figure at the size you asked for. Text does not scale with the
figure, so a figure saved at 20 cm wide has relatively smaller text than the
same figure saved at 10 cm. Always check the saved file, not the preview.

## Design guidelines

- **Left-align** most text, including titles.
- **Labels and subtitles** may be centred or right-aligned when there is a
  reason.
- **Do not rotate** axis text. If the labels do not fit, flip the coordinates
  instead.
- **Fonts** should suit the context. A few good ones: Montserrat, Lato, Roboto,
  Open Sans. Use a couple, not a dozen.
- **Text hierarchy** is created with size, colour and weight — and applied
  consistently.
- **Colour** should carry meaning. If colour is decorative only, greyscale is
  usually better.
- **Check accessibility**: roughly 8% of men have some form of colour vision
  deficiency. Viridis and Okabe-Ito palettes are safe choices, and redundant
  encoding (colour *and* shape) is safer still.

::: {.webex-check .webex-box}
**Check yourself**

Which function zooms into a plot without removing observations?
`r mcq(c("scale_x_continuous(limits = ...)", answer = "coord_cartesian(xlim = ...)", "xlim()", "facet_zoom()"))`

`facet_wrap(~ island, scales = "free")` keeps the panels directly comparable.
`r torf(FALSE)`

Which geom would you use for a bar chart of values you have already summarised?
`r mcq(c("geom_bar()", answer = "geom_col()", "geom_histogram()", "geom_count()"))`

To remove the legend you set `legend.position` to
`r fitb(c("none", "\"none\""), ignore_ws = TRUE)`
:::

::: {.callout-note collapse="true"}
## Exercise 2: distributions by group

1. Plot the distribution of `flipper_length_mm` as a histogram with a sensible
   bin width.
2. Now show the same distribution separately for each species, in a way that
   makes them easy to compare.
3. Which approach do you find clearest, and why?

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: fig-ex2a
#| warning: false
ggplot(penguins, aes(flipper_length_mm)) +
  geom_histogram(binwidth = 2, fill = "steelblue", colour = "white") +
  labs(x = "Flipper length (mm)", y = "Count")
```

Three options, each with a trade-off:

```{r}
#| label: fig-ex2b
#| warning: false
#| fig-width: 9
#| fig-height: 7
# faceted: no overlap, but the eye has to travel between panels
facets <- ggplot(penguins, aes(flipper_length_mm)) +
  geom_histogram(binwidth = 2, fill = "steelblue", colour = "white") +
  facet_wrap(~species, ncol = 1) +
  labs(title = "Faceted histograms", x = NULL, y = NULL)

# overlaid densities: direct comparison, but overlap can obscure
densities <- ggplot(penguins, aes(flipper_length_mm, fill = species)) +
  geom_density(alpha = 0.5) +
  labs(title = "Overlaid densities", x = NULL, y = NULL)

# boxplots: compact and precise, but hides bimodality
boxes <- ggplot(penguins, aes(flipper_length_mm, species, fill = species)) +
  geom_boxplot(show.legend = FALSE) +
  labs(title = "Boxplots", x = NULL, y = NULL)

facets | (densities / boxes)
```

There is no single right answer. The densities make the separation between
Gentoo and the other two species most immediate; the facets are the most honest
about sample size; the boxplots are the most compact. Which you choose depends
on whether you are exploring or explaining.
:::
:::

::: {.callout-note collapse="true"}
## Exercise 3: debug the plot

Each of these produces something other than what was intended. Explain why, then
fix it.

```{r}
#| eval: false
# a) why are the points red, not blue?
ggplot(penguins, aes(bill_length_mm, bill_depth_mm)) +
  geom_point(aes(colour = "blue"))

# b) why does this error?
penguins |>
  ggplot(aes(species, body_mass_g)) |>
  geom_boxplot()

# c) why is there only one boxplot?
ggplot(penguins, aes(body_mass_g, fill = species)) +
  geom_boxplot()
```

::: {.callout-tip collapse="true"}
## Solution

**a)** `colour = "blue"` inside `aes()` maps a constant *string* to the colour
scale, creating a single-level variable whose first default colour is red. Move
it outside `aes()`:

```{r}
#| label: fig-ex3a
ggplot(penguins, aes(bill_length_mm, bill_depth_mm)) +
  geom_point(colour = "blue")
```

**b)** Layers are added with `+`, not piped with `|>`. The pipe is only used to
get the data *into* `ggplot()`:

```{r}
#| label: fig-ex3b
penguins |>
  ggplot(aes(species, body_mass_g)) +
  geom_boxplot()
```

**c)** `fill` colours the boxes but does not position them. Without a
categorical variable on an axis, all three species are drawn in the same place.
Map `species` to `y` as well:

```{r}
#| label: fig-ex3c
ggplot(penguins, aes(body_mass_g, species, fill = species)) +
  geom_boxplot(show.legend = FALSE)
```
:::
:::

## Further resources

- **400 examples with code**: [R Graph Gallery](https://r-graph-gallery.com/)
- **The `ggplot2` book**: [ggplot2-book.org](https://ggplot2-book.org/)
- **Concepts**: Wilke, C. O. (2019). *Fundamentals of Data Visualization*
  ([free online](https://clauswilke.com/dataviz/))
- **People to learn from**: [Cedric Scherer](https://cedricscherer.netlify.app),
  [Georgios Karamanis](https://karaman.is/),
  [Thomas Lin Pedersen](https://www.data-imaginist.com)
- **Choosing a chart type**: [data-to-viz.com](https://www.data-to-viz.com/)
- **Colour**: [viz-palette](https://projects.susielu.com/viz-palette),
  [Datawrapper on beautiful colours](https://blog.datawrapper.de/beautifulcolors/)
