# Version Control {#sec-git}

::: {.callout-note}
## What this chapter covers
Version control is not part of the course lectures, but it is part of working
reproducibly, so it is in the book. The chapter explains the few ideas Git is
built on, sets it up once, and then shows the daily workflow in both RStudio and
Positron: committing, looking back, undoing, branching and resolving conflicts.
It ends with what must never go into a repository.
:::

```{r}
#| label: setup-git
#| include: false
library(webexercises)
```

## Why bother

Without version control, a project's history lives in file names:
`analysis_final.R`, `analysis_final_v2.R`, `analysis_final_v2_DV_comments.R`.
Nobody can tell which one produced the table in the submitted manuscript, and
nobody dares delete the old ones.

Version control replaces that with a single copy of each file and a complete,
annotated history behind it. For research this gives you:

- **A record of what changed, when and why.** When a reviewer asks why the
  sample size went from 1,204 to 1,187 between two drafts, the answer is in the
  history, with the date and your own explanation.
- **Freedom to experiment.** Try a different model or rewrite the cleaning; if
  it does not work out, go back. Nothing is ever really lost.
- **Collaboration without emailing files.** Several people can work on the same
  project and combine their changes.
- **Transparency.** Sharing the repository with a paper lets others see exactly
  what was done.

**Git** is the version control system almost everyone uses. **GitHub** (and
alternatives such as GitLab) hosts copies of Git repositories online.

## The ideas Git is built on

Git has a reputation for being difficult, mostly because of its vocabulary. The
ideas themselves are few:

| Term | What it means |
|------|---------------|
| **Repository** (repo) | A project folder whose history Git tracks. The history lives in a hidden `.git` folder inside it. |
| **Commit** | A saved snapshot of the project, with a message saying what changed and why. The history is a chain of commits. |
| **Staging** | Choosing which changes go into the next commit. You can change five files and commit only two of them. |
| **Diff** | The difference between two versions of a file: lines removed in red, lines added in green. |
| **Remote** | A copy of the repository somewhere else, usually on GitHub. |
| **Push** / **pull** | Send your new commits to the remote / fetch other people's commits from it. |
| **Clone** | Download a complete copy of a repository, history included. |
| **Branch** | A parallel line of commits, for trying something without disturbing the main version. |

A commit is only saved in your local repository until you push it. The remote
is both your backup and the meeting point for collaborators.

## One-time setup

Both editors use the same Git underneath, so this part is identical in RStudio
and Positron.

1. **Install Git** from [git-scm.com](https://git-scm.com/downloads) if you do
   not already have it. On macOS, running `git --version` in a terminal offers
   to install it.
2. **Tell Git who you are.** The name and email end up in every commit you make:

```{r}
#| eval: false
usethis::use_git_config(
  user.name = "Your Name",
  user.email = "you@example.com"
)
```

3. **Connect to GitHub** if you want a remote copy. The `usethis` package
   handles the fiddly parts:

```{r}
#| eval: false
usethis::create_github_token() # opens GitHub to create a personal access token
gitcreds::gitcreds_set() # paste the token in when prompted
```

In Positron you can instead sign in to GitHub through the **Accounts** menu at
the bottom of the activity bar, which is enough for pushing and pulling from the
editor.

::: {.callout-tip}
## When something does not work: `git_sitrep()`
`usethis::git_sitrep()` prints a diagnosis of your whole Git and GitHub setup,
names what is missing, and suggests the fix. It is a plain R function, so it
works the same in both editors.
:::

## Starting a repository

**A project you already have**: open it, then in the console

```{r}
#| eval: false
# keep data out BEFORE Git sees anything
usethis::use_git_ignore(c("data/", "*.csv", "*.sav", "*.parquet", "*.rds"))

usethis::use_git() # start tracking this project
usethis::use_github(private = TRUE) # create a GitHub copy and connect it
```

`use_git()` offers to commit every file already in the folder. That is why the
ignore rules come first: whatever is not ignored at that moment goes into the
first commit. Read the list of files it shows before you answer yes.

::: {.callout-warning}
## `use_github()` makes a public repository unless you say otherwise
The default is **public**: anyone on the internet can read it. For research
projects, pass `private = TRUE` and make it public deliberately, later, once
you have checked what is in the history.
:::

**A project that exists on GitHub already**: clone it.

```{r}
#| eval: false
usethis::create_from_github("owner/repository", destdir = "~/projects")
```

In RStudio, **File → New Project → Version Control → Git** does the same; in
Positron, use **New Folder from Git…** in the **New** menu (or the command
palette).

::: {.callout-important}
## Git and OneDrive do not mix well
A repository inside a folder synced by OneDrive, Dropbox or iCloud can be
damaged when the sync client and Git write to the hidden `.git` folder at the
same time, or when the sync creates "conflicted copy" files inside it. Keep Git
repositories in an ordinary local folder, such as `~/projects/`, and let the
**remote** be your backup. If you use Git, you do not need the cloud sync.
:::

## What goes into the repository

Git is for the files you **write**. Everything that can be recreated from them,
and everything that must stay private, stays out.

| Commit | Do not commit |
|--------|---------------|
| R scripts, `.qmd` files, functions in `R/` | **Data**: `data/`, `*.csv`, `*.sav`, `*.dta`, `*.parquet`, `*.rds` |
| `_targets.R`, `renv.lock` (@sec-pipelines) | Rendered output that the code recreates (`_book/`, `*.html`, figures) |
| `README.md`, `data-dict.yaml` (@sec-data-dictionary) | `.Rhistory`, `.RData`, `.Rproj.user/` |
| `.gitignore` itself | `.Renviron` and anything else holding passwords or API keys |

`.gitignore` is a text file listing patterns Git should not track.
`usethis::use_git()` creates a reasonable one, and `usethis::use_git_ignore()`
adds to it.

::: {.callout-important}
## Add data to `.gitignore` before your first commit
Removing a file in a later commit does **not** remove it from the history: it
is still there for anyone who clones the repository. So the ignore rules have to
be in place before the first commit, not after.

```{r}
#| eval: false
usethis::use_git_ignore(c(
  "data/", "data-raw/",
  "*.csv", "*.sav", "*.dta", "*.parquet", "*.rds",
  ".Renviron"
))
```

Then check the list of changed files before you commit (the Git pane, the
Source Control view, or `git status` in the terminal), and confirm that no data
file is in it.
:::

## The daily workflow

Once the repository exists, the rhythm is always the same:

1. **Pull** first, if others work on the project, so you start from the latest
   version.
2. **Work** as usual: edit, run, render.
3. **Look at the diff** of what you changed. This is a small code review of your
   own work, and it catches the stray `View()` or the accidentally deleted line.
4. **Stage** the changes that belong together and **commit** them with a
   message.
5. **Push**, at the latest at the end of the day.

### Good commits

Commit **small and often**: one commit per logical change, not one per day.
"Exclude participants without baseline" and "Add sensitivity analysis" are two
commits, even if you did both before lunch. If a change later turns out to be
wrong, a small commit is easy to find and easy to undo.

The message should say **why**, because the diff already says what:

```text
Exclude participants without a baseline measurement

The protocol requires baseline for the change score. Removes 17
participants; flow chart updated.
```

A short first line (under about 60 characters), then a blank line, then as much
explanation as a reviewer would need. "Fix", "update" and "changes" tell your
future self nothing.

## Git in RStudio

1. **Initialise**: `usethis::use_git()`, or **Tools → Version Control → Project
   Setup** and choose Git. RStudio restarts and a **Git** pane appears in the
   top-right quadrant.
2. **Stage and commit**: tick the files you want to include, click **Commit**.
   A window opens with the diff of each file; write the message and commit.
3. **Push and pull** with the green and blue arrows in the Git pane.
4. **History**: the clock icon shows the list of commits; click one to see its
   diff.
5. **Revert** a file to its last committed version: select it and click
   **Revert** (in the **More** menu). This discards your uncommitted changes to
   that file.
6. **Branches**: the branch name at the right of the Git pane is a menu for
   switching; the icon beside it creates a new branch.

[More on version control in RStudio](https://docs.posit.co/ide/user/ide/guide/tools/version-control.html)

## Git in Positron

Positron has the Git integration of VS Code, and for day-to-day work it is
clearly the stronger of the two editors. The concepts are identical; the
controls are in different places (@sec-positron).

1. **Open the folder** (**File → Open Folder**). Positron detects an existing
   `.git` folder by itself.
2. **Initialise** if there is no repository yet: open the **Source Control**
   view in the activity bar (<kbd>Ctrl/Cmd</kbd>+<kbd>Shift</kbd>+<kbd>G</kbd>)
   and click **Initialize Repository**, or run `usethis::use_git()`.
3. **See the diff**: click any changed file in the Source Control view to open
   a side-by-side comparison. Changed lines are also marked in the margin of the
   editor while you work, in green (added), blue (modified) and red (deleted).
4. **Stage**: hover a file and click **+**. To stage **part** of a file, open
   its diff, select the lines, and choose **Stage Selected Ranges** from the
   right-click menu. That is the feature RStudio users miss most once they have
   used it: one editing session that produced two unrelated changes can become
   two clean commits.
5. **Commit**: write the message in the box at the top and press
   <kbd>Ctrl/Cmd</kbd>+<kbd>Enter</kbd>.
6. **Push and pull**: the sync button in the status bar at the bottom, or the
   **…** menu in the Source Control view.
7. **Branches**: the branch name in the bottom-left corner of the status bar.
   Click it to switch branch or create a new one.

A few things Positron adds on top:

- **A history graph.** The Source Control view has a **Graph** section showing
  the commits and branches as a timeline.
- **File history.** The **Timeline** section at the bottom of the Explorer shows
  every commit that touched the open file; click one to see what it changed.
- **A three-way merge editor** for conflicts (see below), with buttons to accept
  one side, the other, or both.
- **GitHub built in.** The GitHub Pull Requests and Issues extension comes
  pre-installed, so you can review pull requests and read issues without leaving
  the editor.
- **Suggested commit messages.** If you have configured an AI provider, the
  sparkle icon in the message box drafts a message from the staged diff. Treat
  it like any other AI output (@sec-ai): it can describe *what* changed, but only
  you know *why*, and the why is the part worth writing.

[Git in Positron](https://positron.posit.co/git.html)

## Undoing things

Most of what makes Git feel safe is knowing how to go back. The common cases,
from the least to the most drastic:

| Situation | RStudio | Positron | Terminal |
|-----------|---------|----------|----------|
| Throw away uncommitted changes to a file | **Revert** | **Discard Changes** (↶ icon) | `git restore file.R` |
| Unstage a file, keeping the changes | Untick it | **−** icon | `git restore --staged file.R` |
| Fix the message or add a forgotten file to the last commit, **not yet pushed** | Tick **Amend previous commit** | **Commit → Commit (Amend)** in the **…** menu | `git commit --amend` |
| Look at an old version of a file | History → **View file** | Timeline → click the commit | `git show a1b2c3d:file.R` |
| Undo a commit that is **already pushed** | Terminal | Terminal | `git revert a1b2c3d` |

`a1b2c3d` stands for the commit's ID, the short code shown next to each commit
in the history.

Discarding changes cannot itself be undone, because those changes were never
committed. That is one more reason to commit often.

`git revert` does not delete anything. It adds a **new** commit that reverses
the old one, so the history stays honest and your collaborators' copies stay
valid. Never rewrite history that others have already pulled.

## Branches

A branch lets you try something without touching the working version: a
different model specification, a large restructuring, a reviewer's suggestion
you are not sure about.

```bash
git switch -c try-mixed-model   # create a branch and move to it
# ... work and commit as usual ...
git switch main                 # back to the main version, untouched
git merge try-mixed-model       # keep the experiment: merge it into main
git branch -d try-mixed-model   # delete the branch when it is merged
```

If the experiment fails, switch back to `main` and delete the branch; nothing on
`main` ever knew about it. Both editors do the same from the branch menu. On
GitHub, merging a branch usually happens through a **pull request**, which gives
collaborators a place to review the changes first.

For a project you work on alone, you can get a long way with no branches at
all. They become valuable when you collaborate, or when an experiment will take
more than a day.

## Merge conflicts

When two people change **the same lines** of the same file, Git cannot decide
which version to keep, and it asks you. The file then contains both versions
between markers:

```text
<<<<<<< HEAD
model <- lm(mass ~ flipper + species, data = penguins)
=======
model <- lm(mass ~ flipper * species, data = penguins)
>>>>>>> try-interaction
```

The part above `=======` is yours; the part below is the incoming change.
Resolve it by editing the file to what it *should* be (one version, the other,
or a combination), deleting the three marker lines, and committing. Positron's
merge editor shows the two versions and the result side by side and lets you
accept either with a click; RStudio shows the markers in the ordinary editor.

Conflicts are rarer than their reputation. Changes to different files, or to
different parts of the same file, are merged automatically. Pulling before you
start working and pushing when you finish avoids most of the rest.

## Git and sensitive data

The rules from @sec-import apply with extra force here, because Git remembers
everything:

- **Never commit registry data or any other personal data**, not even to a
  private repository. Private means GitHub decides who can read it, and
  GitHub is a third-party service outside your ethical approval.
- **Secrets count too.** A `.Renviron` with an API key or a database password,
  once pushed, should be treated as leaked: change the key, do not just delete
  the file.
- **If data has been committed**, deleting it in a new commit is not enough.
  Removing it from the history takes specialist tools (`git filter-repo`), and
  any clone made in the meantime still has it. Stop, do not push, and contact
  your data protection officer or IT support. How to report it is decided by
  your institution, not by Git.
- **Commit messages and code are part of what you share.** Do not write
  identifiers, rare diagnoses or individual values into comments or messages.

## Git and GitHub are not the same

- **Git** is the version control system. It runs locally on your computer and
  keeps the history of your project as a series of commits. It works without an
  internet connection and without any account.
- **GitHub** is a service that hosts Git repositories online. It adds a web
  interface, access control, issues, pull requests and GitHub Pages (which
  publishes this book). GitLab and Bitbucket do the same job, and some
  universities run their own GitLab.

You can use Git without GitHub. You cannot use GitHub without Git.

[Happy Git and GitHub for the useR](https://happygitwithr.com/) by Jenny Bryan
is the standard reference for all of this in R, including what to do when it
goes wrong.

::: {.webex-check .webex-box}
**Check yourself**

What is staging?
`r longmcq(c(
  "Uploading your commits to GitHub",
  answer = "Choosing which changes go into the next commit",
  "Creating a new branch",
  "Saving a file in the editor"
))`

`usethis::use_github()` creates a private repository by default. `r torf(FALSE)`

You committed a CSV with registry data, then deleted it in the next commit.
The data is no longer in the repository. `r torf(FALSE)`

A commit you have already pushed turns out to be wrong. What is the safe way to
undo it?
`r mcq(c("Delete the .git folder", answer = "git revert", "git commit --amend", "Discard Changes"))`

What should a commit message explain that the diff cannot?
`r mcq(c("Which files changed", "How many lines changed", answer = "Why the change was made", "Who made the change"))`
:::

::: {.callout-note collapse="true"}
## Exercise 1: your first repository

Do this in a new, throwaway project in an ordinary local folder (not OneDrive),
in whichever editor you use.

1. Create the project and run `usethis::use_git()`. Look at the `.gitignore`
   it created.
2. Add a `data/` folder containing any small CSV file. Add `data/` to
   `.gitignore` with `usethis::use_git_ignore()`, and check that the CSV does
   not show up among the changed files.
3. Create `analysis.R` with three lines of code and commit it with a proper
   message.
4. Change one line, look at the diff, and commit again.
5. Change another line, then discard that change so the file is back to your
   last commit.
6. Find your first commit in the history and look at what the file looked like
   then.

::: {.callout-tip collapse="true"}
## What to look for
- **Step 1**: `use_git()` already ignores `.Rhistory`, `.RData` and
  `.Rproj.user`, but not `data/`. That is why step 2 matters.
- **Step 2**: if the CSV still appears, the ignore pattern does not match its
  path. `git check-ignore -v data/file.csv` in the terminal tells you which
  rule, if any, applies to a file.
- **Step 4**: in the diff, the old line is red and the new one green. This is
  the view to check before every commit.
- **Step 5**: **Revert** in RStudio, **Discard Changes** in Positron. The
  discarded change is gone for good, since it was never committed.
- **Step 6**: the clock icon in RStudio; the **Timeline** in Positron's
  Explorer, or the **Graph** in the Source Control view.
:::
:::

::: {.callout-note collapse="true"}
## Exercise 2: a conflict on purpose

Conflicts are much less frightening once you have made one yourself.

1. In the repository from Exercise 1, create a branch called `experiment` and
   change line 2 of `analysis.R`. Commit.
2. Switch back to `main`, change **the same line** differently, and commit.
3. Merge `experiment` into `main` (`git merge experiment` in the terminal).
4. Resolve the conflict, then commit.

::: {.callout-tip collapse="true"}
## Solution

```bash
git switch -c experiment
# edit line 2, then
git commit -am "Try a different summary"

git switch main
# edit line 2 differently, then
git commit -am "Use the median instead of the mean"

git merge experiment
# CONFLICT (content): Merge conflict in analysis.R
```

`analysis.R` now contains both versions of line 2 between `<<<<<<<`,
`=======` and `>>>>>>>`. Edit the file so it holds the version you want, remove
the three marker lines, then:

```bash
git add analysis.R
git commit -m "Merge experiment: keep the median"
```

In Positron the conflicted file is listed under **Merge Changes** in the Source
Control view; open it and use the merge editor instead of editing the markers
by hand. `git commit -am` stages every tracked file that changed and commits in
one step; it does not add new files.
:::
:::
