# Version Control {#sec-git}

Version control is not a large part of this book, but it is worth a mention as it is a good practice to use in a reproducible workflow. It is an important part of modern software development and data analysis, particularly when projects involve multiple collaborators or making code publicly available (trancparancy). Version control systems help manage changes to source code or data files by keeping track of modifications and allowing users to revert to previous versions if needed. 

### Tracking Changes
Version control systems allow you to track changes made to files over time. This makes it easier to see who changed what and when, which is invaluable in diagnosing and understanding changes that affect the performance or functionality of your projects.

### Collaboration
For projects involving multiple people, version control is crucial as it allows team members to work independently on files without fear of conflicting changes. Tools like Git provide mechanisms such as branches and merges, enabling seamless integration of work from different team members.

### Experimentation
With version control, data scientists and programmers can experiment with different approaches without the risk of losing their original work. You can create branches to experiment with new ideas, and safely merge them back to the main project only if they prove to be beneficial.

### Git
Git is the most widely used version control system today. It is highly flexible and supports both local and remote repositories. GitHub, a web-based service built on Git, extends its functionality by providing a graphical interface and additional tools for project management.

## Using Version Control in R Projects

When working on R projects, integrating version control with your workflow can significantly improve the manageability and reproducibility of your work. RStudio, a popular IDE for R, includes integrated support for Git and Subversion, making it easier to track changes, revert to previous states, and collaborate with others.

### Before either IDE: install and configure Git

Both editors drive the same Git underneath, so this part is identical.

1. **Install Git** from [git-scm.com](https://git-scm.com/downloads) if you do
   not already have it. On macOS, running `git --version` in a terminal will
   offer to install it.
2. **Tell Git who you are.** This only has to be done once per computer, and the
   name and email end up in every commit you make:

```{r}
#| eval: false
usethis::use_git_config(
  user.name = "Your Name",
  user.email = "you@example.com"
)
```

3. **Connect to GitHub** if you want a remote copy. The `usethis` package
   handles the fiddly parts:

```{r}
#| eval: false
usethis::create_github_token() # opens GitHub to create a personal access token
gitcreds::gitcreds_set() # paste the token in when prompted
```

::: {.callout-tip}
## `usethis` works the same in both IDEs
`usethis::use_git()`, `usethis::use_github()` and `usethis::git_sitrep()` are
plain R functions, so they behave identically in RStudio and Positron. When in
doubt, `git_sitrep()` prints a diagnosis of your whole Git and GitHub setup and
tells you what is missing.
:::

### Setting Up Git with RStudio

1. **Initialise a repository**: in an existing project, run
   `usethis::use_git()`, or use **Tools → Version Control → Project Setup** and
   choose Git. RStudio will offer to restart.
2. **Use the Git pane** that appears in the top-right quadrant. Tick the files
   you want to include (this is *staging*), click **Commit**, write a message,
   and commit.
3. **Push and pull** with the arrows in the same pane.
4. **View history** with the clock icon, and see what changed in a file with
   **Diff**.

[More on version control in RStudio](https://docs.posit.co/ide/user/ide/guide/tools/version-control.html)

### Setting Up Git with Positron

Positron inherits VS Code's Git integration, so the concepts are the same but
the buttons are in different places (@sec-positron).

1. **Open the folder** as a workspace — **File → Open Folder**. Positron detects
   an existing `.git` directory automatically.
2. **Initialise a repository** if there is not one yet: open the **Source
   Control** view in the activity bar on the left (or `Ctrl/Cmd + Shift + G`)
   and click **Initialize Repository**. `usethis::use_git()` in the console does
   the same thing and also sets up a sensible `.gitignore`.
3. **Stage and commit** in the Source Control view. Hover a changed file and
   click **+** to stage it, type a commit message in the box at the top, and
   press `Ctrl/Cmd + Enter` to commit.
4. **Push and pull** from the **...** menu in the Source Control view, or from
   the sync button in the status bar at the bottom.
5. **View the diff** by clicking a file name in the Source Control list — it
   opens a side-by-side comparison, which is one of the things Positron does
   noticeably better than RStudio.

::: {.callout-note}
## Differences worth knowing
- **Staging is explicit.** RStudio's checkbox and Positron's **+** button do the
  same thing (`git add`), but in Positron you can also stage *part* of a file by
  selecting lines in the diff view and choosing "Stage Selected Ranges". That is
  genuinely useful when one editing session produced two unrelated changes.
- **A Positron-only project has no `.Rproj` to commit**, unless you added one
  with `usethis::use_rstudio()`. Either way is fine: `.git` is itself a project
  root marker, so `here()` works regardless (@sec-positron).
- **The terminal is always there.** Anything the interface cannot do,
  `git` in the terminal can. Positron's integrated terminal opens with
  `Ctrl + backtick`.
:::

::: {.callout-important}
## Add data to `.gitignore` before your first commit
This matters more than which IDE you use. Removing a file in a later commit does
**not** remove it from the history — the data is still there for anyone who
clones the repository.

`usethis::use_git_ignore(c("data/", "data-raw/", "*.csv", "*.sav", "*.rds"))`
adds the entries, and `usethis::use_git()` sets up a reasonable default
`.gitignore` when it initialises the repository. Check `git status` before your
first commit and confirm that no data file is listed.
:::

[Happy Git and GitHub for the useR](https://happygitwithr.com/) is the standard
reference for all of this, and it covers both the setup and what to do when it
goes wrong.

::: {.callout-tip}  
# Git and GitHub are not the same.

- **Git** is a distributed version control system, which means it helps you manage and track versions of your project. It operates locally on your computer, allowing you to manage your code history in a series of snapshots called commits. Git provides the tools necessary for you to create branches, merge changes, and revert to previous versions of your project, all from your local environment.

- **GitHub**, on the other hand, is a hosting service for Git repositories. It extends Git's functionality by adding a web-based graphical interface, along with collaboration features like bug tracking, feature requests, task management, and wikis for every project. GitHub makes it easier to collaborate with others on projects that are managed using Git by storing a copy of your Git repository on the internet, facilitating shared access and team collaboration.
:::
