# Importing and Exporting Data {#sec-import}

::: {.callout-note}
## What this chapter covers
This chapter follows the lecture *Importera och exportera data*. It covers the
four ways to get data into R, why `setwd()` should be avoided, how the `here`
package solves the file-path problem, and which format to save your results in.
:::

```{r}
#| label: setup-import
#| include: false
library(webexercises)
```

```{r}
#| label: import-libs
#| message: false
library(tidyverse)
library(here)
```

Reading data in is the first line of almost every analysis script, and it is
where a surprising number of problems begin. A file path that works on your
laptop and nowhere else, a column of numbers imported as text, a Swedish decimal
comma silently turning 3,5 into a character string — all of these are import
problems, and all of them are cheaper to fix here than three steps later.

## Four ways to import

### 1. By clicking

RStudio has a graphical import wizard:

- **File → Import Dataset**
- Choose the file type, for example CSV or Excel.
- Follow the dialogue. You get a preview of the data, which is genuinely useful
  for checking delimiters and whether the first row is a header.
- RStudio generates the R code in the bottom-right of the dialogue.

**Copy that generated code into your script.** Clicking through the wizard again
next week is not reproducible; the line of code it produces is.

### 2. With base R

```{r}
#| eval: false
data_base <- read.csv("data/my_file.csv", header = TRUE)
head(data_base)
```

- **Advantage**: built in, no packages needed.
- **Disadvantage**: slower on large files, and older R versions converted text
  to factors automatically, which surprised a generation of users.

### 3. With `readr` (tidyverse)

```{r}
#| eval: false
library(readr)

data_tidy <- read_csv("data/my_file.csv")
```

- Faster and more predictable.
- Returns a tibble rather than a data frame.
- Guesses column types and, importantly, **tells you what it guessed** in a
  message. Read that message.

::: {.callout-important}
## Swedish CSV files
Swedish Excel exports usually use a **semicolon** as the field separator and a
**comma** as the decimal mark. `read_csv()` expects commas and periods, so it
will read the whole file into one column, or turn your numbers into text.

Use `read_csv2()` instead — it expects exactly the Swedish convention.

```{r}
#| eval: false
data_comma <- read_csv(here("data", "data_comma.csv")) # , separator, . decimal
data_semicolon <- read_csv2(here("data", "data_sv.csv")) # ; separator, , decimal
```

If you are unsure, open the file in a plain text editor and look at the first
two lines. This takes ten seconds and saves an hour.
:::

### 4. Other file types

```{r}
#| eval: false
# Excel
library(readxl)
data_excel <- read_excel(here("data", "my_file.xlsx"), sheet = 1)

# SPSS, Stata, SAS
library(haven)
data_spss <- read_sav(here("data", "my_file.sav"))
data_stata <- read_dta(here("data", "my_file.dta"))
data_sas <- read_sas(here("data", "my_file.sas7bdat"))

# tab-separated text
data_txt <- read_tsv(here("data", "my_file.txt"))

# any delimiter you like
data_odd <- read_delim(here("data", "my_file.txt"), delim = "|")
```

::: {.callout-tip}
## `haven` keeps the SPSS labels
When you import an `.sav` file, `haven` preserves SPSS value labels as a
`labelled` class. That is useful — the labels are your codebook — but `labelled`
columns behave oddly in `ggplot2` and some models. Convert them explicitly when
you are ready: `haven::as_factor()` for categorical variables,
`haven::zap_labels()` to strip labels from numeric ones.
:::

## Useful arguments for reading files

Real files are rarely clean. These arguments handle most of what goes wrong:

```{r}
#| eval: false
read_csv(
  file = here("data", "messy.csv"),
  skip = 3, # skip 3 junk rows above the header
  na = c("", "NA", "999", "missing"), # values that mean "missing"
  col_types = cols( # override the guesses
    participant_id = col_character(), # keep leading zeros!
    date = col_date(format = "%Y-%m-%d"),
    weight_kg = col_double()
  ),
  locale = locale(encoding = "latin1") # for old files with broken a, a, o
)
```

::: {.callout-warning}
## ID numbers are text, not numbers
An ID like `00734` read as a number becomes `734`, and it will never join
correctly against the register file that still has the leading zeros. Personal
identity numbers have the same problem, plus they overflow the precision of a
double. Always read identifiers as `col_character()`.
:::

## Do not use `setwd()`

```{r}
#| eval: false
setwd("C:/Users/daniel/OneDrive/projects/study_2024/data")
data <- read.csv("my_file.csv")
```

Why this is a bad idea:

- **Fragile**: works only on one machine, in one folder.
- **Hard to maintain**: rename or move the folder and the script breaks.
- **Not portable**: a new computer, a co-author, or the same project on a server
  all require editing the path.
- **Not reproducible**: the reviewer who downloads your code cannot run it.

## The `here` package

`here()` builds paths relative to the **root of your R project**, which it finds
by looking for the `.Rproj` file. The same line of code then works on every
machine.

```{r}
#| eval: false
library(here)

data_here <- read_csv(here("data", "my_file.csv"))
```

Advantages:

- Readable and reproducible.
- Integrates with RStudio projects.
- The project can be shared, moved, or synced through OneDrive and still work.
- You can step out of the project with `".."`, e.g.
  `here("..", "shared_data", "register.csv")`.

```{r}
#| label: here-demo
# here() reports the project root it has found
here()
```

::: {.callout-tip}
## The recommended setup
1. Always create an R project (**File → New Project**).
2. Use `here()` for every file path.
3. Use `readr` functions for text files.
4. Never use an absolute path in a script you intend to keep.

If you work on several computers and keep the project in a cloud folder, this
combination is the difference between code that runs everywhere and code that
runs on one laptop.
:::

## Saving data

You save data to avoid redoing work, to share results, and to separate slow
steps from fast ones. In a multi-step project, saving the cleaned dataset means
you never have to re-run the cleaning to make one more figure.

### CSV

```{r}
#| eval: false
# base R
write.csv(data, "data_output.csv", row.names = FALSE)

# readr - faster, and no row-names nonsense
write_csv(data, here("data", "data_output.csv"))
```

### RDS

RDS saves a single R object exactly as it is, preserving factor levels, column
types, attributes and the tibble class.

```{r}
#| eval: false
saveRDS(data, here("data", "data_output.rds"))
data <- readRDS(here("data", "data_output.rds"))
```

This is the right format for intermediate results — a cleaned dataset, or a
model that took twenty minutes to fit.

### Excel

```{r}
#| eval: false
library(writexl)
write_xlsx(data, here("output", "data_output.xlsx"))
```

## Which format should you choose?

| Format | Advantages | Disadvantages |
|--------|------------|---------------|
| **CSV** | Readable by every program; plain text; diffs nicely in Git | Loses all metadata — factor levels, date formats, column types |
| **RDS** | Preserves everything R-specific exactly | Only readable in R; binary, so Git cannot diff it |
| **XLSX** | Easy to share with Excel users; multiple sheets | Inefficient for large data; invites manual editing |
| **Parquet** | Fast, compressed, keeps types; readable from R and Python | Needs the `arrow` package; not human-readable |

Recommendations:

- **To share with others**: CSV, or Excel if the recipient insists.
- **To preserve types, or to save a model object**: RDS.
- **For large datasets**: Parquet via the `arrow` package.
- Use `readr` and `here` throughout.

::: {.callout-important}
## Never overwrite your raw data
Raw data is written once and read forever. Keep it in a separate folder, treat
it as read-only, and write every cleaned version to a different file. If a
cleaning step turns out to be wrong, you want to be able to go back — and you
cannot go back to a file you have already overwritten.

For registry data under ethical approval, this is not just good practice. Add
`data/` and `*.csv`, `*.sav`, `*.rds` to `.gitignore` **before** your first
commit, so that identifiable data never enters version control.
:::

## `::` without `library()`

```{r}
#| eval: false
write_csv(data, here::here("data", "data_output.csv"))
```

Writing `here::here()` calls the function directly from the package without
attaching the whole package with `library()`. Two reasons to do it:

- It makes the source of an unfamiliar function obvious to your reader.
- It avoids conflicts when two packages export the same function name — the
  classic case being `dplyr::filter()` versus `stats::filter()`.

::: {.webex-check .webex-box}
**Check yourself**

Which function reads a Swedish CSV file with `;` separators and `,` decimals?
`r mcq(c("read_csv()", answer = "read_csv2()", "read.csv()", "read_tsv()"))`

`here()` builds paths relative to your project root. `r torf(TRUE)`

Which format preserves factor levels and column types exactly?
`r mcq(c("CSV", answer = "RDS", "XLSX", "TXT"))`

Why should a participant ID be read as character rather than numeric?
`r longmcq(c(
  "Because character columns take less memory",
  answer = "Because leading zeros are lost when an ID is read as a number, which breaks later joins",
  "Because dplyr cannot join on numeric columns",
  "There is no reason; numeric is preferable"
))`
:::

::: {.callout-note collapse="true"}
## Exercise 1: a full round trip

Practise the import-modify-export cycle without needing any external file.

1. Create a folder `data/` in your project.
2. Write the `penguins` dataset to `data/penguins_raw.csv` using `write_csv()`
   and `here()`.
3. Read it back in with `read_csv()`. Look at the message about column types.
4. Compare `class(penguins$species)` before saving with the class after reading
   back in. What happened, and why?
5. Save the same data as RDS instead and repeat step 4.

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| eval: false
library(tidyverse)
library(palmerpenguins)
library(here)

dir.create(here("data"), showWarnings = FALSE)

write_csv(penguins, here("data", "penguins_raw.csv"))
penguins_csv <- read_csv(here("data", "penguins_raw.csv"))

class(penguins$species) # "factor"
class(penguins_csv$species) # "character"

saveRDS(penguins, here("data", "penguins_raw.rds"))
penguins_rds <- readRDS(here("data", "penguins_raw.rds"))

class(penguins_rds$species) # "factor"
```

A CSV file is plain text. It has no way to record that `species` was a factor
with three specific levels in a specific order, so `read_csv()` reads it back as
character. RDS stores the R object itself, so everything survives.

This matters in practice: if you carefully ordered your factor levels during
cleaning and then round-trip through CSV, that ordering is silently gone, and
your figures and model reference groups change.
:::
:::

::: {.callout-note collapse="true"}
## Exercise 2: diagnose the import

A colleague sends you a file and complains that "R has broken the numbers".
Their code and the first lines of the file look like this:

```
id;weight;height;group
001;72,5;181;kontroll
002;68,0;175;intervention
```

```{r}
#| eval: false
d <- read_csv("data/study.csv")
```

1. Predict what goes wrong. There are at least three separate problems.
2. Write the corrected import call.

::: {.callout-tip collapse="true"}
## Solution

The problems:

1. The separator is `;`, not `,`. `read_csv()` will read every line into a
   single column.
2. The decimal mark is `,`. Even with the right separator, `72,5` would be read
   as text.
3. `id` has leading zeros. Read as a number it becomes `1`, `2` and will not
   join to anything.

```{r}
#| eval: false
d <- read_csv2(
  here("data", "study.csv"),
  col_types = cols(
    id = col_character(), # keep the leading zeros
    weight = col_double(),
    height = col_double(),
    group = col_factor()
  )
)
```

`read_csv2()` handles the first two problems in one function; `col_types` fixes
the third. Always run `glimpse()` straight after importing and confirm that
every column has the type you expected.
:::
:::
