# Data Transformation with dplyr {#sec-dplyr}

::: {.callout-note}
## What this chapter covers
This chapter follows the lecture *Tidyverse verb*. It works through the core
`dplyr` verbs — `arrange()`, `filter()`, `slice()`, `select()`, `mutate()`,
`summarise()` — and grouped operations with `group_by()` and `.by`.
:::

```{r}
#| label: setup-dplyr
#| include: false
library(webexercises)
```

```{r}
#| label: dplyr-libs
#| message: false
library(tidyverse)
library(palmerpenguins)
```

## Two ways to think about transformation

Data transformation can be grouped along two dimensions, and it is worth being
able to place any operation on both.

**Reshaping versus reducing:**

- **Reshaping** changes the *form* of the data without changing the information.
  You can go back and forth. Converting a data frame to a tibble, or pivoting
  between wide and long, are reshaping operations. Form matters because some
  functions require a particular shape.
- **Reducing** changes the data by removing information. Both structure and
  content change, and you usually **cannot** reconstruct the original. Filtering
  rows and summarising to group means are reducing operations.

**Orientation:** row-wise, column-wise, grouped, or whole-table.

| Function | What it does | Type | Orientation |
|----------|--------------|------|-------------|
| `arrange()` | sorts rows | reshapes | row-wise |
| `filter()` | keeps rows meeting a condition | reduces | row-wise |
| `slice()` | keeps rows by position | reduces | row-wise |
| `select()` | keeps and reorders columns | reshapes | column-wise |
| `rename()` | renames columns | reshapes | column-wise |
| `mutate()` | creates or modifies columns | reshapes | column-wise |
| `summarise()` | collapses to one row per group | reduces | grouped |
| `group_by()` / `.by` | defines groups | both | grouped |

::: {.callout-tip}
## Why this distinction is worth keeping
Every reducing step is a decision that a reader might question. Filtering out
participants, dropping missing values, collapsing categories — these change what
your results can support. Reshaping steps rarely need justification; reducing
steps always do. Comment them.
:::

## The tidyverse

The tidyverse is a collection of packages designed to work together:

- **ggplot2** — visualisation
- **dplyr** — data manipulation
- **tidyr** — reshaping
- **readr** — importing files
- **tibble** — modern data frames
- **stringr** — text
- **forcats** — factors
- **purrr** — functional programming

`library(tidyverse)` loads all of them at once.

## The pipe

A pipe takes the object on its left and passes it as the first argument to the
function on its right. Read it as **"and then"**.

There are two pipes:

- `%>%` — the magrittr pipe, introduced in 2014.
- `|>` — the native pipe, added in R 4.1 (2021).

They behave the same way in almost all everyday use. This book uses `|>`,
because it is part of R itself and needs no package. The keyboard shortcut is
`Ctrl/Cmd + Shift + M` in both RStudio and Positron; which pipe it inserts
depends on a setting in each.

```{r}
#| label: pipe-compare
# without a pipe: read from the inside out
head(arrange(filter(penguins, species == "Gentoo"), body_mass_g), 3)

# with a pipe: read from top to bottom
penguins |>
  filter(species == "Gentoo") |>
  arrange(body_mass_g) |>
  head(3)
```

The second version reads as instructions: take the penguins, **and then** keep
the Gentoos, **and then** sort by body mass, **and then** show the first three.

::: {.callout-tip}
## One verb per line
Put each verb on its own line, ending with the pipe. This makes the chain
readable, makes it obvious where a step is missing, and — most usefully — lets
you comment out a single line to see what the pipeline looks like without it.
:::

## `arrange()`: sorting rows

```{r}
#| label: arrange-basic
penguins |>
  arrange(body_mass_g) |>
  head(5)
```

`desc()` sorts in descending order:

```{r}
#| label: arrange-desc
penguins |>
  arrange(desc(body_mass_g)) |>
  head(5)
```

Multiple variables sort hierarchically — first by the first, ties broken by the
second:

```{r}
#| label: arrange-multi
penguins |>
  arrange(species, desc(body_mass_g)) |>
  head(5)
```

Missing values are always sorted to the end, regardless of direction.

## `filter()`: keeping rows

`filter()` keeps the rows for which a condition is `TRUE`. Rows where the
condition is `NA` are **dropped**, which is worth remembering.

::: {.panel-tabset}
## Equal to

```{r}
#| label: filter-eq
penguins |>
  filter(species == "Gentoo") |>
  head(3)
```

## Greater than

```{r}
#| label: filter-gt
penguins |>
  filter(body_mass_g > 4000) |>
  head(3)
```

## AND

A comma and `&` mean the same thing: both conditions must hold.

```{r}
#| label: filter-and
penguins |>
  filter(species != "Gentoo", body_mass_g >= 4000) |>
  head(3)
```

## OR

```{r}
#| label: filter-or
penguins |>
  filter(species == "Chinstrap" | island == "Dream") |>
  head(3)
```

## Combined

```{r}
#| label: filter-combined
penguins |>
  filter(
    (species == "Gentoo" & body_mass_g > 5500) |
      (species == "Adelie" & body_mass_g > 4500)
  )
```

## Missing values

```{r}
#| label: filter-na
penguins |>
  filter(is.na(sex)) |>
  head(5)
```

## `%in%` and `between()`

```{r}
#| label: filter-in
penguins |>
  filter(
    species %in% c("Adelie", "Chinstrap"),
    between(body_mass_g, 4000, 4250)
  )
```
:::

::: {.callout-warning}
## `filter(x == NA)` never works
`NA == NA` is `NA`, not `TRUE` — R cannot confirm that two unknown values are
equal. To find missing values use `is.na(x)`; to exclude them use `!is.na(x)` or
`drop_na(x)`.

Equally important in the other direction: `filter(sex == "female")` silently
drops the eleven penguins whose sex is missing. If you then report a total, it
will not match the original sample size. Count before and after.
:::

### `filter_out()`: say what to drop

New in **dplyr 1.2.0**. Use `filter()` to say which rows to *keep*, and
`filter_out()` to say which rows to *drop*.

That sounds like a cosmetic difference. It is not, because of how `NA` behaves.

```{r}
#| label: filter-out-motivation
counts <- tibble(id = 1:5, count = c(3, 0, NA, 7, 0))

# with filter(): you must negate the condition AND remember to protect the NA
counts |>
  filter(count != 0 | is.na(count))

# with filter_out(): say what you want gone, and NA is kept automatically
counts |>
  filter_out(count == 0)
```

The `filter()` version needs two things to be right at once: the negation
(`!= 0`) and the explicit `| is.na(count)`. Forget the second and the missing
row disappears silently, which is exactly the class of bug that changes a sample
size without anyone noticing.

```{r}
#| label: filter-out-penguins
# keep everything except Gentoo
penguins |>
  filter_out(species == "Gentoo") |>
  count(species)
```

::: {.callout-tip}
## When to reach for `filter_out()`
Whenever your `filter()` condition contains a `!`, a `!=`, or a `%in%` you were
about to negate. If you are describing what you want *gone*, write it that way.
Exclusions are also easier for a reviewer to check when they read as exclusions.
:::

### `when_any()` and `when_all()`

Also new in dplyr 1.2.0. These are elementwise versions of `any()` and `all()` —
think of them as `|` and `&` applied across any number of conditions.

```{r}
#| label: when-any
# the old way: explicit |, extra parentheses, uneven indentation
penguins |>
  filter(
    (species == "Gentoo" & body_mass_g > 5500) |
      (species == "Adelie" & body_mass_g > 4500)
  ) |>
  nrow()

# the same thing, with every condition at the same indentation level
penguins |>
  filter(when_any(
    species == "Gentoo" & body_mass_g > 5500,
    species == "Adelie" & body_mass_g > 4500
  )) |>
  nrow()
```

`when_all()` does the same for `&`. Both combine with `filter_out()`, which is
where they earn their keep — a list of exclusion criteria reads as a list:

```{r}
#| label: when-any-out
penguins |>
  filter_out(when_any(
    is.na(sex), # sex not recorded
    is.na(body_mass_g), # no body mass measurement
    body_mass_g < 3000 # below the pre-specified inclusion threshold
  )) |>
  nrow()
```

That is a study's exclusion criteria, written once, each on its own commented
line. Compare it with the equivalent chain of negated `filter()` calls.

## `slice()`: rows by position

::: {.panel-tabset}
## slice()

```{r}
#| label: slice-basic
penguins |>
  slice(1:5)
```

## slice_head() / slice_tail()

```{r}
#| label: slice-head
penguins |>
  slice_head(n = 3)

penguins |>
  slice_tail(n = 3)
```

## slice_min() / slice_max()

```{r}
#| label: slice-min
penguins |>
  slice_min(body_mass_g, n = 3)

penguins |>
  slice_max(body_mass_g, n = 3)
```

## slice_sample()

```{r}
#| label: slice-sample
set.seed(42) # a seed makes the random draw reproducible

penguins |>
  slice_sample(n = 5)
```
:::

`slice_min()` and `slice_max()` combine with grouping to answer questions like
"the heaviest penguin of each species":

```{r}
#| label: slice-grouped
penguins |>
  slice_max(body_mass_g, n = 1, by = species) |>
  select(species, island, sex, body_mass_g)
```

## `select()`: choosing columns

::: {.panel-tabset}
## By name

```{r}
#| label: select-name
penguins |>
  select(species, island, body_mass_g) |>
  head(3)
```

## By position

```{r}
#| label: select-pos
# works, but breaks silently if the column order changes upstream
penguins |>
  select(5, 1) |>
  head(3)
```

## A range

```{r}
#| label: select-range
penguins |>
  select(species:bill_depth_mm) |>
  head(3)
```

## Reordering

```{r}
#| label: select-everything
# put year first, keep everything else in its current order
penguins |>
  select(year, everything()) |>
  head(3)
```

## Helpers

```{r}
#| label: select-helpers
penguins |>
  select(starts_with("bill")) |>
  head(3)

penguins |>
  select(ends_with("_mm")) |>
  head(3)

penguins |>
  select(contains("length")) |>
  head(3)
```

## By type

```{r}
#| label: select-where
penguins |>
  select(where(is.numeric)) |>
  head(3)

penguins |>
  select(where(is.factor)) |>
  head(3)
```

## Dropping

```{r}
#| label: select-drop
penguins |>
  select(-year, -island) |>
  head(3)
```
:::

`rename()` renames without dropping anything, and `relocate()` moves columns:

```{r}
#| label: rename-relocate
penguins |>
  rename(mass_g = body_mass_g) |>
  relocate(mass_g, .after = species) |>
  head(3)
```

::: {.callout-tip}
## Select by name, not by number
`select(5, 1)` works today and breaks the day someone adds a column upstream —
without an error, just wrong results. Names are self-documenting and survive
changes to the source file.
:::

## `mutate()`: creating columns

`mutate()` adds new columns or modifies existing ones. It works element by
element, using the vectorised arithmetic from @sec-classes.

::: {.panel-tabset}
## A new variable

```{r}
#| label: mutate-new
penguins |>
  mutate(body_mass_kg = body_mass_g / 1000) |>
  select(species, body_mass_g, body_mass_kg) |>
  head(3)
```

## Using a summary

```{r}
#| label: mutate-mean
# a summary function inside mutate() is recycled across all rows,
# which is exactly what you want for centring a variable
penguins |>
  mutate(mass_centred = body_mass_g - mean(body_mass_g, na.rm = TRUE)) |>
  select(body_mass_g, mass_centred) |>
  head(3)
```

## Row numbers

```{r}
#| label: mutate-rownum
penguins |>
  mutate(row_id = row_number()) |>
  select(row_id, species, island) |>
  head(3)
```

## Several columns at once

```{r}
#| label: mutate-across
# z-standardise every numeric column
penguins |>
  mutate(across(where(is.numeric), \(x) as.numeric(scale(x)))) |>
  select(where(is.numeric)) |>
  head(3)
```

## `if_else()`

```{r}
#| label: mutate-ifelse
penguins |>
  mutate(size_category = if_else(body_mass_g > 4000, "Large", "Small")) |>
  select(species, body_mass_g, size_category) |>
  head(3)
```

## `case_when()`

```{r}
#| label: mutate-casewhen
penguins |>
  mutate(
    size_category = case_when(
      body_mass_g < 3500 ~ "Small",
      body_mass_g < 4500 ~ "Medium",
      body_mass_g >= 4500 ~ "Large",
      .default = NA_character_ # anything unmatched, including NA
    )
  ) |>
  count(size_category)
```
:::

::: {.callout-warning}
## `case_when()` evaluates in order and stops at the first match
That is why the conditions above can be written as `< 3500`, `< 4500`,
`>= 4500` — a penguin of 4000 g fails the first test, matches the second, and
never reaches the third. It also means an accidentally broad first condition
swallows everything below it.

Always finish with `.default =`. Without it, unmatched rows become `NA`
silently, and `NA` is one of the values that will not match any condition.
Verify with `count()` afterwards, every time.
:::

### The recoding family

Until dplyr 1.2.0, `case_when()` did nearly everything and `case_match()`
handled simple value mapping. dplyr 1.2.0 splits the job into four functions
along two axes — *create a new vector* versus *update an existing one*, and
*conditions* versus *value mapping*:

| | Based on **conditions** | Based on **values** |
|---|---|---|
| **Create** a new vector | `case_when()` | `recode_values()` |
| **Update** an existing vector | `replace_when()` | `replace_values()` |

```{r}
#| label: recode-family
penguins |>
  mutate(
    # create a new variable from conditions
    size = case_when(
      body_mass_g < 3500 ~ "Light",
      body_mass_g < 4500 ~ "Medium",
      .default = "Heavy"
    ),
    # create a new variable by mapping values
    species_code = recode_values(
      species,
      "Adelie" ~ "ADE",
      "Chinstrap" ~ "CHI",
      "Gentoo" ~ "GEN"
    ),
    # update an existing variable, leaving the rest untouched
    sex = replace_values(as.character(sex), "female" ~ "F", "male" ~ "M")
  ) |>
  count(species_code, sex, size) |>
  head(5)
```

::: {.callout-warning}
## These functions will not silently widen a factor
`replace_values(island, "Torgersen" ~ "Torgersen Island")` errors, because
`island` is a factor and `"Torgersen Island"` is not one of its levels. That is
deliberate: silently inventing a level is how category lists drift apart between
datasets.

For factors, either convert to character first (as above) or use
`forcats::fct_recode()`, which is built for the job (@sec-strings-factors).
:::

`replace_when()` updates values where a condition holds and leaves everything
else, including `NA`, alone:

```{r}
#| label: replace-when
weights <- c(72.5, 3.0, 68.0, NA, 270.0)

# only the impossible values are touched; the NA stays an NA
replace_when(weights, weights < 35 | weights > 250 ~ NA)
```

Written with `if_else()` that would be
`if_else(weights < 35 | weights > 250, NA, weights)`, which requires you to
repeat the variable name and think about what happens to the `NA`.

`recode_values()` also accepts a lookup table through `from` and `to`, which is
how you should apply a recoding scheme that lives in a file rather than in your
script:

```{r}
#| label: recode-lookup
lookup <- tibble(
  code = c("Adelie", "Chinstrap", "Gentoo"),
  latin = c("Pygoscelis adeliae", "Pygoscelis antarcticus", "Pygoscelis papua")
)

penguins |>
  mutate(
    species_latin = recode_values(
      species,
      from = lookup$code,
      to = lookup$latin
    )
  ) |>
  count(species, species_latin)
```

::: {.callout-note}
## `case_match()` is on its way out
`case_match()` is soft-deprecated in dplyr 1.2.0 and fully replaced by
`recode_values()` and `replace_values()`. It still works, and you will meet it
in existing code and older tutorials — but write the new names in new code.
:::

### `.unmatched = "error"`

The safety feature worth adopting immediately. When you believe you have handled
every possible case, say so, and let R check you:

```{r}
#| label: unmatched-error
#| error: true
mass <- c(3200, 4100, 5300)

case_when(
  mass < 3500 ~ "Light",
  mass < 4500 ~ "Medium",
  .unmatched = "error"
)
```

The error names the exact position that fell through. With `.default =` the same
mistake produces a silent category you did not intend; with
`.unmatched = "error"` it stops the script.

Use `.default =` when there genuinely is a sensible catch-all, and
`.unmatched = "error"` when there should not be one. Both are better than
neither.

## `summarise()`: collapsing rows

```{r}
#| label: summarise-one
penguins |>
  summarise(mean_body_mass = mean(body_mass_g, na.rm = TRUE))
```

Several summaries at once. Note that counting non-missing values alongside the
mean is not optional — a mean of 12 observations and a mean of 300 are very
different things:

```{r}
#| label: summarise-many
penguins |>
  summarise(
    n_rows = n(),
    n_bill_length = sum(!is.na(bill_length_mm)),
    mean_bill_length = mean(bill_length_mm, na.rm = TRUE),
    sd_bill_length = sd(bill_length_mm, na.rm = TRUE),
    n_body_mass = sum(!is.na(body_mass_g)),
    mean_body_mass = mean(body_mass_g, na.rm = TRUE),
    sd_body_mass = sd(body_mass_g, na.rm = TRUE),
    n_species = n_distinct(species)
  )
```

### `across()`: the same summary for many columns

```{r}
#| label: summarise-across1
penguins |>
  summarise(across(where(is.numeric), \(x) mean(x, na.rm = TRUE)))
```

Several functions at once, producing named columns:

```{r}
#| label: summarise-across2
penguins |>
  summarise(
    across(
      where(is.numeric),
      list(
        mean = \(x) mean(x, na.rm = TRUE),
        sd = \(x) sd(x, na.rm = TRUE)
      )
    )
  ) |>
  glimpse()
```

::: {.callout-note}
## The `\(x)` syntax
`\(x) mean(x, na.rm = TRUE)` is R's shorthand for an anonymous function,
available since R 4.1. It is equivalent to `function(x) mean(x, na.rm = TRUE)`.

Passing `na.rm = TRUE` directly to `across()` — as in
`across(where(is.numeric), mean, na.rm = TRUE)` — was deprecated in dplyr 1.1
and will eventually stop working. Use the lambda.
:::

## Grouping

Almost every interesting question is a grouped question: not "what is the mean
body mass" but "how does mean body mass differ between species".

There are two syntaxes, and both are current.

### `group_by()` + `ungroup()`

```{r}
#| label: group-by
penguins |>
  group_by(species) |>
  summarise(
    n = n(),
    mean_body_mass = mean(body_mass_g, na.rm = TRUE),
    sd_body_mass = sd(body_mass_g, na.rm = TRUE)
  ) |>
  ungroup()
```

`group_by()` attaches a grouping to the data that persists until you remove it.
That persistence is the source of a great many confusing bugs.

### `.by`: per-operation grouping

Since dplyr 1.1, most verbs accept a `.by` argument that groups for **that one
operation only**. Nothing persists, so nothing needs ungrouping.

```{r}
#| label: dot-by
penguins |>
  summarise(
    n = n(),
    mean_body_mass = mean(body_mass_g, na.rm = TRUE),
    .by = species
  )
```

Multiple grouping variables:

```{r}
#| label: dot-by-multi
penguins |>
  summarise(
    n = n(),
    mean_body_mass = mean(body_mass_g, na.rm = TRUE),
    .by = c(species, island)
  )
```

::: {.callout-tip}
## Prefer `.by` unless you need persistence
`.by` removes an entire category of bug, keeps the pipeline shorter, and always
returns an ungrouped result. Reach for `group_by()` when you genuinely need the
grouping to survive several steps, or when you are reading older code.
:::

### Grouped `mutate()`

`summarise()` collapses to one row per group. Grouped `mutate()` keeps every row
but computes the summary within each group and recycles it back — the standard
way to centre or standardise within groups.

```{r}
#| label: group-mutate
penguins |>
  mutate(mean_mass_species = mean(body_mass_g, na.rm = TRUE), .by = species) |>
  select(species, body_mass_g, mean_mass_species) |>
  head(5)
```

Z-scores within species:

```{r}
#| label: group-z
penguins |>
  mutate(
    # standardising within species removes between-species size differences
    body_mass_z = (body_mass_g - mean(body_mass_g, na.rm = TRUE)) /
      sd(body_mass_g, na.rm = TRUE),
    .by = species
  ) |>
  select(species, body_mass_g, body_mass_z) |>
  head(5)
```

### Why `ungroup()` matters

With `group_by()`, the grouping stays attached. The second `mutate()` below is
still grouped, so `mean(bill_length_mm)` is a per-species mean, not an overall
one:

```{r}
#| label: ungroup-wrong
penguins |>
  group_by(species) |>
  mutate(mean_mass = mean(body_mass_g, na.rm = TRUE)) |>
  # still grouped here - this is a within-species mean, probably not intended
  mutate(mean_bill = mean(bill_length_mm, na.rm = TRUE)) |>
  select(species, mean_mass, mean_bill) |>
  distinct()
```

With `ungroup()` in the right place, the second mean is over the whole dataset:

```{r}
#| label: ungroup-right
penguins |>
  group_by(species) |>
  mutate(mean_mass = mean(body_mass_g, na.rm = TRUE)) |>
  ungroup() |>
  mutate(mean_bill = mean(bill_length_mm, na.rm = TRUE)) |>
  select(species, mean_mass, mean_bill) |>
  distinct()
```

Both results are legitimate; only one of them is what you meant. With `.by` the
question does not arise.

## `count()` and `distinct()`

```{r}
#| label: count-distinct
penguins |>
  count(species)

penguins |>
  count(species, island, sort = TRUE)

penguins |>
  distinct(species, island)
```

`count()` is the most under-used function in `dplyr`. Use it after every recode
to confirm that the categories are what you think they are.

## Chaining it all together

```{r}
#| label: fig-dplyr-chain
#| fig-cap: "A single pipeline from raw data to figure."
penguins |>
  filter(!is.na(sex)) |>
  mutate(body_mass_kg = body_mass_g / 1000) |>
  ggplot(aes(island, body_mass_kg, colour = sex)) +
  geom_violin(fill = NA) +
  geom_point(position = position_jitterdodge(jitter.width = 0.15), alpha = 0.4) +
  labs(x = "Island", y = "Body mass (kg)", colour = "Sex") +
  theme_minimal(base_size = 13)
```

::: {.webex-check .webex-box}
**Check yourself**

Which verb would you use to keep only rows where age is over 65?
`r mcq(c("select()", answer = "filter()", "slice()", "arrange()"))`

`filter(sex == "female")` keeps rows where sex is missing. `r torf(FALSE)`

Which is the modern way to group for a single operation?
`r mcq(c("group_by() then ungroup()", answer = ".by =", "rowwise()", "aggregate()"))`

After `summarise()` with `group_by(species, island)`, the result is still grouped
by species. `r torf(TRUE)`

To count how many penguins are on each island, you would write
`penguins |> ` `r fitb(c("count(island)", "count( island )"), ignore_ws = TRUE, width = 16)`
:::

::: {.callout-note collapse="true"}
## Exercise 1: the core verbs

Using `penguins`:

1. How many penguins are there of each species on each island?
2. What is the mean and standard deviation of `flipper_length_mm` per species,
   with the number of non-missing observations?
3. Which five penguins have the longest bills? Show species, island and bill
   length only.
4. Create a variable `bill_ratio` (bill length divided by bill depth) and report
   its mean per species.

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: ex1-dplyr
# 1
penguins |>
  count(species, island)

# 2 - reporting n alongside the mean is not optional; a mean of 68
# and a mean of 152 observations carry very different weight
penguins |>
  summarise(
    n = sum(!is.na(flipper_length_mm)),
    mean_flipper = mean(flipper_length_mm, na.rm = TRUE),
    sd_flipper = sd(flipper_length_mm, na.rm = TRUE),
    .by = species
  )

# 3
penguins |>
  slice_max(bill_length_mm, n = 5) |>
  select(species, island, bill_length_mm)

# 4
penguins |>
  mutate(bill_ratio = bill_length_mm / bill_depth_mm) |>
  summarise(mean_ratio = mean(bill_ratio, na.rm = TRUE), .by = species)
```
:::
:::

::: {.callout-note collapse="true"}
## Exercise 2: recoding, carefully

1. Create a categorical variable `mass_group` from `body_mass_g` with the cut
   points below 3500 g, 3500 to under 4500 g, and 4500 g or more.
2. Count how many penguins fall in each group. Do the counts add up to 344?
3. If not, find out why and fix it so that missing body mass becomes an explicit
   "Unknown" category rather than `NA`.

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: ex2-dplyr
penguins |>
  mutate(
    mass_group = case_when(
      body_mass_g < 3500 ~ "Light",
      body_mass_g < 4500 ~ "Medium",
      body_mass_g >= 4500 ~ "Heavy"
    )
  ) |>
  count(mass_group)
```

The counts do add up to 344, but only because `count()` reports the `NA` group
as its own row. Two penguins have no recorded body mass, and none of the three
conditions matches `NA` — so they fall through to the implicit default.

Making that explicit:

```{r}
#| label: ex2-dplyr-fix
penguins |>
  mutate(
    mass_group = case_when(
      is.na(body_mass_g) ~ "Unknown", # handle missing first, deliberately
      body_mass_g < 3500 ~ "Light",
      body_mass_g < 4500 ~ "Medium",
      .default = "Heavy"
    ),
    # an explicit level order controls how this appears in tables and figures
    mass_group = factor(
      mass_group,
      levels = c("Light", "Medium", "Heavy", "Unknown")
    )
  ) |>
  count(mass_group)
```

The lesson generalises: after any recode, `count()` the result and check that
the total matches the number of rows you started with. Recoding is where data
quietly disappears.
:::
:::

::: {.callout-note collapse="true"}
## Exercise 3: grouped thinking

1. For each species, what proportion of penguins are female? (Exclude penguins
   with missing sex, and say so in a comment.)
2. Add a column giving each penguin's body mass as a percentage of the mean body
   mass **of its own species**.
3. Which island has the largest difference in mean body mass between males and
   females?

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: ex3-dplyr
# 1 - the 11 penguins with unknown sex are excluded, so these
# proportions are among sexed birds only
penguins |>
  filter(!is.na(sex)) |>
  summarise(
    n = n(),
    prop_female = mean(sex == "female"),
    .by = species
  )

# 2 - grouped mutate keeps every row while using a per-species summary
penguins |>
  mutate(
    pct_of_species_mean = 100 * body_mass_g / mean(body_mass_g, na.rm = TRUE),
    .by = species
  ) |>
  select(species, body_mass_g, pct_of_species_mean) |>
  head(5)

# 3
penguins |>
  filter(!is.na(sex)) |>
  summarise(mean_mass = mean(body_mass_g, na.rm = TRUE), .by = c(island, sex)) |>
  pivot_wider(names_from = sex, values_from = mean_mass) |>
  mutate(difference = male - female) |>
  arrange(desc(difference))
```

Biscoe shows the largest sex difference — but note that Biscoe is also the only
island with Gentoo penguins, which are the largest species. The apparent island
effect is largely a species effect. This is the kind of confounding that
grouping makes visible and that a single overall mean would hide.
:::
:::

## More information

- [The dplyr website](https://dplyr.tidyverse.org/)
- [R for Data Science, chapter on transformation](https://r4ds.hadley.nz/data-transform)
- [The tidyverse style guide](https://style.tidyverse.org/)
