# Reshaping Data: Pivoting {#sec-pivot}

::: {.callout-note}
## What this chapter covers
This chapter follows the lecture *Pivoting*. It revisits the tidy data rules and
then works through `pivot_longer()` and `pivot_wider()`, including the awkward
cases where column names contain several variables at once.
:::

```{r}
#| label: setup-pivot
#| include: false
library(webexercises)
```

```{r}
#| label: pivot-libs
#| message: false
library(tidyverse)
```

## The same data in many shapes

The same information can be stored in many different layouts, and most of them
are perfectly readable to a human. They are not all equally usable in R.

```{r}
#| label: pivot-shapes
# wide: one row per person, one column per measurement occasion
wide_data <- tibble(
  id = 1:3,
  sex = c("F", "M", "F"),
  weight_pre = c(65, 80, 70),
  weight_post = c(63, 79, 71)
)

wide_data

# long: one row per measurement
long_data <- wide_data |>
  pivot_longer(
    cols = starts_with("weight"),
    names_to = "occasion",
    values_to = "weight_kg"
  )

long_data
```

Neither is wrong. Which you need depends entirely on what you are about to do:

- **`ggplot2` wants long data.** One row per point to be drawn.
- **Most statistical models want long data**, especially anything repeated
  measures or mixed effects.
- **Tables for humans want wide data.** A reader wants pre and post side by
  side.
- **Some calculations want wide data**, for example a change score
  `post - pre`.

You will move back and forth between these constantly. Pivoting is a *reshaping*
operation, so nothing is lost — you can always go back.

## Tidy data

Recall the three rules from @sec-tidy:

1. Each variable has its own column.
2. Each observation has its own row.
3. Each value has its own cell.

Hadley Wickham identified five common ways real data violates these rules:

1. Column headers are values, not variable names.
2. Multiple variables are stored in one column.
3. Variables are stored in both rows and columns.
4. Multiple types of observational units are stored in the same table.
5. A single observational unit is stored in multiple tables.

Problems 1 to 3 are solved by pivoting and separating; problems 4 and 5 are
solved by joins (@sec-joins).

::: {.callout-tip}
## Tidy is not the same as clean
"Tidy" is about *shape* — one variable per column, one observation per row.
"Clean" is about *content* — no typos, sensible units, correct types, no
duplicate rows. Data can be tidy and filthy, or immaculately clean and
completely untidy. You need both, and they are different chapters
(@sec-cleaning covers clean).
:::

## `pivot_longer()`

Long is the default direction: you almost always pivot longer first, fix things,
and only pivot wider at the very end for presentation.

```{r}
#| label: pivot-longer-basic
df <- tribble(
  ~id, ~bp1, ~bp2,
  "A", 100, 120,
  "B", 140, 115,
  "C", 120, 125
)

df

df |>
  pivot_longer(
    cols = bp1:bp2, # which columns to pivot
    names_to = "measurement", # where the old column names go
    values_to = "value" # where the cell values go
  )
```

What happened, step by step:

- Columns that were **not** pivoted (`id`) are repeated once for each pivoted
  column.
- The **names** of the pivoted columns become values in the new `measurement`
  column.
- The **values** in those columns are preserved and split across rows.

### Choosing columns

`cols` accepts the same selection language as `select()`:

```{r}
#| label: pivot-longer-select
#| eval: false
pivot_longer(cols = bp1:bp2) # a range
pivot_longer(cols = starts_with("bp")) # a helper
pivot_longer(cols = !id) # everything except id
pivot_longer(cols = where(is.numeric)) # by type
```

`cols = !id` is often the most robust: it survives the addition of a new
measurement column, whereas `bp1:bp2` does not.

### Several variables in one column name

Real column names often encode more than one thing. The `who2` dataset from
`tidyr` is a good example: a name like `sp_m_014` means diagnosis method `sp`,
sex `m`, age group `0-14`.

```{r}
#| label: who2-raw
who_small <- who2 |> slice_head(n = 6)
who_small |> select(1:6)
```

Give `names_to` a *vector* of names and tell `pivot_longer()` how the parts are
separated:

```{r}
#| label: who2-pivot
who_small |>
  pivot_longer(
    cols = !c(country, year),
    names_to = c("diagnosis", "gender", "age"),
    names_sep = "_",
    values_to = "count"
  ) |>
  head(6)
```

One long column name has become three tidy variables.

### Several variables *and* several values: `.value`

The `household` dataset holds, for each family, the name and date of birth of up
to two children. The column names contain the names of two variables (`dob` and
`name`) plus the value of a third (child 1 or 2).

```{r}
#| label: household-raw
household
```

The special value `.value` tells `pivot_longer()` that this part of the column
name is not data — it is the name of the output column. It replaces the usual
`values_to`:

```{r}
#| label: household-pivot
household |>
  pivot_longer(
    cols = !family,
    names_to = c(".value", "child"),
    names_sep = "_"
  )
```

The result has one row per child, with `dob` and `name` as proper columns.

### Dropping generated missing values

Not every family has two children, so pivoting creates rows that are entirely
empty. `values_drop_na = TRUE` removes them:

```{r}
#| label: household-dropna
household |>
  pivot_longer(
    cols = !family,
    names_to = c(".value", "child"),
    names_sep = "_",
    values_drop_na = TRUE
  )
```

::: {.callout-warning}
## Only drop *structural* missing values
`values_drop_na = TRUE` is right when the missingness is an artefact of the wide
layout — a third child who does not exist. It is wrong when the missingness is
real data, such as a participant who did not attend the follow-up visit. Those
rows are informative: dropping them turns "did not attend" into "was never
scheduled", and your denominators become wrong.

If in doubt, do not drop, and count the missing values instead.
:::

## `pivot_wider()`

`pivot_wider()` is the inverse. You need it for presentation tables, for
calculating change scores, and for correlation matrices.

```{r}
#| label: pivot-wider-basic
df <- tribble(
  ~id, ~measurement, ~value,
  "A", "bp1", 100,
  "B", "bp1", 140,
  "B", "bp2", 115,
  "A", "bp2", 120,
  "A", "bp3", 105
)

df

df |>
  pivot_wider(
    names_from = measurement, # this column supplies the new column names
    values_from = value # this column supplies the cell values
  )
```

Patient B has no `bp3`, so that cell becomes `NA`. This is real information: the
measurement was not made.

### Calculating a change score

The most common practical reason to pivot wider:

```{r}
#| label: pivot-change
long_data |>
  pivot_wider(names_from = occasion, values_from = weight_kg) |>
  mutate(weight_change = weight_post - weight_pre)
```

## When there are several rows per cell

This is the pivot problem that catches people out. What if two rows in the input
map to the same cell in the output?

```{r}
#| label: pivot-dupes-data
df_dupes <- tribble(
  ~id, ~measurement, ~value,
  "A", "bp1", 100,
  "A", "bp1", 102, # a second bp1 for patient A
  "A", "bp2", 120,
  "B", "bp1", 140,
  "B", "bp2", 115
)

df_dupes
```

```{r}
#| label: pivot-dupes-warn
#| warning: true
df_dupes |>
  pivot_wider(names_from = measurement, values_from = value)
```

The `bp1` column now contains *lists*, not numbers, and `tidyr` warns you about
it. A list column will break the next step in your pipeline, usually with a
confusing error a long way from the real cause.

### Find the duplicates first

Before pivoting wider, check whether the combination you are pivoting on is
actually unique:

```{r}
#| label: pivot-find-dupes
df_dupes |>
  summarise(n = n(), .by = c(id, measurement)) |>
  filter(n > 1)
```

### Three ways to resolve it

Which one is correct depends entirely on **why** the duplicate exists, and that
is a question about the data collection, not about R.

```{r}
#| label: pivot-fix
# 1. The second reading is an error - remove it.
df_dupes |>
  filter(!(measurement == "bp1" & value == 102)) |>
  pivot_wider(names_from = measurement, values_from = value)

# 2. It was mislabelled - it should have been bp3.
df_dupes |>
  mutate(
    measurement = if_else(
      measurement == "bp1" & value == 102, "bp3", measurement
    )
  ) |>
  pivot_wider(names_from = measurement, values_from = value)

# 3. Both readings are valid replicates - average them.
df_dupes |>
  pivot_wider(
    names_from = measurement,
    values_from = value,
    values_fn = mean
  )
```

::: {.callout-important}
## The warning is the useful part
`values_fn = mean` makes the warning go away, which is why it is tempting. But
if the duplicate was a data entry error, averaging it silently propagates the
error into every downstream result.

Treat the "values are not uniquely identified" warning as a finding about your
data, not as an obstacle. Investigate first, then choose a fix, then write a
comment saying which you chose and why.
:::

::: {.webex-check .webex-box}
**Check yourself**

To go from one column per timepoint to one row per timepoint, you use
`r mcq(c(answer = "pivot_longer()", "pivot_wider()", "separate()", "spread()"))`

`ggplot2` generally wants data in wide format. `r torf(FALSE)`

In `pivot_longer()`, which argument names the new column that receives the old
column names?
`r mcq(c("values_to", answer = "names_to", "cols", "names_sep"))`

What does the special value `.value` do in `names_to`?
`r longmcq(c(
  "It drops missing values from the result",
  "It fills empty cells with zero",
  answer = "It tells pivot_longer() that this part of the column name is the name of an output column, not data",
  "It sorts the resulting columns alphabetically"
))`
:::

::: {.callout-note collapse="true"}
## Exercise 1: long and back again

```{r}
#| label: ex1-pivot-data
fitness <- tribble(
  ~id, ~group, ~vo2max_week0, ~vo2max_week6, ~vo2max_week12,
  1, "training", 42.1, 45.3, 47.8,
  2, "training", 38.5, 41.0, 43.2,
  3, "control", 40.2, 40.5, 39.9,
  4, "control", 44.0, 43.8, 44.1
)
```

1. Pivot to long format with a `week` column and a `vo2max` column.
2. The `week` column will contain text like `vo2max_week0`. Turn it into a
   number.
3. Plot the individual trajectories, coloured by group.
4. Pivot back to wide and calculate the change from week 0 to week 12.

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: ex1-pivot-sol
fitness_long <- fitness |>
  pivot_longer(
    cols = starts_with("vo2max"),
    names_to = "week",
    values_to = "vo2max",
    # strip the shared prefix so only the number remains
    names_prefix = "vo2max_week",
    names_transform = list(week = as.numeric)
  )

fitness_long
```

`names_prefix` and `names_transform` do in the pivot what would otherwise need
two extra `mutate()` steps.

```{r}
#| label: fig-ex1-pivot
#| fig-cap: "Individual VO2max trajectories by group."
fitness_long |>
  ggplot(aes(week, vo2max, colour = group, group = id)) +
  geom_line() +
  geom_point() +
  labs(x = "Week", y = expression(VO[2]~max~(ml/kg/min)), colour = "Group") +
  theme_minimal(base_size = 13)
```

Note `group = id`: without it `ggplot2` does not know which points belong to the
same person and draws one zig-zagging line per group.

```{r}
#| label: ex1-pivot-change
fitness_long |>
  filter(week %in% c(0, 12)) |>
  pivot_wider(
    names_from = week,
    values_from = vo2max,
    names_prefix = "week_"
  ) |>
  mutate(change = week_12 - week_0)
```
:::
:::

::: {.callout-note collapse="true"}
## Exercise 2: names with several variables

```{r}
#| label: ex2-pivot-data
lab <- tribble(
  ~id, ~glucose_pre, ~glucose_post, ~insulin_pre, ~insulin_post,
  1, 5.2, 7.8, 45, 120,
  2, 5.8, 8.4, 52, 140,
  3, 4.9, 6.9, 38, 95
)
```

1. Pivot so that you end up with columns `id`, `time`, `glucose` and `insulin`.
2. Calculate the mean of each marker at each timepoint.

::: {.callout-tip collapse="true"}
## Solution

The column names contain two things: which marker, and which timepoint. The
marker should become a column name, so it maps to `.value`; the timepoint is
data, so it gets its own name.

```{r}
#| label: ex2-pivot-sol
lab_long <- lab |>
  pivot_longer(
    cols = !id,
    names_to = c(".value", "time"),
    names_sep = "_"
  )

lab_long

lab_long |>
  summarise(
    mean_glucose = mean(glucose),
    mean_insulin = mean(insulin),
    .by = time
  )
```

Had you instead written `names_to = c("marker", "time")` with a single
`values_to`, you would have got one long column of mixed glucose and insulin
values — tidy in shape, but with two different measurement scales in one column,
which is rarely what you want.
:::
:::

::: {.callout-note collapse="true"}
## Exercise 3: a pivot that fails

```{r}
#| label: ex3-pivot-data
visits <- tribble(
  ~patient, ~visit, ~systolic,
  "P1", "baseline", 132,
  "P1", "followup", 128,
  "P2", "baseline", 145,
  "P2", "baseline", 141,
  "P2", "followup", 138
)
```

1. Try `pivot_wider(names_from = visit, values_from = systolic)`. What happens?
2. Identify which patient and visit is duplicated.
3. Assume the two baseline readings for P2 are genuine repeated measurements
   taken minutes apart. Produce a wide table with one baseline value per
   patient, and justify your choice in a comment.

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: ex3-pivot-sol
#| warning: true
visits |>
  pivot_wider(names_from = visit, values_from = systolic)
```

The `baseline` column becomes a list column, and `tidyr` warns.

```{r}
#| label: ex3-pivot-find
visits |>
  summarise(n = n(), .by = c(patient, visit)) |>
  filter(n > 1)
```

```{r}
#| label: ex3-pivot-fix
# Two readings minutes apart are replicates of the same underlying value,
# so averaging them is defensible and reduces measurement error.
visits |>
  pivot_wider(
    names_from = visit,
    values_from = systolic,
    values_fn = mean
  )
```

Had the two readings been 141 and 191 rather than 141 and 145, averaging would
be indefensible — that is not measurement noise, it is a data problem, and it
needs investigating before any analysis.
:::
:::

## More information

- [The tidyr pivoting vignette](https://tidyr.tidyverse.org/articles/pivot.html)
- [Tidy Data (Wickham 2014)](https://www.jstatsoft.org/article/view/v059i10)
- [R for Data Science, chapter on tidy data](https://r4ds.hadley.nz/data-tidy)
