# Combining Datasets: Joins {#sec-joins}

::: {.callout-note}
## What this chapter covers
This chapter follows the lecture *Joins*. It covers the four mutating joins, the
two filtering joins, `join_by()`, and what happens when a key matches more than
one row — the single most common source of silent errors in registry work.
:::

```{r}
#| label: setup-joins
#| include: false
library(webexercises)
```

```{r}
#| label: joins-libs
#| message: false
library(tidyverse)
```

Joins combine two datasets using one or more **key** variables that appear in
both — an ID number, a date, a municipality code. This is the everyday reality
of registry research: exposure in one file, outcome in another, covariates in a
third.

Every join function has the same interface: it takes two datasets and returns
one, as in `left_join(x, y)`.

There are two families:

- **Mutating joins** add columns from `y` to `x`: `left_join()`,
  `right_join()`, `inner_join()`, `full_join()`.
- **Filtering joins** keep or drop rows of `x` based on whether they match in
  `y`, without adding any columns: `semi_join()`, `anti_join()`.

```{r}
#| label: joins-data
x <- tribble(
  ~key, ~val_x,
  1, "x1",
  2, "x2",
  3, "x3"
)

y <- tribble(
  ~key, ~val_y,
  1, "y1",
  2, "y2",
  4, "y3"
)

x
y
```

Note that key 3 exists only in `x`, and key 4 only in `y`. That mismatch is what
distinguishes the joins from each other.

## Mutating joins

::: {.panel-tabset}
## left_join()

Every row of `x` is kept. Where there is no match in `y`, the new columns are
`NA`.

```{r}
#| label: join-left
left_join(x, y, by = join_by(key))
```

## right_join()

Every row of `y` is kept.

```{r}
#| label: join-right
right_join(x, y, by = join_by(key))
```

## inner_join()

Only rows with a match in both are kept.

```{r}
#| label: join-inner
inner_join(x, y, by = join_by(key))
```

## full_join()

Every row from both datasets is kept.

```{r}
#| label: join-full
full_join(x, y, by = join_by(key))
```
:::

::: {.callout-tip}
## `left_join()` is the default choice
In almost all analysis work, one dataset defines your sample and the others add
information to it. `left_join()` preserves your sample exactly, and missing
matches show up as `NA` where you can count them.

`inner_join()` silently deletes participants. That is occasionally what you
want, but it should always be a decision you made on purpose, never the
side-effect of a function choice.
:::

## The row-count check

The most valuable habit in this entire chapter takes one line:

```{r}
#| label: join-check
nrow(x)

joined <- left_join(x, y, by = join_by(key))
nrow(joined)
```

For a `left_join()`, the result should have **exactly** as many rows as `x`. If
it has more, your key was not unique in `y` and rows have been duplicated. If it
has fewer, something is very wrong.

```{r}
#| label: join-check-assert
#| eval: false
# make the check part of the code, so it fails loudly rather than silently
stopifnot(nrow(joined) == nrow(x))
```

## When a key matches more than one row

There are three possible outcomes for a row in `x`:

- **No match**: it is dropped (inner join) or filled with `NA` (left join).
- **One match**: it is kept, with the new columns added.
- **More than one match**: it is **duplicated**, once per match.

That last case is where analyses go wrong. Nothing errors, nothing warns loudly;
your dataset just quietly grows, and every subsequent count, mean and model is
computed on the wrong number of observations.

```{r}
#| label: join-many-data
df1 <- tibble(key = c(1, 2, 2), val_x = c("x1", "x2", "x3"))
df2 <- tibble(key = c(1, 2, 2), val_y = c("y1", "y2", "y3"))

df1
df2
```

```{r}
#| label: join-many
#| warning: true
inner_join(df1, df2, by = join_by(key))
```

Three rows in, five rows out. Key 2 appears twice on each side, so it produces
2 × 2 = 4 combinations. `dplyr` warns about this many-to-many relationship
because it is usually a mistake.

### Check before you join

```{r}
#| label: join-dupe-check
# is the key unique in the dataset you are joining in?
df2 |>
  count(key) |>
  filter(n > 1)
```

If this returns any rows, decide what to do **before** joining: deduplicate,
aggregate to one row per key, or add a second key variable that makes the
combination unique.

```{r}
#| label: join-relationship
#| eval: false
# state the relationship you expect; dplyr then errors if it is violated
left_join(df1, df2, by = join_by(key), relationship = "one-to-one")
```

::: {.callout-important}
## Use `relationship =`
`relationship = "one-to-one"` or `"many-to-one"` turns a silent data-corruption
bug into an immediate, informative error. In any pipeline you intend to trust —
and certainly in anything that will end up in a manuscript — state the
relationship you expect. It costs one argument and catches a class of error that
is otherwise almost invisible.
:::

## Filtering joins

Filtering joins never add columns. They answer the question "which rows of `x`
have a counterpart in `y`?"

```{r}
#| label: join-semi-anti
# semi_join: keep rows of x that DO have a match
semi_join(x, y, by = join_by(key))

# anti_join: keep rows of x that do NOT have a match
anti_join(x, y, by = join_by(key))
```

`anti_join()` is a data-quality tool as much as a data-manipulation one. Before
any real join, run it in both directions to see what will fail to match:

```{r}
#| label: join-anti-diagnostic
# which participants in x have no record in y?
anti_join(x, y, by = join_by(key))

# and which records in y belong to no one in x?
anti_join(y, x, by = join_by(key))
```

A handful of unmatched rows is normal. Half your sample failing to match usually
means the two files code the key differently — one with leading zeros, one
without; one as text, one as a number; one with a stray space.

## `join_by()`

`join_by()` is the modern way to specify keys. It handles the common case where
the key has different names in the two datasets.

```{r}
#| label: joinby-data
students <- tibble(
  id = c(1, 2, 3),
  name = c("Alice", "Bob", "Charlie"),
  score = c(85, 90, 78)
)

grades <- tibble(
  student_id = c(1, 2, 4),
  name = c("Alice", "Bob", "David"),
  grade = c("A", "B", "C")
)

students
grades
```

```{r}
#| label: joinby-use
students |>
  left_join(grades, by = join_by(id == student_id, name))
```

Two keys are used here: `id` matched against `student_id`, and `name` matched
against itself. David is not in `students`, so he does not appear; and because
the join is a left join, all three students are kept.

::: {.callout-warning}
## Always name your keys
If you omit `by`, `dplyr` performs a *natural join* using every column the two
datasets share, and prints a message telling you which. That is convenient and
fragile: add a column called `date` or `n` to one dataset and your join silently
starts matching on it too.

Write `by = join_by(...)` explicitly. Always.
:::

### What happens to duplicate column names

If both datasets have a column of the same name that is *not* a key, `dplyr`
adds suffixes:

```{r}
#| label: join-suffix
a <- tibble(id = 1:2, value = c(10, 20))
b <- tibble(id = 1:2, value = c(99, 88))

left_join(a, b, by = join_by(id))

# clearer: say where each came from
left_join(a, b, by = join_by(id), suffix = c("_baseline", "_followup"))
```

## Non-equi joins

Joins do not have to match on equality.

- **Cross joins** (`cross_join()`) match every combination of rows —
  `nrow(x) * nrow(y)` rows out.
- **Inequality joins** use `<`, `<=`, `>` or `>=` inside `join_by()`.
- **Rolling joins** (`closest()`) find the nearest match rather than all
  matches.
- **Overlap joins** (`between()`, `within()`, `overlaps()`) work with intervals,
  for example joining an event to whichever exposure period contains it.

```{r}
#| label: join-nonequi
events <- tibble(
  id = c(1, 1, 2),
  event_date = as.Date(c("2020-03-15", "2021-06-01", "2020-11-20"))
)

exposure_periods <- tibble(
  id = c(1, 1, 2),
  start = as.Date(c("2020-01-01", "2021-01-01", "2020-01-01")),
  end = as.Date(c("2020-12-31", "2021-12-31", "2020-12-31")),
  exposure = c("low", "high", "medium")
)

# match each event to the exposure period that contains its date
events |>
  left_join(
    exposure_periods,
    by = join_by(id, between(event_date, start, end))
  )
```

This pattern — attaching a time-varying exposure to a dated event — is extremely
common in registry work, and doing it with an overlap join is far safer than
doing it by hand.

::: {.webex-check .webex-box}
**Check yourself**

Which join keeps every row of the left dataset?
`r mcq(c(answer = "left_join()", "right_join()", "inner_join()", "semi_join()"))`

`anti_join(x, y)` returns rows of x that have a match in y. `r torf(FALSE)`

After `left_join(x, y)`, the result has more rows than `x`. What does this tell
you?
`r longmcq(c(
  "Nothing unusual; left joins normally add rows",
  "The join failed and should be re-run",
  answer = "The key is not unique in y, so rows in x have been duplicated once per match",
  "There were missing values in the key column"
))`

Which argument makes dplyr error if the join is not one-to-one?
`r mcq(c("by =", "suffix =", answer = "relationship =", "keep ="))`
:::

::: {.callout-note collapse="true"}
## Exercise 1: choosing the right join

```{r}
#| label: ex1-join-data
participants <- tibble(
  id = c(101, 102, 103, 104, 105),
  age = c(34, 51, 28, 63, 45),
  group = c("A", "B", "A", "B", "A")
)

measurements <- tibble(
  id = c(101, 102, 104, 106),
  vo2max = c(45.2, 38.1, 33.7, 51.0)
)
```

1. Which participants have no measurement?
2. Which measurement belongs to nobody in the participant list?
3. Join so that all five participants are kept. How many rows does the result
   have, and how many missing values?
4. Join so that only participants with a measurement are kept. Which participants
   did you lose, and would you report that in a paper?

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: ex1-join-sol
# 1
anti_join(participants, measurements, by = join_by(id))

# 2 - id 106 has a measurement but is not in the participant list.
# This is a data problem worth investigating, not something to ignore.
anti_join(measurements, participants, by = join_by(id))

# 3
all_participants <- participants |>
  left_join(measurements, by = join_by(id))

all_participants
nrow(all_participants)
sum(is.na(all_participants$vo2max))

# 4
matched <- participants |>
  inner_join(measurements, by = join_by(id))

matched
```

The inner join drops participants 103 and 105 — 40% of the sample. In a paper
that is a flow-diagram entry, not a silent step: "of 5 participants, 3 had a
valid VO2max measurement". The left join keeps the information that they existed
and let you report the missingness honestly. The inner join loses it forever.
:::
:::

::: {.callout-note collapse="true"}
## Exercise 2: the duplicated key

```{r}
#| label: ex2-join-data
patients <- tibble(
  id = c(1, 2, 3),
  sex = c("F", "M", "F")
)

visits <- tibble(
  id = c(1, 1, 2, 3, 3, 3),
  visit_date = as.Date(c(
    "2023-01-10", "2023-06-15", "2023-02-20",
    "2023-03-01", "2023-07-12", "2023-11-05"
  )),
  systolic = c(132, 128, 145, 138, 141, 135)
)
```

1. How many rows does `left_join(patients, visits)` produce? Explain why.
2. You want one row per patient with their **mean** systolic pressure. Do it
   correctly.
3. You want one row per patient with their **most recent** visit. Do it
   correctly.

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: ex2-join-sol
# 1 - six rows, not three: each patient is duplicated once per visit.
# This is correct behaviour, but it is no longer "one row per patient".
patients |>
  left_join(visits, by = join_by(id)) |>
  nrow()

# 2 - aggregate FIRST, then join, so the key is unique before joining
visit_means <- visits |>
  summarise(
    n_visits = n(),
    mean_systolic = mean(systolic),
    .by = id
  )

patients |>
  left_join(visit_means, by = join_by(id), relationship = "one-to-one")

# 3 - reduce to one row per patient before joining
latest_visit <- visits |>
  slice_max(visit_date, n = 1, by = id)

patients |>
  left_join(latest_visit, by = join_by(id), relationship = "one-to-one")
```

The general rule: **reduce to one row per key, then join.** Joining first and
aggregating afterwards gives the same answer here, but as soon as a third table
enters the pipeline the duplication multiplies and the row count becomes
meaningless.
:::
:::

::: {.callout-note collapse="true"}
## Exercise 3: keys that do not match

```{r}
#| label: ex3-join-data
register_a <- tibble(
  pnr = c("19750101", "19820315", "19910722"),
  exposure = c("high", "low", "medium")
)

register_b <- tibble(
  pnr = c(19750101, 19820315, 19910722),
  outcome = c(1, 0, 1)
)
```

1. Try to join these two registers. What happens?
2. Diagnose the problem.
3. Fix it, and explain which of the two files you would rather change and why.

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: ex3-join-sol
#| error: true
left_join(register_a, register_b, by = join_by(pnr))
```

```{r}
#| label: ex3-join-diagnose
class(register_a$pnr)
class(register_b$pnr)
```

The key is character in one file and numeric in the other, so `dplyr` refuses to
join them. This is one of the more helpful errors in the tidyverse — the
alternative, matching them anyway, would be worse.

```{r}
#| label: ex3-join-fix
register_a |>
  left_join(
    # identifiers are labels, not quantities, so character is the correct type
    register_b |> mutate(pnr = as.character(pnr)),
    by = join_by(pnr),
    relationship = "one-to-one"
  )
```

Convert the numeric file to character, not the other way round. An identifier is
never arithmetic: you will never add two personal identity numbers together, but
you will lose leading zeros the moment one is stored as a number. Here the dates
happen to start with 1 or 2 so nothing was lost — a person born in the year 0075
would already have been silently corrupted in `register_b`.
:::
:::

## More information

- [The dplyr joins reference](https://dplyr.tidyverse.org/reference/mutate-joins.html)
- [R for Data Science, chapter on joins](https://r4ds.hadley.nz/joins)
