# Text and Categories: stringr and forcats {#sec-strings-factors}

::: {.callout-note}
## What this chapter covers
This chapter follows the lecture *stringr and forcats*. `stringr` handles text;
`forcats` handles categorical variables. Between them they solve most of what
goes wrong with the non-numeric columns of a real dataset.
:::

```{r}
#| label: setup-strings
#| include: false
library(webexercises)
```

```{r}
#| label: strings-libs
#| message: false
library(tidyverse)
library(palmerpenguins)
```

## stringr: working with text

Every `stringr` function starts with `str_` and takes the string as its first
argument, which makes them pipe-friendly and easy to find by typing `str_` and
pressing Tab.

| Function | What it does | Example |
|----------|--------------|---------|
| `str_detect()` | Does the pattern occur? Returns `TRUE`/`FALSE` | `str_detect("Hello world", "world")` → `TRUE` |
| `str_subset()` | Keep only elements that match | `str_subset(c("apple", "pear"), "a")` → `"apple", "pear"` |
| `str_replace()` | Replace the **first** match | `str_replace("cat bat", "at", "o")` → `"co bat"` |
| `str_replace_all()` | Replace **all** matches | `str_replace_all("cat bat", "at", "o")` → `"co bo"` |
| `str_split()` | Split on a pattern | `str_split("a,b,c", ",")` |
| `str_c()` | Join strings together | `str_c("R", "is", "great", sep = " ")` |
| `str_length()` | Number of characters | `str_length("Hello")` → `5` |
| `str_trim()` | Remove leading/trailing whitespace | `str_trim("  Hi  ")` → `"Hi"` |
| `str_squish()` | Also collapse internal whitespace | `str_squish("Hi    there")` → `"Hi there"` |
| `str_to_lower()` / `str_to_upper()` / `str_to_title()` | Change case | `str_to_title("hello world")` → `"Hello World"` |
| `str_extract()` | Pull out the matching part | `str_extract("BP 132 mmHg", "\\d+")` → `"132"` |
| `str_pad()` | Pad to a fixed width | `str_pad("7", 3, pad = "0")` → `"007"` |

```{r}
#| label: str-basics
messy <- c("  Stockholm ", "GÖTEBORG", "malmö  ", "  Sthlm")

str_trim(messy)
str_to_title(str_trim(messy))
str_length(messy)
str_detect(messy, "holm")
```

::: {.callout-tip}
## `str_trim()` versus `str_squish()`
`str_trim()` removes whitespace at the ends only. `str_squish()` also collapses
runs of internal whitespace to a single space. For data that has been through
Excel and manual entry, `str_squish()` is almost always the one you want —
`"Erik  Mattsson"` with two spaces will not match `"Erik Mattsson"` in a join.
:::

### Combining strings

```{r}
#| label: str-combine
str_c("Participant", 1:3, sep = " ")

# glue() is often more readable for interpolation
library(glue)
n <- 344
glue("The dataset contains {n} observations.")
```

### Extracting numbers from text

A recurring problem: a measurement stored as `"165 cm"` or `"72,5 kg"`.

```{r}
#| label: str-numbers
heights <- c("165 cm", "175 cm", "162 cm")

# readr::parse_number() handles the common cases directly
parse_number(heights)

# str_extract with a regular expression, when you need more control
str_extract(heights, "\\d+") |> as.numeric()
```

`parse_number()` is more forgiving — it copes with thousands separators and
currency symbols — and should be your first choice.

### Regular expressions, briefly

A regular expression is a pattern language. You do not need to master it, but a
handful of pieces cover most data cleaning:

| Pattern | Matches |
|---------|---------|
| `\\d` | any digit |
| `\\d+` | one or more digits |
| `\\s` | any whitespace |
| `[A-Za-z]` | any letter |
| `^` | start of the string |
| `$` | end of the string |
| `.` | any single character |
| `|` | either / or |

```{r}
#| label: str-regex
ids <- c("SUB001", "SUB002", "CTRL001", "sub003")

str_detect(ids, "^SUB") # starts with SUB (case sensitive)
str_detect(ids, regex("^sub", ignore_case = TRUE)) # case insensitive
str_extract(ids, "\\d+") # the numeric part
```

::: {.callout-tip}
## Regex is a good use of AI
Regular expressions are famously write-only. Describing what you want in plain
language and asking an assistant to produce the pattern works well — but always
test the result against cases you know should match *and* cases you know should
not. See @sec-ai.
:::

## forcats: working with factors

Factors were introduced in @sec-classes. `forcats` makes them comfortable to
work with. Every function starts with `fct_`.

| Function | What it does |
|----------|--------------|
| `fct_relevel()` | Move specified levels to the front |
| `fct_recode()` | Rename levels |
| `fct_collapse()` | Merge several levels into one |
| `fct_lump()` | Collapse rare levels into "Other" |
| `fct_infreq()` | Order levels by frequency, most common first |
| `fct_inorder()` | Order levels by first appearance in the data |
| `fct_reorder()` | Order levels by another variable |
| `fct_rev()` | Reverse the level order |
| `fct_drop()` | Drop unused levels |
| `fct_na_value_to_level()` | Turn `NA` into an explicit level |

### Why level order matters

The level order determines two things: the reference group in a model, and the
order of groups in a table or figure.

```{r}
#| label: fct-order
education <- factor(c(
  "University", "Primary", "Secondary", "University", "Primary"
))

levels(education) # alphabetical, and clinically meaningless

education <- fct_relevel(education, "Primary", "Secondary", "University")
levels(education)
```

### Ordering a figure by a value

The most immediately useful `forcats` function is `fct_reorder()`, which sorts a
categorical axis by a numeric variable.

```{r}
#| label: fig-fct-reorder
#| fig-cap: "Left: alphabetical order. Right: ordered by the value being plotted."
#| fig-width: 9
#| fig-height: 3.5
library(patchwork)

mass_summary <- penguins |>
  filter(!is.na(body_mass_g)) |>
  summarise(mean_mass = mean(body_mass_g), .by = c(species, island)) |>
  mutate(group = str_c(species, island, sep = " / "))

unordered <- mass_summary |>
  ggplot(aes(mean_mass, group)) +
  geom_col(fill = "steelblue") +
  labs(title = "Alphabetical", x = NULL, y = NULL)

ordered <- mass_summary |>
  # order the bars by the value they show, so the ranking is readable
  ggplot(aes(mean_mass, fct_reorder(group, mean_mass))) +
  geom_col(fill = "steelblue") +
  labs(title = "Ordered by value", x = NULL, y = NULL)

unordered + ordered
```

The right-hand panel is the same data and takes one extra function. Alphabetical
ordering of a categorical axis is almost never informative.

### Recoding and collapsing levels

```{r}
#| label: fct-recode
diagnosis <- factor(c(
  "Asthma", "COPD", "Diabetes", "Asthma", "Broken foot",
  "COPD", "Migraine", "Eczema", "Asthma"
))

# rename levels: new = old
fct_recode(diagnosis,
  "Chronic obstructive pulmonary disease" = "COPD",
  "Type 2 diabetes" = "Diabetes"
) |>
  levels()

# merge several levels into one
fct_collapse(diagnosis,
  "Respiratory" = c("Asthma", "COPD"),
  "Metabolic" = c("Diabetes")
) |>
  table()

# keep the 2 most common, everything else becomes Other
fct_lump_n(diagnosis, n = 2) |>
  table()
```

::: {.callout-warning}
## Collapsing categories is an analytical decision
Merging "Asthma" and "COPD" into "Respiratory" changes what your results can
say. It is often the right call — small cell counts make estimates unstable, and
combining them may be more clinically meaningful — but it needs a stated
rationale, and it should be decided before you look at the outcome, not after.

`fct_lump_n()` in particular is convenient and dangerous: it chooses categories
based on frequency in *your* sample, which is a data-dependent decision.
:::

### Making missing values explicit

```{r}
#| label: fct-na
sex_with_na <- factor(c("female", "male", NA, "female", NA))

table(sex_with_na) # NA is invisible
table(sex_with_na, useNA = "ifany") # NA is counted but is not a level

fct_na_value_to_level(sex_with_na, level = "Unknown") |>
  table()
```

Making missingness an explicit level means it appears in every table and figure
automatically, rather than quietly vanishing. In descriptive tables this is
usually what you want.

::: {.callout-note}
## `fct_explicit_na()` is deprecated
Older code and older tutorials use `fct_explicit_na()`. It still works but is
superseded by `fct_na_value_to_level()`. Use the new name in new code.
:::

::: {.webex-check .webex-box}
**Check yourself**

Which function removes whitespace at the ends *and* inside a string?
`r mcq(c("str_trim()", answer = "str_squish()", "str_replace()", "str_pad()"))`

`fct_reorder(group, value)` orders factor levels alphabetically.
`r torf(FALSE)`

To extract the number from `"weight: 72.5 kg"` you could use
`r mcq(c("str_detect()", "str_length()", answer = "parse_number()", "str_pad()"))`

The order of factor levels determines the reference group in a regression model.
`r torf(TRUE)`
:::

::: {.callout-note collapse="true"}
## Exercise 1: cleaning up text

```{r}
#| label: ex1-str-data
sites <- c(
  " Stockholm", "STOCKHOLM ", "stockholm", "  Göteborg  ",
  "goteborg", "GÖTEBORG", "Malmö", "malmo  "
)
```

1. How many distinct values are there as it stands?
2. Clean the vector so that whitespace and capitalisation no longer create false
   distinctions. How many distinct values now?
3. Two spellings of Gothenburg and Malmö remain (with and without the Swedish
   characters). Merge them.

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: ex1-str-sol
n_distinct(sites) # 8 - every one looks different to R

sites_clean <- sites |>
  str_squish() |> # remove leading, trailing and internal whitespace
  str_to_title() # normalise capitalisation

n_distinct(sites_clean) # 5
sort(unique(sites_clean))

# the remaining differences are real spelling variants, so they need
# an explicit decision rather than an automatic rule
sites_final <- sites_clean |>
  fct_collapse(
    "Göteborg" = c("Göteborg", "Goteborg"),
    "Malmö" = c("Malmö", "Malmo")
  )

table(sites_final)
```

The order matters: mechanical fixes (whitespace, case) first, then human
judgement (which spellings mean the same place). Never collapse categories
before you have looked at the full list of distinct values.
:::
:::

::: {.callout-note collapse="true"}
## Exercise 2: extracting structured information

```{r}
#| label: ex2-str-data
records <- tibble(
  raw = c(
    "SUB001_baseline_2023-01-15",
    "SUB002_baseline_2023-01-18",
    "SUB001_followup_2023-07-15",
    "SUB003_baseline_2023-02-02"
  )
)
```

1. Split `raw` into three columns: `id`, `visit` and `date`.
2. Make `date` a proper date column.
3. How many participants are there, and how many have a follow-up visit?

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: ex2-str-sol
records_clean <- records |>
  separate_wider_delim(
    raw,
    delim = "_",
    names = c("id", "visit", "date")
  ) |>
  mutate(
    date = as.Date(date),
    visit = factor(visit, levels = c("baseline", "followup"))
  )

records_clean

n_distinct(records_clean$id)

records_clean |>
  count(visit)
```

`separate_wider_delim()` is the current function; older code uses `separate()`,
which is superseded. The newer function has a clearer name and better error
messages when a row has the wrong number of pieces — which is exactly the case
you want to be told about.
:::
:::

::: {.callout-note collapse="true"}
## Exercise 3: ordering a figure

Using `penguins`:

1. Make a bar chart of the number of penguins per island.
2. Order the bars from most to fewest.
3. Now make a boxplot of `body_mass_g` by species, with species ordered by
   median body mass.

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: fig-ex3-fct
#| fig-cap: "Bars ordered by count rather than alphabetically."
penguins |>
  # fct_infreq() orders by frequency; fct_rev() puts the largest at the top
  ggplot(aes(y = fct_rev(fct_infreq(island)))) +
  geom_bar(fill = "steelblue") +
  labs(x = "Number of penguins", y = NULL) +
  theme_minimal(base_size = 13)
```

```{r}
#| label: fig-ex3-fct2
#| fig-cap: "Species ordered by median body mass."
#| warning: false
penguins |>
  filter(!is.na(body_mass_g)) |>
  ggplot(aes(body_mass_g, fct_reorder(species, body_mass_g, .fun = median))) +
  geom_boxplot(fill = "steelblue", alpha = 0.6) +
  labs(x = "Body mass (g)", y = NULL) +
  theme_minimal(base_size = 13)
```

`fct_reorder()` uses the median by default, so `.fun = median` is written here
only to make the choice explicit for the reader. If you order by the mean
instead, say so — with skewed data the two can give a different ranking.
:::
:::

## More information

- [The stringr website](https://stringr.tidyverse.org/) and its
  [regex vignette](https://stringr.tidyverse.org/articles/regular-expressions.html)
- [The forcats website](https://forcats.tidyverse.org/)
- [R for Data Science: strings](https://r4ds.hadley.nz/strings),
  [factors](https://r4ds.hadley.nz/factors)
