# Cleaning Data {#sec-cleaning}

::: {.callout-note}
## What this chapter covers
This chapter follows the lecture *Städa data*. It works through a fourteen-step
cleaning workflow on a deliberately messy dataset, using the
problem → solution → confirm pattern for every step.
:::

```{r}
#| label: setup-cleaning
#| include: false
library(webexercises)
```

```{r}
#| label: cleaning-libs
#| message: false
library(tidyverse)
library(janitor)
library(lubridate)
```

The goal of cleaning is a dataset that is **correct and consistent**, with every
individual still present. This is not the same as *tidy* data, which is about
shape (@sec-pivot). Data can be perfectly tidy and full of errors.

Three rules before you start:

1. **Save the original raw data somewhere safe** and never write over it.
2. **Every change is a line of code**, never a manual edit. If you fix a typo by
   hand in Excel, that fix is invisible, unrepeatable and undocumented.
3. **Confirm every step.** After each change, look at the result. A cleaning
   step that silently does nothing is worse than no step at all, because you
   believe the problem is fixed.

::: {.callout-important}
## Cleaning is where data quietly disappears
Almost every step below can remove observations. `remove_empty()`,
`distinct()`, and any recode that produces `NA` all change your denominator.
Track `nrow()` from start to finish and be able to account for every row you
lost. In a paper this becomes your flow diagram.
:::

## The workflow

1. Column names — no spaces, clear, consistent
2. Whitespace
3. Capitalisation
4. Remove entirely empty rows and columns
5. Duplicates
6. Separate columns
7. De-identify
8. Extract numbers from text
9. Implausible values
10. Misspelled categories
11. Date formats (ISO 8601)
12. Variable classes
13. Empty strings
14. Derived variables

## The raw data

```{r}
#| label: raw-data
raw_data <- tribble(
  ~"Name ", ~" Age ", ~"SEX", ~"City ", ~"Height", ~"Weight_kg",
  ~"Diagnosis_Date", ~"Diagnosis ", ~"  Info  ", ~"Extra info",
  "Maja Larsson", "23", "Woman", "Stockholm", "165 cm", "65 kg",
  "01/20/2020", "Asthma ", " Info A ", NA,
  "Per Gustafsson", "45", "Man", "Gothenburg ", "175 cm", "80 kg",
  "2020-03-05", "COPD", " Info B ", NA,
  "Sara Lindberg", "37", "Woman", "Umea", "162 cm", "70 kg",
  "05/20/20", "DIABETES", " Info C ", NA,
  "Erik  Mattsson", "29", "Men", " Malmo ", "180 cm", " 75 kg ",
  "2020-07-15", "Unknown", " Extra Info ", NA,
  "Johanna Berg", "", "WOMAN", " Halmstad", "168 cm", "60 kg",
  "07/09/2020", "Asthma", "  ", NA,
  "Klas Oberg", "66", "Man", " Kalmar ", "174 cm", "90 kg",
  "2020-12-20", " COPD", " ", NA,
  "Klas Oberg", "66", "Man", " Kalmar ", "174 cm", "90 kg",
  "2020-12-20", " COPD", " ", NA,
  "Asa Holm", "54", "Wooman", "Boras", "170 cm", "3 kg",
  "2020-11-01", "Broken Foot", "", NA,
  "Mikael Nilsson", "40", "Man", "  Gavle", "183 cm", "90 kg",
  "2020-02-14", "Asthma", " Info D ", NA,
  "Louise Hansen", "18", "Woman", " Sundsvall ", "160 cm", "55 kg",
  "2019-09-07", "COPD", "ExtraInfo", NA,
  "Oskar  Dahl", "37", "Man", "Karlstad", "178 cm", "",
  "09/07/19", "Asthma", "info", NA,
  "Felicia Bjork", "25 ", "Woman", "Karlstad", "162 cm", "65 kg",
  "", "Diabetes", "", NA,
  "Bjorn Karlsson", "55", "Man", "  Orebro ", "190 cm", "105 kg",
  "2020-03-15", "  ", " Extra ", NA,
  "Emmy Sandberg", " 40", "Woman", "  Eskilstuna", "164 cm", "70 kg",
  "2020-03-07", "Asthma", "infoExtra", NA,
  "Amanda Strom", " ", "Unknown", "   ", "158 cm", "62 kg",
  "2020-03-07", "Asthma", "N/A", NA,
  "Ingrid Palsson", "157", "Woman", "Gothenburg", "170 cm", "270 kg",
  "2021-09-07", "COPD ", "Extra Info", NA,
  "Viktor Gran", " 25", "Man", "  Lund", "183 cm", "68 kg",
  "04/01/96", "Asthma", "info G", NA,
  "Linus Dahlqvist", "17", "Man", "Halmstad ", "170 cm", "80 kg",
  "2020-11-11", " COPD", " Extra ", NA,
  "Elin Hellstrom", "55", "Woman", "Orebro", "166 cm", "78 kg",
  "2019-11-11", "  ", " Extra ", NA,
  "Cecilia Sand", "33", "Woman", " Sthlm", "168 cm", "",
  "2020-07-15", "Asthma", "infoExtra", NA,
  NA, NA, NA, NA, NA, NA, NA, NA, NA, NA
)

glimpse(raw_data)
```

Even from `glimpse()` you can see several problems at once: column names with
spaces and inconsistent case, every column read as character, an entirely empty
column, and an entirely empty last row.

### Functions for a first look

Several packages give you an overview. Try more than one — they surface
different things.

```{r}
#| eval: false
dplyr::glimpse(data) # types and first values
skimr::skim(data) # types, missingness, distributions
base::summary(data) # classic summary
visdat::vis_dat(data) # visual map of types and missingness
naniar::gg_miss_var(data) # missingness per variable
summarytools::dfSummary(data) # a full HTML summary
```

## 1. Column names

::: {.panel-tabset}
## Problem

```{r}
#| label: clean-1-problem
colnames(raw_data)
```

Spaces, capitals, and inconsistent naming. Every one of these requires
backticks to refer to, and backticks in analysis code are a constant irritation.

## Solution

```{r}
#| label: clean-1-solution
data_clean <- raw_data |>
  janitor::clean_names() |> # to snake_case, no spaces
  rename(
    height_cm = height, # add the unit to the name
    info_extra = extra_info
  )
```

## Confirm

```{r}
#| label: clean-1-confirm
colnames(data_clean)
```
:::

::: {.callout-tip}
## Put units in the column name
`height_cm` and `weight_kg` are self-documenting. `height` is not, and every
person who uses your data later will have to guess or ask. This is the cheapest
piece of documentation available.
:::

## 2. Whitespace

::: {.panel-tabset}
## Problem

```{r}
#| label: clean-2-problem
data_clean |>
  count(city) |>
  head(8)
```

`" Malmo "` and `"Malmo"` are two different values to R.

## Solution

```{r}
#| label: clean-2-solution
data_clean <- data_clean |>
  mutate(
    # str_squish removes leading, trailing AND repeated internal whitespace
    across(where(is.character), str_squish)
  )
```

## Confirm

```{r}
#| label: clean-2-confirm
data_clean |>
  count(city) |>
  head(8)
```
:::

## 3. Capitalisation

::: {.panel-tabset}
## Problem

```{r}
#| label: clean-3-problem
data_clean |>
  count(sex)
```

## Solution

```{r}
#| label: clean-3-solution
data_clean <- data_clean |>
  mutate(
    name = str_to_title(name),
    diagnosis = str_to_title(diagnosis),
    city = str_to_title(city),
    sex = str_to_title(sex),
    info = str_to_lower(info)
  )
```

## Confirm

```{r}
#| label: clean-3-confirm
data_clean |>
  count(sex)
```
:::

Normalising case collapses `"WOMAN"` and `"Woman"` but leaves `"Wooman"` — a
real typo needs a real decision, which is step 10.

## 4. Empty rows and columns

::: {.panel-tabset}
## Problem

```{r}
#| label: clean-4-problem
data_clean |>
  count(info_extra) # a column with nothing in it

data_clean |>
  slice_tail(n = 2) # and a completely empty last row
```

## Solution

```{r}
#| label: clean-4-solution
n_before <- nrow(data_clean)

data_clean <- data_clean |>
  janitor::remove_empty(which = c("rows", "cols"))
```

## Confirm

```{r}
#| label: clean-4-confirm
n_before
nrow(data_clean) # one empty row removed
colnames(data_clean) # info_extra is gone
```
:::

::: {.callout-warning}
## `remove_empty()` removes only *entirely* empty rows
A row with an ID and nothing else is not empty, and it will survive. That is
correct — such a row is usually a real participant with missing data, not junk.
Always check `nrow()` before and after so you know exactly what was removed.
:::

## 5. Duplicates

::: {.panel-tabset}
## Problem

```{r}
#| label: clean-5-problem
data_clean |>
  janitor::get_dupes()
```

## Solution

```{r}
#| label: clean-5-solution
data_clean <- data_clean |>
  distinct(.keep_all = TRUE)
```

## Confirm

```{r}
#| label: clean-5-confirm
data_clean |>
  janitor::get_dupes()

nrow(data_clean)
```
:::

::: {.callout-important}
## Exact duplicates versus duplicated keys
`distinct()` removes rows that are identical in every column — almost always a
data-entry artefact, safe to remove.

A *duplicated ID* with different values in other columns is a different problem
entirely. That might be two visits for the same person (keep both, and add a
visit variable), or two people who happen to share an identifier (a serious data
problem). Check with `get_dupes(data, id)` and never resolve it automatically.
:::

## 6. Separating columns

::: {.panel-tabset}
## Problem

```{r}
#| label: clean-6-problem
data_clean |>
  select(name) |>
  head(4)
```

Two variables in one column: first name and surname.

## Solution

```{r}
#| label: clean-6-solution
data_clean <- data_clean |>
  separate_wider_delim(
    name,
    delim = " ",
    names = c("first_name", "last_name"),
    too_many = "merge" # a middle name is merged into last_name
  )
```

## Confirm

```{r}
#| label: clean-6-confirm
data_clean |>
  select(first_name, last_name) |>
  head(4)
```
:::

::: {.callout-note}
## `separate()` is superseded
The lecture uses `separate()`, which still works. `separate_wider_delim()` is
the current function and is better in one important way: it errors when a row
does not split into the expected number of pieces, instead of silently
truncating. The `too_many` and `too_few` arguments let you say what should
happen — which forces you to think about the edge cases.
:::

## 7. De-identification

::: {.panel-tabset}
## Problem

```{r}
#| label: clean-7-problem
data_clean |>
  select(first_name, last_name) |>
  head(3)
```

Names are directly identifying and have no analytical value.

## Solution

```{r}
#| label: clean-7-solution
data_clean <- data_clean |>
  mutate(id = row_number()) |>
  select(id, everything(), -first_name, -last_name)
```

## Confirm

```{r}
#| label: clean-7-confirm
colnames(data_clean)
```
:::

::: {.callout-important}
## De-identify early, and keep the key separate
Do this as early in the pipeline as you can. Once names are removed, the working
dataset is far less sensitive, and any figure, table or intermediate file you
produce cannot leak identities.

If you need to link back to the source later, save the ID-to-name key as a
separate file, in a separate location, with separate access control — never in
the analysis folder and never in Git. Also note that a sequential `row_number()`
is a *pseudonym*, not anonymisation: with the original file, anyone can reverse
it. Under GDPR, pseudonymised data is still personal data.
:::

## 8. Extracting numbers from text

::: {.panel-tabset}
## Problem

```{r}
#| label: clean-8-problem
data_clean |>
  select(height_cm, weight_kg) |>
  head(4)
```

## Solution

```{r}
#| label: clean-8-solution
data_clean <- data_clean |>
  mutate(
    height_cm = parse_number(height_cm),
    weight_kg = parse_number(weight_kg)
  )
```

## Confirm

```{r}
#| label: clean-8-confirm
data_clean |>
  select(height_cm, weight_kg) |>
  head(4)
```
:::

## 9. Implausible values

If you find values that should not exist, work out **why** before deciding what
to do. A weight of 3 kg in an adult sample is not a person; it is a typo, a unit
error, or a stray entry. But a weight of 45 kg is unusual and entirely possible.

::: {.panel-tabset}
## Problem

```{r}
#| label: fig-clean-9-problem
#| fig-cap: "Two points are clearly impossible for adult humans."
#| warning: false
data_clean |>
  ggplot(aes(height_cm, weight_kg)) +
  geom_point() +
  labs(x = "Height (cm)", y = "Weight (kg)") +
  theme_minimal(base_size = 13)
```

```{r}
#| label: clean-9-problem2
data_clean |>
  summarise(
    across(
      c(age, height_cm, weight_kg),
      list(
        min = \(x) min(as.numeric(x), na.rm = TRUE),
        max = \(x) max(as.numeric(x), na.rm = TRUE)
      )
    )
  ) |>
  glimpse()
```

## Solution

```{r}
#| label: clean-9-solution
data_clean <- data_clean |>
  mutate(
    # values outside physiologically possible ranges for adults are set to
    # NA rather than deleted, so the individual stays in the dataset
    weight_kg = if_else(weight_kg <= 35 | weight_kg > 250, NA, weight_kg),
    height_cm = if_else(height_cm <= 100 | height_cm > 230, NA, height_cm)
  )
```

## Confirm

```{r}
#| label: fig-clean-9-confirm
#| fig-cap: "After setting implausible values to missing."
#| warning: false
data_clean |>
  ggplot(aes(height_cm, weight_kg)) +
  geom_point() +
  labs(x = "Height (cm)", y = "Weight (kg)") +
  theme_minimal(base_size = 13)
```
:::

::: {.callout-important}
## Set to `NA`, do not delete the row
The person is still a participant; only one measurement is unusable. Deleting
the whole row throws away their age, sex, diagnosis and everything else, and
changes your sample size for reasons that have nothing to do with your research
question.

Also: document your cut-offs and choose them from external knowledge, not from
the data. "More than 3 SD from the mean" is a data-dependent rule that removes
observations *because they are extreme*, which is exactly the kind of value that
is sometimes real and important.
:::

## 10. Misspelled categories

::: {.panel-tabset}
## Problem

```{r}
#| label: clean-10-problem
data_clean |>
  count(sex)

data_clean |>
  count(diagnosis)
```

## Solution

```{r}
#| label: clean-10-solution
data_clean <- data_clean |>
  mutate(
    sex = case_when(
      sex %in% c("Man", "Men") ~ "Man",
      sex %in% c("Woman", "Wooman") ~ "Woman",
      # "Unknown" is genuinely missing information, not a third category
      sex == "Unknown" ~ NA_character_,
      .default = sex
    ),
    diagnosis = if_else(diagnosis == "Unknown", NA_character_, diagnosis)
  )
```

## Confirm

```{r}
#| label: clean-10-confirm
data_clean |>
  count(sex)

data_clean |>
  count(diagnosis)
```
:::

::: {.callout-tip}
## `count()` before and after, always
Recoding is where categories silently disappear. Compare the counts before and
after: the totals should still add up, and every original category should have
gone somewhere you intended. A `.default` that catches more than you expected is
invisible unless you look.
:::

## 11. Dates

ISO 8601 (`YYYY-MM-DD`) is the international standard, and the only date format
that sorts correctly as text and is unambiguous between countries.

::: {.panel-tabset}
## Problem

```{r}
#| label: clean-11-problem
data_clean |>
  select(diagnosis_date) |>
  head(8)
```

Three different formats: ISO, American `MM/DD/YYYY`, and two-digit years.

## Solution

```{r}
#| label: clean-11-solution
data_clean <- data_clean |>
  mutate(
    # orders = tries each format in turn; put the most likely first
    diagnosis_date = as.Date(
      lubridate::parse_date_time(diagnosis_date, orders = c("ymd", "dmy", "mdy"))
    )
  )
```

## Confirm

```{r}
#| label: clean-11-confirm
data_clean |>
  select(diagnosis_date) |>
  head(8)

class(data_clean$diagnosis_date)
```
:::

Alternatives when you know the format is consistent:

- `mutate(date = lubridate::ymd(date))`
- `janitor::convert_to_date()` for Excel serial numbers

::: {.callout-warning}
## `01/02/2020` is genuinely ambiguous
Is that 1 February or 2 January? Nothing in the data can tell you. `orders`
resolves it by whichever format you list first, which means **you** are making
the choice, on the basis of where the file came from.

Two-digit years are worse: `parse_date_time` maps `96` to 1996 and `19` to 2019
using a cut-off rule. Always sanity-check the resulting range — a diagnosis date
in 2068 tells you the rule guessed wrong.
:::

## 12. Variable classes

::: {.panel-tabset}
## Problem

```{r}
#| label: clean-12-problem
glimpse(data_clean)
```

`age` is still character, and `sex` should be a factor for modelling.

## Solution

```{r}
#| label: clean-12-solution
data_clean <- data_clean |>
  mutate(
    age = as.numeric(age),
    sex = as.factor(sex)
  )
```

## Confirm

```{r}
#| label: clean-12-confirm
glimpse(data_clean)
```
:::

## 13. Empty strings

::: {.panel-tabset}
## Problem

```{r}
#| label: clean-13-problem
data_clean |>
  count(diagnosis)
```

An empty string `""` is not `NA`. It is a value of zero length, it does not
count as missing, and it will not be caught by `is.na()`, `drop_na()` or
`na.rm = TRUE`.

## Solution

```{r}
#| label: clean-13-solution
data_clean <- data_clean |>
  mutate(across(where(is.character), \(x) na_if(x, "")))
```

## Confirm

```{r}
#| label: clean-13-confirm
data_clean |>
  count(diagnosis)
```
:::

::: {.callout-important}
## `""` is the most under-detected form of missingness
It looks empty in every table and print-out, but every summary function treats
it as a real value. A variable that appears to be 100% complete can be a third
empty strings. Convert `""` to `NA` early, and check the missingness *after*
that conversion, not before.

The same applies to sentinel values: `999`, `-1`, `"N/A"`, `"missing"`. Handle
these at import with the `na =` argument of `read_csv()`.
:::

## 14. Derived variables

::: {.panel-tabset}
## Problem

```{r}
#| label: clean-14-problem
data_clean |>
  select(height_cm, weight_kg) |>
  head(4)
```

## Solution

```{r}
#| label: clean-14-solution
data_clean <- data_clean |>
  mutate(
    # BMI in kg/m2; height is in cm so it is divided by 100
    bmi = weight_kg / ((height_cm / 100)^2),
    .after = weight_kg
  )
```

## Confirm

```{r}
#| label: clean-14-confirm
data_clean |>
  select(height_cm, weight_kg, bmi) |>
  head(4)

# sanity-check the derived variable against known plausible limits
data_clean |>
  summarise(
    min_bmi = min(bmi, na.rm = TRUE),
    max_bmi = max(bmi, na.rm = TRUE),
    n_missing = sum(is.na(bmi))
  )
```
:::

## The whole pipeline

Once each step works, collect them into one readable pipeline. This is what
belongs in your script — with the exploratory checks kept separately.

```{r}
#| label: clean-full-pipeline
data_clean <- raw_data |>
  # 1. Column names
  janitor::clean_names() |>
  rename(
    height_cm = height,
    info_extra = extra_info
  ) |>
  # 2-3. Whitespace and capitalisation
  mutate(
    across(where(is.character), str_squish),
    name = str_to_title(name),
    diagnosis = str_to_title(diagnosis),
    city = str_to_title(city),
    sex = str_to_title(sex),
    info = str_to_lower(info)
  ) |>
  # 4. Entirely empty rows and columns
  janitor::remove_empty(which = c("rows", "cols")) |>
  # 5. Exact duplicate rows
  distinct(.keep_all = TRUE) |>
  # 6-7. Split the name, then replace it with a pseudonymous id
  separate_wider_delim(
    name,
    delim = " ",
    names = c("first_name", "last_name"),
    too_many = "merge"
  ) |>
  mutate(id = row_number()) |>
  select(id, everything(), -first_name, -last_name) |>
  mutate(
    # 8. Numbers out of text
    weight_kg = parse_number(weight_kg),
    height_cm = parse_number(height_cm),
    # 9. Implausible values to NA (limits from adult physiology, not the data)
    weight_kg = if_else(weight_kg <= 35 | weight_kg > 250, NA, weight_kg),
    height_cm = if_else(height_cm <= 100 | height_cm > 230, NA, height_cm),
    # 10. Misspelled categories
    sex = case_when(
      sex %in% c("Man", "Men") ~ "Man",
      sex %in% c("Woman", "Wooman") ~ "Woman",
      sex == "Unknown" ~ NA_character_,
      .default = sex
    ),
    diagnosis = if_else(diagnosis == "Unknown", NA_character_, diagnosis),
    # 11. Dates to ISO 8601
    diagnosis_date = as.Date(
      parse_date_time(diagnosis_date, orders = c("ymd", "dmy", "mdy"))
    ),
    # 12. Classes
    age = as.numeric(age),
    sex = as.factor(sex),
    # 13. Empty strings to NA
    across(where(is.character), \(x) na_if(x, "")),
    # 14. Derived variable
    bmi = weight_kg / ((height_cm / 100)^2)
  ) |>
  relocate(bmi, .after = weight_kg)

glimpse(data_clean)
```

```{r}
#| label: clean-final-check
# account for every row: 21 raw rows, 1 empty, 1 exact duplicate
nrow(raw_data)
nrow(data_clean)
```

## A codebook

When the cleaning is done, document what the variables are. Your future self is
the main beneficiary.

```{r}
#| label: codebook
codebook <- tibble(
  variable = names(data_clean),
  class = map_chr(data_clean, \(x) class(x)[1]),
  n_missing = map_int(data_clean, \(x) sum(is.na(x))),
  n_unique = map_int(data_clean, n_distinct)
) |>
  mutate(
    pct_missing = round(100 * n_missing / nrow(data_clean), 1)
  )

codebook
```

For a fuller codebook, the `codebookr` package generates a Word document with
value labels, ranges and frequencies.

## Saving the cleaned data

```{r}
#| eval: false
library(here)

# RDS preserves factor levels and date classes exactly
saveRDS(data_clean, here("data", "data_clean_2026_09_08.rds"))

# CSV for sharing with people who do not use R
write_csv(data_clean, here("data", "data_clean_2026_09_08.csv"))
```

::: {.callout-tip}
## Date the file, do not overwrite it
Putting the date in the filename means you can always go back to the version
your results were based on. Overwriting `data_clean.csv` every time you re-run
the script means that when a figure changes, you have no way of knowing which
change caused it.
:::

::: {.webex-check .webex-box}
**Check yourself**

An empty string `""` counts as missing to `is.na()`. `r torf(FALSE)`

Which function converts column names to snake_case?
`r mcq(c("str_to_lower()", answer = "janitor::clean_names()", "rename()", "make.names()"))`

You find an adult with a recorded weight of 3 kg. What should you do?
`r longmcq(c(
  "Delete the whole row",
  "Leave it, since removing data is never justified",
  answer = "Set the weight to NA, keep the individual, and document the rule you used",
  "Replace it with the sample mean"
))`

Which format should dates be stored in?
`r mcq(c("DD/MM/YYYY", "MM/DD/YYYY", answer = "YYYY-MM-DD", "as text, unchanged"))`
:::

::: {.callout-note collapse="true"}
## Exercise 1: clean a small messy dataset

```{r}
#| label: ex1-clean-data
messy <- tribble(
  ~"Study ID", ~"  Group", ~"Baseline Weight", ~"Test date", ~"Notes",
  "001", "Control ", "72,5 kg", "2023-01-15", " ok ",
  "002", "control", "68 kg", "15/01/2023", "",
  "003", "INTERVENTION", "0 kg", "2023-01-16", "NA",
  "002", "control", "68 kg", "15/01/2023", "",
  "004", "Intervention ", "81,2 kg", "2023-01-17", "note"
)
```

1. Clean the column names.
2. Fix whitespace and capitalisation in `group`.
3. Remove the exact duplicate.
4. Convert weight to a number, handling the decimal comma.
5. Set the implausible weight to `NA`.
6. Parse the dates.
7. Convert `""` and `"NA"` to real missing values.
8. Report how many rows you started and ended with.

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: ex1-clean-sol
clean <- messy |>
  janitor::clean_names() |>
  mutate(
    across(where(is.character), str_squish),
    group = str_to_title(group)
  ) |>
  distinct(.keep_all = TRUE) |>
  mutate(
    # the decimal comma has to be replaced before parse_number(),
    # which otherwise reads 72,5 as 725
    baseline_weight = parse_number(
      str_replace(baseline_weight, ",", ".")
    ),
    baseline_weight = if_else(baseline_weight <= 20, NA, baseline_weight),
    test_date = as.Date(parse_date_time(test_date, orders = c("ymd", "dmy"))),
    across(where(is.character), \(x) na_if(x, "")),
    across(where(is.character), \(x) na_if(x, "NA")),
    group = as.factor(group)
  )

clean

nrow(messy) # 5
nrow(clean) # 4 - one exact duplicate removed
```

Two things worth noting. First, `parse_number("72,5")` returns 725, because it
treats the comma as a thousands separator — a silent tenfold error, and exactly
the kind of bug that survives to publication. Second, the string `"NA"` is text,
not missing; `na_if()` has to be told about it explicitly.
:::
:::

::: {.callout-note collapse="true"}
## Exercise 2: what would you check?

You are handed a dataset of 1,200 participants with accelerometer data and asked
to analyse it. Before running anything, list the checks you would perform.

::: {.callout-tip collapse="true"}
## Solution

A reasonable checklist:

**Structure**

- `glimpse()` — is every column the class it should be? Numbers stored as text
  are the single most common import problem.
- `nrow()` and `n_distinct(id)` — is there one row per participant, or several?
- `janitor::get_dupes(data, id)` — duplicated identifiers?

**Missingness**

- `naniar::gg_miss_var()` — how much is missing, per variable?
- Are there sentinel values (`999`, `-1`, `""`, `"N/A"`) masquerading as real
  data?
- Is missingness patterned? Missing outcomes concentrated in one group is a very
  different problem from scattered missingness.

**Values**

- `summary()` on every numeric variable — are the minima and maxima possible?
- Wear time: does anyone have 26 hours in a day, or 3 minutes?
- `count()` on every categorical variable — how many distinct levels are there,
  and are any of them obvious typos or near-duplicates?

**Dates**

- Are all dates in a plausible range? Any dates before the study started or in
  the future?
- Is follow-up always after baseline?

**Consistency**

- Does BMI match height and weight where all three exist?
- Do the group sizes match the study protocol?

The point of the list is that all of this is done **before** any analysis, and
all of it is code. Anything you find becomes a documented cleaning step; anything
you cannot explain becomes a question for whoever collected the data.
:::
:::

## More information

- [The janitor package](https://sfirke.github.io/janitor/)
- [The lubridate package](https://lubridate.tidyverse.org/)
- [R for Data Science: dates and times](https://r4ds.hadley.nz/datetimes)
