# Data Dictionaries {#sec-data-dictionary}

::: {.callout-note}
## What this chapter covers
What a data dictionary is, what to put in one, and how to write it in a format
that both people and AI tools can read. You will build one for `penguins` in
the `data-dict.yaml` format, check the data against it from R, and see how it
lets you give an AI assistant the *description* of your data instead of the
rows.
:::

```{r}
#| label: setup-dict
#| include: false
library(webexercises)
```

```{r}
#| label: dict-libs
#| message: false
library(tidyverse)
library(palmerpenguins)
```

## Why write one

Your code says **how** the data was handled. It does not say **what** the data
is. Look at a column called `sex` with the values 1 and 2, or `vikt` with values
between 40 and 180, or a table with 14,000 rows. Which value is which sex? Is
weight in kilograms, and measured or self-reported? Is it one row per person, or
one per visit? The answers are usually in someone's head, an email or a PDF from
the data provider, and they get lost.

A **data dictionary** writes these answers down in one place, next to the data.
In registry research it is usually called a **codebook**, and it is the same
idea. It describes:

- **the dataset**: where it comes from, how and when it was collected, and
  which version you have;
- **each table**: the *grain* (what one row is) and the *population* (who is
  included);
- **each column**: name, type, units, allowed values or plausible range, how
  missing values are coded, and a description a newcomer can understand;
- **how tables connect**: which keys join them;
- **a glossary** of domain terms an outsider could not guess.

Documentation like this has always helped collaborators, reviewers and your
future self. What is new is that there is now a second reader. Hadley Wickham's
argument in [*Why code when AI?*](https://tidydesign.substack.com/p/y-code-when-ai)
is that the same document gives an AI assistant the context it otherwise lacks,
and that this makes it more motivating to write. An AI that knows `vikt` is in
kilograms, that one row is one visit, and that `sex` is missing for some people
writes better code than one that guesses from the column names.

::: {.callout-important}
## You share the description, not the rows
For sensitive data, this is the main benefit. A dictionary tells an AI tool
everything it needs to write correct code, without a single row of the data
leaving your machine. @sec-ai covers when and how to use AI with data; the
dictionary is what makes the safe option practical.
:::

## Three levels

How far to go depends on how much the data matters and how long it will live:

1. **Show** the structure in the moment: `glimpse()`, `count()`, a summary.
   Good enough for a quick exploration.
2. **Write it down** as a simple table: one row per variable, with a few
   columns. The codebook at the end of @sec-cleaning is this level.
3. **Standardise it** in a machine-readable format that can be validated
   against the data and read by any tool. This chapter is about this level.

## The `data-dict.yaml` format

The tidyverse team has published an open specification for data dictionaries,
[data-dict](https://data-dict.tidyverse.org), along with a command-line tool
that drafts, validates and renders them. A dictionary is a YAML file, the same
format as the `---` header of a Quarto document (@sec-quarto): plain text that
you can read, edit, diff in Git, and hand to any program.

::: {.callout-warning}
## An early preview
When this was written (September 2026), the specification was at version 0.1.0
and the tool at 0.0.3, described as an early preview. Details may change. The
ideas in this chapter do not depend on them: even if you never install the tool,
a well-written YAML dictionary is useful on its own.
:::

Here is a complete dictionary for `penguins`:

```{r}
#| label: dict-show
#| echo: false
#| output: asis
cat("```yaml", readLines("examples/penguins-data-dict.yaml"), "```", sep = "\n")
```

Read it from the top: the dataset, then one table, then one column at a time.
The keys you will use most:

| Key | Where | What it records |
|-----|-------|-----------------|
| `description` | dataset, table, column | What a newcomer needs to know. For a table, state the grain and the population. |
| `source` | table | Where the data lives, e.g. `parquet: data/penguins.parquet` (@sec-parquet). |
| `type` | column | What kind of values the column holds (see below). |
| `units` | column | The unit of a measured quantity: `mm`, `g`, `kg`, `min`. |
| `values` | column | The allowed values of an `enum` column. |
| `range` | column | The plausible or observed `[min, max]`. |
| `constraints` | column, table | Rules: `required` (no missing), `unique`, `primary_key`, `foreign_key`. |
| `glossary` | dataset | Definitions of domain terms. |
| `todo` | anywhere | What is still unknown. |

Almost everything is optional. `name` and `columns` are enough to start, and a
short dictionary that exists beats a complete one that does not.

### Types say what you may do with a column

The type vocabulary is richer than R's column classes, and the difference is the
useful part:

| Type | Meaning | Example |
|------|---------|---------|
| `number` | An ordinary number | a test score |
| `number(quantity)` | A measured amount; needs `units` | body mass in g, steps per day |
| `number(id)` | A number that is really a label: never do arithmetic on it | participant ID, municipality code |
| `number(ordinal)` | A ranking: order it, but do not add it | education level 1-5, study year |
| `enum` | One of a known set of values | species, sex, region |
| `string` | Free text | comments |
| `date`, `datetime` | Dates and times (`datetime` records a time zone) | visit date, accelerometer timestamp |
| `boolean` | `true` or `false` | smoker |

`number(id)` and `number(ordinal)` encode exactly the mistakes this book warns
about: taking the mean of an ID, or treating a 1-5 scale as if the steps were
equal (@sec-modelling). Written into the dictionary, the warning travels with
the data to every person and every tool that reads it.

### What the penguin dictionary tells you that the data does not

- **Units.** `body_mass_g` has a unit in its name; the dictionary states
  it explicitly. Columns like `vikt` or `langd` usually do not.
- **The grain**: one row per penguin, measured once. Not stated anywhere in the
  data itself.
- **What missing means.** `sex` is missing for 11 birds where it could not be
  determined, and missing is not a third category. That changes how you model
  it.
- **Which columns must be complete.** `species`, `island` and `year` are
  `required`; the measurements are not.

## Start from the data, finish by hand

Writing every column by hand is tedious and error-prone. Let R draft the
mechanical parts, and write the parts only a person can know yourself.

The function below looks at each column and records its type, its allowed
values or range, and whether it has missing values. Everything it cannot know
is marked `todo`.

```{r}
#| label: dict-skeleton
# draft one column entry from the data; meaning is left as todo
describe_column <- function(x, name) {
  type_part <- if (is.factor(x)) {
    list(type = "enum", values = levels(x))
  } else if (inherits(x, "Date")) {
    list(type = "date", range = as.character(range(x, na.rm = TRUE)))
  } else if (is.numeric(x)) {
    list(type = "number", range = range(x, na.rm = TRUE))
  } else if (is.logical(x)) {
    list(type = "boolean")
  } else {
    list(type = "string")
  }

  # only mark a column required if the data has no missing values
  constraints <- if (!anyNA(x)) list(constraints = list("required")) else list()

  c(list(name = name), type_part, constraints, list(todo = "describe"))
}

skeleton <- list(
  `$version` = "0.1.0",
  name = "penguins",
  tables = list(list(
    name = "penguins",
    todo = "grain and population",
    columns = imap(penguins, describe_column) |> unname()
  ))
)

draft_file <- tempfile(fileext = ".yaml")
yaml::write_yaml(skeleton, draft_file)

# the first two columns of the draft
readLines(draft_file) |>
  head(24) |>
  cat(sep = "\n")
```

`imap()` calls the function once per column, passing both the column and its
name (@sec-functions). The result is a nested list, and `yaml::write_yaml()`
turns a nested list into YAML.

From here you edit the file by hand: replace every `todo` with a description,
add `units` and `label`, change `number` to `number(quantity)`, `number(id)` or
`number(ordinal)` as appropriate, and write the table's grain and population.
That is the work that needs a person who understands the data.

::: {.callout-warning}
## Ranges and values come from the data
The draft copies `range` and `values` from the data, and the `data-dict` format
also has an `examples` field for a few typical values. For sensitive data, these
can themselves be identifying: the maximum age in a small cohort, a rare
diagnosis code, a municipality with twelve residents. Before a dictionary for
sensitive data is shared or given to an AI tool, **replace ranges with plausible
bounds** (`age: [18, 100]` rather than the observed `[19, 97]`), **leave out
rare values** or write them as a rule, and **use invented examples** or none.

The format also has a `display: restricted` field for marking columns that hold
personal data. It is a label that tools can respect, not a protection in
itself.
:::

## Check the data against the dictionary

A dictionary that nobody checks drifts away from the data: a new column is
added, a unit changes, a new category appears. The strength of a
machine-readable dictionary is that you can **validate** the data against it,
so that the two cannot silently disagree.

The `data-dict` tool does this properly (see below). To see what validation
means, here is a small version in R. It reads the dictionary and checks each
column: required columns have no missing values, `enum` columns contain only the
listed values, and numbers stay inside the range.

```{r}
#| label: dict-validate-fn
# return one row per problem found in one column
check_column <- function(spec, data) {
  if (!spec$name %in% names(data)) {
    return(tibble(column = spec$name, problem = "in the dictionary but not in the data", n = NA_integer_))
  }
  x <- data[[spec$name]]
  counts <- list()

  if ("required" %in% spec$constraints) {
    counts[["missing values"]] <- sum(is.na(x))
  }
  if (spec$type == "enum") {
    counts[["value not in the dictionary"]] <- sum(!is.na(x) & !as.character(x) %in% spec$values)
  }
  if (!is.null(spec$range)) {
    counts[["outside the range"]] <- sum(x < spec$range[[1]] | x > spec$range[[2]], na.rm = TRUE)
  }

  tibble(column = spec$name, problem = names(counts), n = unlist(counts, use.names = FALSE)) |>
    filter(n > 0)
}

check_against_dictionary <- function(data, dict_path) {
  columns <- yaml::read_yaml(dict_path)$tables[[1]]$columns
  undocumented <- setdiff(names(data), map_chr(columns, "name"))

  bind_rows(
    map(columns, \(spec) check_column(spec, data)) |> list_rbind(),
    tibble(column = undocumented, problem = "in the data but not in the dictionary", n = NA_integer_)
  )
}
```

The real data agrees with the dictionary, so the list of problems is empty:

```{r}
#| label: dict-validate-ok
dict_file <- "examples/penguins-data-dict.yaml"

problems <- check_against_dictionary(penguins, dict_file)
nrow(problems)
```

Now introduce the kind of problems that turn up in real data: one body mass
entered in kilograms, three species typed in lower case, a new column nobody
documented, and a column that has disappeared.

```{r}
#| label: dict-validate-bad
# simulate typical errors in a new delivery of the data
penguins_new <- penguins |>
  mutate(
    body_mass_g = if_else(row_number() == 5, body_mass_g / 1000, body_mass_g),
    species = if_else(row_number() %in% 10:12, "adelie", as.character(species)),
    comment = NA_character_
  ) |>
  select(-year)

check_against_dictionary(penguins_new, dict_file)
```

Every problem is found, and none of them would have stopped your code from
running. `"adelie"` would have become a fourth species in every table and plot;
the bird weighing 3.45 g would have pulled down the Adelie mean. Put a check
like this straight after import (step 2, *validate*, in the workflow from
@sec-cleaning), and stop if it finds anything.

For serious use, reach for a package rather than your own function:
`pointblank` and `validate` (@sec-packages) write validation rules in R, and
the `data-dict` tool validates against the dictionary directly.

::: {.callout-note}
## What validation cannot catch
Validation catches **mechanical** errors: wrong type, forbidden value, number
out of range, missing column. It cannot tell you whether you described the
variable correctly in the first place. If the dictionary says `vikt` is in
kilograms and it is really in pounds, the data will agree with the dictionary
and both will be wrong.
:::

## The `data-dict` tool

The command-line tool does three things: **draft** a dictionary from a data
file, **validate** data against a dictionary, and **render** the dictionary as
a browsable HTML page. It can be installed from R:

```{r}
#| eval: false
pak::pak("tidyverse/data-dict/r")
datadict::dd_install()
```

or with Python's `uv` (`uv tool install data-dict-yaml`). Then, in the
terminal:

```bash
data-dict draft data/penguins.parquet       # writes data-dict.yaml with todo items
data-dict validate-data data-dict.yaml      # checks names, types, ranges and constraints
data-dict render-spec data-dict.yaml        # writes an HTML version for people
```

The tool reads the data from the `source` given for each table, which is one
reason Parquet (@sec-parquet) pairs well with it: the file stores its own column
types, so there is nothing to guess.

It is also built to be driven by an AI agent. The quickstart simply asks an
agent to *use the data-dict CLI to document the dataset*, pointing it to
existing documentation; the agent runs `draft` and fills in the descriptions
from that documentation. That is convenient for public data. For sensitive data
it means the agent sees the data, so draft locally, as above, and write the
descriptions yourself.

## Using the dictionary with AI

Once the dictionary exists, it is the first thing to give an AI tool:

- **In a chat**, paste the dictionary (with any sensitive ranges and values
  generalised) together with your question, instead of pasting rows.
- **With an agent**, put a line in the project's instruction file such as
  *"Read `data-dict.yaml` before writing any code that touches the data"*, and
  name it as context in each task (@sec-ai).
- **In reviews**, ask the AI to check your code against the dictionary: "does
  this pipeline treat `year` as ordinal and `sex` missing as missing?"

Wickham points out one more benefit when the dictionary lives in Git with the
code: a change shows up in **three places at once**, in the code, in the data,
and in the dictionary diff. If all three agree after a change, it is easier to
trust that the right thing happened than if you only look at one of them.

How much a dictionary improves an AI's code is so far mostly practical
experience rather than research. That it helps the humans on the project is
well established.

## Where dictionaries come from in practice

You rarely start from nothing. Swedish registers come with variable
descriptions from the data holder (Statistics Sweden, the National Board of
Health and Welfare), REDCap exports a data dictionary for every project, SPSS
and Stata files carry variable and value labels (which `haven` keeps, see
@sec-import), and older studies often have a codebook as a PDF or Word file.

The work is to bring that information into one file that sits next to the data,
is under version control, and can be checked. The R packages `codebook` and
`codebookr` produce readable codebooks from labelled data; the social sciences
have the DDI standard; the FAIR principles and *Datasheets for Datasets* (see
@sec-ai for links) describe what good documentation should contain.

::: {.webex-check .webex-box}
**Check yourself**

What should a table's `description` state, according to the `data-dict`
specification?
`r longmcq(c(
  "The file size and number of rows",
  answer = "The grain (what one row is) and the population (who is included)",
  "The name of the analyst",
  "The R code that created it"
))`

A column of municipality codes should have type
`r mcq(c("number(quantity)", answer = "number(id)", "number(ordinal)", "boolean"))`

A dictionary's `range` and `values` can never reveal anything about individuals.
`r torf(FALSE)`

Validation against a dictionary confirms that each variable is described
correctly. `r torf(FALSE)`

Why give an AI assistant the dictionary rather than a sample of rows?
`r longmcq(c(
  "Because AI tools cannot read rows",
  answer = "It gives the assistant the meaning, types and units it needs without any data leaving your machine",
  "Because it is shorter",
  "There is no reason; rows are always better"
))`
:::

::: {.callout-note collapse="true"}
## Exercise 1: write a dictionary for `mpg`

The `mpg` dataset from `ggplot2` has 234 cars and 11 columns.

1. Use `describe_column()` and `imap()` from this chapter to draft a dictionary
   for `mpg` and write it to `mpg-data-dict.yaml`.
2. Open the file and complete it by hand. Look up the columns in `?mpg`. At a
   minimum, write the table's grain, give `displ`, `cty` and `hwy` their units,
   and decide the type of `year`, `cyl` and `drv`.
3. Run `check_against_dictionary()` on `mpg`. Then break the data on purpose so
   that each kind of problem is reported at least once.

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| eval: false
skeleton <- list(
  `$version` = "0.1.0",
  name = "mpg",
  tables = list(list(
    name = "mpg",
    todo = "grain and population",
    columns = imap(mpg, describe_column) |> unname()
  ))
)
yaml::write_yaml(skeleton, here::here("mpg-data-dict.yaml"))
```

A completed extract might look like this:

```yaml
tables:
  - name: mpg
    description: >
      Grain: one row per car model, year and configuration. Population:
      models from fueleconomy.gov that had a new release every year
      between 1999 and 2008.
    columns:
      - name: displ
        label: Engine displacement
        type: number(quantity)
        units: litres
        range: [1.6, 7.0]
      - name: year
        type: enum
        values: [1999, 2008]
        description: Year of manufacture. Only two years are included.
      - name: cyl
        type: number(ordinal)
        range: [4, 8]
        description: Number of cylinders (4, 5, 6 or 8).
      - name: drv
        type: enum
        values: ["f", "r", "4"]
        description: Drive train. f = front, r = rear, 4 = four-wheel.
      - name: hwy
        label: Highway fuel economy
        type: number(quantity)
        units: miles per gallon
```

Two decisions worth noticing. `year` holds numbers, but with only two values it
is really a category, so `enum` describes it better than `number`. And `drv`
needs `"4"` in quotes: unquoted, YAML reads `4` as a number, while in the data
it is text. Our small check happens to tolerate that, because `%in%` converts
both sides to text, but a stricter tool will not.

```{r}
#| eval: false
mpg_broken <- mpg |>
  mutate(
    hwy = if_else(row_number() == 1, 290, hwy), # out of range
    drv = if_else(row_number() == 2, "front", drv), # not an allowed value
    manufacturer = if_else(row_number() == 3, NA, manufacturer) # missing in a required column
  )

check_against_dictionary(mpg_broken, here::here("mpg-data-dict.yaml"))
```
:::
:::

::: {.callout-note collapse="true"}
## Exercise 2: make a dictionary safe to share

You have drafted a dictionary for a registry extract. It contains, among others:

```yaml
- name: age
  type: number(quantity)
  units: years
  range: [19, 103]
- name: municipality
  type: enum
  values: ["0180", "1480", "2521", ...]   # all 290 codes that occur
- name: diagnosis
  type: enum
  values: [I21, I63, Q87.4, ...]         # every code that occurs
  examples: [I21, I21, Q87.4]
```

You want to give it to an AI assistant to help write the analysis code. What
would you change, and why?

::: {.callout-tip collapse="true"}
## Solution

- **`age`**: the observed maximum of 103 may single out one person. Use a
  plausible bound such as `[18, 110]`, or state the inclusion criterion in the
  description instead.
- **`municipality`**: the list itself is public, but the *set that occurs* in
  the extract, together with other fields, can narrow down who is included. Use
  `type: number(id)` with a description ("four-digit municipality code, SCB
  standard") instead of the list.
- **`diagnosis`**: rare codes such as Q87.4 (Marfan syndrome) are identifying in
  small groups. Describe the coding system ("ICD-10-SE, main diagnosis") and list
  only the codes your analysis uses, if any.
- **`examples`**: they are taken from real rows. Delete them, or invent values
  that are clearly examples.

The assistant does not need any of the removed detail to write correct code.
It needs the types, units, coding systems and the grain, and those are not
sensitive.
:::
:::
