# Exploratory Data Analysis {#sec-eda}

::: {.callout-note}
## What this chapter covers
This chapter follows the lecture *Explorativ dataanalys*. It covers what EDA is
and why it comes before modelling, then works through univariate, bivariate and
multivariate exploration, missing-data patterns, and the packages that make it
quicker.
:::

```{r}
#| label: setup-eda
#| include: false
library(webexercises)
```

## What EDA is

John Tukey, often regarded as the father of exploratory data analysis,
introduced the idea in *Exploratory Data Analysis* [@tukey1977]. EDA is related to
data cleaning but serves a different purpose. Cleaning prepares the data —
handling missing values, correcting errors, ensuring consistency. EDA is about
*understanding* the data's structure through visual and statistical methods. In
practice the two are iterative: EDA reveals what still needs cleaning.

To describe EDA I borrow a paragraph from @jebb2017:

> Putting all these perspectives together, EDA is best described as an
> overarching analytic attitude characterized as "detective work designed to
> reveal the structure or patterns in the data" [@haig2005, p. 375; see also
> @tukey1977]. Put simply, when confronted with a data set, EDA tries to answer the
> simple question, "What is going on here?" [@behrens1997, p. 132] and aims to
> build a rich mental model of the data. The goal is simply to understand the
> structure of the data, and this understanding goes on to serve all kinds of
> analytic goals: **indicating whether statistical assumptions are met,
> identifying outliers, suggesting future hypotheses, uncovering empirical
> relationships, identifying potential transformations, and suggesting
> appropriate models for the data** [@behrens1997]. In this way, **EDA is the
> statistical embodiment of inductive research; through its visualization and
> quantitative techniques**, EDA comprises the research practices that allow
> researchers to detect empirical phenomena.

In simple terms, EDA is how you assess the quality of your data and build a
feel for what it can support. It is a prerequisite for any analysis.

Key characteristics:

- **Visual, intuitive, iterative.** There are no strict rules.
- **Question-driven.** Start with a question; the answer generates more.
- **Not confirmatory.** You are getting a sense of what to expect, not testing
  hypotheses.

### Three levels

- **Univariate** — the distribution and variation *within* one variable.
  Histograms, density curves, Q-Q plots; mean, median, standard deviation.
- **Bivariate** — covariation *between* two variables. Scatterplots,
  correlations; group comparisons with boxplots and violin plots.
- **Multivariate** — patterns among several variables at once. Pairs plots,
  heatmaps, dimension reduction such as PCA.

::: {.callout-important}
## EDA is exploratory, and that has consequences
Everything you learn here shapes what you do next, which means that a hypothesis
generated by staring at your data cannot then be tested on the same data as
though it were pre-specified. The p-value from a comparison you chose *because
it looked interesting* does not mean what a p-value normally means.

This is not an argument against EDA — you cannot analyse data you do not
understand. It is an argument for being explicit in the write-up about which
analyses were planned in advance and which were suggested by the data.
:::

## Starting with questions

Broad questions:

- What variation exists within the variables?
- What covariation exists between them?

Specific questions for the penguin data:

- What differences are there between species?
- Is there overlap between species, islands, sexes?
- Which species is most common?
- Are the species evenly distributed across the islands?
- Are there extreme values?
- How strong is the relationship between bill length and flipper length, or
  body mass?
- Is that relationship linear?
- Are there variables that modify the relationship between other variables?
- Which variables have missing values, and is there a pattern?

## Setup

```{r}
#| label: eda-libs
#| message: false
library(tidyverse)
library(palmerpenguins)
library(skimr)
library(naniar)
library(patchwork)

theme_set(
  theme_minimal(base_size = 13) +
    theme(panel.grid.minor = element_blank())
)
```

## First look

`skim()` from `skimr` is the fastest way to see class, missingness and
distribution for every variable at once.

```{r}
#| label: eda-skim
skimr::skim(penguins)
```

Read this output carefully before doing anything else. It tells you:

- **344 rows, 8 columns.**
- `sex` has 11 missing values; the four measurement variables have 2 each.
- The mean and median of each numeric variable are close, suggesting no severe
  skew.
- The little `hist` column gives a first hint of shape — `body_mass_g` looks
  right-skewed, `flipper_length_mm` looks possibly bimodal.

That last observation is the sort of thing worth following up.

```{r}
#| label: eda-glimpse
glimpse(penguins)
```

## Overview table

```{r}
#| label: eda-overview
penguins |>
  summarise(
    n = n(),
    across(
      where(is.numeric),
      list(
        mean = \(x) round(mean(x, na.rm = TRUE), 1),
        sd = \(x) round(sd(x, na.rm = TRUE), 1)
      )
    ),
    .by = c(species, sex)
  ) |>
  select(-starts_with("year"))
```

## Univariate analysis

**Question: what does the distribution of each variable look like?**

### Categorical variables

```{r}
#| label: fig-eda-cat
#| fig-cap: "Distribution of the categorical variables."
penguins |>
  select(where(is.factor)) |>
  pivot_longer(everything(), names_to = "variable", values_to = "category") |>
  ggplot(aes(y = category)) +
  geom_bar(fill = "steelblue") +
  facet_wrap(~variable, scales = "free") +
  labs(x = "Count", y = NULL)
```

Three immediate observations: the species are unbalanced (Adelie is the most
common), Biscoe holds the most penguins, and sex is close to balanced.

### Continuous variables

Three views of the same variable, each showing something the others do not.

::: {.panel-tabset}
## Histogram

```{r}
#| label: fig-eda-hist
#| warning: false
penguins |>
  ggplot(aes(bill_length_mm)) +
  geom_histogram(binwidth = 2, fill = "steelblue", colour = "white") +
  labs(x = "Bill length (mm)", y = "Count")
```

## Density

```{r}
#| label: fig-eda-density
#| warning: false
penguins |>
  ggplot(aes(bill_depth_mm)) +
  geom_density(fill = "gold3", colour = "gold4", alpha = 0.7) +
  labs(x = "Bill depth (mm)", y = "Density")
```

## All at once

```{r}
#| label: fig-eda-all-dens
#| warning: false
penguins |>
  select(where(is.numeric), -year) |>
  pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  ggplot(aes(value)) +
  geom_density(fill = "indianred", alpha = 0.7) +
  facet_wrap(~variable, scales = "free") +
  labs(x = "Value", y = "Density")
```

## Boxplots

```{r}
#| label: fig-eda-all-box
#| warning: false
penguins |>
  select(where(is.numeric), -year) |>
  pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  ggplot(aes(y = value)) +
  geom_boxplot(fill = "steelblue", alpha = 0.6) +
  facet_wrap(~variable, scales = "free") +
  labs(y = "Value", x = NULL)
```
:::

Notice that `bill_depth_mm` and `flipper_length_mm` are clearly **bimodal** —
two humps. A boxplot of the same variable shows none of this; it would report a
median and a reasonable-looking spread. This is the single best argument for
always looking at a distribution before summarising it.

### Histogram and Q-Q plot together

A Q-Q plot compares your data against a theoretical normal distribution. Points
on the line mean the data is approximately normal.

```{r}
#| label: fig-eda-qq
#| fig-cap: "Distribution and normality of flipper length."
#| fig-width: 9
#| fig-height: 4
#| warning: false
hist_plot <- penguins |>
  filter(!is.na(flipper_length_mm)) |>
  ggplot(aes(flipper_length_mm)) +
  geom_histogram(binwidth = 5, fill = "steelblue", colour = "white") +
  labs(title = "Histogram", x = "Flipper length (mm)", y = "Count")

qq_plot <- penguins |>
  filter(!is.na(flipper_length_mm)) |>
  ggplot(aes(sample = flipper_length_mm)) +
  geom_qq(colour = "navy", alpha = 0.6) +
  geom_qq_line(colour = "red") +
  labs(title = "Q-Q plot", x = "Theoretical quantiles", y = "Sample quantiles")

hist_plot + qq_plot
```

The S-shape in the Q-Q plot is what bimodality looks like there. It is *not*
evidence that a transformation is needed — it is evidence that the sample
contains more than one group. Splitting by species would show three
approximately normal distributions.

::: {.callout-tip}
## Normality of what?
A common misunderstanding: linear regression does not assume that your outcome
variable is normally distributed. It assumes the *residuals* are. A bimodal
outcome, as here, is entirely compatible with a well-behaved model once the
grouping variable is included.
:::

## Bivariate analysis

**Question: how do distributions differ between groups, and how do variables
relate to each other?**

### Continuous by categorical

```{r}
#| label: fig-eda-bivar1
#| fig-cap: "Body mass by species, showing summary and raw data together."
#| warning: false
penguins |>
  filter(!is.na(body_mass_g)) |>
  ggplot(aes(species, body_mass_g, fill = species)) +
  geom_violin(alpha = 0.4, trim = FALSE, show.legend = FALSE) +
  geom_boxplot(width = 0.15, alpha = 0.8, outlier.shape = NA, show.legend = FALSE) +
  geom_point(
    position = position_jitter(width = 0.12, seed = 1),
    alpha = 0.3, size = 1, show.legend = FALSE
  ) +
  labs(x = NULL, y = "Body mass (g)")
```

Adding a grouping variable:

```{r}
#| label: fig-eda-bivar2
#| fig-cap: "Body mass by species and sex."
#| warning: false
penguins |>
  filter(!is.na(sex)) |>
  ggplot(aes(species, body_mass_g, fill = sex)) +
  geom_boxplot(alpha = 0.7) +
  labs(x = NULL, y = "Body mass (g)", fill = "Sex")
```

### Continuous by continuous

```{r}
#| label: fig-eda-scatter
#| fig-cap: "Flipper length against body mass."
#| warning: false
penguins |>
  ggplot(aes(flipper_length_mm, body_mass_g)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", formula = y ~ x) +
  labs(x = "Flipper length (mm)", y = "Body mass (g)")
```

### Correlations

```{r}
#| label: eda-cor
penguins |>
  select(where(is.numeric), -year) |>
  cor(use = "pairwise.complete.obs") |>
  round(2)
```

::: {.callout-warning}
## `use = "pairwise.complete.obs"`
By default `cor()` returns `NA` if there is any missing value anywhere.
`"pairwise.complete.obs"` uses every available pair, which means each
correlation may be based on a different subset of participants. That is usually
acceptable for exploration, but it is worth knowing that the correlations in
such a matrix are not all computed on the same sample, and the matrix is not
guaranteed to be internally consistent.
:::

## Multivariate analysis

### Simpson's paradox

This is the most important idea in this chapter. A relationship in the pooled
data can **reverse** when you account for a grouping variable.

```{r}
#| label: fig-eda-simpson
#| fig-cap: "Same data, with and without accounting for species."
#| fig-width: 9
#| fig-height: 4
#| warning: false
pooled <- penguins |>
  ggplot(aes(bill_length_mm, bill_depth_mm)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", formula = y ~ x, colour = "black") +
  labs(title = "Ignoring species", x = "Bill length (mm)", y = "Bill depth (mm)")

by_species <- penguins |>
  ggplot(aes(bill_length_mm, bill_depth_mm, colour = species)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", formula = y ~ x) +
  scale_colour_manual(values = c("darkorange", "purple", "cyan4")) +
  labs(title = "Within species", x = "Bill length (mm)", y = NULL, colour = "Species")

pooled + by_species
```

Pooled, the relationship is **negative**: longer bills appear shallower. Within
each species it is **positive**. Both are computed correctly from the same data;
the pooled slope is simply an artefact of Gentoo penguins having long, shallow
bills while Adelie have short, deep ones.

::: {.callout-important}
## The practical lesson
There is no statistical test that tells you which grouping variable you forgot.
This is a subject-matter problem, not a computational one — the only defence is
knowing your data and plotting it by every grouping you have. In epidemiology
this same phenomenon appears as confounding, and it is why a scatterplot of raw
data is never sufficient evidence of a relationship.
:::

### Non-linear relationships

```{r}
#| label: fig-eda-nonlinear
#| fig-cap: "A loess smoother makes no assumption of linearity."
#| warning: false
penguins |>
  ggplot(aes(bill_length_mm, bill_depth_mm, colour = species)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE) +
  scale_colour_manual(values = c("darkorange", "purple", "cyan4")) +
  labs(x = "Bill length (mm)", y = "Bill depth (mm)", colour = "Species")
```

Fitting a smoother before a straight line is a cheap way to check whether a
linear model is defensible.

### Pairs plot

A pairs plot shows every bivariate relationship at once — distributions on the
diagonal, scatterplots below, correlations above.

```{r}
#| label: fig-eda-pairs
#| fig-cap: "All pairwise relationships, coloured by species."
#| fig-width: 9
#| fig-height: 7
#| message: false
#| warning: false
penguins |>
  select(species, body_mass_g, ends_with("_mm")) |>
  GGally::ggpairs(
    aes(colour = species, alpha = 0.5),
    columns = c(
      "flipper_length_mm", "body_mass_g",
      "bill_length_mm", "bill_depth_mm"
    )
  ) +
  scale_colour_manual(values = c("darkorange", "purple", "cyan4")) +
  scale_fill_manual(values = c("darkorange", "purple", "cyan4"))
```

This single figure contains most of what the preceding pages established. Pairs
plots become unreadable beyond about eight variables, but for a first look at a
modest set they are hard to beat.

### Heatmap of group means

```{r}
#| label: fig-eda-heatmap
#| fig-cap: "Standardised mean measurements by species."
penguins |>
  drop_na(species, bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g) |>
  summarise(
    across(
      c(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g),
      mean
    ),
    .by = species
  ) |>
  # variables are on different scales, so standardise before comparing them
  mutate(across(where(is.numeric), \(x) as.numeric(scale(x)))) |>
  pivot_longer(!species, names_to = "variable", values_to = "z") |>
  ggplot(aes(variable, species, fill = z)) +
  geom_tile(colour = "white") +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "indianred") +
  labs(x = NULL, y = NULL, fill = "z-score") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

### Summary of what we found

- Gentoo penguins are generally the largest, with longer flippers and higher
  body mass.
- Chinstrap and Adelie are similar in size but differ in bill measurements.
- The bill length / bill depth relationship reverses when species is ignored.
- Bill depth and flipper length are bimodal in the pooled data, which is a
  consequence of the species structure rather than a property of any species.

## Missing data

```{r}
#| label: fig-eda-miss-var
#| fig-cap: "Missing values per variable."
naniar::gg_miss_var(penguins)
```

```{r}
#| label: fig-eda-miss-upset
#| fig-cap: "Which combinations of variables are missing together."
naniar::gg_miss_upset(penguins)
```

The upset plot is the more informative of the two: it shows that the two
penguins missing body mass are missing *everything*, while nine others are
missing only sex. Those are two different problems with two different causes.

### Is the missingness random?

Little's test has the null hypothesis that data are **missing completely at
random** (MCAR).

```{r}
#| label: eda-mcar
naniar::mcar_test(penguins)
```

::: {.callout-warning}
## What the MCAR test can and cannot tell you
A high p-value is consistent with MCAR; it does not prove it, and with a small
sample the test has little power to detect anything.

More importantly, the three missingness mechanisms are not equally
distinguishable. MCAR and MAR can be examined from the observed data; **MNAR —
missing because of the unobserved value itself** — cannot be detected
statistically at all. A participant who dropped out because they got worse is
MNAR, and no test on your dataset will reveal it. That question is answered by
knowing how the data were collected, not by running a test.
:::

## Packages worth knowing

**For exploring and summarising**

- [**skimr**](https://github.com/ropensci/skimr) — fast, informative summaries
- [**naniar**](https://naniar.njtierney.com/) — missing data from every angle
- [**visdat**](https://docs.ropensci.org/visdat/) — a visual map of your data
- [**GGally**](https://ggobi.github.io/ggally/) — pairs plots and more
- [**DataExplorer**](https://boxuancui.github.io/DataExplorer/) — automated
  exploratory reports
- [**summarytools**](https://github.com/dcomtois/summarytools) — `dfSummary()`

**For presenting**

- [**gtsummary**](https://www.danieldsjoberg.com/gtsummary/) — publication-ready
  summary tables (@sec-tables)
- [**gt**](https://gt.rstudio.com/) — full control over table appearance
- [**patchwork**](https://patchwork.data-imaginist.com/) — combining figures

::: {.webex-check .webex-box}
**Check yourself**

Which function gives a quick overview of class, missingness and distribution for
every variable?
`r mcq(c("glimpse()", answer = "skim()", "head()", "str()"))`

A boxplot will reveal whether a distribution is bimodal. `r torf(FALSE)`

Simpson's paradox is when
`r longmcq(c(
  "a variable is missing not at random",
  "two variables are perfectly correlated",
  answer = "the direction of a relationship reverses once a grouping variable is accounted for",
  "a p-value is significant but the effect is small"
))`

A non-significant Little's MCAR test proves your data are missing completely at
random. `r torf(FALSE)`
:::

::: {.callout-note collapse="true"}
## Exercise 1: explore a new dataset

Use `msleep` from `ggplot2` — mammal sleep data, deliberately messier than
`penguins`.

1. Get an overview. How many rows and columns, and which variables have the most
   missing values?
2. Plot the distribution of `sleep_total`. Is it symmetric?
3. Is there a relationship between `bodywt` and `sleep_total`? Try it on the raw
   scale and on a log scale for body weight.
4. Does the relationship differ by `vore` (diet)?

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: ex1-eda-sol
#| message: false
skimr::skim(msleep)
```

```{r}
#| label: fig-ex1-eda1
#| fig-cap: "Total sleep time is roughly symmetric."
msleep |>
  ggplot(aes(sleep_total)) +
  geom_histogram(binwidth = 1.5, fill = "steelblue", colour = "white") +
  labs(x = "Total sleep (hours)", y = "Count")
```

```{r}
#| label: fig-ex1-eda2
#| fig-cap: "Body weight spans five orders of magnitude, so a log scale is essential."
#| fig-width: 9
#| fig-height: 4
raw_scale <- msleep |>
  ggplot(aes(bodywt, sleep_total)) +
  geom_point(alpha = 0.6) +
  labs(title = "Raw scale", x = "Body weight (kg)", y = "Total sleep (h)")

log_scale <- msleep |>
  ggplot(aes(bodywt, sleep_total)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", formula = y ~ x) +
  scale_x_log10() +
  labs(title = "Log scale", x = "Body weight (kg, log10)", y = NULL)

raw_scale + log_scale
```

On the raw scale almost every point is crushed against the left axis and the
elephant dominates everything. On the log scale a clear negative relationship
appears: larger mammals sleep less.

```{r}
#| label: fig-ex1-eda3
#| fig-cap: "The relationship by diet, with sample sizes."
#| warning: false
msleep |>
  filter(!is.na(vore)) |>
  ggplot(aes(bodywt, sleep_total, colour = vore)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE) +
  scale_x_log10() +
  labs(x = "Body weight (kg, log10)", y = "Total sleep (h)", colour = "Diet")

msleep |>
  count(vore)
```

The slopes differ, but note the group sizes: five insectivores is not enough to
estimate a slope with any confidence. Always plot the counts alongside the
trends — a fitted line through five points looks exactly as authoritative as one
through fifty.
:::
:::

::: {.callout-note collapse="true"}
## Exercise 2: find the hidden group

```{r}
#| label: ex2-eda-data
set.seed(2024)

hidden <- tibble(
  group = rep(c("A", "B"), each = 60),
  x = c(rnorm(60, 50, 8), rnorm(60, 70, 8)),
  y = c(rnorm(60, 100, 10), rnorm(60, 80, 10)) + 0.4 * x
)
```

1. Plot `y` against `x` for the whole dataset and fit a straight line. What is
   the direction of the relationship?
2. Now colour by `group` and fit a line per group. What changes?
3. Calculate the correlation overall and within each group.
4. What would you have concluded if `group` had not been recorded?

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: fig-ex2-eda
#| fig-cap: "Pooled and within-group relationships."
#| fig-width: 9
#| fig-height: 4
pooled <- hidden |>
  ggplot(aes(x, y)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", formula = y ~ x, colour = "black") +
  labs(title = "Pooled")

grouped <- hidden |>
  ggplot(aes(x, y, colour = group)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", formula = y ~ x) +
  labs(title = "By group", y = NULL)

pooled + grouped
```

```{r}
#| label: ex2-eda-cor
# overall
cor(hidden$x, hidden$y) |> round(2)

# within each group
hidden |>
  summarise(r = round(cor(x, y), 2), n = n(), .by = group)
```

Pooled, the correlation is negative (about -0.25); within each group it is
positive (about 0.3 and 0.24). The two groups sit at different points on both
axes, and pooling them not only cancels out the real within-group relationship
but reverses its sign.

Had `group` not been recorded, you would have concluded there was no
relationship between `x` and `y` — and you would have had no way, from this data
alone, of knowing you were wrong. That is the honest answer to part 4, and it is
why EDA cannot be separated from knowing how the data were collected.
:::
:::

::: {.callout-note collapse="true"}
## Exercise 3: a missingness investigation

Using `penguins`:

1. Which rows have missing values, and in which variables?
2. Are the penguins with missing `sex` distributed evenly across species and
   islands?
3. Does body mass differ between penguins with and without a recorded sex?
4. What would you conclude about the missingness mechanism, and what could you
   *not* conclude?

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: ex3-eda-sol
# 1
penguins |>
  filter(if_any(everything(), is.na))

# 2
penguins |>
  mutate(sex_missing = is.na(sex)) |>
  summarise(n = n(), .by = c(species, sex_missing)) |>
  pivot_wider(
    names_from = sex_missing,
    values_from = n,
    names_prefix = "missing_",
    # Chinstrap has no missing-sex penguins, so that combination has no row
    # at all. Without values_fill it would appear as NA; the true count is 0.
    values_fill = 0
  )

# 3
penguins |>
  filter(!is.na(body_mass_g)) |>
  mutate(sex_missing = is.na(sex)) |>
  summarise(
    n = n(),
    mean_mass = round(mean(body_mass_g), 0),
    sd_mass = round(sd(body_mass_g), 0),
    .by = sex_missing
  )
```

Missing sex is spread across all three species, with a slight concentration in
Adelie. The nine penguins with missing sex and a recorded body mass have a mean
mass close to the overall mean, which is at least consistent with the
missingness being unrelated to size.

What you **cannot** conclude is that the data are MCAR. If sex was unrecorded
because the field team could not determine it in the field, and determinability
relates to size or condition, the missingness depends on unobserved
characteristics. Nothing in the dataset can rule that out — only knowledge of
the field protocol can.
:::
:::

## More information

- [R for Data Science: EDA](https://r4ds.hadley.nz/eda)
- [The interactive missingness explorer](https://allisonhorst.shinyapps.io/missingexplorer/)
- @tukey1977, the book that named the field
