# Writing Functions and Iterating {#sec-functions}

::: {.callout-note}
## What this chapter covers
How to write your own functions, in both base R and tidyverse style, and how to
repeat an operation over many things with `purrr::map()`. This is the step from
*using* R to *programming* in R, and it is the point at which copy-paste stops
being your main tool.
:::

```{r}
#| label: setup-functions
#| include: false
library(webexercises)
```

```{r}
#| label: functions-libs
#| message: false
library(tidyverse)
library(palmerpenguins)
```

## Why write a function

The rule of thumb: **if you have copied and pasted the same code three times,
write a function.**

Copy-paste has three failure modes, and all three are common in analysis
scripts:

```{r}
#| label: copy-paste-problem
# standardise three variables by copy-paste
penguins |>
  mutate(
    bill_length_z = (bill_length_mm - mean(bill_length_mm, na.rm = TRUE)) /
      sd(bill_length_mm, na.rm = TRUE),
    bill_depth_z = (bill_depth_mm - mean(bill_depth_mm, na.rm = TRUE)) /
      sd(bill_depth_mm, na.rm = TRUE),
    flipper_z = (flipper_length_mm - mean(bill_length_mm, na.rm = TRUE)) /
      sd(flipper_length_mm, na.rm = TRUE)
  ) |>
  select(ends_with("_z")) |>
  head(3)
```

There is a bug in that code. The third line standardises `flipper_length_mm`
using the mean of `bill_length_mm`. It is the kind of error that copy-paste
produces reliably, that no error message will ever report, and that is nearly
invisible on a screen full of similar-looking lines.

A function fixes all three problems: the logic is written once, named once, and
tested once.

```{r}
#| label: first-function
z_score <- function(x) {
  (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)
}

penguins |>
  mutate(across(ends_with("_mm"), z_score, .names = "{.col}_z")) |>
  select(ends_with("_z")) |>
  head(3)
```

## Anatomy of a function

```{r}
#| label: function-anatomy
#| eval: false
name <- function(argument, optional_argument = default_value) {
  # body: what the function does
  result <- argument * 2

  # the last expression is returned automatically
  result
}
```

Four parts:

- **Name** — a verb if the function does something, a noun if it computes
  something. `calculate_bmi()`, `z_score()`.
- **Arguments** — the inputs. Required arguments have no default; optional ones
  do.
- **Body** — the code, between `{` and `}`.
- **Return value** — the value of the last expression. `return()` exists but is
  usually unnecessary; use it only to exit early.

```{r}
#| label: bmi-function
calculate_bmi <- function(weight_kg, height_cm) {
  # height is converted from cm to m because BMI is defined in kg/m2
  height_m <- height_cm / 100
  weight_kg / height_m^2
}

calculate_bmi(weight_kg = 72, height_cm = 180)
calculate_bmi(c(72, 68, 90), c(180, 165, 183)) # vectorised for free
```

::: {.callout-tip}
## Name the function after what it returns or does
`f()`, `myfun()` and `temp2()` are not names, they are placeholders that will
still be there in two years. A good name means the calling code reads as a
sentence: `data |> mutate(bmi = calculate_bmi(weight_kg, height_cm))`.
:::

### Default arguments

```{r}
#| label: function-defaults
summarise_variable <- function(x, digits = 1, na_rm = TRUE) {
  c(
    n = sum(!is.na(x)),
    mean = round(mean(x, na.rm = na_rm), digits),
    sd = round(sd(x, na.rm = na_rm), digits)
  )
}

summarise_variable(penguins$body_mass_g)
summarise_variable(penguins$body_mass_g, digits = 3)
```

Choose defaults that are safe rather than convenient. `na_rm = TRUE` above is a
defensible default for a descriptive summary; it would be a poor default for a
function that computes a study's primary outcome, where silently ignoring
missing data is exactly what you do not want.

### Checking the inputs

A function that fails clearly is worth far more than one that fails
mysteriously three steps later.

```{r}
#| label: function-checks
#| error: true
calculate_bmi <- function(weight_kg, height_cm) {
  if (!is.numeric(weight_kg) || !is.numeric(height_cm)) {
    stop("`weight_kg` and `height_cm` must be numeric.")
  }
  if (any(height_cm < 50, na.rm = TRUE)) {
    warning("Height below 50 cm - is `height_cm` really in centimetres?")
  }

  weight_kg / (height_cm / 100)^2
}

calculate_bmi("72", 180)
```

```{r}
#| label: function-checks2
#| warning: true
calculate_bmi(72, 1.8) # metres by mistake
```

That warning catches one of the most common unit errors in anthropometric data,
and it costs two lines.

::: {.callout-note}
## `stop()`, `warning()`, `message()`
- `stop()` — the function cannot continue. Use it when proceeding would produce
  a wrong answer.
- `warning()` — the function continues, but something looks off. Use it
  sparingly; warnings that are always ignored are worse than none.
- `message()` — informational, for progress or reporting a choice the function
  made.

The `cli` package produces nicer versions of all three, and `rlang::abort()` is
what the tidyverse uses internally.
:::

## Three ways to write an anonymous function

Not every function needs a name. When you pass a function to `across()`,
`map()`, `filter()` or `sapply()`, it is often used once and never again — an
**anonymous function**.

R gives you three ways to write one, and you will meet all three in real code.

```{r}
#| label: lambda-three-ways
values <- list(c(1, 2, NA), c(4, 5, 6))

# 1. the classic: function(x)
map_dbl(values, function(x) mean(x, na.rm = TRUE))

# 2. the backslash shorthand, added in R 4.1 (2021)
map_dbl(values, \(x) mean(x, na.rm = TRUE))

# 3. the purrr formula, using .x for the argument
map_dbl(values, ~ mean(.x, na.rm = TRUE))
```

All three produce the same result. They are not equally good choices.

### `\(x)`: the one to use

`\(x)` is exactly `function(x)` with a shorter spelling. The backslash is
chosen because it resembles the λ of lambda calculus, where the idea of an
anonymous function comes from. It is **base R**, so it works everywhere — in
`sapply()`, in `across()`, in `map()`, in your own functions, in a package.

```{r}
#| label: lambda-equivalence
identical(
  body(\(x) x + 1),
  body(function(x) x + 1)
)
```

This is not a tidyverse feature, and it is not new any more: R 4.1 was released
in 2021, and `purrr` itself now requires R >= 4.1 specifically so that it can
rely on the base pipe and this shorthand.

### `~ .x`: the purrr formula

Before R 4.1 there was no short way to write an anonymous function, so `purrr`
invented one: a one-sided formula, where the argument is `.x` (and `.y` for a
second argument).

```{r}
#| label: lambda-formula
map2_dbl(1:3, 4:6, ~ .x + .y) # formula
map2_dbl(1:3, 4:6, \(x, y) x + y) # the same thing
```

It still works, it is not deprecated, and you will see it constantly — in
Stack Overflow answers, in older tutorials, and in your lecture slides. But it
has two real drawbacks:

- **It only works inside functions that understand it** — `purrr`, and
  tidyverse functions built on `rlang::as_function()`. Pass `~ .x + 1` to base
  R's `sapply()` and you get a formula object, not a function, and a confusing
  error.
- **`.x` and `.y` say nothing.** `\(participant, visit)` documents itself;
  `~ .x + .y` requires the reader to work out what each placeholder refers to.

### When to use which

| Situation | Use |
|-----------|-----|
| Any anonymous function, tidyverse or base | `\(x)` |
| Very short expression where the argument is obvious | `\(x)` is still fine |
| The function is used more than once, or is longer than a line or two | **Give it a name** |
| Reading or maintaining older code | Recognise `~ .x`, leave it alone unless you are already editing that line |
| Writing for R < 4.1 (rare now) | `function(x)` |

::: {.callout-tip}
## The rule in one line
**Write `\(x)`. Read `~ .x`.**

New code gets the backslash. Old code that already works gets left alone — a
mass find-and-replace across a working script is a change with real risk and no
benefit.
:::

::: {.callout-warning}
## Where the difference actually bites
Two places where the choice is not cosmetic.

**Passing extra arguments to `across()`.** This was the old way, and it is
deprecated since dplyr 1.1:

```{r}
#| label: lambda-across-old
#| eval: false
# deprecated: passing ... through across()
penguins |>
  summarise(across(where(is.numeric), mean, na.rm = TRUE))
```

```{r}
#| label: lambda-across-new
penguins |>
  summarise(across(where(is.numeric), \(x) mean(x, na.rm = TRUE))) |>
  glimpse()
```

The lambda makes it explicit which function receives `na.rm`, which is exactly
why the old form was removed: with two nested functions it was genuinely
ambiguous.

**Base R functions do not understand formulas:**

```{r}
#| label: lambda-base-fails
#| error: true
sapply(list(1:3, 4:6), ~ mean(.x))
```

```{r}
#| label: lambda-base-works
sapply(list(1:3, 4:6), \(x) mean(x))
```
:::

### One more shorthand: the extraction shortcut

`purrr` also lets you pass a **string or a number** instead of a function, to
pull out an element by name or position. This one has no base R equivalent and
is worth knowing:

```{r}
#| label: lambda-extract
records <- list(
  list(id = 1, score = 88),
  list(id = 2, score = 92)
)

map_dbl(records, "score") # by name
map_dbl(records, 2) # by position
```

Useful when unpacking the nested lists that come back from an API or a JSON
file.

## The tidyverse way: functions that take column names

Base R functions take *values*. Tidyverse verbs take *bare column names* —
`filter(penguins, species == "Gentoo")` refers to `species` without quotes. That
convenience is called **tidy evaluation**, and it means a naive wrapper does not
work:

```{r}
#| label: tidyeval-broken
#| error: true
# looks reasonable, fails
group_mean_broken <- function(data, group_var, value_var) {
  data |>
    summarise(mean = mean(value_var, na.rm = TRUE), .by = group_var)
}

group_mean_broken(penguins, species, body_mass_g)
```

R tries to find an object called `species` in your environment rather than a
column called `species` in the data. The fix is the **embrace operator**,
`{{ }}`, which says "look this up in the data frame":

```{r}
#| label: tidyeval-fixed
group_mean <- function(data, group_var, value_var) {
  data |>
    summarise(
      n = n(),
      mean = mean({{ value_var }}, na.rm = TRUE),
      sd = sd({{ value_var }}, na.rm = TRUE),
      .by = {{ group_var }}
    )
}

group_mean(penguins, species, body_mass_g)
group_mean(penguins, island, flipper_length_mm)
```

One function, any dataset, any grouping variable, any outcome.

### Naming the output column after the input

`{{ }}` handles the input side. To use an argument as an output *name*, use the
walrus operator `:=` together with `{{ }}`:

```{r}
#| label: tidyeval-naming
add_z_score <- function(data, var) {
  data |>
    mutate("{{ var }}_z" := z_score({{ var }}))
}

penguins |>
  add_z_score(body_mass_g) |>
  select(species, body_mass_g, body_mass_g_z) |>
  head(3)
```

### Several columns at once

For an argument that should accept `select()`-style input — several columns,
helpers like `starts_with()` — pass it straight to `pick()` or `across()`:

```{r}
#| label: tidyeval-multi
summarise_many <- function(data, cols, by) {
  data |>
    summarise(
      across({{ cols }}, \(x) mean(x, na.rm = TRUE)),
      .by = {{ by }}
    )
}

penguins |>
  summarise_many(cols = ends_with("_mm"), by = species)
```

### Functions that make plots

The same rules apply to `ggplot2`:

```{r}
#| label: fig-tidyeval-plot
#| fig-cap: "One function, reusable for any pair of variables."
scatter_by_species <- function(data, x, y) {
  data |>
    ggplot(aes({{ x }}, {{ y }}, colour = species)) +
    geom_point(alpha = 0.6) +
    geom_smooth(method = "lm", formula = y ~ x, se = FALSE) +
    scale_colour_manual(values = c("darkorange", "purple", "cyan4")) +
    theme_minimal(base_size = 13)
}

penguins |>
  scatter_by_species(flipper_length_mm, body_mass_g)
```

::: {.callout-tip}
## When you need `{{ }}`
Only when an argument names a **column**. If the argument is an ordinary value —
a number, a string, a threshold — pass it normally. `{{ }}` around a plain value
is not an error but it is noise, and it signals to the reader that a column is
expected when none is.
:::

## Iteration with `purrr`

Once you have a function, the next question is how to apply it to many things.
`purrr::map()` is the tidyverse answer.

`map(x, f)` applies `f` to every element of `x` and returns a **list**. The
variants return a specific type instead, and error if they cannot:

| Function | Returns |
|----------|---------|
| `map()` | a list |
| `map_dbl()` | a double vector |
| `map_int()` | an integer vector |
| `map_chr()` | a character vector |
| `map_lgl()` | a logical vector |
| `map_vec()` | a vector of whatever type fits (dates, factors) |
| `map2()` | maps over two inputs in parallel |
| `pmap()` | maps over any number of inputs |
| `walk()` | returns nothing; for side effects such as writing files |

```{r}
#| label: map-basics
# the mean of every numeric column
penguins |>
  select(where(is.numeric)) |>
  map_dbl(\(x) mean(x, na.rm = TRUE))

# the number of missing values in every column
penguins |>
  map_int(\(x) sum(is.na(x)))
```

A data frame is a list of columns, which is why `map()` over a data frame
iterates over columns.

::: {.callout-note}
## `map_dbl()` versus `map()`
Always prefer the typed variant when you know the type. `map_dbl()` errors
immediately if a function returns something that is not a single number;
`map()` returns a list containing the surprise, and you find out later, in a
less obvious place.
:::

### `map()` versus the loop you would have written

```{r}
#| label: map-vs-loop
species_list <- unique(penguins$species)

# base R: a loop
means_loop <- numeric(length(species_list))
for (i in seq_along(species_list)) {
  subset_data <- penguins[penguins$species == species_list[i], ]
  means_loop[i] <- mean(subset_data$body_mass_g, na.rm = TRUE)
}
names(means_loop) <- species_list
means_loop

# purrr: a map
means_map <- species_list |>
  set_names() |>
  map_dbl(\(sp) {
    penguins |>
      filter(species == sp) |>
      pull(body_mass_g) |>
      mean(na.rm = TRUE)
  })
means_map
```

Loops are not wrong, and R has not deprecated them. But the loop needs you to
create the output container, get its length right, and index it correctly —
three chances to make a mistake that `map()` removes.

::: {.callout-important}
## And often neither is needed
```{r}
#| label: map-vs-groupby
penguins |>
  summarise(mean_mass = mean(body_mass_g, na.rm = TRUE), .by = species)
```

For "do something per group of rows", `.by` is clearer than both. Reach for
`map()` when you are iterating over **things that are not rows** — files,
models, plots, variables, parameter values.
:::

### Where `map()` genuinely earns its place

**Reading many files at once:**

```{r}
#| eval: false
library(here)

files <- list.files(here("data-raw"), pattern = "\\.csv$", full.names = TRUE)

all_data <- files |>
  set_names(basename) |>
  map(read_csv) |>
  # keep track of which file each row came from
  list_rbind(names_to = "source_file")
```

Four lines, however many files there are, with the provenance of every row
recorded. Doing this by hand is where "I forgot to include site 4" comes from.

**Fitting the same model to several outcomes:**

```{r}
#| label: map-models
outcomes <- c("body_mass_g", "flipper_length_mm", "bill_length_mm")

model_results <- outcomes |>
  set_names() |>
  map(\(outcome) {
    # build the formula from the outcome name
    lm(reformulate("species + sex", response = outcome), data = penguins)
  }) |>
  map(broom::tidy) |>
  list_rbind(names_to = "outcome")

model_results |>
  filter(term != "(Intercept)") |>
  mutate(across(where(is.numeric), \(x) round(x, 3)))
```

Three models, one tidy table, no copy-paste.

**Writing many files:**

```{r}
#| eval: false
penguins |>
  group_split(species) |>
  walk(\(d) {
    species_name <- unique(d$species)
    write_csv(d, here("output", paste0(species_name, ".csv")))
  })
```

`walk()` rather than `map()` because you want the side effect, not the return
value.

### `map2()` and `pmap()`

```{r}
#| label: map2-pmap
weights <- c(72, 68, 90)
heights <- c(180, 165, 183)

# two inputs in parallel
map2_dbl(weights, heights, calculate_bmi)

# any number of inputs, supplied as a data frame or list
params <- tibble(
  weight_kg = c(72, 68, 90),
  height_cm = c(180, 165, 183)
)

pmap_dbl(params, calculate_bmi)
```

`pmap()` matches the column names of `params` to the argument names of the
function, which makes it the natural tool for running an analysis across a grid
of parameter combinations.

### When a function might fail

```{r}
#| label: possibly
#| error: true
strict_mean <- function(x) {
  if (!is.numeric(x)) {
    stop("`x` must be numeric.", call. = FALSE)
  }
  mean(x, na.rm = TRUE)
}

inputs <- list(c(1, 2, 3), "not a number", c(4, 5))

# without possibly(): the second element stops everything
inputs |>
  map_dbl(strict_mean)
```

```{r}
#| label: possibly2
# with possibly(): the failure becomes a value you can find later
safe_mean <- possibly(strict_mean, otherwise = NA_real_)

inputs |>
  map_dbl(safe_mean)
```

Without `possibly()`, one bad element stops the whole run — after twenty minutes
of processing, on file 47 of 50. `possibly()` returns a fallback instead;
`safely()` returns both the result and the error so you can inspect what went
wrong afterwards.

::: {.webex-check .webex-box}
**Check yourself**

What is the rule of thumb for when to write a function?
`r mcq(c("When the code is more than 10 lines", answer = "When you have copied and pasted the same code three times", "When you need a loop", "Never; functions are for package developers"))`

Inside a function, `{{ var }}` is needed when the argument is
`r mcq(c("a number", "a string", answer = "a column name", "a data frame"))`

`map_dbl()` returns a list. `r torf(FALSE)`

Which function applies a function for its side effect and returns nothing
useful?
`r mcq(c("map()", "map_chr()", answer = "walk()", "possibly()"))`

To standardise a per-group summary of your data, the clearest tool is usually
`r mcq(c("a for loop", "map()", answer = "summarise() with .by", "pmap()"))`
:::

::: {.callout-note collapse="true"}
## Exercise 1: your first functions

1. Write a function `cm_to_inches()` that converts centimetres to inches
   (1 inch = 2.54 cm).
2. Write `summarise_numeric(x)` that returns the n, mean, median, SD, minimum
   and maximum of a numeric vector, rounded to a number of digits you can
   control.
3. Add a check that errors with a clear message if `x` is not numeric.
4. Test it on `penguins$body_mass_g` and on `penguins$species`.

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: ex1-functions-sol
#| error: true
cm_to_inches <- function(cm) {
  cm / 2.54
}

cm_to_inches(c(165, 180))

summarise_numeric <- function(x, digits = 1) {
  if (!is.numeric(x)) {
    stop(
      "`x` must be numeric, not ", class(x)[1], ".",
      call. = FALSE
    )
  }

  round(
    c(
      n = sum(!is.na(x)),
      mean = mean(x, na.rm = TRUE),
      median = median(x, na.rm = TRUE),
      sd = sd(x, na.rm = TRUE),
      min = min(x, na.rm = TRUE),
      max = max(x, na.rm = TRUE)
    ),
    digits
  )
}

summarise_numeric(penguins$body_mass_g)
summarise_numeric(penguins$species)
```

Note `call. = FALSE`: it suppresses the internal call in the error message, so
the user sees your sentence rather than a line of code they did not write.
Compare that with the message R gives you from `mean()` on a factor, which tells
you what went wrong but not where.
:::
:::

::: {.callout-note collapse="true"}
## Exercise 2: a tidyverse function

Write `plot_distribution(data, var, group)` that:

1. Takes a data frame, a continuous variable and a grouping variable.
2. Draws a violin plot with a boxplot inside it.
3. Labels the axes with the variable names.
4. Works for any combination of variables in `penguins`.

Then use it twice, on different variables.

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: fig-ex2-functions
#| fig-cap: "The same function, two different questions."
#| fig-width: 9
#| fig-height: 4
#| warning: false
plot_distribution <- function(data, var, group) {
  data |>
    filter(!is.na({{ var }}), !is.na({{ group }})) |>
    ggplot(aes({{ group }}, {{ var }}, fill = {{ group }})) +
    geom_violin(alpha = 0.4, trim = FALSE, show.legend = FALSE) +
    geom_boxplot(width = 0.15, alpha = 0.8, show.legend = FALSE) +
    theme_minimal(base_size = 13)
}

p1 <- penguins |> plot_distribution(body_mass_g, species)
p2 <- penguins |> plot_distribution(flipper_length_mm, island)

library(patchwork)
p1 + p2
```

`ggplot2` picks up the axis labels from the mapped expressions automatically, so
`labs()` is not needed for basic labelling. For a figure that will be published
you would still add a `labs()` argument to the function so units can be
supplied — an exercise worth doing.

Note the `filter()` at the start: dropping missing values inside the plotting
function keeps the console free of "removed N rows" warnings, but it also hides
them. For a quick exploratory function that is a reasonable trade; for anything
you report from, count the exclusions separately.
:::
:::

::: {.callout-note collapse="true"}
## Exercise 3: iteration

1. Use `map_dbl()` to calculate the median of every numeric column in
   `penguins`.
2. Use `map_chr()` to get the class of every column.
3. Combine both into a small data frame with one row per variable.
4. Fit a linear model of `body_mass_g` on `flipper_length_mm` separately for
   each species, and extract the slope from each.

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: ex3-functions-sol
# 1 and 2
penguins |>
  select(where(is.numeric)) |>
  map_dbl(\(x) median(x, na.rm = TRUE))

penguins |>
  map_chr(\(x) class(x)[1])

# 3 - a small codebook, built by iterating over columns
tibble(
  variable = names(penguins),
  class = map_chr(penguins, \(x) class(x)[1]),
  n_missing = map_int(penguins, \(x) sum(is.na(x))),
  n_unique = map_int(penguins, n_distinct)
)

# 4 - one model per species, using nest() + map() + unnest()
penguins |>
  drop_na(body_mass_g, flipper_length_mm) |>
  # nest() puts each species' rows into a list column called `data`
  nest(.by = species) |>
  mutate(
    model = map(data, \(d) lm(body_mass_g ~ flipper_length_mm, data = d)),
    coefficients = map(model, broom::tidy)
  ) |>
  select(species, coefficients) |>
  unnest(coefficients) |>
  filter(term == "flipper_length_mm") |>
  mutate(across(where(is.numeric), \(x) round(x, 2)))
```

`nest()` + `map()` + `unnest()` is the standard tidyverse pattern for
"one model per group". The intermediate object is a data frame with a list
column holding a model in each row, which keeps the model and the group label
attached to each other — much harder to mismatch than a separate list of models
and a separate vector of names.

The slopes differ between species: an extra millimetre of flipper is worth more
grams in a Gentoo than in an Adelie. That is a fitted interaction, arrived at
without fitting an interaction — and it is worth noticing that three separate
models is *not* the same analysis as one model with an interaction term. The
separate models each use only their own species' data to estimate the residual
variance; the interaction model pools it. Which you want depends on the
question, and `map()` makes the wrong one just as easy to produce as the right
one.
:::
:::

## More information

- [R for Data Science: functions](https://r4ds.hadley.nz/functions)
- [R for Data Science: iteration](https://r4ds.hadley.nz/iteration)
- [Programming with dplyr](https://dplyr.tidyverse.org/articles/programming.html)
  — the reference for `{{ }}` and `:=`
- [The purrr website](https://purrr.tidyverse.org/)
- [Advanced R](https://adv-r.hadley.nz/) — when you want the underlying model
