# Causal Analysis and Advanced Models {#sec-causal}

::: {.callout-note}
## What this chapter covers
@sec-modelling was about fitting a model and reporting it. This chapter is about
the harder questions that come after: what the model is *for*, what to do when
the observations are not independent, how to get a confidence interval when no
formula exists, and where predictive modelling fits in.

It assumes the previous chapter — `broom`, `marginaleffects` and list columns —
and continues with the same `penguins` data.
:::

```{r}
#| label: setup-causal
#| include: false
library(webexercises)
```

```{r}
#| label: causal-libs
#| message: false
library(tidyverse)
library(palmerpenguins)
library(broom)
library(marginaleffects)

model_data <- penguins |> drop_na()

d_int <- penguins |>
  drop_na(body_mass_g, flipper_length_mm, species, sex)
```

## Causal analysis

Everything so far has been about *fitting* models. This section is about a
question you have to answer before fitting anything: **what is the model for?**

### Three different jobs

| Job | Question | What matters |
|-----|----------|--------------|
| **Description** | What does this population look like? | Representative sample, honest summaries |
| **Prediction** | What is the likely outcome for a new case? | Out-of-sample accuracy; whether a predictor is "causal" is irrelevant |
| **Causal inference** | What would happen if we intervened? | The right adjustment set, and assumptions you must state |

The same `lm()` call serves all three, which is exactly why they get confused.
A variable that improves prediction can wreck a causal estimate, and vice versa.
Decide which job you are doing before you choose variables.

::: {.callout-important}
## "We are not claiming causality" does not make a model descriptive
An adjusted regression *is* a causal analysis in intent. Nobody adjusts for
confounders to describe a population; they do it to isolate an effect. Writing
"associations should not be interpreted causally" in the discussion while
having selected covariates on causal grounds is having it both ways.

The honest version is to state the causal question, state the assumptions it
rests on, and say plainly that those assumptions are untestable. That is
stronger than a disclaimer, not weaker.
:::

### Define the estimand first

Say what you are estimating before touching the outcome:

- **ATE** — the average treatment effect if *everyone* were treated versus no
  one.
- **ATT** — the effect among those who *actually received* the treatment. Often
  the policy-relevant one, and what matching usually targets.

A useful discipline is the **target trial**: what randomised trial would this
analysis emulate? Answering it forces you to be explicit about eligibility, when
follow-up starts (time zero) and what is being compared — which is how
immortal-time and selection bias get caught before they happen.

### Choose covariates from a causal model, not by algorithm

Stepwise selection, "everything that was significant at p < 0.20", and "adjust
for everything available" are all ways of choosing covariates without a causal
argument. They can each make the estimate worse.

A **DAG** — a directed acyclic graph — writes down what you believe causes what,
so the adjustment set follows from the assumptions rather than from the data.

```{r}
#| label: dag-setup
#| message: false
library(dagitty)
library(ggdag)

dag <- dagitty('dag {
  training -> fitness
  age -> training
  age -> fitness
  motivation -> training
  motivation -> fitness
}')

# which variables must be adjusted for to identify training -> fitness?
adjustmentSets(dag, exposure = "training", outcome = "fitness")
```

```{r}
#| label: fig-dag
#| fig-cap: "A causal diagram. Age and motivation cause both the exposure and the outcome."
#| fig-height: 3.2
#| message: false
ggdag(dag, text_size = 3) +
  theme_dag()
```

Three roles a third variable can play, and they demand opposite treatment:

- A **confounder** causes both exposure and outcome. **Adjust for it** — leaving
  it out biases the estimate.
- A **mediator** sits on the path from exposure to outcome. **Do not adjust for
  it** if you want the total effect; adjusting removes the part of the effect
  that travels through it.
- A **collider** is caused by both exposure and outcome (or by two variables of
  interest). **Do not adjust for it** — adjusting *creates* an association that
  does not exist.

### What happens if you get it wrong

Because we can simulate data where the truth is known, we can watch each mistake
happen. Here the true effect of training on fitness is exactly **3**.

```{r}
#| label: causal-sim
set.seed(5)

n <- 4000

sim <- tibble(
  age = rnorm(n, 45, 12),
  motivation = rnorm(n),
  # both age and motivation influence how much people train ...
  training = 2 * motivation - 0.05 * age + rnorm(n),
  # ... and both also influence fitness directly. True training effect = 3.
  fitness = 3 * training + 1.5 * motivation - 0.2 * age + rnorm(n, 0, 3)
) |>
  # gym membership is caused by BOTH training and fitness: a collider
  mutate(gym_member = as.numeric(scale(training + fitness + rnorm(n))) > 0)
```

```{r}
#| label: causal-compare
models <- list(
  "Unadjusted"            = fitness ~ training,
  "Adjusted (correct)"    = fitness ~ training + motivation + age,
  "Also adjusted for the collider" = fitness ~ training + motivation + age + gym_member
)

map(models, \(f) lm(f, data = sim) |> tidy(conf.int = TRUE)) |>
  list_rbind(names_to = "model") |>
  filter(term == "training") |>
  select(model, estimate, conf.low, conf.high) |>
  mutate(across(where(is.numeric), \(x) round(x, 2)))
```

Read the three rows against the truth of 3.00:

- **Unadjusted: 3.84.** Too high. Motivated people both train more and are
  fitter anyway, so some of motivation's effect is being credited to training.
  This is ordinary confounding.
- **Correctly adjusted: 3.00.** Right, because the adjustment set from the DAG
  closes both back-door paths.
- **With the collider added: 2.58.** Wrong again — and note the direction. This
  model adjusts for *more* variables than the correct one, and is *further* from
  the truth. Gym membership is a consequence of training and fitness, and
  conditioning on it induces a spurious negative association between them.

That third row is the whole argument against "adjust for everything you have".
None of the three models warns you; all three have narrow confidence intervals;
only the DAG tells you which is right.

### Mediators remove part of the effect

```{r}
#| label: causal-mediator
set.seed(11)

med <- tibble(
  training = rnorm(2000),
  # training builds muscle ...
  muscle = 2 * training + rnorm(2000),
  # ... and both training and muscle improve fitness
  fitness = 1.5 * training + 1.0 * muscle + rnorm(2000, 0, 2)
)

# total effect: 1.5 direct + (2 x 1.0) through muscle = 3.5
lm(fitness ~ training, data = med) |> tidy() |> filter(term == "training")

# adjusting for the mediator leaves only the direct effect, 1.5
lm(fitness ~ training + muscle, data = med) |> tidy() |> filter(term == "training")
```

Neither is wrong — they answer different questions. The first is "how much does
training improve fitness, by any route", the second "how much does training
improve fitness *other than* by building muscle". Decide which you want, and say
which you report. Adjusting for a mediator by accident, because it was in the
dataset and seemed relevant, is a common way to make a real effect disappear.

::: {.callout-warning}
## The Table 2 fallacy
When you fit `outcome ~ exposure + age + sex + smoking`, only the exposure
coefficient has been designed to be interpretable. The adjustment set was chosen
to identify the *exposure* effect, not the effect of age, and not the effect of
smoking — each of those would need its own DAG and its own adjustment set.

Presenting the full coefficient table as though every row were an effect
estimate is the **Table 2 fallacy** (Westreich and Greenland). It is extremely
common in published papers.

The fix is presentational and easy: report the exposure effect as the result,
and either omit the covariate coefficients or label them explicitly as
adjustment terms that should not be interpreted.
:::

### Design before outcome

For a serious causal analysis, the recommended order is to build and check the
design *before* looking at the outcome, so balance is judged blind to the
effect. In outline:

```{r}
#| label: causal-design
#| eval: false
library(MatchIt)
library(cobalt)

matched_design <- matchit(
  treat ~ age + sex + bmi + comorbidity,
  data = cohort,
  method = "full",
  estimand = "ATT"
)

# check balance BEFORE estimating anything
bal.tab(matched_design, un = TRUE)
love.plot(matched_design, thresholds = c(m = 0.1)) # aim for |SMD| < 0.1

matched <- match.data(matched_design)
```

Then estimate on the balanced design, combining weights with an outcome model —
which is **doubly robust**, meaning the estimate stays consistent if *either*
the propensity model or the outcome model is right:

```{r}
#| label: causal-estimate
#| eval: false
fit <- glm(outcome ~ treat + age + sex + bmi,
  data = matched, weights = weights, family = binomial()
)

avg_comparisons(
  fit,
  variables = "treat",
  wts = "weights",
  vcov = ~subclass # cluster on the matched sets
)
```

`WeightIt` does the weighting equivalent (IPTW, entropy balancing) and can
target the ATE as well as the ATT.

### How fragile is the conclusion?

Every observational estimate rests on **no unmeasured confounding**, which is
untestable. You cannot prove it holds; you can quantify how much it would take
to overturn your result.

```{r}
#| label: causal-evalue
#| message: false
# how strong would an unmeasured confounder have to be, with both the
# exposure and the outcome, to explain away this estimate?
EValue::evalues.OLS(
  est = 3.0,
  se = 0.1,
  sd = sd(sim$fitness)
)
```

The E-value is the minimum association an unmeasured confounder would need with
*both* exposure and outcome to explain the effect away. A large E-value means
the finding is robust; a small one means a modest unmeasured confounder would
suffice — and you should say so plainly. `tipr` does the same job as a
tipping-point analysis across a range of plausible confounders.

::: {.callout-note}
## A reporting checklist
For a causal analysis, report in this order:

1. The **estimand** — ATE or ATT, and the target trial it emulates.
2. The **assumed DAG** and the adjustment set that follows from it.
3. The **design method** and which estimand it targets.
4. **Balance** before and after, with the threshold you used.
5. The **effect estimate** with a confidence interval, and how the standard
   errors were computed.
6. A **sensitivity analysis** for unmeasured confounding.

And state the no-unmeasured-confounding assumption as a limitation, in those
words. We estimate effects under stated assumptions; we do not prove causation.
:::

::: {.callout-tip}
## Going further
This is an orientation, not a course. For the real thing:

- [Causal Inference: What If](https://miguelhernan.org/whatifbook) — Hernán and
  Robins, free online, the standard reference
- [Causal Inference in R](https://www.r-causal.org/) — the same ideas worked
  through in R with `ggdag`, `MatchIt` and `WeightIt`
- [dagitty.net](https://dagitty.net/) — draw a DAG in the browser and get the
  adjustment set
:::

## Bootstrapping

### When and why

Every confidence interval so far came from a formula. Those formulas exist for
means, regression coefficients and a handful of other quantities, and they rely
on assumptions — usually that the estimate is approximately normally distributed
in repeated sampling.

The **bootstrap** replaces the formula with brute force: resample your data with
replacement many times, recompute the statistic each time, and use the spread of
those values as the sampling distribution.

Reach for it when:

- **There is no standard error formula** for what you computed — a median, a
  ratio of two coefficients, the difference between two correlations, a value
  derived through several steps.
- **The formula exists but its assumptions look shaky** — small samples, badly
  skewed outcomes, heavy tails.
- **Your quantity comes out of a multi-step pipeline**, so analytic propagation
  of uncertainty would be painful. Bootstrap the whole pipeline instead.

::: {.callout-warning}
## What the bootstrap does not fix
- **It does not remove bias.** If your estimate is systematically wrong,
  resampling reproduces that bias faithfully in every replicate.
- **It does not fix dependence.** If your data has repeated measures or
  clustering, resampling individual *rows* breaks the clustering and gives
  intervals that are too narrow. Resample the **clusters** instead —
  `rsample::group_bootstraps(group = id)`.
- **It needs a reasonable sample.** With 12 observations the bootstrap can only
  reshuffle 12 values; it cannot invent information that is not there.
- **It is not required for ordinary regression.** For a linear model with
  well-behaved residuals, the textbook interval is fine and faster.
:::

`rsample` handles the resampling, and the results are — again — list columns.

```{r}
#| label: bootstrap
#| message: false
library(rsample)

set.seed(2026)

boot_results <- model_data |>
  bootstraps(times = 200) |>
  mutate(
    coefficients = map(splits, \(s) {
      lm(body_mass_g ~ flipper_length_mm, data = analysis(s)) |> tidy()
    })
  ) |>
  select(id, coefficients) |>
  unnest(coefficients) |>
  filter(term == "flipper_length_mm")

# the percentile bootstrap interval
quantile(boot_results$estimate, c(0.025, 0.5, 0.975)) |> round(2)
```

```{r}
#| label: fig-bootstrap
#| fig-cap: "The bootstrap distribution of the slope."
#| fig-height: 3
boot_results |>
  ggplot(aes(estimate)) +
  geom_histogram(bins = 30, fill = "steelblue", colour = "white") +
  labs(x = "Slope (g per mm)", y = "Bootstrap replicates") +
  theme_minimal(base_size = 13)
```

`analysis(s)` extracts the resampled data from a split; `assessment(s)` gives
the rows left out. Useful when a standard error is hard to derive analytically —
a median, a ratio of coefficients, a complicated derived statistic.

## Mixed models

### When and why

Every model so far assumed the observations are **independent** — that knowing
one participant's outcome tells you nothing about another's. That assumption
fails whenever the data has structure:

- **Repeated measurements** on the same person: baseline and follow-up, a
  weekly diary, an accelerometer worn over seven days.
- **Clustering**: patients within clinics, pupils within schools, athletes
  within teams, siblings within families.
- **Crossed structures**: participants each rating several items.

Why it matters: two measurements from the same person carry less information
than two measurements from two people, because they resemble each other. Fit an
ordinary `lm()` to such data and it counts every row as fresh evidence, so the
standard errors come out **too small**, the confidence intervals too narrow and
the p-values too small. The estimate may be fine; the uncertainty is wrong, in
the direction that produces false positives.

A mixed model handles this by giving each cluster its own **random** intercept
(and optionally slope), drawn from a distribution:

- `(1 | id)` — each person gets their own baseline level.
- `(1 + time | id)` — each person gets their own baseline *and* their own slope
  over time.

::: {.callout-tip}
## The alternatives, and when they are enough
Mixed models are not the only answer to clustering:

- **Cluster-robust standard errors** — keep the `lm()` and correct only the
  uncertainty (`sandwich`/`clubSandwich`, or `vcov = ~id` in `marginaleffects`).
  Simpler, and often sufficient when the clustering is a nuisance rather than
  something you want to describe. Needs a reasonable number of clusters.
- **GEE** (`geepack`) — population-average effects rather than
  cluster-specific ones, which is sometimes the quantity you actually want.
- **Aggregate first** — one summary value per person, then an ordinary model.
  Crude, but honest and easy to explain, and hard to beat for a simple
  pre/post comparison.

Choose a mixed model when the variation *between* clusters is itself of
interest, when clusters have different numbers of observations, or when you
need person-specific predictions.
:::

`lme4` fits mixed models, and `broom.mixed` tidies them:

```{r}
#| label: mixed-model
#| message: false
library(lme4)

# a random intercept per species: each species gets its own baseline,
# and the slope is shared
mixed <- lmer(body_mass_g ~ flipper_length_mm + (1 | species), data = model_data)

broom.mixed::tidy(mixed, effects = "fixed", conf.int = TRUE)
```

```{r}
#| label: mixed-model-random
broom.mixed::tidy(mixed, effects = "ran_pars")
```

`glmmTMB` handles the cases `lme4` cannot — zero inflation, negative binomial,
beta responses — with a similar formula interface.

## Predictive modelling with tidymodels

Everything above is **inference**: what is the effect, and how certain are we?
`tidymodels` is built for a different question — **prediction**: how well does
this generalise to data it has not seen?

::: {.callout-warning}
## Load tidymodels last, and qualify your data
`tidymodels` attaches `modeldata`, which contains its own `penguins`. After
`library(tidymodels)`, a bare `penguins` silently resolves to
`modeldata::penguins` rather than the one you have been using:

```{r}
#| label: penguins-masking
#| message: false
library(tidymodels)

utils::getAnywhere("penguins")$where[1]

setdiff(names(palmerpenguins::penguins), names(penguins))
```

Same column names, but no `year` — a different data frame, chosen for you
without a word. Since R 4.5 there is a third `penguins` in `datasets`, with
shorter column names again (@sec-classes).

Two defences. Write `palmerpenguins::penguins` explicitly, which is what this
section does. Or call `tidymodels_prefer()`, which activates `conflicted` and
converts silent masking into a loud error:

```{r}
#| label: conflict-demo
#| eval: false
tidymodels_prefer()

penguins |> drop_na()
#> Error: [conflicted] penguins found in 3 packages.
#> Either pick the one you want with `::`:
#> * modeldata::penguins
#> * palmerpenguins::penguins
#> * datasets::penguins
```

An error is the better outcome. This section is placed at the end of the chapter
precisely because attaching `tidymodels` changes what later code means.
:::

```{r}
#| label: tidymodels
#| message: false
penguins_tm <- palmerpenguins::penguins |> drop_na()

set.seed(123)
split <- initial_split(penguins_tm, prop = 0.75, strata = species)

recipe_spec <- recipe(
  body_mass_g ~ flipper_length_mm + species + sex,
  data = training(split)
) |>
  step_dummy(all_nominal_predictors())

workflow_spec <- workflow() |>
  add_recipe(recipe_spec) |>
  add_model(linear_reg() |> set_engine("lm"))

fitted_wf <- fit(workflow_spec, data = training(split))

# performance on data the model has never seen
augment(fitted_wf, new_data = testing(split)) |>
  metrics(truth = body_mass_g, estimate = .pred)
```

Cross-validation, in three lines:

```{r}
#| label: tidymodels-cv
#| message: false
set.seed(42)

fit_resamples(workflow_spec, vfold_cv(training(split), v = 5)) |>
  collect_metrics() |>
  select(.metric, mean, n, std_err)
```

The pieces: **`rsample`** splits the data, **`recipes`** defines the
preprocessing, **`parsnip`** specifies the model independently of which engine
fits it, **`workflows`** binds preprocessing and model together so both are
applied identically to every resample, and **`yardstick`** computes the metrics.

::: {.callout-tip}
## Which framework?
Use **`lm()`/`glm()` + `broom` + `marginaleffects`** when you want to estimate
and report an effect — almost all epidemiological and clinical research.

Use **`tidymodels`** when the goal is prediction and you need resampling, tuning
and honest out-of-sample performance.

The two are not competitors, and using a predictive framework for an inferential
question is a common mistake: a cross-validated R² does not tell you whether an
exposure causes an outcome.
:::

::: {.webex-check .webex-box}
**Check yourself**

Adjusting for a collider
`r longmcq(c(
  "reduces bias, like adjusting for a confounder",
  "has no effect either way",
  answer = "introduces bias, by creating an association that was not there",
  "is required whenever the variable predicts the outcome"
))`

Adjusting for a mediator gives you the total effect of the exposure.
`r torf(FALSE)`

Which quantity says how strong an unmeasured confounder would have to be to
explain away your result?
`r mcq(c("the p-value", "the R squared", answer = "the E-value", "the variance inflation factor"))`

Ignoring clustering in the data usually makes confidence intervals
`r mcq(c("too wide", answer = "too narrow", "unbiased", "impossible to compute"))`

The bootstrap corrects a biased estimate. `r torf(FALSE)`

In `(1 + time | id)`, what varies between participants?
`r mcq(c("only the intercept", "only the residual variance", answer = "both the intercept and the slope over time", "nothing; it is a fixed effect"))`
:::

::: {.callout-note collapse="true"}
## Exercise 1: draw the DAG first

You want to estimate the effect of **regular training** on **resting heart
rate** in an observational cohort. You have age, sex, smoking, BMI, VO2max and
whether the person owns a fitness watch.

1. Write down which variables you think are confounders, which are mediators,
   and which might be colliders.
2. Build the DAG with `dagitty` and ask it for the adjustment set.
3. Does the answer match your intuition? Where does it differ?

::: {.callout-tip collapse="true"}
## Solution

One defensible reading — yours may differ, and that is the point, because the
DAG makes the disagreement explicit rather than hiding it in a variable list:

```{r}
#| label: ex1-causal-sol
#| message: false
library(dagitty)

hr_dag <- dagitty('dag {
  training -> heart_rate
  age -> training
  age -> heart_rate
  sex -> training
  sex -> heart_rate
  smoking -> training
  smoking -> heart_rate
  training -> vo2max
  vo2max -> heart_rate
  training -> watch
  heart_rate -> watch
}')

adjustmentSets(hr_dag, exposure = "training", outcome = "heart_rate")
```

- **Age, sex and smoking** are confounders: they influence both training and
  heart rate. Adjust.
- **VO2max is a mediator** — training improves fitness, which lowers heart rate.
  Adjusting for it would remove most of the effect you are trying to measure.
- **Owning a fitness watch is a collider**: people who train buy watches, and so
  do people worried about their heart rate. Adjusting for it induces bias.
- **BMI** is the genuinely ambiguous one. Confounder, mediator, or both,
  depending on the time ordering you assume. Say which you assumed.

Note that `adjustmentSets()` returns the *minimal sufficient* set given your
assumptions. It cannot tell you the assumptions are right — it only works out
their consequences correctly, which is still a great deal more than a variable
list does.
:::
:::

::: {.callout-note collapse="true"}
## Exercise 2: does clustering change your answer?

Simulate a study with 20 clinics, 20 patients each, where the outcome varies
between clinics but the treatment effect is zero.

1. Fit an ordinary `lm()` ignoring clinic. How often is the treatment
   "significant"?
2. Fit a mixed model with `(1 | clinic)`. How does the standard error change?
3. What does this tell you about analysing clustered data as if it were
   independent?

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: ex2-causal-sol
#| message: false
library(lme4)

set.seed(99)

simulate_once <- function() {
  clinics <- tibble(
    clinic = factor(1:20),
    # clinics genuinely differ in their average outcome
    clinic_effect = rnorm(20, 0, 5)
  )

  tibble(
    clinic = rep(factor(1:20), each = 20),
    treated = rep(c(0, 1), length.out = 400)
  ) |>
    left_join(clinics, by = "clinic") |>
    # note: NO treatment effect in the data-generating process
    mutate(outcome = clinic_effect + rnorm(400, 0, 3))
}

one <- simulate_once()

# ignoring the clinics
tidy(lm(outcome ~ treated, data = one)) |> filter(term == "treated")

# accounting for them
broom.mixed::tidy(
  lmer(outcome ~ treated + (1 | clinic), data = one),
  effects = "fixed"
) |>
  filter(term == "treated")
```

In this particular draw both give a similar answer, because the treatment was
balanced *within* every clinic — which is exactly the design that protects you.
The damage appears when treatment is unbalanced across clusters:

```{r}
#| label: ex2-causal-sol2
set.seed(3)

unbalanced <- simulate_once() |>
  # now whole clinics tend to be treated or not
  mutate(treated = as.numeric(as.integer(clinic) > 10))

tidy(lm(outcome ~ treated, data = unbalanced)) |>
  filter(term == "treated") |>
  select(estimate, std.error, p.value)

broom.mixed::tidy(
  lmer(outcome ~ treated + (1 | clinic), data = unbalanced),
  effects = "fixed"
) |>
  filter(term == "treated") |>
  select(estimate, std.error)
```

The estimates agree exactly (−0.84 both) and the **standard errors do not**:
0.48 from the `lm()` against 1.74 from the mixed model, nearly **four times
larger**. Ignoring the clustering treats 400 patients as 400 independent pieces
of evidence, when the treatment really varies across only 20 clinics — so the
effective sample size is closer to 20 than 400. The mixed model's larger
standard error is the honest one.

Notice what that does to a p-value. The same estimate, divided by a standard
error four times too small, gives a t statistic four times too large.

That is the general lesson: clustering rarely biases the estimate much, but it
routinely makes the uncertainty look smaller than it is — and a p-value computed
from a standard error that is too small is a false positive waiting to happen.
:::
:::

## More information

- [Causal Inference: What If](https://miguelhernan.org/whatifbook) — Hernán and
  Robins, free online
- [Causal Inference in R](https://www.r-causal.org/) — the same ideas in R
- [dagitty.net](https://dagitty.net/) — draw a DAG and get the adjustment set
- [marginaleffects](https://marginaleffects.com/) — including the chapters on
  g-computation and causal estimands
- [Tidy Modeling with R](https://www.tmwr.org/) — the tidymodels book
- [rsample](https://rsample.tidymodels.org/) and
  [lme4](https://github.com/lme4/lme4) documentation

