# Statistics and Modelling {#sec-modelling}

::: {.callout-note}
## What this chapter covers
How to fit models and get their results back as tidy data you can filter, plot
and tabulate — starting from a single `lm()` and building up to fitting many
models at once and keeping them in list columns. This is not a statistics
course: it assumes you know which model you want and shows you how to run it
without losing control of the output.
:::

```{r}
#| label: setup-modelling
#| include: false
library(webexercises)
```

```{r}
#| label: modelling-libs
#| message: false
library(tidyverse)
library(palmerpenguins)
library(broom)
```

## Before you model: shape and types

Two things decide whether a model does what you meant, and both are settled
before you call `lm()`.

### One row per observation

Modelling functions want **long, tidy data** (@sec-tidy): one row per
observation, one column per variable.

That matters most with repeated measurements. A wide table with `weight_pre`
and `weight_post` is convenient to read, but a model cannot use it — there is no
column that says *when* the measurement was taken.

```{r}
#| label: shape-wide-long
wide <- tibble(
  id = 1:3,
  group = c("training", "training", "control"),
  weight_pre = c(72, 68, 80),
  weight_post = c(70, 67, 81)
)

wide
```

```{r}
#| label: shape-long
long <- wide |>
  pivot_longer(
    starts_with("weight_"),
    names_to = "time",
    values_to = "weight_kg",
    names_prefix = "weight_"
  ) |>
  mutate(
    # an explicit level order makes "pre" the reference, which is what
    # you almost always want for a before/after comparison
    time = factor(time, levels = c("pre", "post")),
    id = factor(id)
  )

long
```

Now `weight_kg ~ time * group + (1 | id)` is expressible. In the wide form it is
not.

The exception is a **change score**: `post - pre` as a single outcome needs the
wide form. That is why you move between the two (@sec-pivot) — wide to compute
the difference, long to model the repeated measurements.

### Variable types

The class of a column decides how the model treats it, silently.

| Class | How a model treats it | Use it when |
|-------|----------------------|-------------|
| `numeric` / `integer` | continuous; one coefficient per unit | the variable really is a quantity |
| `factor` | categorical; one coefficient per level, against the first | ordinary categorical variables |
| `ordered` factor | **polynomial contrasts** — usually not what you want | rarely; see the warning below |
| `character` | converted to a factor on the fly, levels alphabetical | works, but you lose control of the reference |
| `logical` | `FALSE` = 0, `TRUE` = 1 | binary indicators |

Character and unordered factor give **identical fits** — R converts on the way
in:

```{r}
#| label: types-chr-vs-fct
d_types <- penguins |> drop_na(body_mass_g, species)

chr_fit <- lm(body_mass_g ~ as.character(species), data = d_types)
fct_fit <- lm(body_mass_g ~ species, data = d_types)

all.equal(unname(fitted(chr_fit)), unname(fitted(fct_fit)))
```

The difference is control. A character column orders its levels alphabetically,
so the reference group is whichever category comes first in the alphabet — a
decision made by spelling rather than by you. Convert to a factor and set the
reference deliberately (@sec-strings-factors).

::: {.callout-warning}
## Ordered factors do something you probably did not ask for
An ordered factor gets **polynomial contrasts**, not the level-by-level
comparisons everyone expects:

```{r}
#| label: types-ordered
d_types |>
  mutate(species_ord = factor(species, ordered = TRUE)) |>
  lm(body_mass_g ~ species_ord, data = _) |>
  tidy() |>
  select(term, estimate)
```

`.L` and `.Q` are the linear and quadratic trends across the levels, and the
intercept is the grand mean rather than a reference group. That is a sensible
default for a genuinely ordered dose — low, medium, high — where a trend is the
question. It is nonsense for species, and the output gives no warning that you
have asked for something unusual.

Since `ordered = TRUE` is exactly what you *should* use for sorting and plotting
(@sec-classes), it is easy to arrive here by accident. If you want ordinary
comparisons, drop the ordering for the model:

```{r}
#| label: types-ordered-fix
#| eval: false
data |> mutate(dose = factor(dose, ordered = FALSE))
```

Or set the contrasts explicitly with
`contrasts(data$dose) <- contr.treatment(3)`.
:::

::: {.callout-warning}
## A coded categorical is the silent one
If a categorical variable arrives from SPSS or a register as `1`, `2`, `3`, R
has no way to know it is not a quantity:

```{r}
#| label: types-numeric-code
d_types |>
  mutate(species_code = as.numeric(species)) |>
  lm(body_mass_g ~ species_code, data = _) |>
  tidy() |>
  select(term, estimate)
```

One coefficient instead of two, and it claims that going from species 1 to 2 has
the same effect as going from 2 to 3 — an assumption nobody made deliberately.
Nothing errors, nothing warns.

Check every categorical variable's class before modelling. `glimpse()` takes a
second, and `haven::as_factor()` converts labelled SPSS columns properly
(@sec-import).
:::

### A quick checklist

```{r}
#| label: types-check
penguins |>
  select(species, island, sex, body_mass_g, year) |>
  map_chr(\(x) class(x)[1])
```

Before fitting, confirm: is every categorical variable a factor with the
reference level you want, is every quantity numeric, and is the data long enough
that each observation has its own row?

## The problem broom solves

R's modelling functions are older than the tidyverse, and they return objects
designed to be *printed*, not manipulated.

```{r}
#| label: model-print
model <- lm(body_mass_g ~ flipper_length_mm + species + sex, data = penguins)

summary(model)
```

That output is readable and almost useless programmatically. You cannot filter
it, plot it, join it to anything, or put it in a table without copying numbers
by hand — which is how wrong numbers reach manuscripts.

`broom` converts model objects into data frames. Three verbs cover nearly
everything.

### `tidy()`: one row per term

```{r}
#| label: broom-tidy
tidy(model, conf.int = TRUE)
```

Now it is a tibble. It can be filtered, arranged, plotted and joined like any
other data.

```{r}
#| label: fig-coef-plot
#| fig-cap: "A coefficient plot, straight from tidy() output."
#| fig-height: 3
tidy(model, conf.int = TRUE) |>
  filter(term != "(Intercept)") |>
  ggplot(aes(estimate, fct_rev(term))) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high)) +
  labs(x = "Estimate (g)", y = NULL) +
  theme_minimal(base_size = 13)
```

For models on a log scale — logistic, Poisson, Cox — `exponentiate = TRUE` gives
you odds ratios, rate ratios or hazard ratios, with the confidence interval
exponentiated too:

```{r}
#| label: broom-exponentiate
penguins |>
  drop_na() |>
  glm(sex ~ body_mass_g + species, data = _, family = binomial()) |>
  tidy(exponentiate = TRUE, conf.int = TRUE) |>
  filter(term != "(Intercept)")
```

::: {.callout-note}
## The `_` placeholder
`data = _` passes the piped value to a named argument rather than the first one.
`glm()` takes the formula first and the data second, so the plain pipe would put
the penguins in the wrong slot. The placeholder needs a *named* argument — this
is one of the few places the base pipe is stricter than `%>%`, which uses `.`.
:::

### `glance()`: one row per model

```{r}
#| label: broom-glance
glance(model)
```

Model-level statistics: R², AIC, BIC, residual degrees of freedom, and `nobs` —
the number of observations actually used, which matters more than people expect
(see the warning below).

### `augment()`: one row per observation

```{r}
#| label: broom-augment
augment(model) |>
  select(body_mass_g, .fitted, .resid, .std.resid, .cooksd) |>
  head(4)
```

`augment()` adds fitted values, residuals, leverage and Cook's distance to the
data. Since the result is a data frame, diagnostics become ordinary `ggplot2`:

```{r}
#| label: fig-resid
#| fig-cap: "Residuals against fitted values, drawn with ggplot2."
#| fig-height: 3.2
augment(model) |>
  ggplot(aes(.fitted, .resid)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE) +
  labs(x = "Fitted (g)", y = "Residual (g)") +
  theme_minimal(base_size = 13)
```

### It works on hypothesis tests too

`tidy()` is not only for regression. Most `htest` objects tidy as well, which is
handy when you want a t-test result in a table rather than on screen:

```{r}
#| label: broom-htest
d <- penguins |> drop_na(body_mass_g, sex)

tidy(t.test(body_mass_g ~ sex, data = d))

tidy(cor.test(d$body_mass_g, d$flipper_length_mm))
```

::: {.webex-check .webex-box}
**Check yourself**

Which broom verb returns one row per model term?
`r mcq(c(answer = "tidy()", "glance()", "augment()", "summary()"))`

To get odds ratios from a logistic regression you add
`r mcq(c("family = binomial()", answer = "exponentiate = TRUE", "log = FALSE", "conf.int = TRUE"))`

`augment()` returns one row per
`r mcq(c("model term", "model", answer = "observation", "group"))`
:::

## Two gotchas worth knowing early

::: {.callout-warning}
## Models silently drop rows with missing values
`lm()` and friends default to `na.action = na.omit`: any row with a missing
value in *any* variable in the formula is dropped, without a message.

```{r}
#| label: na-drop
m1 <- lm(body_mass_g ~ flipper_length_mm, data = penguins)
m2 <- lm(body_mass_g ~ flipper_length_mm + sex, data = penguins)

nrow(penguins)
nobs(m1)
nobs(m2)
```

344 rows in, 342 in the first model, 333 in the second — because adding `sex`
brought its 11 missing values into the formula.

This has a consequence people miss: **`m1` and `m2` are fitted to different
data, so their AIC, BIC and R² are not comparable.** To compare models, fit them
all to the same complete-case set:

```{r}
#| label: na-drop-fix
model_data <- penguins |>
  drop_na(body_mass_g, flipper_length_mm, sex)

m1 <- lm(body_mass_g ~ flipper_length_mm, data = model_data)
m2 <- lm(body_mass_g ~ flipper_length_mm + sex, data = model_data)

c(nobs(m1), nobs(m2))
```

Always report `nobs()`, and always say how missing data was handled.
:::

::: {.callout-warning}
## The reference level is the first factor level
A categorical predictor is coded relative to its first level, so every
coefficient is a comparison *against that level*. Change the level order and the
whole table changes meaning:

```{r}
#| label: reference-level
d <- penguins |> drop_na()

# default: Adelie is the reference, because it is the first level
lm(body_mass_g ~ species, data = d) |>
  tidy() |>
  select(term, estimate)

# make Gentoo the reference instead
d |>
  mutate(species = fct_relevel(species, "Gentoo")) |>
  lm(body_mass_g ~ species, data = _) |>
  tidy() |>
  select(term, estimate)
```

Both are the same model — the fit, the predictions and the R² are identical —
but the numbers you would report are completely different. Set the reference
level deliberately with `fct_relevel()` (@sec-strings-factors), choosing the
group that makes clinical sense as the comparator, and say in the text which it
is.
:::

## Checking the model

`performance` runs the standard diagnostics and returns them as one figure:

```{r}
#| label: fig-check-model
#| fig-cap: "Diagnostic plots from performance::check_model()."
#| fig-width: 9
#| fig-height: 8
#| message: false
#| warning: false
performance::check_model(model)
```

```{r}
#| label: performance-numbers
performance::model_performance(model)

performance::check_collinearity(model)
```

::: {.callout-important}
## A diagnostic plot cannot tell you the model was appropriate
`check_model()` will happily draw six panels for a model that should never have
been fitted — one that ignores clustering, or that adjusts for a variable on the
causal pathway. It checks the model's *internal* assumptions, not whether the
model answers your question. That part is yours (@sec-ai).
:::

## Interpreting the model

A fitted model can answer three different questions, and most confusion about
"marginal effects" comes from not noticing which one is being asked.

| Question | Plain English | Function | Reported as |
|----------|---------------|----------|-------------|
| **Prediction** | What value do we expect? | `avg_predictions()` | the outcome's own units, or a probability |
| **Comparison** | How much difference between these two groups? | `avg_comparisons()` | a difference, ratio, odds ratio |
| **Slope** | How much does the outcome change per unit? | `avg_slopes()` | outcome units per predictor unit |

`marginaleffects` provides all three, and has become the standard tool for this.

```{r}
#| label: marginaleffects-setup
#| message: false
library(marginaleffects)
```

### What a slope is

A **slope** is the answer to: *if this predictor increased by one unit, how much
would the outcome change?*

```{r}
#| label: slope-basics
avg_slopes(model, variables = "flipper_length_mm")
```

Read as: **each additional millimetre of flipper is associated with about 20 g
more body mass**, holding species and sex constant. The units are always
"outcome units per predictor unit" — here grams per millimetre.

Three things follow from that definition, and each causes trouble in practice:

**The unit matters more than the number.** A slope of 20 g/mm and a slope of
20,000 g/m are the same fact. If a coefficient looks absurdly small, it is
usually because the predictor's unit is small. Report a step people can picture
— per 10 mm, per 100 g, per standard deviation:

```{r}
#| label: slope-scaled
# the effect of a 10 mm difference in flipper length
avg_comparisons(model, variables = list(flipper_length_mm = 10))
```

**A slope only exists for a continuous predictor.** "The slope of species" is
meaningless — species has no units and no in-between values. For a categorical
predictor you want a *comparison*, not a slope.

**In a non-linear model the slope is not constant.** In logistic regression the
effect of one more gram on the probability of being male depends on where you
start, because the curve is steeper in the middle than at the ends. That is the
whole reason the next section exists.

### When to report which

- **Report a slope** when the predictor is continuous and the relationship is
  roughly linear over the range that matters. "Each additional 10 mm of flipper
  is associated with 200 g more mass."
- **Report a comparison** when the predictor is categorical, or when a specific
  contrast is the question. "Males were on average 530 g heavier than females."
  This is what most clinical and epidemiological papers want.
- **Report predictions** when the absolute level matters, not just the
  difference — a risk, a predicted score, an expected count. "Predicted
  probability of the outcome was 12% in the control group and 7% in the
  treatment group."
- **Report the coefficient table too**, but as supporting material. In a linear
  model without interactions it says the same thing; in anything more
  complicated it does not.

```{r}
#| label: comparison-basics
# a comparison: the male-female difference, on the outcome's own scale
avg_comparisons(model, variables = "sex")
```

```{r}
#| label: prediction-basics
# predictions: the expected body mass of each species
avg_predictions(model, by = "species")
```

### Averaging over your data, or evaluating at the average

This is the distinction that trips people up, and it is worth understanding
because in a non-linear model the two give different answers.

To summarise a model in one number you have to decide **where** to evaluate it:

1. **Average over the data.** Compute the quantity for every penguin you
   actually observed, then average those. `marginaleffects` does this by
   default — hence the `avg_` prefix.
2. **Evaluate at the average.** Build one hypothetical average penguin — mean
   flipper length, mean body mass — and compute the quantity once, there.

In a **linear** model these are identical, and both equal the coefficient:

```{r}
#| label: ame-vs-mem-linear
c(
  coefficient = unname(coef(model)["flipper_length_mm"]),
  averaged_over_data = avg_slopes(model, variables = "flipper_length_mm")$estimate,
  at_the_average = slopes(model, variables = "flipper_length_mm",
                          newdata = "mean")$estimate
) |> round(3)
```

In a **non-linear** model they are not:

```{r}
#| label: ame-vs-mem-logistic
sex_model <- penguins |>
  drop_na() |>
  glm(sex ~ body_mass_g + flipper_length_mm + species, data = _,
      family = binomial())

# averaged over the real penguins
avg_slopes(sex_model, variables = "body_mass_g")
```

```{r}
#| label: ame-vs-mem-logistic2
# evaluated at one hypothetical average penguin
slopes(sex_model, variables = "body_mass_g", newdata = "mean")
```

Same model, same data. Averaged over the penguins, an extra gram raises the
probability of being male by 0.00067 and the result is unambiguous
(p < 0.001). Evaluated at the average penguin it is 0.000075, nine times
smaller, and p = 0.06.

::: {.callout-important}
## Why they differ, and which to report
The logistic curve is steep in the middle and flat at the ends. The "average
penguin" here has an average body mass *and* an average flipper length — a
combination that sits on a flat part of the curve for most species, so the slope
measured there is small. Averaging over real penguins includes many that sit on
the steep part, and their larger slopes pull the average up.

The deeper problem is that **the average penguin does not exist**. Watch what
happens when you ask for predicted probabilities at averaged covariates:

```{r}
#| label: average-penguin
predictions(sex_model, newdata = datagrid(species = unique)) |>
  as_tibble() |>
  select(species, estimate, conf.low, conf.high) |>
  mutate(across(where(is.numeric), \(x) round(x, 3)))
```

A Gentoo of *average* body mass is a very small Gentoo, so the model says it is
almost certainly female — 0.001. Compare with averaging over the real Gentoos:

```{r}
#| label: average-penguin2
avg_predictions(sex_model, by = "species") |>
  as_tibble() |>
  select(species, estimate) |>
  mutate(estimate = round(estimate, 3))
```

About 0.51, which matches the actual sex ratio. The first answer describes a
penguin nobody measured; the second describes the sample.

**Default to averaging over your data** — the `avg_*()` functions. Evaluate at
specific values when you deliberately want a "what if" for a described person,
and then say which values you fixed.
:::

::: {.callout-note}
## The jargon, so you recognise it
The literature calls the first approach **average marginal effects (AME)** and
the second **marginal effects at the mean (MEM)**; fixing covariates at chosen
values gives **marginal effects at representative values (MER)**. "Marginal" is
borrowed from economics, where it means *incremental* — the effect of one more
unit — not the everyday sense of "minor".

`marginaleffects` defaults to AME, `emmeans` to MEM, which is one reason the two
packages can report different numbers for the same model.

Andrew Heiss's [guide to marginal
effects](https://www.andrewheiss.com/blog/2022/05/20/marginalia/) works through
this in far more depth than a course book can, and is the reference to reach for
when the distinction actually bites.
:::

### emmeans

`emmeans` answers the same kinds of question with a longer history and stronger
support for multiplicity adjustment:

```{r}
#| label: emmeans
#| message: false
library(emmeans)

means <- emmeans(model, ~species)
means

pairs(means)
```

`pairs()` applies a Tukey adjustment by default — worth knowing you are getting
it, and worth stating in the methods. Note also that these are estimated at
fixed covariate values rather than averaged over the data, so they will not
always match `avg_predictions()`.

## Many models at once

This is where list columns earn their place. The pattern from @sec-functions —
`nest()`, `map()`, `unnest()` — is the tidyverse answer to "fit this model
separately for every group".

### The pattern

```{r}
#| label: many-models
by_species <- penguins |>
  drop_na(body_mass_g, flipper_length_mm) |>
  # nest() collapses each species' rows into a data frame in a list column
  nest(.by = species)

# print() rather than plain `by_species`: this book prints data frames as
# formatted tables, which would try to flatten the whole nested data frame
# into a cell. The tibble print method abbreviates it instead.
print(by_species)
```

`data` is a **list column**: each cell holds an entire data frame. That is legal
because a tibble column is a list, and a list can hold anything.

```{r}
#| label: many-models-fit
results <- by_species |>
  mutate(
    model = map(data, \(d) lm(body_mass_g ~ flipper_length_mm, data = d)),
    coefficients = map(model, \(m) tidy(m, conf.int = TRUE)),
    fit = map(model, glance)
  )

print(results)
```

Three models, each sitting next to its own data and its own label. Nothing can
get out of order, because the model and the group name are in the same row —
which is the real advantage over keeping a separate list of models and a
separate vector of names.

Then unnest whichever part you want:

```{r}
#| label: many-models-coefs
results |>
  select(species, coefficients) |>
  unnest(coefficients) |>
  filter(term == "flipper_length_mm") |>
  mutate(across(where(is.numeric), \(x) round(x, 2)))
```

```{r}
#| label: many-models-fit-stats
results |>
  select(species, fit) |>
  unnest(fit) |>
  select(species, r.squared, sigma, nobs)
```

::: {.callout-important}
## Three separate models are not one model with an interaction
The slopes above differ between species, which looks like an interaction — and
it is tempting to treat it as one. But the two analyses are not the same:

- **Three separate models** each estimate their own residual variance from their
  own species only.
- **One interaction model** pools the residual variance across all species and
  gives you a formal test of whether the slopes differ.

```{r}
#| label: interaction-model
interaction_model <- penguins |>
  drop_na(body_mass_g, flipper_length_mm) |>
  lm(body_mass_g ~ flipper_length_mm * species, data = _)

tidy(interaction_model) |>
  filter(str_detect(term, ":")) |>
  select(term, estimate, p.value)
```

Which you want depends on your question. Stratified models suit "describe each
group separately"; the interaction model suits "do these groups differ?". Only
the second gives you a p-value for the difference. `map()` makes both equally
easy to produce, which is exactly why you have to decide deliberately.
:::

### Several outcomes at once

The same pattern, rotated: instead of one model per group, one model per
outcome.

```{r}
#| label: many-outcomes
outcomes <- c("body_mass_g", "flipper_length_mm", "bill_length_mm")

outcome_results <- tibble(outcome = outcomes) |>
  mutate(
    model = map(outcome, \(y) {
      # reformulate() builds a formula from strings
      lm(reformulate(c("species", "sex"), response = y), data = penguins)
    }),
    coefficients = map(model, \(m) tidy(m, conf.int = TRUE))
  ) |>
  select(outcome, coefficients) |>
  unnest(coefficients) |>
  filter(term == "sexmale")

outcome_results |>
  mutate(across(where(is.numeric), \(x) round(x, 2)))
```

::: {.callout-warning}
## Many models means many p-values
Fitting a model per outcome, or per subgroup, multiplies your tests. With twenty
outcomes at the 0.05 level you expect one "significant" result from noise alone.

If the analyses were pre-specified, adjust for multiplicity or report them as a
set. If they were not, describe them as exploratory and say so. The `map()`
pattern makes it trivially easy to fit fifty models; the responsibility for what
that does to your error rate is unchanged.
:::

### Unadjusted and adjusted side by side

A staple of epidemiological papers, and a natural fit for list columns:

```{r}
#| label: adjusted-unadjusted
model_data <- penguins |> drop_na()

specifications <- tibble(
  label = c("Unadjusted", "Adjusted for species", "Adjusted for species and sex"),
  formula = c(
    "body_mass_g ~ flipper_length_mm",
    "body_mass_g ~ flipper_length_mm + species",
    "body_mass_g ~ flipper_length_mm + species + sex"
  )
)

specifications |>
  mutate(
    model = map(formula, \(f) lm(as.formula(f), data = model_data)),
    coefficients = map(model, \(m) tidy(m, conf.int = TRUE))
  ) |>
  select(label, coefficients) |>
  unnest(coefficients) |>
  filter(term == "flipper_length_mm") |>
  select(label, estimate, conf.low, conf.high) |>
  mutate(across(where(is.numeric), \(x) round(x, 1)))
```

The estimate falls from 50.2 to 40.6 once species is adjusted for, and to 20.0
once sex is added as well — less than half the unadjusted value. Because every
model is fitted to the same `model_data`, the three estimates are directly
comparable.

Why does adjustment change it so much? Because species differs in both
variables at once:

```{r}
#| label: confounding-visible
model_data |>
  summarise(
    mean_flipper = round(mean(flipper_length_mm), 0),
    mean_mass = round(mean(body_mass_g), 0),
    .by = species
  ) |>
  arrange(mean_flipper)
```

Gentoos have both the longest flippers and the largest mass. Pooled, some of
what looks like "longer flipper, therefore heavier" is really "Gentoo, therefore
both". Holding species constant removes that path and leaves the
within-species association — which @fig-interaction-predictions shows directly:
the within-species lines are flatter than a line drawn through the whole cloud
would be.

This is the same phenomenon as Simpson's paradox in @sec-eda, in its milder and
far more common form: adjustment does not reverse the association here, it
shrinks it.

## Interactions

An interaction says that the effect of one variable **depends on** another: the
slope differs between groups, or a treatment works better for some people than
others. It is the formal version of the stratified analysis above.

### Writing one

```{r}
#| label: interaction-syntax
#| eval: false
# a * b expands to a + b + a:b - both main effects and the interaction
lm(body_mass_g ~ flipper_length_mm * species, data = d)

# a:b is the interaction term only. Almost always a mistake on its own:
# an interaction without its main effects is very hard to interpret
lm(body_mass_g ~ flipper_length_mm:species, data = d)
```

Use `*`. Reach for `:` only when you have a specific reason and can say what the
model means without the main effects.

### Why the coefficients are hard to read

```{r}
#| label: interaction-coefs
d_int <- penguins |> drop_na(body_mass_g, flipper_length_mm, species, sex)

interaction_fit <- lm(body_mass_g ~ flipper_length_mm * species, data = d_int)

tidy(interaction_fit) |>
  select(term, estimate, p.value) |>
  mutate(across(where(is.numeric), \(x) round(x, 3)))
```

Read that carefully and notice how little it tells you directly:

- `flipper_length_mm` (32.7) is **not** the effect of flipper length. It is the
  slope **for Adelie only** — the reference level.
- `speciesGentoo` (-4166) is **not** how much heavier Gentoos are. It is the
  difference in intercepts at `flipper_length_mm = 0`, a flipper length no
  penguin has. The number is large and negative and means almost nothing on its
  own.
- `flipper_length_mm:speciesGentoo` (21.5) is the *difference in slopes* between
  Gentoo and Adelie. To get the Gentoo slope you add: 32.7 + 21.5 = 54.2.

This is why interaction models get misreported. Every coefficient is conditional
on the others being zero, and "zero" is often impossible.

### The easy way: ask for the quantity you want

Rather than adding coefficients by hand, ask `marginaleffects` directly.

**The slope in each group:**

```{r}
#| label: interaction-slopes
avg_slopes(interaction_fit, variables = "flipper_length_mm", by = "species")
```

54.2 for Gentoo, matching the hand calculation — with a correct standard error,
which adding coefficients would not have given you.

**Do the slopes actually differ?**

```{r}
#| label: interaction-hypothesis
avg_slopes(
  interaction_fit,
  variables = "flipper_length_mm",
  by = "species",
  hypothesis = ~pairwise
)
```

### How to read that, and where to see it

Three rows, and each is a difference between two slopes:

- **Chinstrap − Adelie = 1.88** (95% CI −13.5 to 17.3). The interval comfortably
  includes zero, so the data are consistent with these two species having the
  same slope. Note what this does *not* say: it is not evidence that the slopes
  are equal, only that this sample cannot distinguish them — and the interval is
  wide enough to contain differences that would matter.
- **Gentoo − Adelie = 21.5** (7.8 to 35.1) and **Gentoo − Chinstrap = 19.6**
  (3.6 to 35.6). Both intervals exclude zero, so Gentoo's slope is
  distinguishable from both.

You can see the same thing in three places, which is a good way to build
confidence that you have read it correctly.

**In the coefficient table.** Compare the pairwise output with the interaction
terms from the model above: `flipper_length_mm:speciesGentoo` was 21.5, and
"(Gentoo) − (Adelie)" is 21.5. They are the same number, because an interaction
coefficient *is* the difference in slopes against the reference level. What the
coefficient table cannot give you directly is Gentoo − Chinstrap, since neither
is the reference — you would have to subtract two coefficients and then work out
the standard error of that difference. `hypothesis = ~pairwise` does exactly
that, correctly.

**In the figure.** In @fig-interaction-predictions the Gentoo line is visibly
steeper than the other two, which run close to parallel. Non-parallel lines are
a difference in slopes; how confident you can be about it is what the intervals
above quantify.

**In the slope estimates themselves.**

```{r}
#| label: interaction-slopes-table
avg_slopes(interaction_fit, variables = "flipper_length_mm", by = "species") |>
  as_tibble() |>
  select(species, estimate, conf.low, conf.high) |>
  mutate(across(where(is.numeric), \(x) round(x, 1)))
```

Adelie 32.7, Chinstrap 34.6, Gentoo 54.2 — the first two are close, the third is
much larger.

::: {.callout-warning}
## Do not read the difference off two separate intervals
It is tempting to compare the two confidence intervals above and conclude
"they overlap, so no difference". That reasoning is wrong, and it is wrong in
the direction that loses findings: two estimates whose 95% intervals overlap can
still differ significantly, because the interval for a *difference* is not the
union of the two individual intervals.

Here the two approaches happen to agree, but they need not. The test of a
difference is the contrast — the `hypothesis = ~pairwise` output — not a visual
comparison of error bars. When you want to show the comparison in a figure, plot
the *differences* with their intervals, not the two groups side by side.
:::

Written up, this is one sentence: *the association between flipper length and
body mass was stronger in Gentoo (54.2 g/mm, 95% CI 44.1–64.3) than in Adelie
(32.7, 23.5–41.9; difference 21.5, 7.8–35.1) or Chinstrap (34.6, 22.2–46.9;
difference 19.6, 3.6–35.6), while Adelie and Chinstrap did not differ detectably
(1.9, −13.5–17.3).*

::: {.callout-note}
## `hypothesis` takes a formula
In older code and tutorials you will see `hypothesis = "pairwise"` as a string.
Current `marginaleffects` wants a formula: `~pairwise`, `~reference` (each level
against the first) or `~sequential` (each against the previous). The string form
errors, and the message is not obvious — worth recognising when you meet old
code.
:::

### Plot it

For an interaction, a figure is almost always clearer than a table.

```{r}
#| label: fig-interaction-predictions
#| fig-cap: "Predicted body mass across flipper length, one line per species."
#| fig-height: 3.6
plot_predictions(
  interaction_fit,
  condition = c("flipper_length_mm", "species")
) +
  labs(x = "Flipper length (mm)", y = "Predicted body mass (g)") +
  theme_minimal(base_size = 13)
```

Non-parallel lines *are* the interaction. If the lines were parallel, the
interaction term would be near zero.

`plot_predictions()` returns a `ggplot` object, so you can add layers, labels
and themes to it like any other figure — including the raw data underneath:

```{r}
#| label: fig-interaction-with-data
#| fig-cap: "The same predictions with the observations behind them."
#| fig-height: 3.6
plot_predictions(
  interaction_fit,
  condition = c("flipper_length_mm", "species"),
  points = 0.4
) +
  labs(x = "Flipper length (mm)", y = "Body mass (g)") +
  theme_minimal(base_size = 13)
```

`plot_slopes()` shows the effect itself rather than the prediction, which
answers "how much does the slope differ, and is the difference distinguishable
from zero?" directly:

```{r}
#| label: fig-interaction-slopes
#| fig-cap: "The slope of body mass on flipper length, per species, with 95% intervals."
#| fig-height: 3
plot_slopes(
  interaction_fit,
  variables = "flipper_length_mm",
  condition = "species"
) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  labs(x = NULL, y = "Slope (g per mm)") +
  theme_minimal(base_size = 13)
```

### Categorical by categorical

The same tools handle a two-factor interaction, where the question is usually
whether a group difference is itself different between groups.

```{r}
#| label: interaction-cat
cat_fit <- lm(body_mass_g ~ species * sex, data = d_int)

# the male - female difference, within each species
avg_comparisons(cat_fit, variables = "sex", by = "species")
```

```{r}
#| label: interaction-did
# and whether those differences differ: a difference in differences
avg_comparisons(
  cat_fit,
  variables = "sex",
  by = "species",
  hypothesis = ~pairwise
)
```

Males are heavier in every species, but by different amounts: about 805 g in
Gentoo, 675 g in Adelie and 412 g in Chinstrap.

The pairwise output then says which of those gaps differ from each other.
Reading it the same way as before: Chinstrap − Adelie is about −263 g with an
interval excluding zero, so the sex gap really is smaller in Chinstrap;
Gentoo − Adelie is about +130 g with an interval that includes zero, so those
two are not distinguishable. The finding is that **the sex difference is roughly
half as large in Chinstrap as in the other two species** — a statement about a
difference between differences, which is what an interaction between two factors
means.

You can see it directly in the group means, without any model:

```{r}
#| label: interaction-did-raw
d_int |>
  summarise(mean_mass = mean(body_mass_g), .by = c(species, sex)) |>
  pivot_wider(names_from = sex, values_from = mean_mass) |>
  mutate(sex_gap = round(male - female, 0))
```

The `sex_gap` column reproduces the model's comparisons almost exactly, because
this model is saturated — species, sex and their interaction between them
describe every cell, so the fitted values *are* the cell means. That will not be
true once you add covariates, which is when you need the model to do the
averaging for you.

::: {.callout-tip}
## Centre continuous predictors and the main effects become readable
The `speciesGentoo` coefficient was uninterpretable because it described a
penguin with a 0 mm flipper. Centre the predictor and it describes a penguin of
average flipper length instead:

```{r}
#| label: interaction-centred
d_int |>
  mutate(flipper_c = flipper_length_mm - mean(flipper_length_mm)) |>
  lm(body_mass_g ~ flipper_c * species, data = _) |>
  tidy() |>
  select(term, estimate) |>
  mutate(estimate = round(estimate, 1))
```

The intercept is now the predicted mass of an average-flippered Adelie, and
`speciesGentoo` is the Gentoo–Adelie difference at average flipper length — both
quantities that exist. The fit, the predictions and the interaction term are
completely unchanged; only the parameterisation is easier to read.
:::

::: {.callout-warning}
## Interactions need much more data than main effects
Detecting an interaction reliably takes roughly **four times** the sample size
needed for a main effect of the same magnitude. Two consequences:

- A non-significant interaction is weak evidence of no interaction. Report the
  estimate and its interval, not just the p-value.
- A *significant* interaction found by testing many subgroups is very likely
  noise. Pre-specify the interactions you will test, and treat the rest as
  exploratory (@sec-eda).

The tools above make it easy to fit an interaction for every variable you have.
That ease is not evidence that you should.
:::

## Bootstrapping

`rsample` handles resampling, and the results are — again — list columns.

```{r}
#| label: bootstrap
#| message: false
library(rsample)

set.seed(2026)

boot_results <- model_data |>
  bootstraps(times = 200) |>
  mutate(
    coefficients = map(splits, \(s) {
      lm(body_mass_g ~ flipper_length_mm, data = analysis(s)) |> tidy()
    })
  ) |>
  select(id, coefficients) |>
  unnest(coefficients) |>
  filter(term == "flipper_length_mm")

# the percentile bootstrap interval
quantile(boot_results$estimate, c(0.025, 0.5, 0.975)) |> round(2)
```

```{r}
#| label: fig-bootstrap
#| fig-cap: "The bootstrap distribution of the slope."
#| fig-height: 3
boot_results |>
  ggplot(aes(estimate)) +
  geom_histogram(bins = 30, fill = "steelblue", colour = "white") +
  labs(x = "Slope (g per mm)", y = "Bootstrap replicates") +
  theme_minimal(base_size = 13)
```

`analysis(s)` extracts the resampled data from a split; `assessment(s)` gives
the rows left out. Useful when a standard error is hard to derive analytically —
a median, a ratio of coefficients, a complicated derived statistic.

## Reporting the model

`gtsummary` (@sec-tables) builds the table directly from the model object, so
the numbers cannot drift out of sync with the analysis:

```{r}
#| label: tbl-regression
#| message: false
library(gtsummary)

unadjusted <- lm(body_mass_g ~ flipper_length_mm, data = model_data) |>
  tbl_regression()

adjusted <- lm(
  body_mass_g ~ flipper_length_mm + species + sex,
  data = model_data
) |>
  tbl_regression()

tbl_merge(
  list(unadjusted, adjusted),
  tab_spanner = c("**Unadjusted**", "**Adjusted**")
)
```

## Mixed models

Repeated measurements on the same participant violate independence. `lme4`
fits mixed models, and `broom.mixed` tidies them:

```{r}
#| label: mixed-model
#| message: false
library(lme4)

# a random intercept per species: each species gets its own baseline,
# and the slope is shared
mixed <- lmer(body_mass_g ~ flipper_length_mm + (1 | species), data = model_data)

broom.mixed::tidy(mixed, effects = "fixed", conf.int = TRUE)
```

```{r}
#| label: mixed-model-random
broom.mixed::tidy(mixed, effects = "ran_pars")
```

`glmmTMB` handles the cases `lme4` cannot — zero inflation, negative binomial,
beta responses — with a similar formula interface.

::: {.webex-check .webex-box}
**Check yourself**

`m1` uses 342 observations and `m2` uses 333, because `m2` has an extra
predictor with missing values. Can you compare their AIC?
`r mcq(c("Yes, AIC is always comparable", answer = "No, they are fitted to different data", "Only if the R2 is similar", "Only for linear models"))`

A list column holds
`r mcq(c("only numbers", "only strings", answer = "any R object, including data frames and models", "nothing; tibbles cannot do this"))`

Which function gives the average difference between two groups from a fitted
model?
`r mcq(c("avg_predictions()", answer = "avg_comparisons()", "avg_slopes()", "glance()"))`

Fitting one model per group is the same analysis as one model with an
interaction. `r torf(FALSE)`

Changing the reference level of a factor changes the model's predictions.
`r torf(FALSE)`
:::

::: {.callout-note collapse="true"}
## Exercise 1: from model to table

Using `penguins`:

1. Fit a model of `bill_length_mm` on `body_mass_g`, `species` and `sex`.
2. Report the coefficients with confidence intervals, rounded sensibly.
3. How many observations did the model actually use, and why is that not 344?
4. Make a coefficient plot.
5. Report the average difference between males and females.

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: ex1-model-sol
model_ex <- lm(
  bill_length_mm ~ body_mass_g + species + sex,
  data = penguins
)

tidy(model_ex, conf.int = TRUE) |>
  mutate(across(where(is.numeric), \(x) round(x, 3)))

nobs(model_ex)
```

333, not 344: two penguins have no bill measurement and eleven have no recorded
sex, and `lm()` drops any row missing any variable in the formula.

```{r}
#| label: fig-ex1-model
#| fig-cap: "Coefficients with 95% confidence intervals."
#| fig-height: 3
tidy(model_ex, conf.int = TRUE) |>
  filter(term != "(Intercept)") |>
  ggplot(aes(estimate, fct_rev(term))) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high)) +
  labs(x = "Estimate (mm)", y = NULL) +
  theme_minimal(base_size = 13)
```

```{r}
#| label: ex1-model-sol2
avg_comparisons(model_ex, variables = "sex")
```

Note the coefficient for `body_mass_g` is tiny — but the unit is one gram. An
effect per 100 g or per kilogram is far more readable, which is a good argument
for scaling predictors before fitting, or for using `avg_comparisons()` with an
explicit contrast.
:::
:::

::: {.callout-note collapse="true"}
## Exercise 2: many models

1. Fit the model from Exercise 1 separately for each island.
2. Produce a table of the `sexmale` coefficient with its confidence interval,
   one row per island, alongside the number of observations.
3. Plot the estimates with their intervals.
4. Which island has the least precise estimate? Check whether the obvious
   explanation is the right one.

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: ex2-model-sol
island_models <- penguins |>
  drop_na(bill_length_mm, body_mass_g, sex) |>
  nest(.by = island) |>
  mutate(
    model = map(data, \(d) lm(bill_length_mm ~ body_mass_g + sex, data = d)),
    coefficients = map(model, \(m) tidy(m, conf.int = TRUE)),
    n = map_int(data, nrow)
  )

island_estimates <- island_models |>
  select(island, n, coefficients) |>
  unnest(coefficients) |>
  filter(term == "sexmale") |>
  select(island, n, estimate, conf.low, conf.high) |>
  mutate(across(where(is.numeric), \(x) round(x, 2)))

island_estimates
```

```{r}
#| label: fig-ex2-model
#| fig-cap: "Sex difference in bill length by island, with 95% intervals."
#| fig-height: 2.8
island_estimates |>
  ggplot(aes(estimate, island)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high)) +
  labs(x = "Male - female (mm)", y = NULL) +
  theme_minimal(base_size = 13)
```

The obvious guess is Torgersen, which has by far the fewest penguins (47 against
123 and 163). It is wrong: **Dream** has the widest interval.

```{r}
#| label: ex2-model-width
island_estimates |>
  mutate(width = conf.high - conf.low) |>
  arrange(desc(width)) |>
  select(island, n, width)
```

Precision depends on the residual variance and the spread of the predictors, not
only on the sample size. Dream holds two species, Adelie and Chinstrap, whose
bills differ markedly — and this model does not adjust for species, so that
between-species variation lands in the residual and widens the interval.
Torgersen is Adelie only, so it is far more homogeneous and gets more precision
from fewer birds.

Two lessons. First, sample size belongs *in* the table, because three separate
models produce three estimates that look equally solid on a plot. Second, a
stratified analysis inherits whatever the model omits: stratifying by island did
not remove the species confounding, it just hid it inside one stratum.

:::
:::

::: {.callout-note collapse="true"}
## Exercise 3: does it matter which species is the reference?

1. Fit `body_mass_g ~ species` and note the coefficients.
2. Refit with Chinstrap as the reference level.
3. Compare the two tables. What changed, and what did not?
4. Confirm your answer by comparing the fitted values from both models.

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: ex3-model-sol
d3 <- penguins |> drop_na(body_mass_g)

fit_a <- lm(body_mass_g ~ species, data = d3)
fit_b <- d3 |>
  mutate(species = fct_relevel(species, "Chinstrap")) |>
  lm(body_mass_g ~ species, data = _)

tidy(fit_a) |> select(term, estimate)
tidy(fit_b) |> select(term, estimate)
```

The coefficients change completely, because they are comparisons against a
different group. What does **not** change:

```{r}
#| label: ex3-model-check
# identical fit
c(glance(fit_a)$r.squared, glance(fit_b)$r.squared)

# identical predictions, to within floating-point noise
all.equal(unname(fitted(fit_a)), unname(fitted(fit_b)))
```

Same model, same fit, same predictions — only the parameterisation differs. The
practical lesson: choose the reference level so that the comparisons in your
table are the ones you actually want to discuss, and state it in the methods.
Readers cannot tell from the table alone which group is the comparator unless
you say.
:::
:::

## Predictive modelling with tidymodels

Everything above is **inference**: what is the effect, and how certain are we?
`tidymodels` is built for a different question — **prediction**: how well does
this generalise to data it has not seen?

::: {.callout-warning}
## Load tidymodels last, and qualify your data
`tidymodels` attaches `modeldata`, which contains its own `penguins`. After
`library(tidymodels)`, a bare `penguins` silently resolves to
`modeldata::penguins` rather than the one you have been using:

```{r}
#| label: penguins-masking
#| message: false
library(tidymodels)

utils::getAnywhere("penguins")$where[1]

setdiff(names(palmerpenguins::penguins), names(penguins))
```

Same column names, but no `year` — a different data frame, chosen for you
without a word. Since R 4.5 there is a third `penguins` in `datasets`, with
shorter column names again (@sec-classes).

Two defences. Write `palmerpenguins::penguins` explicitly, which is what this
section does. Or call `tidymodels_prefer()`, which activates `conflicted` and
converts silent masking into a loud error:

```{r}
#| label: conflict-demo
#| eval: false
tidymodels_prefer()

penguins |> drop_na()
#> Error: [conflicted] penguins found in 3 packages.
#> Either pick the one you want with `::`:
#> * modeldata::penguins
#> * palmerpenguins::penguins
#> * datasets::penguins
```

An error is the better outcome. This section is placed at the end of the chapter
precisely because attaching `tidymodels` changes what later code means.
:::

```{r}
#| label: tidymodels
#| message: false
penguins_tm <- palmerpenguins::penguins |> drop_na()

set.seed(123)
split <- initial_split(penguins_tm, prop = 0.75, strata = species)

recipe_spec <- recipe(
  body_mass_g ~ flipper_length_mm + species + sex,
  data = training(split)
) |>
  step_dummy(all_nominal_predictors())

workflow_spec <- workflow() |>
  add_recipe(recipe_spec) |>
  add_model(linear_reg() |> set_engine("lm"))

fitted_wf <- fit(workflow_spec, data = training(split))

# performance on data the model has never seen
augment(fitted_wf, new_data = testing(split)) |>
  metrics(truth = body_mass_g, estimate = .pred)
```

Cross-validation, in three lines:

```{r}
#| label: tidymodels-cv
#| message: false
set.seed(42)

fit_resamples(workflow_spec, vfold_cv(training(split), v = 5)) |>
  collect_metrics() |>
  select(.metric, mean, n, std_err)
```

The pieces: **`rsample`** splits the data, **`recipes`** defines the
preprocessing, **`parsnip`** specifies the model independently of which engine
fits it, **`workflows`** binds preprocessing and model together so both are
applied identically to every resample, and **`yardstick`** computes the metrics.

::: {.callout-tip}
## Which framework?
Use **`lm()`/`glm()` + `broom` + `marginaleffects`** when you want to estimate
and report an effect — almost all epidemiological and clinical research.

Use **`tidymodels`** when the goal is prediction and you need resampling, tuning
and honest out-of-sample performance.

The two are not competitors, and using a predictive framework for an inferential
question is a common mistake: a cross-validated R² does not tell you whether an
exposure causes an outcome.
:::

## More information

- [broom](https://broom.tidymodels.org/) and its
  [vignette on many models](https://broom.tidymodels.org/articles/broom.html)
- [R for Data Science: many models with list columns](https://r4ds.hadley.nz/rectangling)
- [marginaleffects](https://marginaleffects.com/) — the accompanying book is the
  best available explanation of predictions, comparisons and slopes
- [emmeans vignettes](https://rvlenth.github.io/emmeans/articles/)
- [easystats](https://easystats.github.io/easystats/) — `performance`,
  `parameters`, `effectsize`, `report`
- [Tidy Modeling with R](https://www.tmwr.org/) — the tidymodels book
- [Regression and Other Stories](https://avehtari.github.io/ROS-Examples/) — for
  the statistics rather than the code
