# Statistics and Modelling {#sec-modelling}

::: {.callout-note}
## What this chapter covers
How to fit models and get their results back as tidy data you can filter, plot
and tabulate — starting from a single `lm()` and building up to fitting many
models at once and keeping them in list columns. This is not a statistics
course: it assumes you know which model you want and shows you how to run it
without losing control of the output.
:::

```{r}
#| label: setup-modelling
#| include: false
library(webexercises)
```

```{r}
#| label: modelling-libs
#| message: false
library(tidyverse)
library(palmerpenguins)
library(broom)
```

## The problem broom solves

R's modelling functions are older than the tidyverse, and they return objects
designed to be *printed*, not manipulated.

```{r}
#| label: model-print
model <- lm(body_mass_g ~ flipper_length_mm + species + sex, data = penguins)

summary(model)
```

That output is readable and almost useless programmatically. You cannot filter
it, plot it, join it to anything, or put it in a table without copying numbers
by hand — which is how wrong numbers reach manuscripts.

`broom` converts model objects into data frames. Three verbs cover nearly
everything.

### `tidy()`: one row per term

```{r}
#| label: broom-tidy
tidy(model, conf.int = TRUE)
```

Now it is a tibble. It can be filtered, arranged, plotted and joined like any
other data.

```{r}
#| label: fig-coef-plot
#| fig-cap: "A coefficient plot, straight from tidy() output."
#| fig-height: 3
tidy(model, conf.int = TRUE) |>
  filter(term != "(Intercept)") |>
  ggplot(aes(estimate, fct_rev(term))) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high)) +
  labs(x = "Estimate (g)", y = NULL) +
  theme_minimal(base_size = 13)
```

For models on a log scale — logistic, Poisson, Cox — `exponentiate = TRUE` gives
you odds ratios, rate ratios or hazard ratios, with the confidence interval
exponentiated too:

```{r}
#| label: broom-exponentiate
penguins |>
  drop_na() |>
  glm(sex ~ body_mass_g + species, data = _, family = binomial()) |>
  tidy(exponentiate = TRUE, conf.int = TRUE) |>
  filter(term != "(Intercept)")
```

::: {.callout-note}
## The `_` placeholder
`data = _` passes the piped value to a named argument rather than the first one.
`glm()` takes the formula first and the data second, so the plain pipe would put
the penguins in the wrong slot. The placeholder needs a *named* argument — this
is one of the few places the base pipe is stricter than `%>%`, which uses `.`.
:::

### `glance()`: one row per model

```{r}
#| label: broom-glance
glance(model)
```

Model-level statistics: R², AIC, BIC, residual degrees of freedom, and `nobs` —
the number of observations actually used, which matters more than people expect
(see the warning below).

### `augment()`: one row per observation

```{r}
#| label: broom-augment
augment(model) |>
  select(body_mass_g, .fitted, .resid, .std.resid, .cooksd) |>
  head(4)
```

`augment()` adds fitted values, residuals, leverage and Cook's distance to the
data. Since the result is a data frame, diagnostics become ordinary `ggplot2`:

```{r}
#| label: fig-resid
#| fig-cap: "Residuals against fitted values, drawn with ggplot2."
#| fig-height: 3.2
augment(model) |>
  ggplot(aes(.fitted, .resid)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE) +
  labs(x = "Fitted (g)", y = "Residual (g)") +
  theme_minimal(base_size = 13)
```

### It works on hypothesis tests too

`tidy()` is not only for regression. Most `htest` objects tidy as well, which is
handy when you want a t-test result in a table rather than on screen:

```{r}
#| label: broom-htest
d <- penguins |> drop_na(body_mass_g, sex)

tidy(t.test(body_mass_g ~ sex, data = d))

tidy(cor.test(d$body_mass_g, d$flipper_length_mm))
```

::: {.webex-check .webex-box}
**Check yourself**

Which broom verb returns one row per model term?
`r mcq(c(answer = "tidy()", "glance()", "augment()", "summary()"))`

To get odds ratios from a logistic regression you add
`r mcq(c("family = binomial()", answer = "exponentiate = TRUE", "log = FALSE", "conf.int = TRUE"))`

`augment()` returns one row per
`r mcq(c("model term", "model", answer = "observation", "group"))`
:::

## Two gotchas worth knowing early

::: {.callout-warning}
## Models silently drop rows with missing values
`lm()` and friends default to `na.action = na.omit`: any row with a missing
value in *any* variable in the formula is dropped, without a message.

```{r}
#| label: na-drop
m1 <- lm(body_mass_g ~ flipper_length_mm, data = penguins)
m2 <- lm(body_mass_g ~ flipper_length_mm + sex, data = penguins)

nrow(penguins)
nobs(m1)
nobs(m2)
```

344 rows in, 342 in the first model, 333 in the second — because adding `sex`
brought its 11 missing values into the formula.

This has a consequence people miss: **`m1` and `m2` are fitted to different
data, so their AIC, BIC and R² are not comparable.** To compare models, fit them
all to the same complete-case set:

```{r}
#| label: na-drop-fix
model_data <- penguins |>
  drop_na(body_mass_g, flipper_length_mm, sex)

m1 <- lm(body_mass_g ~ flipper_length_mm, data = model_data)
m2 <- lm(body_mass_g ~ flipper_length_mm + sex, data = model_data)

c(nobs(m1), nobs(m2))
```

Always report `nobs()`, and always say how missing data was handled.
:::

::: {.callout-warning}
## The reference level is the first factor level
A categorical predictor is coded relative to its first level, so every
coefficient is a comparison *against that level*. Change the level order and the
whole table changes meaning:

```{r}
#| label: reference-level
d <- penguins |> drop_na()

# default: Adelie is the reference, because it is the first level
lm(body_mass_g ~ species, data = d) |>
  tidy() |>
  select(term, estimate)

# make Gentoo the reference instead
d |>
  mutate(species = fct_relevel(species, "Gentoo")) |>
  lm(body_mass_g ~ species, data = _) |>
  tidy() |>
  select(term, estimate)
```

Both are the same model — the fit, the predictions and the R² are identical —
but the numbers you would report are completely different. Set the reference
level deliberately with `fct_relevel()` (@sec-strings-factors), choosing the
group that makes clinical sense as the comparator, and say in the text which it
is.
:::

## Checking the model

`performance` runs the standard diagnostics and returns them as one figure:

```{r}
#| label: fig-check-model
#| fig-cap: "Diagnostic plots from performance::check_model()."
#| fig-width: 9
#| fig-height: 8
#| message: false
#| warning: false
performance::check_model(model)
```

```{r}
#| label: performance-numbers
performance::model_performance(model)

performance::check_collinearity(model)
```

::: {.callout-important}
## A diagnostic plot cannot tell you the model was appropriate
`check_model()` will happily draw six panels for a model that should never have
been fitted — one that ignores clustering, or that adjusts for a variable on the
causal pathway. It checks the model's *internal* assumptions, not whether the
model answers your question. That part is yours (@sec-ai).
:::

## Interpreting the model

Coefficients are often not the quantity you want to report, especially on a
non-identity link. `marginaleffects` turns a model into the estimates a reader
actually understands, and it has become the standard tool for this.

```{r}
#| label: marginaleffects
#| message: false
library(marginaleffects)

# average predicted body mass for each species
avg_predictions(model, by = "species")
```

```{r}
#| label: marginaleffects-comparisons
# the average difference between males and females, holding everything else
# at its observed values
avg_comparisons(model, variables = "sex")
```

```{r}
#| label: marginaleffects-slopes
# the average slope: grams per millimetre of flipper
avg_slopes(model, variables = "flipper_length_mm")
```

The three verbs cover most needs: **`avg_predictions()`** for predicted values,
**`avg_comparisons()`** for differences between groups or levels, and
**`avg_slopes()`** for the effect of a one-unit change in a continuous variable.
All return data frames, so they go straight into a plot or a table.

`emmeans` does a related job with a longer history and strong support for
multiplicity adjustment:

```{r}
#| label: emmeans
#| message: false
library(emmeans)

means <- emmeans(model, ~species)
means

pairs(means)
```

Note that `pairs()` applies Tukey adjustment by default — worth knowing you are
getting it, and worth reporting.

## Many models at once

This is where list columns earn their place. The pattern from @sec-functions —
`nest()`, `map()`, `unnest()` — is the tidyverse answer to "fit this model
separately for every group".

### The pattern

```{r}
#| label: many-models
by_species <- penguins |>
  drop_na(body_mass_g, flipper_length_mm) |>
  # nest() collapses each species' rows into a data frame in a list column
  nest(.by = species)

by_species
```

`data` is a **list column**: each cell holds an entire data frame. That is legal
because a tibble column is a list, and a list can hold anything.

```{r}
#| label: many-models-fit
results <- by_species |>
  mutate(
    model = map(data, \(d) lm(body_mass_g ~ flipper_length_mm, data = d)),
    coefficients = map(model, \(m) tidy(m, conf.int = TRUE)),
    fit = map(model, glance)
  )

results
```

Three models, each sitting next to its own data and its own label. Nothing can
get out of order, because the model and the group name are in the same row —
which is the real advantage over keeping a separate list of models and a
separate vector of names.

Then unnest whichever part you want:

```{r}
#| label: many-models-coefs
results |>
  select(species, coefficients) |>
  unnest(coefficients) |>
  filter(term == "flipper_length_mm") |>
  mutate(across(where(is.numeric), \(x) round(x, 2)))
```

```{r}
#| label: many-models-fit-stats
results |>
  select(species, fit) |>
  unnest(fit) |>
  select(species, r.squared, sigma, nobs)
```

::: {.callout-important}
## Three separate models are not one model with an interaction
The slopes above differ between species, which looks like an interaction — and
it is tempting to treat it as one. But the two analyses are not the same:

- **Three separate models** each estimate their own residual variance from their
  own species only.
- **One interaction model** pools the residual variance across all species and
  gives you a formal test of whether the slopes differ.

```{r}
#| label: interaction-model
interaction_model <- penguins |>
  drop_na(body_mass_g, flipper_length_mm) |>
  lm(body_mass_g ~ flipper_length_mm * species, data = _)

tidy(interaction_model) |>
  filter(str_detect(term, ":")) |>
  select(term, estimate, p.value)
```

Which you want depends on your question. Stratified models suit "describe each
group separately"; the interaction model suits "do these groups differ?". Only
the second gives you a p-value for the difference. `map()` makes both equally
easy to produce, which is exactly why you have to decide deliberately.
:::

### Several outcomes at once

The same pattern, rotated: instead of one model per group, one model per
outcome.

```{r}
#| label: many-outcomes
outcomes <- c("body_mass_g", "flipper_length_mm", "bill_length_mm")

outcome_results <- tibble(outcome = outcomes) |>
  mutate(
    model = map(outcome, \(y) {
      # reformulate() builds a formula from strings
      lm(reformulate(c("species", "sex"), response = y), data = penguins)
    }),
    coefficients = map(model, \(m) tidy(m, conf.int = TRUE))
  ) |>
  select(outcome, coefficients) |>
  unnest(coefficients) |>
  filter(term == "sexmale")

outcome_results |>
  mutate(across(where(is.numeric), \(x) round(x, 2)))
```

::: {.callout-warning}
## Many models means many p-values
Fitting a model per outcome, or per subgroup, multiplies your tests. With twenty
outcomes at the 0.05 level you expect one "significant" result from noise alone.

If the analyses were pre-specified, adjust for multiplicity or report them as a
set. If they were not, describe them as exploratory and say so. The `map()`
pattern makes it trivially easy to fit fifty models; the responsibility for what
that does to your error rate is unchanged.
:::

### Unadjusted and adjusted side by side

A staple of epidemiological papers, and a natural fit for list columns:

```{r}
#| label: adjusted-unadjusted
model_data <- penguins |> drop_na()

specifications <- tibble(
  label = c("Unadjusted", "Adjusted for species", "Adjusted for species and sex"),
  formula = c(
    "body_mass_g ~ flipper_length_mm",
    "body_mass_g ~ flipper_length_mm + species",
    "body_mass_g ~ flipper_length_mm + species + sex"
  )
)

specifications |>
  mutate(
    model = map(formula, \(f) lm(as.formula(f), data = model_data)),
    coefficients = map(model, \(m) tidy(m, conf.int = TRUE))
  ) |>
  select(label, coefficients) |>
  unnest(coefficients) |>
  filter(term == "flipper_length_mm") |>
  select(label, estimate, conf.low, conf.high) |>
  mutate(across(where(is.numeric), \(x) round(x, 1)))
```

The estimate falls from 50.2 to 40.6 once species is adjusted for, and to 20.0
once sex is added as well — less than half the unadjusted value. That is
confounding made visible in three lines: most of the apparent relationship
between flipper length and body mass is explained by the fact that larger
species have both. Because every model is fitted to the same `model_data`, the
three estimates are directly comparable.

## Bootstrapping

`rsample` handles resampling, and the results are — again — list columns.

```{r}
#| label: bootstrap
#| message: false
library(rsample)

set.seed(2026)

boot_results <- model_data |>
  bootstraps(times = 200) |>
  mutate(
    coefficients = map(splits, \(s) {
      lm(body_mass_g ~ flipper_length_mm, data = analysis(s)) |> tidy()
    })
  ) |>
  select(id, coefficients) |>
  unnest(coefficients) |>
  filter(term == "flipper_length_mm")

# the percentile bootstrap interval
quantile(boot_results$estimate, c(0.025, 0.5, 0.975)) |> round(2)
```

```{r}
#| label: fig-bootstrap
#| fig-cap: "The bootstrap distribution of the slope."
#| fig-height: 3
boot_results |>
  ggplot(aes(estimate)) +
  geom_histogram(bins = 30, fill = "steelblue", colour = "white") +
  labs(x = "Slope (g per mm)", y = "Bootstrap replicates") +
  theme_minimal(base_size = 13)
```

`analysis(s)` extracts the resampled data from a split; `assessment(s)` gives
the rows left out. Useful when a standard error is hard to derive analytically —
a median, a ratio of coefficients, a complicated derived statistic.

## Reporting the model

`gtsummary` (@sec-tables) builds the table directly from the model object, so
the numbers cannot drift out of sync with the analysis:

```{r}
#| label: tbl-regression
#| message: false
library(gtsummary)

unadjusted <- lm(body_mass_g ~ flipper_length_mm, data = model_data) |>
  tbl_regression()

adjusted <- lm(
  body_mass_g ~ flipper_length_mm + species + sex,
  data = model_data
) |>
  tbl_regression()

tbl_merge(
  list(unadjusted, adjusted),
  tab_spanner = c("**Unadjusted**", "**Adjusted**")
)
```

## Mixed models

Repeated measurements on the same participant violate independence. `lme4`
fits mixed models, and `broom.mixed` tidies them:

```{r}
#| label: mixed-model
#| message: false
library(lme4)

# a random intercept per species: each species gets its own baseline,
# and the slope is shared
mixed <- lmer(body_mass_g ~ flipper_length_mm + (1 | species), data = model_data)

broom.mixed::tidy(mixed, effects = "fixed", conf.int = TRUE)
```

```{r}
#| label: mixed-model-random
broom.mixed::tidy(mixed, effects = "ran_pars")
```

`glmmTMB` handles the cases `lme4` cannot — zero inflation, negative binomial,
beta responses — with a similar formula interface.

::: {.webex-check .webex-box}
**Check yourself**

`m1` uses 342 observations and `m2` uses 333, because `m2` has an extra
predictor with missing values. Can you compare their AIC?
`r mcq(c("Yes, AIC is always comparable", answer = "No, they are fitted to different data", "Only if the R2 is similar", "Only for linear models"))`

A list column holds
`r mcq(c("only numbers", "only strings", answer = "any R object, including data frames and models", "nothing; tibbles cannot do this"))`

Which function gives the average difference between two groups from a fitted
model?
`r mcq(c("avg_predictions()", answer = "avg_comparisons()", "avg_slopes()", "glance()"))`

Fitting one model per group is the same analysis as one model with an
interaction. `r torf(FALSE)`

Changing the reference level of a factor changes the model's predictions.
`r torf(FALSE)`
:::

::: {.callout-note collapse="true"}
## Exercise 1: from model to table

Using `penguins`:

1. Fit a model of `bill_length_mm` on `body_mass_g`, `species` and `sex`.
2. Report the coefficients with confidence intervals, rounded sensibly.
3. How many observations did the model actually use, and why is that not 344?
4. Make a coefficient plot.
5. Report the average difference between males and females.

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: ex1-model-sol
model_ex <- lm(
  bill_length_mm ~ body_mass_g + species + sex,
  data = penguins
)

tidy(model_ex, conf.int = TRUE) |>
  mutate(across(where(is.numeric), \(x) round(x, 3)))

nobs(model_ex)
```

333, not 344: two penguins have no bill measurement and eleven have no recorded
sex, and `lm()` drops any row missing any variable in the formula.

```{r}
#| label: fig-ex1-model
#| fig-cap: "Coefficients with 95% confidence intervals."
#| fig-height: 3
tidy(model_ex, conf.int = TRUE) |>
  filter(term != "(Intercept)") |>
  ggplot(aes(estimate, fct_rev(term))) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high)) +
  labs(x = "Estimate (mm)", y = NULL) +
  theme_minimal(base_size = 13)
```

```{r}
#| label: ex1-model-sol2
avg_comparisons(model_ex, variables = "sex")
```

Note the coefficient for `body_mass_g` is tiny — but the unit is one gram. An
effect per 100 g or per kilogram is far more readable, which is a good argument
for scaling predictors before fitting, or for using `avg_comparisons()` with an
explicit contrast.
:::
:::

::: {.callout-note collapse="true"}
## Exercise 2: many models

1. Fit the model from Exercise 1 separately for each island.
2. Produce a table of the `sexmale` coefficient with its confidence interval,
   one row per island, alongside the number of observations.
3. Plot the estimates with their intervals.
4. Which island has the least precise estimate? Check whether the obvious
   explanation is the right one.

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: ex2-model-sol
island_models <- penguins |>
  drop_na(bill_length_mm, body_mass_g, sex) |>
  nest(.by = island) |>
  mutate(
    model = map(data, \(d) lm(bill_length_mm ~ body_mass_g + sex, data = d)),
    coefficients = map(model, \(m) tidy(m, conf.int = TRUE)),
    n = map_int(data, nrow)
  )

island_estimates <- island_models |>
  select(island, n, coefficients) |>
  unnest(coefficients) |>
  filter(term == "sexmale") |>
  select(island, n, estimate, conf.low, conf.high) |>
  mutate(across(where(is.numeric), \(x) round(x, 2)))

island_estimates
```

```{r}
#| label: fig-ex2-model
#| fig-cap: "Sex difference in bill length by island, with 95% intervals."
#| fig-height: 2.8
island_estimates |>
  ggplot(aes(estimate, island)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high)) +
  labs(x = "Male - female (mm)", y = NULL) +
  theme_minimal(base_size = 13)
```

The obvious guess is Torgersen, which has by far the fewest penguins (47 against
123 and 163). It is wrong: **Dream** has the widest interval.

```{r}
#| label: ex2-model-width
island_estimates |>
  mutate(width = conf.high - conf.low) |>
  arrange(desc(width)) |>
  select(island, n, width)
```

Precision depends on the residual variance and the spread of the predictors, not
only on the sample size. Dream holds two species, Adelie and Chinstrap, whose
bills differ markedly — and this model does not adjust for species, so that
between-species variation lands in the residual and widens the interval.
Torgersen is Adelie only, so it is far more homogeneous and gets more precision
from fewer birds.

Two lessons. First, sample size belongs *in* the table, because three separate
models produce three estimates that look equally solid on a plot. Second, a
stratified analysis inherits whatever the model omits: stratifying by island did
not remove the species confounding, it just hid it inside one stratum.

:::
:::

::: {.callout-note collapse="true"}
## Exercise 3: does it matter which species is the reference?

1. Fit `body_mass_g ~ species` and note the coefficients.
2. Refit with Chinstrap as the reference level.
3. Compare the two tables. What changed, and what did not?
4. Confirm your answer by comparing the fitted values from both models.

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: ex3-model-sol
d3 <- penguins |> drop_na(body_mass_g)

fit_a <- lm(body_mass_g ~ species, data = d3)
fit_b <- d3 |>
  mutate(species = fct_relevel(species, "Chinstrap")) |>
  lm(body_mass_g ~ species, data = _)

tidy(fit_a) |> select(term, estimate)
tidy(fit_b) |> select(term, estimate)
```

The coefficients change completely, because they are comparisons against a
different group. What does **not** change:

```{r}
#| label: ex3-model-check
# identical fit
c(glance(fit_a)$r.squared, glance(fit_b)$r.squared)

# identical predictions, to within floating-point noise
all.equal(unname(fitted(fit_a)), unname(fitted(fit_b)))
```

Same model, same fit, same predictions — only the parameterisation differs. The
practical lesson: choose the reference level so that the comparisons in your
table are the ones you actually want to discuss, and state it in the methods.
Readers cannot tell from the table alone which group is the comparator unless
you say.
:::
:::

## Predictive modelling with tidymodels

Everything above is **inference**: what is the effect, and how certain are we?
`tidymodels` is built for a different question — **prediction**: how well does
this generalise to data it has not seen?

::: {.callout-warning}
## Load tidymodels last, and qualify your data
`tidymodels` attaches `modeldata`, which contains its own `penguins`. After
`library(tidymodels)`, a bare `penguins` silently resolves to
`modeldata::penguins` rather than the one you have been using:

```{r}
#| label: penguins-masking
#| message: false
library(tidymodels)

utils::getAnywhere("penguins")$where[1]

setdiff(names(palmerpenguins::penguins), names(penguins))
```

Same column names, but no `year` — a different data frame, chosen for you
without a word. Since R 4.5 there is a third `penguins` in `datasets`, with
shorter column names again (@sec-classes).

Two defences. Write `palmerpenguins::penguins` explicitly, which is what this
section does. Or call `tidymodels_prefer()`, which activates `conflicted` and
converts silent masking into a loud error:

```{r}
#| label: conflict-demo
#| eval: false
tidymodels_prefer()

penguins |> drop_na()
#> Error: [conflicted] penguins found in 3 packages.
#> Either pick the one you want with `::`:
#> * modeldata::penguins
#> * palmerpenguins::penguins
#> * datasets::penguins
```

An error is the better outcome. This section is placed at the end of the chapter
precisely because attaching `tidymodels` changes what later code means.
:::

```{r}
#| label: tidymodels
#| message: false
penguins_tm <- palmerpenguins::penguins |> drop_na()

set.seed(123)
split <- initial_split(penguins_tm, prop = 0.75, strata = species)

recipe_spec <- recipe(
  body_mass_g ~ flipper_length_mm + species + sex,
  data = training(split)
) |>
  step_dummy(all_nominal_predictors())

workflow_spec <- workflow() |>
  add_recipe(recipe_spec) |>
  add_model(linear_reg() |> set_engine("lm"))

fitted_wf <- fit(workflow_spec, data = training(split))

# performance on data the model has never seen
augment(fitted_wf, new_data = testing(split)) |>
  metrics(truth = body_mass_g, estimate = .pred)
```

Cross-validation, in three lines:

```{r}
#| label: tidymodels-cv
#| message: false
set.seed(42)

fit_resamples(workflow_spec, vfold_cv(training(split), v = 5)) |>
  collect_metrics() |>
  select(.metric, mean, n, std_err)
```

The pieces: **`rsample`** splits the data, **`recipes`** defines the
preprocessing, **`parsnip`** specifies the model independently of which engine
fits it, **`workflows`** binds preprocessing and model together so both are
applied identically to every resample, and **`yardstick`** computes the metrics.

::: {.callout-tip}
## Which framework?
Use **`lm()`/`glm()` + `broom` + `marginaleffects`** when you want to estimate
and report an effect — almost all epidemiological and clinical research.

Use **`tidymodels`** when the goal is prediction and you need resampling, tuning
and honest out-of-sample performance.

The two are not competitors, and using a predictive framework for an inferential
question is a common mistake: a cross-validated R² does not tell you whether an
exposure causes an outcome.
:::

## More information

- [broom](https://broom.tidymodels.org/) and its
  [vignette on many models](https://broom.tidymodels.org/articles/broom.html)
- [R for Data Science: many models with list columns](https://r4ds.hadley.nz/rectangling)
- [marginaleffects](https://marginaleffects.com/) — the accompanying book is the
  best available explanation of predictions, comparisons and slopes
- [emmeans vignettes](https://rvlenth.github.io/emmeans/articles/)
- [easystats](https://easystats.github.io/easystats/) — `performance`,
  `parameters`, `effectsize`, `report`
- [Tidy Modeling with R](https://www.tmwr.org/) — the tidymodels book
- [Regression and Other Stories](https://avehtari.github.io/ROS-Examples/) — for
  the statistics rather than the code
