# Reproducible Workflows {#sec-pipelines}

::: {.callout-note}
## What this chapter covers
@sec-workflow describes the *steps* of an analysis — import, clean, explore,
model, report. This chapter is about the machinery that makes those steps run
reliably: turning a folder of scripts into a pipeline with `targets`, pinning
your package versions with `renv`, and knowing when the extra machinery is
worth it.
:::

```{r}
#| label: setup-pipelines
#| include: false
library(webexercises)
```

## The problem with a folder of scripts

A typical project reaches this state within a few weeks:

```
R/
├── 01_import.R
├── 02_clean.R
├── 03_explore.R
├── 04_model.R
└── 05_report.qmd
```

This is already good practice (@sec-workflow) and it works. Until it does not:

- You change one line in `02_clean.R`. Which of the later scripts now need
  re-running? You cannot be certain, so you either re-run everything — twenty
  minutes — or you re-run some of it and hope.
- You come back after three months. Does `04_model.R` still work, or did it
  depend on an object you created by hand in the console?
- A co-author asks whether the figure in the manuscript came from the current
  data. You do not actually know.
- You re-run everything to be safe, and something breaks in step 2, and you
  cannot tell whether it was broken before.

The underlying problem is that **the dependencies between the steps exist only
in your head**. The numbers in the filenames are a hint, not a mechanism.

## `targets`

[`targets`](https://docs.ropensci.org/targets/) makes those dependencies
explicit and then does the bookkeeping for you. You declare the steps; it works
out what depends on what, runs only what is out of date, and stores the results.

Three consequences, and the middle one is the one that changes how you work:

- **Skip what has not changed.** Re-running a pipeline where nothing changed
  costs seconds.
- **Trust the result.** If the pipeline is up to date, the outputs *were*
  produced by the current code and the current data. Not "probably".
- **Never lose the objects.** Every result is cached and readable at any time
  with `tar_read()` — no more re-running an analysis to look at a number.

### The shape of a project

```
my_project/
├── _targets.R          <- the pipeline definition
├── _targets/           <- the cache (gitignored)
├── R/
│   └── functions.R     <- your functions
├── data-raw/
└── report.qmd
```

Two things differ from the script-based layout. There is a **`_targets.R`** file
describing the pipeline, and `R/` holds **functions** rather than top-to-bottom
scripts. That second point is the real prerequisite: `targets` orchestrates
function calls, so the work of @sec-functions is what makes this possible.

### Functions first

```{r}
#| eval: false
# R/functions.R

clean_penguins <- function(data) {
  data |>
    drop_na(body_mass_g, flipper_length_mm, sex) |>
    mutate(body_mass_kg = body_mass_g / 1000)
}

fit_model <- function(data) {
  lm(body_mass_g ~ flipper_length_mm + species + sex, data = data)
}

model_table <- function(model) {
  broom::tidy(model, conf.int = TRUE)
}
```

Each function takes data in and returns something. No file reading, no
`<-` at the top level, no side effects.

### The pipeline

```{r}
#| eval: false
# _targets.R

library(targets)
library(tarchetypes)

# packages every target needs, declared once
tar_option_set(
  packages = c("dplyr", "tidyr", "ggplot2", "broom", "palmerpenguins")
)

# source everything in R/
tar_source()

list(
  tar_target(raw_data, palmerpenguins::penguins),
  tar_target(clean_data, clean_penguins(raw_data)),
  tar_target(model, fit_model(clean_data)),
  tar_target(coefficients, model_table(model))
)
```

`tar_target(name, command)` says "there is a thing called *name*, and this is
how to make it". You never say what depends on what — `targets` reads the code
and works it out: `clean_data` mentions `raw_data`, so it depends on it.

`targets::use_targets()` writes a starter `_targets.R` for you to fill in.

### Running it

Here is that pipeline actually running. `tar_dir()` builds it in a temporary
directory, which is how the `targets` documentation runs self-contained
examples:

```{r}
#| label: targets-run
#| message: false
library(targets)

tar_dir({
  tar_script({
    library(targets)
    tar_option_set(
      packages = c("dplyr", "tidyr", "broom", "palmerpenguins")
    )
    list(
      tar_target(raw_data, palmerpenguins::penguins),
      tar_target(clean_data, tidyr::drop_na(raw_data, body_mass_g, flipper_length_mm)),
      tar_target(model, lm(body_mass_g ~ flipper_length_mm, data = clean_data)),
      tar_target(coefficients, broom::tidy(model))
    )
  })

  tar_make()
})
```

Every target was built. Now run it again with nothing changed:

```{r}
#| label: targets-skip
#| message: false
tar_dir({
  tar_script({
    library(targets)
    tar_option_set(
      packages = c("dplyr", "tidyr", "broom", "palmerpenguins")
    )
    list(
      tar_target(raw_data, palmerpenguins::penguins),
      tar_target(clean_data, tidyr::drop_na(raw_data, body_mass_g, flipper_length_mm)),
      tar_target(model, lm(body_mass_g ~ flipper_length_mm, data = clean_data)),
      tar_target(coefficients, broom::tidy(model))
    )
  })

  tar_make()
  cat("\n--- running a second time, nothing changed ---\n")
  tar_make()

  cat("\n--- the cached result, read back without re-running ---\n")
  print(tar_read(coefficients))
})
```

**"skipped pipeline"** is the whole point. Nothing was recomputed, because
nothing changed.

::: {.callout-important}
## What "changed" means
`targets` invalidates a target when its **command**, its **dependencies**, or
the **functions it calls** change. Edit `clean_penguins()` and `clean_data`,
`model` and `coefficients` all rebuild — but `raw_data` does not, because it
does not use that function.

That is stronger than a timestamp check. It hashes the code, so reformatting
that does not change meaning will not trigger a rebuild, and a changed function
buried three files away will.

What it does **not** track is anything invisible to R: a database that changed
underneath you, a file you edited outside the pipeline, or randomness without a
seed. Use `tar_target(format = "file")` to make file dependencies explicit.
:::

### The commands you will actually use

| Command | What it does |
|---------|--------------|
| `tar_make()` | run the pipeline, skipping what is up to date |
| `tar_read(name)` | read one result into the console |
| `tar_load(name)` | load one or more results into your environment |
| `tar_outdated()` | which targets *would* rebuild, without running them |
| `tar_visnetwork()` | interactive diagram of the pipeline and its status |
| `tar_manifest()` | the pipeline as a table |
| `tar_prune()` | delete cached results no longer in the pipeline |
| `tar_destroy()` | delete the whole cache and start clean |

`tar_visnetwork()` is worth running early — it draws the dependency graph with
each target coloured by whether it is up to date, which is often the fastest way
to see that two things you thought were connected are not.

::: {.callout-tip}
## Working interactively with a pipeline
The common worry is "I still want to poke at my data in the console". You do:

```{r}
#| eval: false
tar_load(clean_data) # now clean_data is in your environment
tar_load_everything() # or load the lot
```

Develop a function interactively, get it right, put it in `R/`, add a target.
The pipeline is where finished work goes, not where you do exploratory work.
:::

### Branching: one target, many results

Chapter @sec-modelling fitted a model per species with `map()`. In a pipeline
you can do the same with **dynamic branching**, and each branch is cached
separately — so if one species' data changes, only that branch reruns:

```{r}
#| eval: false
list(
  tar_target(species_list, c("Adelie", "Chinstrap", "Gentoo")),
  tar_target(clean_data, clean_penguins(raw_data)),
  tar_target(
    fits,
    fit_one(clean_data, species_list),
    pattern = map(species_list) # one branch per species
  ),
  tar_target(combined, bind_rows(fits))
)
```

`tarchetypes::tar_map()` does the *static* equivalent, creating separately named
targets at build time — better when you want to see each one in the graph.

### Rendering the report as a target

```{r}
#| eval: false
library(tarchetypes)

list(
  # ... upstream targets ...
  tar_quarto(report, "report.qmd")
)
```

Inside `report.qmd` you read results with `tar_read()` rather than recomputing
them:

````markdown
```{{r}}
library(targets)
coefficients <- tar_read(coefficients)
```
````

Now the report is part of the dependency graph. Change the cleaning function and
`tar_make()` rebuilds the model *and* re-renders the report — so the manuscript
cannot silently disagree with the analysis. That is the single biggest practical
benefit for research work.

### When targets is worth it

**Use it when** a full re-run takes more than a few minutes, the project has
more than a handful of steps, several outputs depend on the same cleaning, you
will come back to it after months, or a report must stay in sync with the
analysis.

**Skip it when** the analysis is one script that runs in seconds, or it is a
one-off exploration. The overhead — restructuring into functions, learning the
commands — is real, and a small project does not repay it.

::: {.callout-warning}
## The cache holds your data
`_targets/` stores every result, which means it stores your data. `targets`
handles this for you: the first `tar_make()` writes a `_targets/.gitignore` that
excludes the cached objects, keeping everything except a small metadata file out
of Git.

Do not take that on trust for sensitive data. Run `git status` before your first
commit and confirm that nothing from `_targets/objects/` is staged. The
generated file itself carries a warning that `_targets/meta/process` can contain
sensitive information — for registry data, the safest choice is to commit
nothing from `_targets/` at all (@sec-cleaning).
:::

## `renv`: pinning package versions

A pipeline guarantees your code runs in the right order. It does not guarantee
the packages behave the same way next year. `dplyr` 1.2.0 deprecated
`case_match()`; `ggplot2` 4.0 changed its internals. Code written against one
version may not run, or may run differently, against another.

[`renv`](https://rstudio.github.io/renv/) gives each project its own library and
records exactly what is in it.

```{r}
#| eval: false
renv::init() # project library + renv.lock, once per project

# ... work normally, install packages as usual ...

renv::status() # what has drifted from the lockfile?
renv::snapshot() # record the current versions in renv.lock
```

`renv.lock` is a small text file listing every package and its version. **Commit
it.** Anyone who clones the project — including you on a new machine — runs:

```{r}
#| eval: false
renv::restore() # install exactly the recorded versions
```

::: {.callout-tip}
## Snapshot at the points that matter
`renv::snapshot()` when you finish a piece of work, before submitting a
manuscript, and before a long break. The lockfile from the day you generated the
results is what makes them reproducible later.

Two habits that avoid the common frustrations: run `renv::status()` before
snapshotting so you know what you are recording, and do not snapshot in the
middle of debugging a package installation.
:::

::: {.callout-note}
## renv is not a container
`renv` pins **R packages**. It does not pin R itself, your operating system, or
system libraries such as GDAL or a Java runtime. For most analyses that is
enough. When it is not — long-lived projects, regulated work, anything that must
run identically in ten years — the next step is a container (Docker, with the
[Rocker](https://rocker-project.org/) images), which pins the whole environment.
That is beyond this book, but `renv` and Docker compose well: the container
provides R, and `renv.lock` provides the packages.
:::

## Guarding against silent breakage

A pipeline reruns reliably. That is not the same as producing *correct* results
— it will happily rebuild a wrong answer. Two cheap habits catch most of it.

**Assert what must be true**, so the pipeline fails loudly rather than
propagating nonsense:

```{r}
#| eval: false
clean_penguins <- function(data) {
  out <- data |>
    drop_na(body_mass_g, flipper_length_mm, sex)

  # fail here rather than three targets later
  stopifnot(
    nrow(out) > 300,
    !any(is.na(out$body_mass_g)),
    all(out$body_mass_g > 0)
  )

  out
}
```

**Test the functions**, since they are ordinary functions now:

```{r}
#| eval: false
# tests/testthat/test-clean.R
test_that("cleaning drops incomplete rows and keeps the rest", {
  result <- clean_penguins(palmerpenguins::penguins)

  expect_false(any(is.na(result$body_mass_g)))
  expect_gt(nrow(result), 300)
  expect_s3_class(result, "tbl_df")
})
```

Restructuring into functions has a payoff beyond `targets`: functions are
testable, and a script is not.

::: {.callout-tip}
## The reproducibility ladder
Each rung costs more effort and buys more certainty. Most research work belongs
on rung 3 or 4.

1. **A script that runs top to bottom in a clean session** (@sec-workflow). If
   you do nothing else, do this.
2. **Project-relative paths** with `here()` (@sec-projects), and data that never
   enters Git.
3. **Recorded package versions** with `renv`.
4. **A pipeline** with `targets`, so the outputs provably match the code.
5. **A container** pinning R and the operating system too.

Climb only as far as the work warrants. A rung skipped lower down is worth more
than a rung added at the top — `renv` on a project whose script does not run in
a clean session solves nothing.
:::

::: {.webex-check .webex-box}
**Check yourself**

What does `tar_make()` do when nothing has changed?
`r mcq(c("re-runs everything", answer = "skips every target", "errors", "rebuilds only the last target"))`

Which file records your package versions?
`r mcq(c("_targets.R", answer = "renv.lock", "DESCRIPTION", ".Rprofile"))`

You edit one function. `targets` rebuilds
`r longmcq(c(
  "the whole pipeline, to be safe",
  "only the target that calls it",
  answer = "that target and everything downstream of it, but nothing upstream",
  "nothing until you clear the cache"
))`

`renv` guarantees your analysis will run identically in ten years.
`r torf(FALSE)`

To use a pipeline result in the console you would write
`r mcq(c("readRDS()", answer = "tar_read()", "tar_make()", "load()"))`
:::

::: {.callout-note collapse="true"}
## Exercise 1: convert a script into a pipeline

Take a short analysis of your own — or write one: import `penguins`, drop
missing values, fit a model, produce a coefficient table.

1. Rewrite each step as a function that takes data and returns a result.
2. Create `_targets.R` with `targets::use_targets()` and fill in the target
   list.
3. Run `tar_make()`. Then run it again — how long does the second run take?
4. Run `tar_visnetwork()` and look at the graph.
5. Change one function, run `tar_outdated()` *before* rebuilding. Are the
   targets it lists the ones you expected?

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| eval: false
# R/functions.R
clean_penguins <- function(data) {
  data |> tidyr::drop_na(body_mass_g, flipper_length_mm, sex)
}

fit_model <- function(data) {
  lm(body_mass_g ~ flipper_length_mm + species + sex, data = data)
}

model_table <- function(model) {
  broom::tidy(model, conf.int = TRUE)
}
```

```{r}
#| eval: false
# _targets.R
library(targets)
tar_option_set(packages = c("dplyr", "tidyr", "broom", "palmerpenguins"))
tar_source()

list(
  tar_target(raw_data, palmerpenguins::penguins),
  tar_target(clean_data, clean_penguins(raw_data)),
  tar_target(model, fit_model(clean_data)),
  tar_target(coefficients, model_table(model))
)
```

Step 5 is the one that teaches the idea. Editing `clean_penguins()` should mark
`clean_data`, `model` and `coefficients` as outdated, but **not** `raw_data` —
it does not call that function. If `tar_outdated()` lists something you did not
expect, that is a real dependency you had not noticed, and worth understanding
before you trust the pipeline.

A common surprise: a target that reads a file rebuilds every time, or never
rebuilds when the file changes. Declare it properly with
`tar_target(my_file, "data-raw/thing.csv", format = "file")`, which tracks the
file's contents rather than its name.
:::
:::

::: {.callout-note collapse="true"}
## Exercise 2: decide whether you need this

For a project you actually have, answer honestly:

1. How long does a full re-run take?
2. If you changed one cleaning rule, could you say with certainty which outputs
   would change?
3. Could someone else reproduce your last set of results from what is in the
   repository today?
4. Which rung of the reproducibility ladder are you on?
5. What is the *cheapest* next rung, and is it worth it this week?

::: {.callout-tip collapse="true"}
## What people usually find

Question 3 is where most projects fail, and rarely for exotic reasons. The
recurring causes:

- An object created in the console months ago that the script assumes exists.
- A file path pointing at a folder only you have.
- A package that has since changed behaviour.
- A manual step — "then I fixed the two typos in Excel" — that exists in nobody's
  code.

Note that `targets` fixes only the first of those, `here()` the second, `renv`
the third, and **nothing** fixes the fourth except not doing it.

That ordering is the honest answer to question 5. If your script does not
currently run in a clean session, adding a pipeline on top will not help; fix
that first. If it does run cleanly and takes twenty minutes, `targets` will
repay itself in a fortnight.
:::
:::

## More information

- [The targets manual](https://books.ropensci.org/targets/) — thorough and
  well written, with a good chapter on debugging pipelines
- [targets reference](https://docs.ropensci.org/targets/reference/) and
  [tarchetypes](https://docs.ropensci.org/tarchetypes/) for the ready-made
  target factories
- [renv documentation](https://rstudio.github.io/renv/)
- [The Rocker project](https://rocker-project.org/) — Docker images for R
- [rOpenSci: reproducibility in R](https://ropensci.org/) — packages and
  practices, peer reviewed
