# Useful Packages {#sec-packages}

::: {.callout-note}
## What this chapter covers
A curated map of the R package ecosystem, organised by the stage of the workflow
where you would reach for each one, plus how to judge whether an unfamiliar
package is worth adopting. Tidyverse packages are mostly left out, since the
rest of the book covers them.
:::

```{r}
#| label: setup-packages
#| include: false
library(webexercises)
```

There are nearly 25,000 packages on CRAN — run `nrow(available.packages())` to
see the number today. That is the great strength of R and also a real problem:
for any task there are five plausible packages, three of them abandoned. This
chapter is a starting point, not a complete list.

Packages marked **[used in this book]** appear in the code you have already run,
so you can see them working in context.

::: {.callout-tip}
## Installing several at once
```{r}
#| eval: false
install.packages(c("janitor", "skimr", "naniar", "gtsummary"))
```

You only install a package once per R version, but you load it with `library()`
in every session. If a package is used once in a script, `package::function()`
is often clearer than attaching the whole thing (@sec-import).
:::

## Getting data in and out

- **`haven`** — import and export SPSS, Stata and SAS files, preserving value
  labels. [Site](https://haven.tidyverse.org/) **[used in this book]**
- **`readxl`** — read Excel files without any external dependency.
  [Site](https://readxl.tidyverse.org/) **[used in this book]**
- **`writexl`** — write Excel files, quickly and without Java.
  [Site](https://docs.ropensci.org/writexl/)
- **`openxlsx`** — read *and* write Excel, with control over formatting,
  formulas and multiple sheets. Heavier than the two above, and worth it when
  you must produce a formatted workbook.
  [Site](https://ycphs.github.io/openxlsx/)
- **`here`** — build file paths from the project root so they work on any
  machine. [Site](https://here.r-lib.org/) **[used in this book]**
- **`fs`** — a consistent, cross-platform replacement for base R's file
  functions (`file.path()`, `list.files()`, `dir.create()`).
  [Site](https://fs.r-lib.org/)
- **`arrow`** — read and write Parquet and Feather. Parquet keeps column types,
  compresses well, and is read by both R and Python — a good format for a
  cleaned dataset that is too large for CSV.
  [Site](https://arrow.apache.org/docs/r/)

## Cleaning and validating

- **`janitor`** — `clean_names()`, `get_dupes()`, `remove_empty()`,
  `tabyl()`. The first package to install after the tidyverse.
  [Site](https://sfirke.github.io/janitor/) **[used in this book]**
- **`naniar`** — explore and visualise missing data: `gg_miss_var()`,
  `gg_miss_upset()`, `mcar_test()`.
  [Site](https://naniar.njtierney.com/) **[used in this book]**
- **`visdat`** — `vis_dat()` draws a picture of your whole data frame: types and
  missingness at a glance. [Site](https://docs.ropensci.org/visdat/)
- **`labelled`** — work with variable and value labels, which is how SPSS and
  Stata files arrive via `haven`.
  [Site](https://larmarange.github.io/labelled/)
- **`mice`** — multiple imputation for missing data. Powerful, and requires you
  to understand the missingness mechanism first (@sec-eda).
  [Site](https://amices.org/mice/)
- **`validate`** and **`pointblank`** — write down the rules your data must
  satisfy (age between 18 and 100, no duplicate IDs) and check them
  automatically. [validate](https://github.com/data-cleaning/validate) ·
  [pointblank](https://rstudio.github.io/pointblank/)
- **`dtrackr`** — tracks what your `dplyr` pipeline does to the data and draws
  the participant flow diagram from it. Solves a genuinely tedious problem:
  keeping the numbers in your flow chart in sync with your code.
  [Site](https://terminological.github.io/dtrackr/)
- **`flowchart`** — draws participant flow diagrams from a data frame, in
  tidyverse style. [Site](https://bruigtp.github.io/flowchart/)
- **`codebookr`** — generates a Word codebook documenting every variable in a
  dataset. [Site](https://brad-cannell.github.io/codebookr/)

::: {.callout-tip}
## Documenting your data pays for itself
`dtrackr` and `codebookr` solve the same problem from two directions: the flow
diagram and the codebook are the two documents a reviewer asks for, and both are
tedious to maintain by hand. Generating them from the code means they cannot
drift out of date. See also the data dictionary discussion in @sec-ai.
:::

## Exploring and summarising

- **`skimr`** — `skim()` gives types, missingness, quartiles and a sparkline
  histogram for every column at once.
  [Site](https://docs.ropensci.org/skimr/) **[used in this book]**
- **`summarytools`** — `dfSummary()` produces a fuller HTML overview.
  [Site](https://github.com/dcomtois/summarytools)
- **`DataExplorer`** — `create_report()` writes an entire exploratory report as
  an HTML file. Useful as a first pass; no substitute for looking yourself.
  [Site](https://boxuancui.github.io/DataExplorer/)
- **`GGally`** — `ggpairs()` draws every pairwise relationship in one figure.
  [Site](https://ggobi.github.io/ggally/) **[used in this book]**

## Tables

- **`gt`** — the most control over appearance, and the engine behind
  `gtsummary`. HTML, LaTeX, RTF and Word.
  [Site](https://gt.rstudio.com/) **[used in this book]**
- **`gtsummary`** — publication-ready summary and regression tables with almost
  no code. The fastest route to a Table 1 (@sec-tables).
  [Site](https://www.danieldsjoberg.com/gtsummary/) **[used in this book]**
- **`flextable`** — the best choice when the destination is Word or PowerPoint.
  [Site](https://davidgohel.github.io/flextable/)
- **`tinytable`** — small, few dependencies, outputs HTML, LaTeX, Word, PDF,
  Markdown and Typst. [Site](https://vincentarelbundock.github.io/tinytable/)
- **`DT`** — interactive HTML tables the reader can sort, filter and search.
  HTML only. [Site](https://rstudio.github.io/DT/)
- **`reactable`** — interactive tables with more layout control than `DT`,
  including expandable rows and nested tables.
  [Site](https://glin.github.io/reactable/)
- **`kableExtra`** — extends `knitr::kable()` with styling, grouped headers and
  footnotes. [Site](https://haozhu233.github.io/kableExtra/)

## Visualisation

Extensions to `ggplot2` (@sec-visualisation):

- **`patchwork`** — combine several plots into one figure with `+`, `|` and `/`.
  [Site](https://patchwork.data-imaginist.com/) **[used in this book]**
- **`scales`** — format axis labels as percentages, currency or dates, and
  control breaks and transformations. [Site](https://scales.r-lib.org/)
- **`ggrepel`** — text labels that push each other apart instead of overlapping.
  Indispensable for labelled scatterplots.
  [Site](https://ggrepel.slowkow.com/)
- **`ggtext`** — Markdown and HTML in plot text, so you can put **bold** or
  coloured words in a title and drop the legend.
  [Site](https://wilkelab.org/ggtext/)
- **`ggforce`** — extra geoms and a set of tools for annotating: circle or
  outline groups of points, zoom into a facet, draw arcs and links.
  [Site](https://ggforce.data-imaginist.com/)
- **`ggdist`** — visualise distributions and uncertainty properly: rain-cloud
  plots, gradient intervals, dot plots. Well suited to showing an effect with
  its uncertainty rather than a bar with an error bar.
  [Site](https://mjskay.github.io/ggdist/)
- **`ggridges`** — ridgeline plots, for comparing many distributions at once.
  [Site](https://wilkelab.org/ggridges/)
- **`ggbeeswarm`** — points offset so none overlap, more readable than jitter
  for modest sample sizes. [Site](https://github.com/eclarke/ggbeeswarm)
- **`ggsurvfit`** — publication-ready Kaplan-Meier curves with risk tables, from
  the `gtsummary` author. [Site](https://www.danieldsjoberg.com/ggsurvfit/)
- **`plotly`** — turn a `ggplot2` figure interactive with `ggplotly()`.
  [Site](https://plotly.com/r/)
- **`gganimate`** — animate a plot over a variable, usually time.
  [Site](https://gganimate.com/)

### Colour

- **`viridis`** — perceptually uniform palettes that survive greyscale printing
  and colour vision deficiency. A safe default.
  [Site](https://sjmgarnier.github.io/viridis/)
- **`colorspace`** — build and inspect palettes, and simulate how yours looks to
  a colour-blind reader. [Site](https://colorspace.r-forge.r-project.org/)
- **`paletteer`** — one interface to several thousand palettes from dozens of
  packages. [Site](https://emilhvitfeldt.github.io/paletteer/)
- **`scico`** and **`MetBrewer`** — scientific and museum-derived palettes.
  [scico](https://github.com/thomasp85/scico) ·
  [MetBrewer](https://github.com/BlakeRMills/MetBrewer)
- [R Graph Gallery palette finder](https://r-graph-gallery.com/color-palette-finder)
  — browse palettes visually.

## Modelling and model output

- **`broom`** — turn a model object into a tidy data frame with `tidy()`,
  `glance()` and `augment()`. The bridge between modelling and everything else.
  [Site](https://broom.tidymodels.org/) **[used in this book]**
- **`marginaleffects`** — predictions, comparisons, contrasts and marginal
  effects from almost any model, on the scale you actually want to report.
  Increasingly the standard way to interpret a model.
  [Site](https://marginaleffects.com/)
- **`emmeans`** — estimated marginal means and pairwise contrasts, with
  multiplicity adjustment. [Site](https://rvlenth.github.io/emmeans/)
- **`modelsummary`** — regression and summary tables for many models side by
  side. [Site](https://modelsummary.com/)
- **`easystats`** — a family of packages with a consistent interface:
  `parameters` (coefficients), `performance` (model checks and R²),
  `effectsize` (standardised effects), `see` (diagnostic plots), `report`
  (a written summary of a model). [Site](https://easystats.github.io/easystats/)
- **`sjPlot`** — plots and tables for regression models, popular in the social
  and health sciences. [Site](https://strengejacke.github.io/sjPlot/)
- **`tidymodels`** — a tidyverse-consistent framework for predictive modelling:
  resampling, tuning, workflows. For prediction rather than inference.
  [Site](https://www.tidymodels.org/)

::: {.callout-important}
## A package does not choose the analysis for you
`performance::check_model()` will draw six diagnostic plots for any model you
give it, including one that should never have been fitted. These packages make
it much faster to do the right analysis and no harder to do the wrong one. The
statistical decision — which model, which comparison, which adjustment — is
still yours, and it is the part a reviewer will ask about.
:::

## Health and sport science

Packages that come up repeatedly in this field:

- **`survival`** — the reference implementation for time-to-event analysis,
  Cox models included. [CRAN](https://cran.r-project.org/package=survival)
- **`ggsurvfit`** / **`survminer`** — survival curves worth publishing.
  [ggsurvfit](https://www.danieldsjoberg.com/ggsurvfit/) ·
  [survminer](https://rpkgs.datanovia.com/survminer/)
- **`lme4`**, **`nlme`**, **`glmmTMB`** — mixed-effects models, for repeated
  measurements and clustered data. `glmmTMB` handles zero inflation and a wider
  range of distributions. [glmmTMB](https://glmmTMB.github.io/glmmTMB/)
- **`mgcv`** — generalised additive models, for relationships that are clearly
  not linear. [CRAN](https://cran.r-project.org/package=mgcv)
- **`epiR`** and **`epitools`** — epidemiological measures: risk ratios, odds
  ratios, standardised rates, sample size.
  [epiR](https://cran.r-project.org/package=epiR)
- **`tableone`** — the classic Table 1 package, still widely used in
  epidemiology. `gtsummary` is the more flexible modern option.
  [CRAN](https://cran.r-project.org/package=tableone)
- **`GGIR`** — processes raw accelerometer files (ActiGraph, GENEActiv, Axivity)
  into physical activity and sleep outcomes. The standard tool in the field.
  [Site](https://wadpac.github.io/GGIR/)
- **`slider`** and **`zoo`** — rolling windows and moving averages, which is
  what most wearable-data processing comes down to once the files are read.
  [slider](https://slider.r-lib.org/)
- **`lubridate`**, **`hms`** and **`clock`** — dates, times of day, and
  timezone-safe arithmetic. Essential for anything with a timestamp.
  [lubridate](https://lubridate.tidyverse.org/) **[used in this book]**

## Larger data

When your data stops fitting comfortably in memory, you do not have to abandon
`dplyr` syntax:

- **`data.table`** — very fast, very memory-efficient, with its own concise
  syntax. The long-standing alternative to `dplyr`.
  [Site](https://rdatatable.gitlab.io/data.table/)
- **`dtplyr`** — write `dplyr` code, execute it as `data.table`.
  [Site](https://dtplyr.tidyverse.org/)
- **`duckplyr`** — a drop-in `dplyr` backend powered by DuckDB, which spills to
  disk when the data does not fit in memory.
  [Site](https://duckplyr.tidyverse.org/)
- **`arrow`** — query Parquet files larger than memory with `dplyr` verbs.
  [Site](https://arrow.apache.org/docs/r/)
- **`collapse`** — very fast grouped statistics.
  [Site](https://sebkrantz.github.io/collapse/)

::: {.callout-tip}
## Try the boring fix first
Before reaching for any of these, check whether you actually need every column
and every row. `select()` on import and filtering early often solves the problem
outright, and keeps your code readable.
:::

## Reproducibility and project hygiene

- **`renv`** — records the exact package versions your project uses in a
  `renv.lock` file, so it can be restored later or on another machine. The
  single most effective step for long-term reproducibility (@sec-pipelines).
  [Site](https://rstudio.github.io/renv/)
- **`targets`** — a pipeline tool: define the steps, and it re-runs only what
  the change affects. Worth it once a full re-run takes more than a few minutes.
  See @sec-pipelines. [Site](https://docs.ropensci.org/targets/)
- **`tarchetypes`** — ready-made target factories for `targets`, including
  `tar_quarto()` for rendering a report as part of the pipeline.
  [Site](https://docs.ropensci.org/tarchetypes/)
- **`usethis`** — automates project setup: `use_git()`, `use_readme_md()`,
  `use_git_ignore()`, `create_project()` (@sec-git).
  [Site](https://usethis.r-lib.org/)
- **`testthat`** — write checks that fail loudly when an assumption breaks. Not
  only for package developers: a handful of `expect_equal()` calls on a cleaning
  script catches a lot (@sec-ai). [Site](https://testthat.r-lib.org/)
- **`quarto`** — render Quarto documents from R.
  [Site](https://quarto-dev.github.io/quarto-r/) **[used in this book]**
- **`glue`** — string interpolation: `glue("n = {nrow(data)}")`.
  [Site](https://glue.tidyverse.org/) **[used in this book]**

## AI from R

Covered in @sec-ai:

- **`ellmer`** — call language models from R, with structured output and tool
  calling. [Site](https://ellmer.tidyverse.org/)
- **`mall`** — run LLM predictions row-wise over a data frame.
  [Site](https://mlverse.github.io/mall/)
- **`btw`** — describe your R environment to an assistant so it can answer
  questions about your actual objects.
  [Site](https://posit-dev.github.io/btw/)
- **`chattr`** — a chat interface inside the RStudio IDE.
  [Site](https://mlverse.github.io/chattr/)

## How to judge a package

You will constantly meet packages that are not on this list. A quick assessment:

**Is it maintained?** Look at the CRAN page: when was it last updated? A package
untouched for five years may be finished, or abandoned — check the GitHub issues
to see which.

**Is it documented?** A pkgdown website with vignettes and a reference index is
a strong signal. A CRAN page with no examples is a weak one.

**Who uses it?** Reverse dependencies on CRAN, and whether papers in your field
cite it.

**How many dependencies does it add?** A package pulling in forty others is
forty more things that can break your project in two years. Check with:

```{r}
#| eval: false
tools::package_dependencies("gtsummary", recursive = TRUE)
```

**Does it fit how you already work?** A package that takes and returns data
frames drops into your pipeline. One that requires its own object type means
converting in and out at every step.

::: {.callout-warning}
## Do not add a dependency for one function
If you need one small thing, consider writing it yourself (@sec-functions) or
calling it once with `package::function()`. Every attached package is a
potential name conflict — `MASS::select()` masking `dplyr::select()` is a
classic afternoon lost — and a potential future breakage.

This cuts both ways. Do not reimplement `janitor::clean_names()` either. The
question is whether the package earns its place, not whether dependencies are
bad.
:::

::: {.webex-check .webex-box}
**Check yourself**

Which package would you use to make a Table 1 for a paper?
`r mcq(c("janitor", "skimr", answer = "gtsummary", "broom"))`

Which package records the exact package versions a project uses?
`r mcq(c("targets", answer = "renv", "usethis", "here"))`

You need a table in a Word document. The best starting point is
`r mcq(c("DT", "reactable", answer = "flextable", "kableExtra"))`

Which turns a model object into a tidy data frame?
`r mcq(c("modelsummary", answer = "broom", "performance", "emmeans"))`

Attaching many packages is harmless because R resolves conflicts automatically.
`r torf(FALSE)`
:::

::: {.callout-note collapse="true"}
## Exercise: assess a package you have not used

Pick a package from this chapter that is new to you and relevant to your work.

1. Open its website. Is there a "Get started" vignette?
2. When was it last updated on CRAN?
3. How many recursive dependencies does it have?
4. Find one example in its documentation and run it.
5. Would you use it? Write one sentence saying why or why not.

::: {.callout-tip collapse="true"}
## Solution

There is no single right answer, but here is how to get the facts:

```{r}
#| eval: false
# 1 and 2: the CRAN page and the pkgdown site
utils::browseURL("https://cran.r-project.org/package=naniar")

# how many packages does it drag in?
length(tools::package_dependencies("naniar", recursive = TRUE)[[1]])

# what vignettes ship with it, once installed?
vignette(package = "naniar")

# and the function index
help(package = "naniar")
```

Step 4 is the one people skip and the one that matters. Documentation examples
are run by CRAN on every submission, so they are guaranteed to work — which
makes them the fastest honest test of whether the package does what you thought
it did.

A useful habit: when you adopt a package, note in your project README *why* you
chose it. In two years, when it is abandoned and you have to replace it, that
sentence tells you what you actually needed.
:::
:::

## Where to learn more

**Books**

- [R for Data Science (2nd ed.)](https://r4ds.hadley.nz/) — the standard
  reference for everything in this book
- [R for Non-Programmers](https://bookdown.org/daniel_dauber_io/r4np_book/) — a
  gentler introduction
- [Fundamentals of Data Visualization](https://clauswilke.com/dataviz/) — the
  concepts behind good figures
- [Advanced R](https://adv-r.hadley.nz/) — how the language actually works
- [R Packages](https://r-pkgs.org/) — when your functions outgrow a script and
  should become a package
- [Tidy Modeling with R](https://www.tmwr.org/) — the tidymodels book
  (@sec-modelling)
- [Happy Git and GitHub for the useR](https://happygitwithr.com/) — the
  reference for @sec-git, including what to do when Git goes wrong
- [Reporting with R](https://utdata.github.io/rwdir/) — data journalism, with a
  strong practical workflow
- [AI TutoR](https://psyteachr.github.io/AITutoR/) — coding with generative AI
- [The Big Book of R](https://www.bigbookofr.com/) — a catalogue of several
  hundred free R books, sorted by topic. Start here when you need something this
  list does not cover.

**Blogs worth following**

- [Hadley Wickham's Substack, *Tidy Design*](https://tidydesign.substack.com/) —
  the thinking behind tidyverse design decisions, and lately a lot on what AI
  changes about writing code. His personal site is [hadley.nz](https://hadley.nz/).
- [The tidyverse blog](https://www.tidyverse.org/blog/) — where new features and
  breaking changes are announced. Worth skimming a few times a year, and the
  fastest way to notice that a function you use has been superseded.
- [The Posit blog](https://posit.co/blog/) — releases and announcements for
  RStudio, Positron and Quarto.
- [R Weekly](https://rweekly.org/) — a weekly roundup of R news, blog posts and
  new packages. One page a week is a good way to stay current without effort.
- [Julia Silge's blog](https://juliasilge.com/) — screencasts working through a
  real analysis end to end, mostly with tidymodels.
- [Simon Willison's blog](https://simonwillison.net/) — not R, but the clearest
  writing on how LLM tools actually work (@sec-ai).
- [R-bloggers](https://www.r-bloggers.com/) — an aggregator of several hundred R
  blogs. High volume, variable quality, good for stumbling on things.

**Reference**

- [Posit cheatsheets](https://posit.co/resources/cheatsheets/) — one-page
  summaries of `dplyr`, `ggplot2`, `tidyr`, `stringr`, `forcats` and more. Print
  the ones you use.
- [Notes on the tidyverse](https://jessesadler.github.io/notes-tidyverse/)
- [Data wrangling functions wiki](https://github.com/Cghlewis/data-wrangling-functions/wiki)
  — a task-oriented index: "how do I do X"
- [Allison Horst's illustrated packages and functions](https://allisonhorst.com/r-packages-functions)
- [Posit IDE documentation](https://docs.posit.co/ide/user/) and
  [Positron documentation](https://positron.posit.co/)

**Visual and practical**

- [R Graph Gallery](https://r-graph-gallery.com/) — several hundred chart
  examples with code
- [Data Science Box](https://datasciencebox.org/) — course materials with videos
- [posit::conf talks](https://www.youtube.com/@PositPBC) — conference
  presentations, many of them short and specific
- [TidyTuesday](https://github.com/rfordatascience/tidytuesday) — a new dataset
  every week, with a community that posts their code. The single best way to
  practise once you have finished a course: real messy data, no grade, and
  hundreds of worked solutions to compare yourself against.

**Getting unstuck**

- [Posit Community](https://forum.posit.co/) — friendly, and the tidyverse
  developers read it
- [Stack Overflow, `r` tag](https://stackoverflow.com/questions/tagged/r) — most
  questions have already been asked
- A **reprex** — a minimal reproducible example — gets you an answer far faster
  than a description. The `reprex` package formats one for pasting:

```{r}
#| eval: false
reprex::reprex({
  library(dplyr)
  starwars |> summarise(mean_height = mean(height))
})
```

**Staying current**

Package documentation is the authority, and it is the thing language models are
worst at (@sec-ai). Two commands are worth the habit:

```{r}
#| eval: false
packageVersion("dplyr") # what am I actually running?
news(package = "dplyr") # what changed, and what was superseded?
```
